### 2017-10-31 ms.tex
coppied from Overleaf, changed template, fixed minor issues by MWF

\documentclass[a4paper,twocolumn]{article}   	% use "amsart" instead of "article" for AMSLaTeX format
\usepackage{geometry}                		% See geometry.pdf to learn the layout options. There are lots.
\geometry{letterpaper}                   		% ... or a4paper or a5paper or ... 
%\geometry{landscape}                		% Activate for rotated page geometry
%\usepackage[parfill]{parskip}    		% Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}				% Use pdf, png, jpg, or eps§ with pdflatex; use eps in DVI mode
								% TeX will automatically convert eps --> pdf in pdflatex		
\usepackage{amssymb,amsmath,amsthm}
\usepackage{longtable,booktabs}
\PassOptionsToPackage{hyphens}{url} % url is loaded by hyperref
\usepackage[unicode=true]{hyperref}
\usepackage[auth-sc]{authblk}

%SetFonts
% newtxtext+newtxmath
\usepackage{newtxtext} %loads helv for ss, txtt for tt
\usepackage{amsmath}
\usepackage[bigdelims]{newtxmath}
\usepackage[T1]{fontenc}
\usepackage{textcomp}
%SetFonts

% Yoav & Lee commands
\newcommand*{\tr}{^\intercal}
\newcommand{\matrx}[1]{{\left[ \stackrel{}{#1}\right]}}
\newcommand{\diag}[1]{\mbox{diag}\matrx{#1}}
\newcommand{\goesto}{\rightarrow}
\newcommand{\dspfrac}[2]{\frac{\displaystyle #1}{\displaystyle #2} }
\newtheorem*{theorem}{Theorem}
\newtheorem*{corollary}{Corollary}
\newtheorem*{lemma}{Lemma}
\renewcommand\qedsymbol{} % no square at end of proof

% NatBib
\usepackage[round,colon,authoryear]{natbib}

% Title page
\title{Necessity is the mother of invention}

\author[a]{Yoav Ram}
\author[b]{Lee Altenberg}
\author[a]{Marcus W. Feldman}
\affil[a]{Department of Biology, Stanford University, Stanford, CA}
\affil[b]{Information and Computer Sciences, University of Hawai`I at M{\=a}noa, Honolulu, HI}

\date{\today} % Activate to display a given date or no date

% Document
\begin{document}
\maketitle

% Abstract
\begin{abstract}
Generation of variation is considered detrimental in well-adapted
populations evolving under constant selection. Therefore, the rate at
which variation is generated through processes such as mutation,
migration, and innovation, is considered to be reduced by natural
selection. However, departures from the \emph{reduction principle} have
been demonstrated before. Here we analyze a general model of evolution
under constant selection where the rate of variation generation can
depend on the individual fitness. We find a general and simple result:
if individuals with below average fitness increase the rate of variation
generation, then the population mean fitness increases. This principle
applies to phenomena such as stress-induced mutagenesis,
condition-dependent dispersal and outcrossing, and even to the famous
proverb, \emph{"Necessity is the mother of invention"}.
\end{abstract}

% Introduction
\section*{Introduction}

According to the \emph{reduction principle}, in populations that evolve near an equilibrium between natural selection and generation of variation (i.e. mutation, migration, or recombination), natural selection favors modifiers that decrease the rate of variation generation. 
The \emph{reduction principle} was first proposed for modifiers of the recombination~\citep{Feldman1972}, mutation, and migration~\citep{Liberman1986a}, and dispersal rates~\citep{Altenberg1987}.

These separate principles were recently unified by~\citet{Altenberg2017}, which laid down the assumptions and conditions for the validity of the \emph{unified reduction principle}: effectively infinite population size, constant-viability selection, a population at an equilibrium, and \emph{linear variation} -- the equal scaling of transition probabilities by the modifier.
A departure from the latter assumption occurs if two variation producing processes interact~\citep{Altenberg2012} [REF: Holsinger?], or if there are biases in the mutation or migration process, such that \ldots{}

Another departure from the \emph{linear variation} assumption of the \emph{reduction principle} for mutation rates involves a mechanism commonly called \emph{stress-induced mutagenesis}, in which the mutation rate increases in individuals with poor fitness.
\citet{Ram2012} demonstrated that modifiers that increase the mutation rate of individuals with below average fitness actually \emph{increase} the population mean fitness, rather than decrease it.
Their analysis included an infinite population size and constant-viability selection, such that the only departure from the \emph{reduction principle} assumptions is due to the differential effect the modifier has on different genotypes. 
\citet{Ram2012} stated that their result represents a departure from the \emph{reduction principle}, but did not explain this departure.
Moreover, their analysis was limited to low mutation rates as it assumed that the number of mutations per individual per generation is very low.

Here, we present a general evolutionary model where the rate of variation generation can be different for each type of individual.
This model generalizes both mutation and migration or dispersal models. 
Our result shows that low fitness individuals should produce more variation then high fitness individuals.

% Model
\section*{Model}

Consider a large haploid population with an arbitrary set of types $A_1, A_2, ...$.
The change in the frequencies of these types is described by the transformation $f \to f'$: 
\begin{equation}\label{eq:model}
\bar{\omega} f' = (I-C+MC)D f,
\end{equation}
where $f_i$ is the frequency if type $A_i$, such that $f$ is a stochastic vector ($\sum_i{f_i} = 1$);
$C_i$ is the probability that an individual of type $A_i$ will transition to another type ($C$ is a diagonal non-negative matrix with diagonal elements $C_i$);
$M_{i,j}$ is the transition probability from type $A_j$ to type $A_i$ given that an individual of type $A_j$ is transitioning ($M$ is a stochastic matrix;
$\sum_i {M_{i,j}} = 1$ for any column \emph{j}); 
$D_i$ is the relative fitness of type $A_i$ ($D$ is a diagonal matrix with diagonal elements $0 < D_k \le 1$); 
$I$ is the identity matrix;
and $\bar{\omega}$ is a normalizing factor such that $\sum_i{f'_i}=1$, which turns out to be the population mean fitness (see~\nameref{sec:AppA}).

The types $A_i$ can be encoded by a single or multiple genetic loci or non-genetic trait (behavior, technology. etc.).
The important constraint is that the type is transmitted from a single parent to the offspring, and the transmission infidelity causes type transition (via mutation,
migration, innovation, etc.).
More formally, type transmission must vertical and uni-parental, independent of the frequencies of the other types, and fully described by the matrix $M$. Therefore, this model precludes recombination, social learning, sexual outcrossing, and any horizontal or oblique transmission~\cite[pg.~54]{Cavalli-Sforza1981}.

Transition between types is determined by a combination of two effects:
(i) the probability to transition \emph{out} of type $A_i$ is determined by $C_i$;
(ii) the distribution of the destination types $A_i$ is given by $M_{i,j}$.
We focus on a modifier~\citep{Altenberg2017} that has no direct effect on fitness and whose sole function is to determine the rate of transition out of the different types.
Importantly, this modifier can determine different rates for different types, therefore $C$ is a diagonal matrix rather than a scalar (the latter is the case in the \emph{reduction principle}~\citep{Altenberg2017}.

\textbf{TODO} Next, we present three interpretations of this model for
mutation, migration, and cultural evolution.

% Results
\section*{Results}
% Mean fitness principle
\subsection*{Mean fitness principle}

The equilibrium of the frequency vector $f$ in Eq.~\ref{eq:model} is given by the equilibrium frequency vector $v$ and the stable population mean fitness
$\bar{\omega}^*$, namely
\begin{equation}\label{eq:model_equilibrium}
\bar{\omega}^* v = (I-C+MC)D v
\end{equation}

The following theorem presents a \emph{mean fitness principle} for the
sensitivity of the population mean fitness $\bar{\omega}^*$ to changes
in $C_k$ the rate of transition out of $A_k$.

\begin{theorem}
Let $\bar{\omega}^*$ be the leading eigenvalue of $(I-C+MC)D$, $u$ and $v$ be the corresponding left and right eigenvectors, and $[M]_k$ the $k$-th column of $M$. Then
\begin{equation}\label{eq:theorem}
\frac{\partial \bar{\omega}^*}{\partial C_k} = 
\frac{u_k v_k}{C_k} (\bar{\omega}^* - D_k),
\end{equation}
or in simpler terms,
$$
sign\frac{\partial \bar{\omega}^*}{\partial C_k} = 
sign(\bar{\omega}^* - D_k), 
$$
and therefore increased transition out of type $k$ will increase the stable population mean fitness if and only if the fitness of type $k$ is below the stable population mean fitness.
\end{theorem}

\begin{proof}
By the \emph{Perron-Frobenius theorem}~\cite[Appendix~A]{Otto2007},
$\bar{\omega}^*$ is the leading real positive eigenvalue of $(I-C+MC)D$,
and the corresponding left and right eigenvectors $u$ and $v$ are both non-negative and uniquely defined by $\sum_i{v_i} = 1$ and $\sum_i{u_i v_i} = 1$.

Using \emph{Caswell's formula} (Eq.~\ref{eq:Caswells_formula} in~\nameref{sec:AppB}),
$$
\frac{\partial \bar{\omega}^*}{\partial C_k} = 
u\tr \frac{\partial (I-C+MC)D}{\partial C_k} v.
$$

Let $e_k$ and $e\tr_k$ be the column and row vectors with 1 at position $k$ and 0 elsewhere, $Z_k = e_k e\tr_k$ the matrix with 1 at position $(k,k)$ and 0 elsewhere, and $[M]_k$ the $k$-th column of $M$.

Then,
\begin{equation}\label{eq:theorem_intermediate}
\begin{aligned}
u\tr \frac{\partial (I-C+MC)D}{\partial C_k} v = 
u\tr (0 - Z_k + M Z_k)D v = \\
-v_k u_k D_k + v_k D_k u\tr[M]_k = \\
v_k D_k (u\tr[M]_k - u_k). 
\end{aligned}
\end{equation}

The corresponding equation to Eq.~\ref{eq:model_equilibrium} for the left
eigenvector $u$ is 
$$
u\tr \bar{\omega}^* = u\tr (I - C + MC) D,
$$
which gives us a relation between $\bar{\omega}^*$ and the $k$
element of $u$:
$$
u_k \bar{\omega}^* = (1-C_k) D_k u_k + C_k D_k u\tr [M]_k.
$$
Multiplying both sides by $v_k$ (which is positive if $D_k>0$) and rearranging, we get $\frac{u_k v_k}{C_k} (\bar{\omega}^* - D_k) = v_k D_k(u\tr [M]_k - u_k)$.
Plugging this in eq. \ref{eq:theorem_intermediate} yields:
$$
\frac{\partial \bar{\omega}^*}{\partial C_k} = 
\frac{u_k v_k}{C_k} (\bar{\omega}^* - D_k),
$$
which completes the proof.
\end{proof}

Note that if the modifier has the same effect on all types (i.e.,
$C_k = \alpha$ for all $k$) then Eq.~\ref{eq:model_equilibrium}
becomes $\bar{\omega}^* v = ((1-\alpha)I + \alpha M)Dv$,
and the \emph{unified reduction principle} is in effect~\cite[eqs. 65, 72]{Altenberg2017}.
Therefore, we can expect modifiers that reduce $\alpha$ to be favored by natural selection.
Indeed, setting $C = \alpha I$ in Eq.~\ref{eq:model_equilibrium} and proceeding as in the proof, we get
$$
\frac{\partial \bar{\omega}^*}{\partial \alpha} = 
\bar{\omega}^* - 1 \le 0,
$$
because $\bar{\omega}^* = \sum_i{v_i D_i}$ (see~\nameref{sec:AppA}) and $D_i \le 1$ for all $i$.

% Reproductive value principle
\subsection*{Reproductive value principle}

A corollary of Eq.~\ref{eq:theorem} appears in Eq.~\ref{eq:theorem_intermediate}:

\begin{multline}\label{eq:corollary_reproductive_value}
\frac{\partial \bar{\omega}^*}{\partial C_k} = 
v_k D_k (u\tr[M]_k - u_k) = \\
D_k (\sum_i{u_i M_{i,k} v_k} - u_k v_k).
\end{multline}

$u_k$ can be regarded as \emph{Fisher's reproductive value} of type $k$ (see~\nameref{sec:AppC}), which gives the relative contribution of type $k$ to the long-term population.
Subsequently, $u_k v_k$ is the \emph{ancestor frequency} \citep{Hermisson2002} of type $k$, which is the fraction of the equilibrium population descended from type $k$.
The sum $\sum_i{u_i M_{i,k} v_k}$ can be similarly interpreted as the fraction of the equilibrium population descended from individuals that transitioned from type $k$ to another type (via the $k$ column of the transition matrix $M$).

Given $0 < D_k \le 1$, from Eq.~\ref{eq:theorem_intermediate} we can simplify the mathematical notation to 
$$
sign \frac{\partial \bar{\omega}^*}{\partial C_k} = 
sign (u\tr [M]_k - u_k),
$$
which means that increasing the overall transition rate $C_k$ from type $k$ to other types will increase the population mean fitness if and only if the fraction of the population descended from type $k$ is expected to increase due to a transition to another type.

% Threshold modifier
\subsection*{Threshold modifier}

We can turn our attention to a modifier that concurrently controls the transition out of a subset of types, $K$.
For example, \citet[Appendix~B]{Ram2012} considered individuals that are grouped by the number of their accumulated deleterious mutations, $k$, and a modifier that regulates the mutation rates in individuals with at least $\pi$ deleterious mutations.

Similarly, if the transition rates are controlled by a threshold rule 
$$
C_k = \begin{cases}
\tau, & k \in K \\
1, & k \not\in K
\end{cases},
$$
then the sensitivity of the population mean fitness to changes in the
rate of transition out of types $K$ is
\begin{equation*}
\begin{aligned}
\frac{\partial \bar{\omega}^*}{\partial \tau} = 
u\tr \Big( \sum_{k \in \mathbb{K{}}}{\frac{\partial (I - C + MC)D}{\partial \tau}} \Big) v = \\
\frac{1}{\tau} \sum_{k \in \mathbb{K}}{u_k v_k (\bar{\omega}^* - D_k)}.
\end{aligned}
\end{equation*}

% Discussion
\section*{Discussion}
% Departure from the reduction principle
\subsection*{Departure from the \emph{reduction principle}}

% Extension of Ram and Hadany, 2012
\subsection*{Extension of \citealt{Ram2012}}

The theorem Eq.~\ref{eq:theorem} is a generalization of a result from \citet[Eq.~4]{Ram2012}.
Translated to the notation presented here, \citeauthor{Ram2012} modeled the accumulation of deleterious mutations in a population at the mutation-selection balance, assuming no more than one mutation occurs per individual per generation: 
$$
M_{i,j} = \begin{cases}
\beta \mu, & i = j-1 \\
(1-\beta-\delta) \mu, & i = j \\
\delta \mu, & i = j+1 \\
(1-\delta \mu), & i=j=0 \\
0, & \text{otherwise}
\end{cases}.
$$

Using Eq.~\ref{eq:Caswells_formula} in \nameref{sec:AppB} and a recursion on the ratios of the reproductive values \cite[see][eqs.~A5-6]{Ram2012}, they concluded
that at the mutation-selection balance, if individuals with below-average fitness ($D_k < \bar{\omega}^*$) increase their mutation rate, then the population mean fitness will increase.

% Evidence
\subsection*{Evidence}

\textbf{TODO} evidence for faster generation of variation in low fitness individuals.

% Conclusions
\subsection*{Conclusions}

\section*{Appendices}
% Appendix A
\subsection*{Appendix A}\label{sec:AppA}
\subsubsection*{The population mean fitness is an eigenvalue}

Given Eq.~\ref{eq:model}, we show that the normalizing factor is the population mean fitness, i.e. $\bar{\omega} = \sum_i{f_i D_i}$.

Denote $e = (1, ..., 1)$ the vector with ones at all positions.
$e\tr M = e\tr$ because $M$ is a stochastic matrix. Together with $e\tr I = e\tr$,
we have
\begin{multline}\label{eq:mean_fitness}
\lambda = 
e\tr (I - C + M C) D f = \\
(e\tr I - e\tr C + e\tr M C) D f = \\
(e\tr - e\tr C + e\tr C) D f = \\
e\tr D f = \\
\diag{D} f = \\
\sum_i{D_i f_i} = \bar{\omega} \;\;\; \blacksquare
\end{multline}

% Appendix B
\subsection*{Appendix B}\label{sec:AppB}
\subsubsection*{Caswell's formula}

\citet{Caswell1978} presented a \emph{formula for the sensitivity of the population growth rate to changes in life history parameters}.
In this formula, the population growth rate is the leading eigenvalue of the population transformation matrix $T$, the life history parameters are elements of $T$, and the \emph{sensitivity} is the derivative of the former with respect
to the latter.
This is a useful formula~\citep{Hermisson2002,Ram2012}.

\begin{lemma}[Caswell's formula~\citep{Caswell1978}]
Let $T$ be a matrix with eigenvalue $\lambda$ and left and right eigenvectors $u$ and $v$ such that $\sum{v_i}=1$ and $u\tr v = \sum{u_i v_i} = 1$,
then the sensitivity of $\lambda$ to changes in any parameter $x$ of the matrix $T$ is
\begin{equation}\label{eq:Caswells_formula}
\frac{\partial \lambda}{\partial x} = 
u\tr \frac{\partial T}{\partial x} v
\end{equation}
\end{lemma}

\begin{proof} 
Using the lemma assumptions,
$\lambda = \lambda u\tr v = u\tr \lambda v = u\tr T v$ and differentiating both sides we get $\partial \lambda = \partial (u\tr T v)$.
Using the product rule (once in each direction),

\begin{multline*}
\partial (u\tr T v) = \\
\partial u\tr T v + u\tr \partial T v + u\tr T \partial v = \\
u\tr \partial T v + \partial u\tr \lambda v  + \lambda u\tr \partial v = \\
u\tr \partial T v + \lambda(\partial u\tr v  + u\tr \partial v) = \\
u\tr \partial T v + \lambda \partial(u\tr v).
\end{multline*}

Because $u\tr v = 1$,
we have $\partial (u\tr v) = 0$ and
$\partial \lambda = u\tr \partial T v$.
\end{proof}

% Appendix C
\subsection*{Appendix C}\label{sec:AppC}
\subsubsection*{Fisher's reproductive value}

Let $M$ be a stochastic matrix and $D$ be a diagonal matrix with positive diagonal elements. 
The elements of th left \emph{Perron} eigenvector $u$ of matrix $MD$ can be regarded as \emph{Fisher's reproductive values}.
These values can be understood as follows.

Let us look at the dynamics not of frequencies but of absolute
population sizes, $n_i(t)$.
Let the vector of the number of each type at time $t$ be $n(t)$.
The dynamics are
$$
n(t) = (MD)^t n(0).
$$

Let $n(j, t)$ be the vector when the initial population is a single individual of type $j$.
The dynamics are
$$
n(j,t) = (MD)^t e_j,
$$
where $e_i$ is a vector with 1 at position $i$ and 0 elsewhere.

The total population size at time $t$ starting with type $j$ is then
$$
N(j,t) = e\tr n(j,t) = e\tr (MD)^t e_j.
$$

Now we can compare the population sizes based on what type they start
with:
$$
\frac{N(j,t)}{N(k,t)} = \frac{e\tr (MD)^t e_j}{e\tr (MD)^t e_k}.
$$

To further evaluate, we write $MD$ as its Jordan canonical form, 
$$
A = V \Lambda U\tr,
$$
where $V$ is the matrix of right column eigenvectors of $MD$,
$U\tr$ is the matrix of left row eigenvectors of $MD$,
$V U\tr = U\tr V = I$, 
and $\Lambda$ is the diagonal matrix of
eigenvalues of $A$ (for a nongeneric set of matrices $M$, the
geometric and algebraic multiplicities of the eigenvalues of $MD$
differ, and $\Lambda$ will not be a diagonal matrix, a case we can
ignore).

Hence,
\begin{multline*}
N(i, t) = 
e\tr n(t) = \\
e\tr (MD)^t e_i =
e\tr  (V \Lambda U\tr)^t e_i = \\
e\tr V \Lambda^t U\tr e_i.
\end{multline*}

For the ratio, we can divide $\Lambda$ by $\lambda_1 = \rho(MD)$,
the spectral radius of $MD$:
\begin{multline*}
\frac{N(j,t)}{N(k,t)} =
\frac{ e\tr V \Lambda^t U\tr e_j}{ e\tr V \Lambda^t U\tr e_k} = \\
\frac{ e\tr V \diag{1, (\frac{\lambda_2}{\lambda_1})^t, \cdots} U\tr e_j} { e\tr V \diag{1,(\frac{\lambda_2}{\lambda_1})^t, \cdots} U\tr e_k}.
\end{multline*}

Now take the long term limit.
By assumption $MD$ is irreducible,
hence $\lambda_i < \lambda_1$ for all $i > 1$.
Hence $\lim_{t \goesto \infty} \lambda_i / \lambda_1 = 0$ for all $i > 1$.
Thus:
\begin{multline*}
\lim_{t \goesto \infty}\frac{N(j,t)}{N(k,t)} =
\lim_{t \goesto \infty} \frac{ e\tr V \Lambda^t U\tr e_j}{ e\tr V \Lambda^t U\tr e_k} = \\
\lim_{t \goesto \infty} \frac{ e\tr V diag {1, (\dspfrac{\lambda_2}{\lambda_1})^t, \cdots} U\tr e_h} { e\tr V diag {1,(\dspfrac{\lambda_2}{\lambda_1})^t, \cdots} U\tr e_k} = \\
\frac{e\tr V (e_1 e_1\tr) U\tr e_j} { e\tr V (e_1 e_1\tr) U\tr e_k} = \\
\frac{ e\tr v u_j} { e\tr v  u_k} =
\frac{u_j} {u_k}.
\end{multline*}

The vector $u$ is the left \emph{Perron} eigenvector of $MD$, and $u_i$ is element $i$ of $u$.
This is why the value $u_i$ can be interpreted as the \emph{reproductive value} of type $i$, as it is a weight for the size of the population generated by a single individual of type $i$.

If we begin with a population at the equilibrium distribution $v$,
and ask what fraction of long-term descendants descended from type $i$ at that time,
we weight $v_i$ by the reproductive value $u_i$, to get $u_i v_i$.
We note that $\{u_i v_i\}_{i}$ is a normalized probability distribution, since
$$
\sum_i {u_i v_i} = u\tr v = 1.
$$ 
\citet{Hermisson2002} called the $u_i v_i$ the \emph{ancestor} or \emph{ancestral distribution}.

\textbf{TODO} cite \citep{Grafen2006}?

% Acknowledgements
{\small
\section*{Acknowledgements}

This manuscript was supported in part by the Stanford Center for Computational, Evolutionary and Human Genomics, and by the Morrison Institute for Population and Resources Studies, Stanford University.
}

\bibliographystyle{plainnat}
\bibliography{/Users/yoavram/Documents/library}

\end{document}

### 2017-11-01 ms.tex
revised all section - formatting, phrasing, typesetting, TODOs

@@
-\documentclass[a4paper,twocolumn]{article}   	% use "amsart" instead of "article" for AMSLaTeX format
-\usepackage{geometry}                		% See geometry.pdf to learn the layout options. There are lots.
+\documentclass[9pt, a4paper, twocolumn]{extarticle}   	% use "amsart" instead of "article" for AMSLaTeX format
+\usepackage[margin=1in]{geometry}                		% See geometry.pdf to learn the layout options. There are lots.
 \geometry{letterpaper}                   		% ... or a4paper or a5paper or ... 
@@
 \newcommand*{\tr}{^\intercal}
+\let\vec\mathbf
 \newcommand{\matrx}[1]{{\left[ \stackrel{}{#1}\right]}}
@@
 \newtheorem*{lemma}{Lemma}
+\newtheorem*{mfp}{Theorem: Mean Fitness Principle}
+\newtheorem*{rvp}{Corollary: Reproductive Value Principle}
+\newtheorem*{caswell}{Lemma: Caswell's Formula}
 \renewcommand\qedsymbol{} % no square at end of proof
@@
 Generation of variation is considered detrimental in well-adapted
-populations evolving under constant selection. Therefore, the rate at
-which variation is generated through processes such as mutation,
-migration, and innovation, is considered to be reduced by natural
-selection. However, departures from the \emph{reduction principle} have
-been demonstrated before. Here we analyze a general model of evolution
-under constant selection where the rate of variation generation can
-depend on the individual fitness. We find a general and simple result:
-if individuals with below average fitness increase the rate of variation
-generation, then the population mean fitness increases. This principle
-applies to phenomena such as stress-induced mutagenesis,
-condition-dependent dispersal and outcrossing, and even to the famous
-proverb, \emph{"Necessity is the mother of invention"}.
+populations evolving under constant selection.
+Therefore, the rate at which variation is generated through processes such as mutation,
+migration, and innovation is reduced by natural selection.
+However, departures from this \emph{reduction principle} have been demonstrated.
+Here we analyze a general model of evolution under constant selection where the rate at which variation is generated can depend on the individual fitness.
+We find a general and simple result:
+if individuals with below-average fitness increase the rate at which variation is generated,
+then the population mean fitness increases.
+This principle applies to phenomena such as stress-induced mutagenesis,
+condition-dependent dispersal and outcrossing,
+and even to the famous proverb: \emph{"Necessity is the mother of invention"}.
 \end{abstract}
@@
 
-According to the \emph{reduction principle}, in populations that evolve near an equilibrium between natural selection and generation of variation (i.e.
-mutation, migration, or recombination), natural selection favors modifiers that decrease the rate of variation generation.
-The \emph{reduction principle} was first proposed for modifiers of the recombination~\citep{Feldman1972}, mutation, and migration~\citep{Liberman1986a}, and dispersal rates~\citep{Altenberg1987}.
+According to the \emph{reduction principle}, in populations that evolve near an equilibrium between natural selection and a process that generates variation (i.e.
+mutation, migration, or recombination), selection favors modifiers that decrease the rate of variation generation.
 
-These separate principles were recently unified by~\citet{Altenberg2017}, which laid down the assumptions and conditions for the validity of the \emph{unified reduction principle}: effectively infinite population size, constant-viability selection, a population at an equilibrium, and \emph{linear variation} -- the equal scaling of transition probabilities by the modifier.
+The \emph{reduction principle} was demonstrated for modifiers of the recombination~\citep{Feldman1972}, mutation, and migration~\citep{Liberman1986a}, and dispersal rates~\citep{Altenberg1987}.
+These separate principles were recently unified by~\citet{Altenberg2017}
+
+The assumptions and conditions for the validity of the \emph{unified reduction principle} are~\citep{Altenberg2017}:
+(i) effectively infinite population size, (ii) constant-viability selection, (iii) a population at an equilibrium, and (iv) \emph{linear variation} -- the equal scaling of transition probabilities by the modifier.
 A departure from the latter assumption occurs if two variation producing processes interact~\citep{Altenberg2012} [REF: Holsinger?], or if there are biases in the mutation or migration process, such that \ldots{}
@@
 Another departure from the \emph{linear variation} assumption of the \emph{reduction principle} for mutation rates involves a mechanism commonly called \emph{stress-induced mutagenesis}, in which the mutation rate increases in individuals with poor fitness.
-\citet{Ram2012} demonstrated that modifiers that increase the mutation rate of individuals with below average fitness actually \emph{increase} the population mean fitness, rather than decrease it.
+\citet{Ram2012} demonstrated that modifiers that increase the mutation rate of individuals with below-average fitness actually \textbf{increase} the population mean fitness, rather than decrease it.
 Their analysis included an infinite population size and constant-viability selection, such that the only departure from the \emph{reduction principle} assumptions is due to the differential effect the modifier has on different genotypes.
 \citet{Ram2012} stated that their result represents a departure from the \emph{reduction principle}, but did not explain this departure.
-Moreover, their analysis was limited to low mutation rates as it assumed that the number of mutations per individual per generation is very low.
+Moreover, their analysis was limited to low mutation rates as it neglected double-mutations (i.e. assuming $U<1/\sqrt{N}$ where $U$ is the genome-wide mutation rate and $N$ is the population size).
 
-Here, we present a general evolutionary model where the rate of variation generation can be different for each type of individual.
-This model generalizes both mutation and migration or dispersal models. 
+Here, we present a general evolutionary model where the rate at which variation is generated depend on the type of individual, and specifically on the its fitness.
 Our result shows that low fitness individuals should produce more variation then high fitness individuals.
@@
 \end{equation}
-where $f_i$ is the frequency if type $A_i$, such that $f$ is a stochastic vector ($\sum_i{f_i} = 1$);
-$C_i$ is the probability that an individual of type $A_i$ will transition to another type ($C$ is a diagonal non-negative matrix with diagonal elements $C_i$);
-$M_{i,j}$ is the transition probability from type $A_j$ to type $A_i$ given that an individual of type $A_j$ is transitioning ($M$ is a stochastic matrix;
-$\sum_i {M_{i,j}} = 1$ for any column \emph{j}); 
-$D_i$ is the relative fitness of type $A_i$ ($D$ is a diagonal matrix with diagonal elements $0 < D_k \le 1$); 
+where $f_k$ is the frequency of type $A_k$, such that $f$ is a stochastic vector ($\sum_k{f_k} = 1$);
+$C_k$ is the probability that an individual of type $A_k$ will transition to another type ($C$ is a diagonal non-negative matrix with diagonal elements $C_k$);
+$M_{i,j}$ is the transition probability from type $A_j$ to type $A_i$ given that an individual of type $A_j$ is transitioning ($M$ is a stochastic matrix: $\sum_j {M_{i,j}} = 1$ for any column \emph{j});
+$D_k$ is the relative fitness of type $A_k$ ($D$ is a diagonal matrix with diagonal elements $0 < D_k \le 1$; fitness values are relative to the fitness of the fittest type); 
 $I$ is the identity matrix;
-and $\bar{\omega}$ is a normalizing factor such that $\sum_i{f'_i}=1$, which turns out to be the population mean fitness (see~\nameref{sec:AppA}).
+and $\bar{\omega}$ is a normalizing factor such that $\sum_k{f'_k}=1$, which turns out to be the population mean fitness (see~\nameref{sec:AppA}).
 
-The types $A_i$ can be encoded by a single or multiple genetic loci or non-genetic trait (behavior, technology. etc.).
-The important constraint is that the type is transmitted from a single parent to the offspring, and the transmission infidelity causes type transition (via mutation,
-migration, innovation, etc.).
-More formally, type transmission must vertical and uni-parental, independent of the frequencies of the other types, and fully described by the matrix $M$.
+The types $A_k$ can be encoded by a single or multiple genetic loci or non-genetic traits.
+Importantly, the type is transmitted from a single parent to the offspring, and transmission infidelity causes type transitions.
+For example, genotypes are transmitted by genetic inheritance, and genetic infidelity, i.e.
+mutation, generates new genotypes.
+More formally, type transmission is vertical and uni-parental, independent of the frequencies of the other types, and is fully described by the matrix $M$.
 Therefore, this model precludes recombination, social learning, sexual outcrossing, and any horizontal or oblique transmission~\cite[pg.~54]{Cavalli-Sforza1981}.
@@
 Transition between types is determined by a combination of two effects:
-(i) the probability to transition \emph{out} of type $A_i$ is determined by $C_i$;
-(ii) the distribution of the destination types $A_i$ is given by $M_{i,j}$.
-We focus on a modifier~\citep{Altenberg2017} that has no direct effect on fitness and whose sole function is to determine the rate of transition out of the different types.
-Importantly, this modifier can determine different rates for different types, therefore $C$ is a diagonal matrix rather than a scalar (the latter is the case in the \emph{reduction principle}~\citep{Altenberg2017}.
-
-\textbf{TODO} Next, we present three interpretations of this model for
-mutation, migration, and cultural evolution.
+(i) the probability to transition \emph{out} of type $A_k$ is determined by $C_k$;
+(ii) given a transition out of type $A_i$, the distribution of the destination types $A_j$ is given by $M_{j,i}$ (note the index order).
+We focus on a modifier that has no direct effect on fitness and whose sole function is to determine the rate of transition out of the different types ($C_k$).
+Importantly, this modifier can determine different rates for different types, therefore $C$ is a diagonal matrix rather than a scalar.
+The case of a scalar in covered by the \emph{reduction principle}, see~\citet{Altenberg2017}.
 
@@
 \section*{Results}
+
 % Mean fitness principle
-\subsection*{Mean fitness principle}
-
 The equilibrium of the frequency vector $f$ in Eq.~\ref{eq:model} is given by the equilibrium frequency vector $v$ and the stable population mean fitness
@@
 The following theorem presents a \emph{mean fitness principle} for the
-sensitivity of the population mean fitness $\bar{\omega}^*$ to changes
-in $C_k$ the rate of transition out of $A_k$.
+sensitivity of the stable population mean fitness $\bar{\omega}^*$ to changes
+in $C_k$ the probability for transition out of $A_k$.
 
-\begin{theorem}
+\begin{mfp}
 Let $\bar{\omega}^*$ be the leading eigenvalue of $(I-C+MC)D$, $u$ and $v$ be the corresponding left and right eigenvectors, and $[M]_k$ the $k$-th column of $M$. Then
@@
 or in simpler terms,
-$$
+\begin{equation}
 sign\frac{\partial \bar{\omega}^*}{\partial C_k} = 
 sign(\bar{\omega}^* - D_k), 
-$$
+\end{equation}
 and therefore increased transition out of type $k$ will increase the stable population mean fitness if and only if the fitness of type $k$ is below the stable population mean fitness.
-\end{theorem}
+\end{mfp}
 
@@
 $\bar{\omega}^*$ is the leading real positive eigenvalue of $(I-C+MC)D$,
-and the corresponding left and right eigenvectors $u$ and $v$ are both non-negative and uniquely defined by $\sum_i{v_i} = 1$ and $\sum_i{u_i v_i} = 1$.
+and the corresponding left and right eigenvectors $u$ and $v$ are both non-negative and uniquely defined by $\sum_k{v_k} = 1$ and $\sum_k{u_k v_k} = 1$.
 
 Using \emph{Caswell's formula} (Eq.~\ref{eq:Caswells_formula} in~\nameref{sec:AppB}),
-$$
+\begin{equation}
 \frac{\partial \bar{\omega}^*}{\partial C_k} = 
 u\tr \frac{\partial (I-C+MC)D}{\partial C_k} v.
-$$
+\end{equation}
 
@@
 eigenvector $u$ is 
-$$
+\begin{equation}
 u\tr \bar{\omega}^* = u\tr (I - C + MC) D,
-$$
+\end{equation}
 which gives us a relation between $\bar{\omega}^*$ and the $k$
 element of $u$:
-$$
+\begin{equation}
 u_k \bar{\omega}^* = (1-C_k) D_k u_k + C_k D_k u\tr [M]_k.
-$$
-Multiplying both sides by $v_k$ (which is positive if $D_k>0$) and rearranging, we get $\frac{u_k v_k}{C_k} (\bar{\omega}^* - D_k) = v_k D_k(u\tr [M]_k - u_k)$.
-Plugging this in eq. \ref{eq:theorem_intermediate} yields:
-$$
+\end{equation}
+Multiplying both sides by $v_k$ (which is positive if $D_k>0$) and rearranging, we get 
+\begin{equation}
+\frac{u_k v_k}{C_k} (\bar{\omega}^* - D_k) = v_k D_k(u\tr [M]_k - u_k).
+\end{equation}
+Plugging this in Eq.~\ref{eq:theorem_intermediate} yields:
+\begin{equation}
 \frac{\partial \bar{\omega}^*}{\partial C_k} = 
 \frac{u_k v_k}{C_k} (\bar{\omega}^* - D_k),
-$$
+\end{equation}
 which completes the proof.
@@
 % Reproductive value principle
-\subsection*{Reproductive value principle}
 
-A corollary of Eq.~\ref{eq:theorem} appears in Eq.~\ref{eq:theorem_intermediate}:
+An interesting intermediate result in Eq.~\ref{eq:theorem_intermediate} is
 
-\begin{multline}\label{eq:corollary_reproductive_value}
+\begin{equation}
+\begin{aligned}
 \frac{\partial \bar{\omega}^*}{\partial C_k} = 
@@
 D_k (\sum_i{u_i M_{i,k} v_k} - u_k v_k).
-\end{multline}
-
-$u_k$ can be regarded as \emph{Fisher's reproductive value} of type $k$ (see~\nameref{sec:AppC}), which gives the relative contribution of type $k$ to the long-term population.
+\end{aligned}
+\end{equation}
+Here, $u_k$ can be regarded as the \emph{reproductive value} of type $k$ (see~\nameref{sec:AppC}), which gives the relative contribution of type $k$ to the long-term population.
 Subsequently, $u_k v_k$ is the \emph{ancestor frequency} \citep{Hermisson2002} of type $k$, which is the fraction of the equilibrium population descended from type $k$.
@@
 
-Given $0 < D_k \le 1$, from Eq.~\ref{eq:theorem_intermediate} we can simplify the mathematical notation to 
-$$
+Given $0 < D_k \le 1$, from Eq.~\ref{eq:theorem_intermediate} we can get the following corollary.
+
+\begin{rvp}
+In the notation of the theorem,
+\begin{equation}
 sign \frac{\partial \bar{\omega}^*}{\partial C_k} = 
 sign (u\tr [M]_k - u_k),
-$$
-which means that increasing the overall transition rate $C_k$ from type $k$ to other types will increase the population mean fitness if and only if the fraction of the population descended from type $k$ is expected to increase due to a transition to another type.
+\end{equation}
+and therefore increased transition out of type $k$ will increase the stable population mean fitness if and only if the fraction of the population descended from type $k$ is expected to increase due to a transition to another type.
+\end{rvp}
 
-% Threshold modifier
-\subsection*{Threshold modifier}
-
-We can turn our attention to a modifier that concurrently controls the transition out of a subset of types, $K$.
+We turn our attention to a modifier that concurrently controls the transition out of a subset of types, $K$.
 For example, \citet[Appendix~B]{Ram2012} considered individuals that are grouped by the number of their accumulated deleterious mutations, $k$, and a modifier that regulates the mutation rates in individuals with at least $\pi$ deleterious mutations.
@@
 Similarly, if the transition rates are controlled by a threshold rule 
-$$
+\begin{equation}
 C_k = \begin{cases}
@@
 \end{cases},
-$$
-then the sensitivity of the population mean fitness to changes in the
+\end{equation}
+then the sensitivity of the stable population mean fitness to changes in the
 rate of transition out of types $K$ is
-\begin{equation*}
+\begin{equation}
 \begin{aligned}
@@
 \end{aligned}
-\end{equation*}
+\end{equation}
+
+\textbf{TODO} summarize result in words.
 
@@
 \section*{Discussion}
-% Departure from the reduction principle
-\subsection*{Departure from the \emph{reduction principle}}
+\textbf{TODO}: Departure from the \emph{reduction principle}
 
-% Extension of Ram and Hadany, 2012
-\subsection*{Extension of \citealt{Ram2012}}
-
-The theorem Eq.~\ref{eq:theorem} is a generalization of a result from \citet[Eq.~4]{Ram2012}.
+Eq.~\ref{eq:theorem} is a generalization of a result from \citet[Eq.~4]{Ram2012}.
 Translated to the notation presented here, \citeauthor{Ram2012} modeled the accumulation of deleterious mutations in a population at the mutation-selection balance, assuming no more than one mutation occurs per individual per generation:
-$$
+\begin{equation}
 M_{i,j} = \begin{cases}
@@
 0, & \text{otherwise}
-\end{cases}.
-$$
+\end{cases}
+\end{equation}
 
-Using Eq.~\ref{eq:Caswells_formula} in \nameref{sec:AppB} and a recursion on the ratios of the reproductive values \cite[see][eqs.~A5-6]{Ram2012}, they concluded
-that at the mutation-selection balance, if individuals with below-average fitness ($D_k < \bar{\omega}^*$) increase their mutation rate, then the population mean fitness will increase.
+Using Eq.~\ref{eq:Caswells_formula} in \nameref{sec:AppB} and a recursion on the ratios of the reproductive values \cite[see][eqs.~A5-6]{Ram2012}, they concluded that at the mutation-selection balance, if individuals with below-average fitness ($D_k < \bar{\omega}^*$) increase their mutation rate, then the population mean fitness will increase.
 
-% Evidence
-\subsection*{Evidence}
-
-\textbf{TODO} evidence for faster generation of variation in low fitness individuals.
+\textbf{TODO} migration model
+\textbf{TODO} cultural evolution model?
 
@@
 \subsection*{Appendix A}\label{sec:AppA}
-\subsubsection*{The population mean fitness is an eigenvalue}
 
-Given Eq.~\ref{eq:model}, we show that the normalizing factor is the population mean fitness, i.e. $\bar{\omega} = \sum_i{f_i D_i}$.
+\begin{lemma}
+The normalizing factor $\bar{\omega}$ in Eq.~\ref{eq:model} is the population mean fitness
+\begin{equation}
+\bar{\omega} = \sum_k{f_k D_k}.
+\end{equation}
+\end{lemma}
 
-Denote $e = (1, ..., 1)$ the vector with ones at all positions.
-$e\tr M = e\tr$ because $M$ is a stochastic matrix. Together with $e\tr I = e\tr$,
-we have
+\begin{proof}
+By definition, the normalizing factor $\bar{\omega}$ ensures that $\sum_k{f'_k}=1$.
+Denote $e = (1, ..., 1)$ the vector with ones at all positions and rewrite $e\tr f' = 1$.
+Therefore, 
 \begin{multline}\label{eq:mean_fitness}
-\lambda = 
+\bar{\omega} = 
+\bar{\omega} e\tr f' = 
 e\tr (I - C + M C) D f = \\
@@
 (e\tr - e\tr C + e\tr C) D f = \\
-e\tr D f = \\
-\diag{D} f = \\
-\sum_i{D_i f_i} = \bar{\omega} \;\;\; \blacksquare
+e\tr D f = 
+\diag{D} f = 
+\sum_k{f_k D_k},
 \end{multline}
+using the fact that $e\tr M = e\tr$ and $e\tr I = e\tr$ because $M$ and $I$ are stochastic matrices.
+\end{proof}
+
+Note that this lemma also ensures that the stable population mean fitness $\bar{\omega}^*=\sum_k{v_k D_k},$ is the leading eigenvalue of $(I - C + M C) D$; see Eq.~\ref{eq:model_equilibrium} and the \emph{Perron-Frobenius theorem}~\cite[Appendix~A]{Otto2007}.
 
@@
 \subsection*{Appendix B}\label{sec:AppB}
-\subsubsection*{Caswell's formula}
 
 \citet{Caswell1978} presented a \emph{formula for the sensitivity of the population growth rate to changes in life history parameters}.
-In this formula, the population growth rate is the leading eigenvalue of the population transformation matrix $T$, the life history parameters are elements of $T$, and the \emph{sensitivity} is the derivative of the former with respect
-to the latter.
-This is a useful formula~\citep{Hermisson2002,Ram2012}.
+In this formula, the \emph{population growth rate} is the leading eigenvalue of the population transformation matrix $T$, the \emph{life history parameters} are elements of $T$, and the \emph{sensitivity} is the derivative of the former with respect to the latter.
+This is a useful formula~\citep{Hermisson2002,Ram2012}, and therefore we reproduce it here.
 
-\begin{lemma}[Caswell's formula~\citep{Caswell1978}]
-Let $T$ be a matrix with eigenvalue $\lambda$ and left and right eigenvectors $u$ and $v$ such that $\sum{v_i}=1$ and $u\tr v = \sum{u_i v_i} = 1$,
-then the sensitivity of $\lambda$ to changes in any parameter $x$ of the matrix $T$ is
+\begin{caswell}
+Let $T$ be a matrix with eigenvalue $\lambda$ and left and right eigenvectors $u$ and $v$ such that $\sum{v_k}=1$ and $u\tr v = \sum{u_k v_k} = 1$.
+Then the sensitivity of $\lambda$ to changes in any element $t$ of the matrix $T$ is
 \begin{equation}\label{eq:Caswells_formula}
-\frac{\partial \lambda}{\partial x} = 
-u\tr \frac{\partial T}{\partial x} v
+\frac{\partial \lambda}{\partial t} = 
+u\tr \frac{\partial T}{\partial t} v
 \end{equation}
-\end{lemma}
+\end{caswell}
 
@@
 
-\begin{multline*}
-\partial (u\tr T v) = \\
+\begin{multline}
+\partial (u\tr T v) = 
 \partial u\tr T v + u\tr \partial T v + u\tr T \partial v = \\
@@
 u\tr \partial T v + \lambda \partial(u\tr v).
-\end{multline*}
+\end{multline}
 
@@
 \subsection*{Appendix C}\label{sec:AppC}
-\subsubsection*{Fisher's reproductive value}
 
 Let $M$ be a stochastic matrix and $D$ be a diagonal matrix with positive diagonal elements. 
-The elements of th left \emph{Perron} eigenvector $u$ of matrix $MD$ can be regarded as \emph{Fisher's reproductive values}.
+The elements of the left \emph{Perron} eigenvector $u$ of the matrix $MD$ can be regarded as \textbf{Fisher's reproductive values}.
 These values can be understood as follows.
 
-Let us look at the dynamics not of frequencies but of absolute
-population sizes, $n_i(t)$.
-Let the vector of the number of each type at time $t$ be $n(t)$.
+Let us look at the dynamics not of frequencies but of absolute population sizes such that the vector of the number of individuals of each type at time $t$ is $n(t)$ and $f_k(t) = n_k(t) / \sum_i{n_i(t)}$.
 The dynamics are
-$$
+\begin{equation}
 n(t) = (MD)^t n(0).
-$$
+\end{equation}
 
-Let $n(j, t)$ be the vector when the initial population is a single individual of type $j$.
+Let $n(k, t)$ be the vector when the initial population is a single individual of type $k$.
 The dynamics are
-$$
-n(j,t) = (MD)^t e_j,
-$$
-where $e_i$ is a vector with 1 at position $i$ and 0 elsewhere.
+\begin{equation}
+n(k,t) = (MD)^t e_k,
+\end{equation}
+where $e_k$ is a vector with 1 at position $k$ and 0 elsewhere.
 
-The total population size at time $t$ starting with type $j$ is then
-$$
-N(j,t) = e\tr n(j,t) = e\tr (MD)^t e_j.
-$$
+The total population size at time $t$ starting with type $k$ is then
+\begin{equation}
+N(k,t) = \sum_i{n_i(k,t)} = e\tr (MD)^t e_k.
+\end{equation}
 
-Now we can compare the population sizes based on what type they start
+Now we can compare the sizes of populations based on what type they started
 with:
-$$
+\begin{equation}
 \frac{N(j,t)}{N(k,t)} = \frac{e\tr (MD)^t e_j}{e\tr (MD)^t e_k}.
-$$
+\end{equation}
 
-To further evaluate, we write $MD$ as its Jordan canonical form, 
-$$
+To further evaluate, we write $MD$ in its Jordan canonical form 
+\begin{equation}
 A = V \Lambda U\tr,
-$$
+\end{equation}
 where $V$ is the matrix of right column eigenvectors of $MD$,
@@
 $V U\tr = U\tr V = I$, 
-and $\Lambda$ is the diagonal matrix of
-eigenvalues of $A$ (for a nongeneric set of matrices $M$, the
-geometric and algebraic multiplicities of the eigenvalues of $MD$
-differ, and $\Lambda$ will not be a diagonal matrix, a case we can
-ignore).
+and $\Lambda$ is the diagonal matrix of eigenvalues of $A$ 
+(for a non-generic set of matrices $M$, the geometric and algebraic multiplicities of the eigenvalues of $MD$ differ, and $\Lambda$ will not be a diagonal matrix, a case we can ignore).
 
 Hence,
-\begin{multline*}
-N(i, t) = 
-e\tr n(t) = \\
-e\tr (MD)^t e_i =
-e\tr  (V \Lambda U\tr)^t e_i = \\
-e\tr V \Lambda^t U\tr e_i.
-\end{multline*}
+\begin{multline}
+N(k, t) = 
+e\tr (MD)^t e_k =
+e\tr  (V \Lambda U\tr)^t e_k = \\
+e\tr V \Lambda^t U\tr e_k.
+\end{multline}
 
@@
 the spectral radius of $MD$:
-\begin{multline*}
+\begin{multline}
 \frac{N(j,t)}{N(k,t)} =
 \frac{ e\tr V \Lambda^t U\tr e_j}{ e\tr V \Lambda^t U\tr e_k} = \\
-\frac{ e\tr V \diag{1, (\frac{\lambda_2}{\lambda_1})^t, \cdots} U\tr e_j} { e\tr V \diag{1,(\frac{\lambda_2}{\lambda_1})^t, \cdots} U\tr e_k}.
-\end{multline*}
+\frac{ e\tr V \diag{1, \Big(\frac{\lambda_2}{\lambda_1}\Big)^t, \cdots} U\tr e_j} { e\tr V \diag{1,\Big(\frac{\lambda_2}{\lambda_1}\Big)^t, \cdots} U\tr e_k}.
+\end{multline}
 
-Now take the long term limit.
-By assumption $MD$ is irreducible,
-hence $\lambda_i < \lambda_1$ for all $i > 1$.
-Hence $\lim_{t \goesto \infty} \lambda_i / \lambda_1 = 0$ for all $i > 1$.
-Thus:
-\begin{multline*}
+Now take the long term limit $t \goesto \infty$.
+By assumption, $MD$ is irreducible, so $\lambda_i < \lambda_1$ for all $i > 1$.
+Therefore,
+\begin{equation}
+\lim_{t \goesto \infty} \Big( \frac{\lambda_k}{\lambda_1} \Big)^t = 0
+\end{equation}
+for all $k > 1$, and
+\begin{equation}
 \lim_{t \goesto \infty}\frac{N(j,t)}{N(k,t)} =
-\lim_{t \goesto \infty} \frac{ e\tr V \Lambda^t U\tr e_j}{ e\tr V \Lambda^t U\tr e_k} = \\
-\lim_{t \goesto \infty} \frac{ e\tr V diag {1, (\dspfrac{\lambda_2}{\lambda_1})^t, \cdots} U\tr e_h} { e\tr V diag {1,(\dspfrac{\lambda_2}{\lambda_1})^t, \cdots} U\tr e_k} = \\
-\frac{e\tr V (e_1 e_1\tr) U\tr e_j} { e\tr V (e_1 e_1\tr) U\tr e_k} = \\
+\frac{e\tr V (e_1 e_1\tr) U\tr e_j} { e\tr V (e_1 e_1\tr) U\tr e_k} = 
 \frac{ e\tr v u_j} { e\tr v  u_k} =
 \frac{u_j} {u_k}.
-\end{multline*}
+\end{equation}
 
-The vector $u$ is the left \emph{Perron} eigenvector of $MD$, and $u_i$ is element $i$ of $u$.
-This is why the value $u_i$ can be interpreted as the \emph{reproductive value} of type $i$, as it is a weight for the size of the population generated by a single individual of type $i$.
+The vector $u$ is the left \emph{Perron} eigenvector of $MD$, and $u_k$ is element $k$ of $u$.
+This is why the value $u_k$ can be interpreted as the \emph{reproductive value} of type $i$: it is a weight for the size of the population generated by a single individual of type $k$.
 
 If we begin with a population at the equilibrium distribution $v$,
-and ask what fraction of long-term descendants descended from type $i$ at that time,
-we weight $v_i$ by the reproductive value $u_i$, to get $u_i v_i$.
-We note that $\{u_i v_i\}_{i}$ is a normalized probability distribution, since
-$$
-\sum_i {u_i v_i} = u\tr v = 1.
-$$ 
-\citet{Hermisson2002} called the $u_i v_i$ the \emph{ancestor} or \emph{ancestral distribution}.
+and ask what fraction of long-term descendants descended from type $k$ at that time,
+we weight the equilibrium frequency $v_k$ by the reproductive value $u_k$, to get $u_k v_k$.
+$\{u_k v_k\}_{Ik}$ is a probability distribution, since
+\begin{equation}
+\sum_k {u_k v_k} = u\tr v = 1.
+\end{equation}
+\citet{Hermisson2002} called this distribution the \emph{ancestor} or \emph{ancestral distribution}.
 

### 2017-11-01 ms.tex
fix hawaii typo; add Holsinger1986 ref

@@
 \affil[a]{Department of Biology, Stanford University, Stanford, CA}
-\affil[b]{Information and Computer Sciences, University of Hawai`I at M{\=a}noa, Honolulu, HI}
+\affil[b]{Information and Computer Sciences, University of Hawai`i at M{\=a}noa, Honolulu, HI}
 
@@
 (i) effectively infinite population size, (ii) constant-viability selection, (iii) a population at an equilibrium, and (iv) \emph{linear variation} -- the equal scaling of transition probabilities by the modifier.
-A departure from the latter assumption occurs if two variation producing processes interact~\citep{Altenberg2012} [REF: Holsinger?], or if there are biases in the mutation or migration process, such that \ldots{}
+A departure from the latter assumption occurs if two variation producing processes interact~\citep{Feldman1980,Altenberg2012}.
+Departures from the \emph{reduction principle} have also been demonstrated when conditions (i)-(iii) are not met, see for example~\citet{Holsinger1986} and references therein.
 

### 2017-11-01 ms.tex
cite grafen and fisher; reformat appendix C

@@
 \newtheorem*{lemma}{Lemma}
+\newtheorem*{remark}{Remark}
 \newtheorem*{mfp}{Theorem: Mean Fitness Principle}
@@
 \newtheorem*{caswell}{Lemma: Caswell's Formula}
+\newtheorem*{frv}{Fisher's reproductive values}
 \renewcommand\qedsymbol{} % no square at end of proof
@@
 \subsection*{Appendix C}\label{sec:AppC}
+\begin{frv}
+Let $M$ be a stochastic matrix and $D$ be a diagonal matrix with positive diagonal elements. 
+The elements of the left \emph{Perron} eigenvector $u$ of the matrix $MD$ can be regarded as \emph{Fisher's reproductive values}~\citep{Fisher1930,Grafen2006}.
+\end{frv}
 
-Let $M$ be a stochastic matrix and $D$ be a diagonal matrix with positive diagonal elements. 
-The elements of the left \emph{Perron} eigenvector $u$ of the matrix $MD$ can be regarded as \textbf{Fisher's reproductive values}.
-These values can be understood as follows.
-
+\emph{Fisher's reproductive values} can be understood as follows.
 Let us look at the dynamics not of frequencies but of absolute population sizes such that the vector of the number of individuals of each type at time $t$ is $n(t)$ and $f_k(t) = n_k(t) / \sum_i{n_i(t)}$.
@@
 
-\textbf{TODO} cite \citep{Grafen2006}?
-
 % Acknowledgements

### 2017-11-02 ms.tex
two mutation models

@@
 \usepackage[auth-sc]{authblk}
+\usepackage{titlesec}
+
+\titleformat{\subsection}[runin]
+  {\bf}{\thesection}{1em}{}
 
@@
 \citet{Ram2012} stated that their result represents a departure from the \emph{reduction principle}, but did not explain this departure.
-Moreover, their analysis was limited to low mutation rates as it neglected double-mutations (i.e. assuming $U<1/\sqrt{N}$ where $U$ is the genome-wide mutation rate and $N$ is the population size).
+Their analysis was specific to a model that classified individuals by the number of mutant genes in their genotype, similar to~\citet{Kimura1966} and~\citet{Haigh1978}.
 
@@
 % Model
-\section*{Model}
+\section*{The Models}
 
+\subsection*{General model.}
 Consider a large haploid population with an arbitrary set of types $A_1, A_2, ...$.
@@
 The types $A_k$ can be encoded by a single or multiple genetic loci or non-genetic traits.
-Importantly, the type is transmitted from a single parent to the offspring, and transmission infidelity causes type transitions.
-For example, genotypes are transmitted by genetic inheritance, and genetic infidelity, i.e.
-mutation, generates new genotypes.
-More formally, type transmission is vertical and uni-parental, independent of the frequencies of the other types, and is fully described by the matrix $M$.
-Therefore, this model precludes recombination, social learning, sexual outcrossing, and any horizontal or oblique transmission~\cite[pg.~54]{Cavalli-Sforza1981}.
+Importantly, type transmission is vertical and uni-parental (the type is transmitted from a single parent to the offspring), independent of the frequencies of the other types, and is fully described by the matrix $M$.
+Specifically, transmission infidelity causes type transitions.
+Therefore, this model precludes processes such as recombination, social learning, sexual outcrossing, and any horizontal or oblique transmission, as these processes are frequency-dependent~\cite[pg.~54]{Cavalli-Sforza1981}.
 
@@
 The case of a scalar in covered by the \emph{reduction principle}, see~\citet{Altenberg2017}.
+
+In the following section we present three particular examples of this model for the modification of mutation, migration, and learning.
+
+\subsection*{Mutation modification model 1.}
+Here we consider a large population of haploids and a trait determined by a major locus with $n$ possible alleles $A_1, A_2, ..., A_n$.
+Linked to to this locus is a modifier locus that determines the mutation rates $\mu_k$ of individuals with allele $A_k$.
+Specifically, with probability $1-mu_k$, the allele $A_k$ does not mutate, and with probability $\frac{\mu_k}{n-1}$, the allele $A_k$ mutates to $A_j$ for any $j \ne k$.
+This is an extension of a model presented by~\citet{Altenberg2017}, allowing for the mutation rate of an individual $\mu_k$ to depend on the identity of the allele $A_k$.
+
+Let the frequency of $A_k$ in the present generation be $f_k$ such that $f_k \ge 0$, $\sum_k{f_k}=1$, $\vec{f}=(f_1, \ldots, f_n)\tr$.
+Then after selection and mutation $\vec{f'}$ in the next generation is given by
+\begin{equation}
+f_k' = (1 - \mu_k) \frac{w_k}{\bar{w}} f_k + \frac{1}{n-1} \sum_{k \ne j}{\mu_j \frac{w_j}{\bar{w}} x_j},
+\label{eq:mutation_model_1}
+\end{equation}
+for all $k=1,2,\ldots,n$, where $\bar{w} = \sum_{k=1}^{n}{w_k f_k}$ is the population mean fitness.
+
+This model can be rephrased in terms of the general model: 
+Eq.~\ref{eq:mutation_model_1} is equivalent to Eq.~\ref{eq:model} if we define
+\begin{equation} 
+C_k = \mu_k, \quad
+D_k = w_k, \quad
+M_{i,j} = \begin{cases}
+\frac{1}{n-1}, & i \ne j \\
+0, & i=j
+\end{cases}.
+\end{equation}
+
+\subsection*{Mutation modification model 2.}
+
+Here, we consider a large population of haploids in which individuals with genotype $A_k$ are characterized by the number of deleterious or mutant alleles in their genotype, $k$.
+A modifier locus then determines the probability that a mutation occurs in individuals with $k$ mutant alleles, $m_k$.
+In this case, if we assume a low mutation rate $m_k$ such that two mutations are unlikely to occur in the same individual in one generation.
+When a mutation occurs it is \emph{deleterious} with probability $\delta$, generating a new mutant allele and moving the individual from $A_k$ to $A_{k+1}$, or it is beneficial with probability $\beta$, moving the individual from $A_k$ to $A_{k-1}$.
+Note that such beneficial mutations can be either compensatory or back-mutations, and that mutations are neutral with probability $1-\delta-\beta$.
+This model has been presented and analyzed by~\citet{Ram2012}.
+
+Let the frequency of $A_k$ in the present generation be $f_k$ such that $f_k \ge 0$, $\sum_k{f_k}=1$, $\vec{f}=(f_1, \ldots, f_n)\tr$.
+Then after selection and mutation $\vec{f'}$ in the next generation is given by
+\begin{equation}
+f_0' = (1 - \delta m_0) \frac{w_0}{\bar{w}} f_0 + \beta m_{1} \frac{w_{1}}{\bar{w}} f_{1},
+\end{equation}
+and
+\begin{multline}
+f_k' = (1 - m_k(\beta+\delta)) \frac{w_k}{\bar{w}} f_k + \\
+\delta m_{k-1} \frac{w_{k-1}}{\bar{w}} f_{k-1} + \beta m_{k+1} \frac{w_{k+1}}{\bar{w}} f_{k+1},
+\label{eq:mutation_model_2}
+\end{multline}
+for all $k=1,2,\ldots$, where $w_k$ is the fitness of an individual with $k$ mutant alleles and $\bar{w} = \sum_{k \ge 0}{w_k f_k}$ is the population mean fitness.
+
+To rephrase this model in the terms of the general model, we define
+
+\begin{align}
+C_k = m_k, \quad
+D_k = w_k, \quad
+M_{i,j} = \begin{cases}
+\beta, & i = j-1 \\
+1-\beta-\delta, & i = j \\
+\delta, & i = j+1 \\
+1-\delta, & i=j=0 \\
+0, & \text{otherwise}
+\end{cases}
+\end{align}
+so that Eq.~\ref{eq:mutation_model_2} is equivalent to Eq.~\ref{eq:model}.
+
+\subsection*{Migration modification model.}
+
+\textbf{TODO}
 
@@
 Eq.~\ref{eq:theorem} is a generalization of a result from \citet[Eq.~4]{Ram2012}.
-Translated to the notation presented here, \citeauthor{Ram2012} modeled the accumulation of deleterious mutations in a population at the mutation-selection balance, assuming no more than one mutation occurs per individual per generation:
-\begin{equation}
-M_{i,j} = \begin{cases}
-\beta \mu, & i = j-1 \\
-(1-\beta-\delta) \mu, & i = j \\
-\delta \mu, & i = j+1 \\
-(1-\delta \mu), & i=j=0 \\
-0, & \text{otherwise}
-\end{cases}
-\end{equation}
+Translated to the notation presented here, \citeauthor{Ram2012} modeled the accumulation of deleterious mutations in a population at the mutation-selection balance, assuming no more than one mutation occurs per individual per generation...
+\textbf{TODO} continue this...
 
@@
 
-\textbf{TODO} migration model
-\textbf{TODO} cultural evolution model?
-
 % Conclusions
 \subsection*{Conclusions}
+
+\textbf{TODO}
 

### 2017-11-02 ms.tex
migration model + MWF comments

@@
 % Title page
-\title{Necessity is the mother of invention}
+\title{Necessity is the mother of genetic invention}
 
@@
 \begin{abstract}
-Generation of variation is considered detrimental in well-adapted
-populations evolving under constant selection.
-Therefore, the rate at which variation is generated through processes such as mutation,
-migration, and innovation is reduced by natural selection.
+Generation of variation is considered detrimental in well-adapted populations evolving under constant selection.
+Therefore, in a constant environment, genetic modifiers that reduce the rate at which variation is generated by processes such as mutation and migration, succeed.
 However, departures from this \emph{reduction principle} have been demonstrated.
-Here we analyze a general model of evolution under constant selection where the rate at which variation is generated can depend on the individual fitness.
+Here we analyze a general model of evolution under constant selection where the rate at which variation is generated can depend on the fitness of the individual.
 We find a general and simple result:
-if individuals with below-average fitness increase the rate at which variation is generated,
-then the population mean fitness increases.
-This principle applies to phenomena such as stress-induced mutagenesis,
-condition-dependent dispersal and outcrossing,
-and even to the famous proverb: \emph{"Necessity is the mother of invention"}.
+if individuals with below-average fitness increase the rate at which variation is generated, then the population mean fitness increases.
+This principle applies to phenomena such as stress-induced mutagenesis, condition-dependent dispersal and outcrossing,
+and even to the famous proverb: \emph{"Necessity is the mother of genetic invention"}.
 \end{abstract}
@@
 According to the \emph{reduction principle}, in populations that evolve near an equilibrium between natural selection and a process that generates variation (i.e.
-mutation, migration, or recombination), selection favors modifiers that decrease the rate of variation generation.
+mutation, migration, or recombination), selection favors neutral modifiers that decrease the rate at which variation is generated.
 
@@
 
-The assumptions and conditions for the validity of the \emph{unified reduction principle} are~\citep{Altenberg2017}:
+The assumptions and conditions for the validity of the \emph{unified reduction principle} by neutral genetic modifiers are~\citep{Altenberg2017}:
 (i) effectively infinite population size, (ii) constant-viability selection, (iii) a population at an equilibrium, and (iv) \emph{linear variation} -- the equal scaling of transition probabilities by the modifier.
@@
 \citet{Ram2012} demonstrated that modifiers that increase the mutation rate of individuals with below-average fitness actually \textbf{increase} the population mean fitness, rather than decrease it.
-Their analysis included an infinite population size and constant-viability selection, such that the only departure from the \emph{reduction principle} assumptions is due to the differential effect the modifier has on different genotypes.
+Their analysis assumed an infinite population size and constant-viability selection, such that the only departure from the \emph{reduction principle} assumptions is due to the differential effect the modifier has on different genotypes.
 \citet{Ram2012} stated that their result represents a departure from the \emph{reduction principle}, but did not explain this departure.
-Their analysis was specific to a model that classified individuals by the number of mutant genes in their genotype, similar to~\citet{Kimura1966} and~\citet{Haigh1978}.
+Their analysis was specific to a model that classified individuals by the number of mutant genes in their genotype, similar to those of~\citet{Kimura1966} and~\citet{Haigh1978}.
 
-Here, we present a general evolutionary model where the rate at which variation is generated depend on the type of individual, and specifically on the its fitness.
-Our result shows that low fitness individuals should produce more variation then high fitness individuals.
+Here, we present a general evolutionary model where the rate at which variation is generated depends on the type of individual, and specifically on the its fitness.
+Our result shows that low fitness individuals should produce more variation than high fitness individuals.
 
@@
 \subsection*{General model.}
-Consider a large haploid population with an arbitrary set of types $A_1, A_2, ...$.
+Consider a large haploid population with an arbitrary (possibly infinite) set of types $A_1, A_2, \ldots$.
 The change in the frequencies of these types is described by the transformation $f \to f'$: 
@@
 \end{equation}
-where $f_k$ is the frequency of type $A_k$, such that $f$ is a stochastic vector ($\sum_k{f_k} = 1$);
+where $f_k$ is the frequency of type $A_k$ ($f=(f_1, f_2, \ldots)$ is a stochastic vector: $\sum_k{f_k} = 1$);
 $C_k$ is the probability that an individual of type $A_k$ will transition to another type ($C$ is a diagonal non-negative matrix with diagonal elements $C_k$);
-$M_{i,j}$ is the transition probability from type $A_j$ to type $A_i$ given that an individual of type $A_j$ is transitioning ($M$ is a stochastic matrix: $\sum_j {M_{i,j}} = 1$ for any column \emph{j});
-$D_k$ is the relative fitness of type $A_k$ ($D$ is a diagonal matrix with diagonal elements $0 < D_k \le 1$; fitness values are relative to the fitness of the fittest type); 
+$M_{i,j}$ is the transition probability from type $A_j$ to type $A_i$ given that an individual of type $A_j$ is transitioning ($M$ is a column-stochastic matrix: $\sum_j {M_{i,j}} = 1$ for any column \emph{j});
+$D_k$ is the relative fitness of type $A_k$ ($D$ is a diagonal matrix with diagonal elements $D_k$, where we can assume $0 < D_k \le 1$, i.e., fitness values are relative to the fitness of the fittest type);
 $I$ is the identity matrix;
-and $\bar{\omega}$ is a normalizing factor such that $\sum_k{f'_k}=1$, which turns out to be the population mean fitness (see~\nameref{sec:AppA}).
+and $\bar{\omega}$ is a normalizing factor such that $\sum_k{f'_k}=1$ and is equal to the population mean fitness (see~\nameref{sec:AppA}).
 
-The types $A_k$ can be encoded by a single or multiple genetic loci or non-genetic traits.
+The types $A_k$ can represent a single or multiple genetic loci or non-genetic traits.
 Importantly, type transmission is vertical and uni-parental (the type is transmitted from a single parent to the offspring), independent of the frequencies of the other types, and is fully described by the matrix $M$.
@@
 We focus on a modifier that has no direct effect on fitness and whose sole function is to determine the rate of transition out of the different types ($C_k$).
-Importantly, this modifier can determine different rates for different types, therefore $C$ is a diagonal matrix rather than a scalar.
-The case of a scalar in covered by the \emph{reduction principle}, see~\citet{Altenberg2017}.
+Importantly, this modifier can determine different rates for different types; that is, $C_j \ne C_i$ for some $i,j$.
+The case $C_j = C_j$ for all $i,j$ is covered by the \emph{reduction principle}, see~\citet{Altenberg2017}.
 
@@
 \subsection*{Mutation modification model 1.}
+
 Here we consider a large population of haploids and a trait determined by a major locus with $n$ possible alleles $A_1, A_2, ..., A_n$.
 Linked to to this locus is a modifier locus that determines the mutation rates $\mu_k$ of individuals with allele $A_k$.
-Specifically, with probability $1-mu_k$, the allele $A_k$ does not mutate, and with probability $\frac{\mu_k}{n-1}$, the allele $A_k$ mutates to $A_j$ for any $j \ne k$.
-This is an extension of a model presented by~\citet{Altenberg2017}, allowing for the mutation rate of an individual $\mu_k$ to depend on the identity of the allele $A_k$.
+Specifically, with probability $1-\mu_k$, the allele $A_k$ does not mutate, and with probability $\frac{\mu_k}{n-1}$, the allele $A_k$ mutates to $A_j$ for any $j \ne k$.
+This is an extension of a model studied by~\citet{Altenberg2017}, that allows for the mutation rate of an individual, $\mu_k$, to depend on the identity of the allele $A_k$.
 
-Let the frequency of $A_k$ in the present generation be $f_k$ such that $f_k \ge 0$, $\sum_k{f_k}=1$, $\vec{f}=(f_1, \ldots, f_n)\tr$.
-Then after selection and mutation $\vec{f'}$ in the next generation is given by
+Let the frequency of $A_k$ in the present generation be $f_k$ with $f_k \ge 0$ and $\sum_k{f_k}=1$.
+Then after selection and mutation, $f'_k$ in the next generation is given by
 \begin{equation}
@@
 
-Here, we consider a large population of haploids in which individuals with genotype $A_k$ are characterized by the number of deleterious or mutant alleles in their genotype, $k$.
-A modifier locus then determines the probability that a mutation occurs in individuals with $k$ mutant alleles, $m_k$.
-In this case, if we assume a low mutation rate $m_k$ such that two mutations are unlikely to occur in the same individual in one generation.
+Here, we consider a large population of haploids in which individuals with genotype $A_k$ are characterized by the number $k$ of deleterious or mutant alleles in their genotype.
+A modifier locus then determines the probability $m_k$ that a mutation occurs in individuals with $k$ mutant alleles.
+In this case, we assume a low mutation rate $m_k$ such that two mutations are unlikely to occur in the same individual in one generation.
 When a mutation occurs it is \emph{deleterious} with probability $\delta$, generating a new mutant allele and moving the individual from $A_k$ to $A_{k+1}$, or it is beneficial with probability $\beta$, moving the individual from $A_k$ to $A_{k-1}$.
 Note that such beneficial mutations can be either compensatory or back-mutations, and that mutations are neutral with probability $1-\delta-\beta$.
-This model has been presented and analyzed by~\citet{Ram2012}.
+This model has been analyzed by~\citet{Ram2012}.
 
-Let the frequency of $A_k$ in the present generation be $f_k$ such that $f_k \ge 0$, $\sum_k{f_k}=1$, $\vec{f}=(f_1, \ldots, f_n)\tr$.
+Let the frequency of $A_k$ in the present generation be $f_k$ such that $f_k \ge 0$ and $\sum_k{f_k}=1$.
 Then after selection and mutation $\vec{f'}$ in the next generation is given by
@@
 \begin{multline}
-f_k' = (1 - m_k(\beta+\delta)) \frac{w_k}{\bar{w}} f_k + \\
+f_k' = (1 - m_k(\delta+\beta)) \frac{w_k}{\bar{w}} f_k + \\
 \delta m_{k-1} \frac{w_{k-1}}{\bar{w}} f_{k-1} + \beta m_{k+1} \frac{w_{k+1}}{\bar{w}} f_{k+1},
@@
 
-\begin{align}
+\begin{equation}
 C_k = m_k, \quad
-D_k = w_k, \quad
-M_{i,j} = \begin{cases}
-\beta, & i = j-1 \\
-1-\beta-\delta, & i = j \\
-\delta, & i = j+1 \\
-1-\delta, & i=j=0 \\
-0, & \text{otherwise}
-\end{cases}
-\end{align}
+D_k = w_k,
+%M_{i,j} = \begin{cases}
+%\beta, & i = j-1 \\
+%1-\beta-\delta, & i = j \\
+%\delta, & i = j+1 \\
+%1-\delta, & i=j=0 \\
+%0, & \text{otherwise}
+%\end{cases}
+\end{equation}
+and
+\begin{equation}
+M_{i,j} = \begin{bmatrix}
+1-\delta & \beta & 0 &  \cdots \\
+\delta & 1-\delta-\beta & \beta &  \ddots \\
+0 & \delta & 1-\delta-\beta & \ddots \\
+\vdots & \ddots & \ddots & \ddots 
+\end{bmatrix}
+\end{equation}
 so that Eq.~\ref{eq:mutation_model_2} is equivalent to Eq.~\ref{eq:model}.
@@
 
-\textbf{TODO}
+Here we assume a large population of haploids that occupies two demes and a modifier locus that determine the migration rates between the demes, $m_k$. 
+Let the fitness of genotypes $A_k$ be $w_k$ in deme 1 and $u_k$ in deme 2 for $k=1,2,\ldots,n$.
+Let the frequencies of $A_k$ in deme 1 be $x_k$ and in deme 2 $y_k$ for $k=1,2,\ldots,n$ with $x_k \ge 0$, $y_k \ge 0$, $\sum_{k=1}^{n}{x_k} = 1$. and $\sum_{k=1}^{n}{y_k} = 1$.
+
+Then following selection and migration, after one generation the new frequencies $x_k'$ and $y_k'$ are given by
+\begin{align}
+x_k' = (1-m_k) \frac{w_k}{\bar{w}} x_k + m_k \frac{u_k}{\bar{u}} y_k \\
+y_k' = m_k \frac{w_k}{\bar{w}} x_k + (1 - m_k) \frac{u_k}{\bar{u}} y_k
+\end{align}
+for $k=1,2,\ldots,n$, where $\bar{w} = \sum_{k \ge 0}{w_k x_k}$ and $\bar{u} = \sum_{k \ge 0}{u_k y_k}$ is the population mean fitness in deme 1 and 2, respectively.
+This is an extension of a model presented by~\citet{Altenberg2017}, allowing for the migration rate of an individual, $m_k$, to depend on the identity of the allele $A_k$.
+
+To rephrase this model in terms of the general model, we define
+\begin{equation}
+\begin{aligned}
+f = (x_1, \ldots, x_n, y_1, \ldots, y_n)\tr, \\
+C = diag[m_1, \ldots, m_n, m_1, \ldots, m_n], \\
+D = diag[w_1, \ldots, w_n, u_1, \ldots, u_n], \\
+M = \begin{bmatrix}
+0 & I_n \\
+I_n & 0
+\end{bmatrix}
+\end{aligned}
+\end{equation}
+where $I_n$ is the n-by-n identity matrix.
 
@@
 % Mean fitness principle
-The equilibrium of the frequency vector $f$ in Eq.~\ref{eq:model} is given by the equilibrium frequency vector $v$ and the stable population mean fitness
-$\bar{\omega}^*$, namely
+With the equilibrium of the frequency vector $f$ in Eq.~\ref{eq:model} as $v$ and the stable population mean fitness as $\bar{\omega}^*$, then
 \begin{equation}\label{eq:model_equilibrium}
@@
 \end{equation}
-and therefore increased transition out of type $k$ will increase the stable population mean fitness if and only if the fitness of type $k$ is below the stable population mean fitness.
+and therefore increased transition from type $k$ will increase the stable population mean fitness if and only if the fitness of type $k$ is below the stable population mean fitness.
 \end{mfp}
@@
 \frac{\partial \bar{\omega}^*}{\partial C_k} = 
-\frac{u_k v_k}{C_k} (\bar{\omega}^* - D_k),
+\frac{u_k v_k}{C_k} (\bar{\omega}^* - D_k).
+\end{equation}
+Finally, note that $u_k, v_k, C_k \ge 0$ and therefore
+\begin{equation}
+sign \frac{\partial \bar{\omega}^*}{\partial C_k} = 
+sign (\bar{\omega}^* - D_k),
 \end{equation}
@@
 Therefore, we can expect modifiers that reduce $\alpha$ to be favored by natural selection.
-Indeed, setting $C = \alpha I$ in Eq.~\ref{eq:model_equilibrium} and proceeding as in the proof, we get
+Indeed, setting $C = \alpha I$ in Eq.~\ref{eq:model_equilibrium} and proceeding as in the above proof, we get
 $$

### 2017-11-03 ms.tex
finished with MWF comments of Nov 2

@@
 Here, $u_k$ can be regarded as the \emph{reproductive value} of type $k$ (see~\nameref{sec:AppC}), which gives the relative contribution of type $k$ to the long-term population.
-Subsequently, $u_k v_k$ is the \emph{ancestor frequency} \citep{Hermisson2002} of type $k$, which is the fraction of the equilibrium population descended from type $k$.
+Consequently, $u_k v_k$ is the \emph{ancestor frequency} \citep{Hermisson2002} of type $k$, which is the fraction of the equilibrium population descended from type $k$.
 The sum $\sum_i{u_i M_{i,k} v_k}$ can be similarly interpreted as the fraction of the equilibrium population descended from individuals that transitioned from type $k$ to another type (via the $k$ column of the transition matrix $M$).
@@
 \section*{Appendices}
-% Appendix A
+% Appendix A: mean fitness
 \subsection*{Appendix A}\label{sec:AppA}
+\textbf{TODO} reference to Fisher 1930 or Fisher 1958 (Caswell 1978 cites the latter in pg. 217 just above eq. 4).
 
@@
 
-% Appendix B
+% Appendix B: Caswell
 \subsection*{Appendix B}\label{sec:AppB}
 
-\citet{Caswell1978} presented a \emph{formula for the sensitivity of the population growth rate to changes in life history parameters}.
+\citet{Caswell1978} gave a \emph{formula for the sensitivity of the population growth rate to changes in life history parameters}.
 In this formula, the \emph{population growth rate} is the leading eigenvalue of the population transformation matrix $T$, the \emph{life history parameters} are elements of $T$, and the \emph{sensitivity} is the derivative of the former with respect to the latter.
-This is a useful formula~\citep{Hermisson2002,Ram2012}, and therefore we reproduce it here.
+This is a useful formula~\citep{Caswell1978,Hermisson2002,Ram2012}, and therefore we reproduce it here.
 
 \begin{caswell}
-Let $T$ be a matrix with eigenvalue $\lambda$ and left and right eigenvectors $u$ and $v$ such that $\sum{v_k}=1$ and $u\tr v = \sum{u_k v_k} = 1$.
+Let $T$ be a non-negative matrix with leading eigenvalue $\lambda$ and left and right eigenvectors $u$ and $v$ such that $\sum{v_k}=1$ and $u\tr v = \sum{u_k v_k} = 1$.
 Then the sensitivity of $\lambda$ to changes in any element $t$ of the matrix $T$ is
@@
 
-% Appendix C
+% Appendix C: Fisher's reproductive value
 \subsection*{Appendix C}\label{sec:AppC}
@@
 \emph{Fisher's reproductive values} can be understood as follows.
-Let us look at the dynamics not of frequencies but of absolute population sizes such that the vector of the number of individuals of each type at time $t$ is $n(t)$ and $f_k(t) = n_k(t) / \sum_i{n_i(t)}$.
+Consider the dynamics not of frequencies but of absolute population sizes such that the vector of the number of individuals of each type at time $t$ is $n(t)$ and $f_k(t) = n_k(t) / \sum_i{n_i(t)}$.
 The dynamics are
@@
 
-To further evaluate, we write $MD$ in its Jordan canonical form 
+Now write $MD$ in its Jordan canonical form 
 \begin{equation}
@@
 
-Now take the long term limit $t \goesto \infty$.
+Now take the long termlimit $t \goesto \infty$.
 By assumption, $MD$ is irreducible, so $\lambda_i < \lambda_1$ for all $i > 1$.
@@
 The vector $u$ is the left \emph{Perron} eigenvector of $MD$, and $u_k$ is element $k$ of $u$.
-This is why the value $u_k$ can be interpreted as the \emph{reproductive value} of type $i$: it is a weight for the size of the population generated by a single individual of type $k$.
+This is why the value $u_k$ can be interpreted as the \emph{reproductive value} of type $i$: it is a weighting for the size of the population generated by a single individual of type $k$.
 
@@
 we weight the equilibrium frequency $v_k$ by the reproductive value $u_k$, to get $u_k v_k$.
-$\{u_k v_k\}_{Ik}$ is a probability distribution, since
+$\{u_k v_k\}_{k}$ is a probability distribution, since
 \begin{equation}

### 2017-11-03 get_bibtex.py
copy required citations from bibtex library to ms.bib

@@ -1,0 +1,35 @@
+from bibtexparser import load as load_bib
+from bibtexparser.bwriter import BibTexWriter
+from bibtexparser.bibdatabase import BibDatabase
+import click
+
+@click.command()
+@click.argument('keys_filename', type=click.Path(exists=True, file_okay=True, dir_okay=False, readable=True))
+@click.argument('bibtex_filename', type=click.Path(exists=True, file_okay=True, dir_okay=False, readable=True))
+@click.argument('output_filename', type=click.Path(file_okay=True, dir_okay=False, writable=True))
+@click.option('-v/-V', '--verbose/--no-verbose', default=False)
+def main(keys_filename, bibtex_filename, output_filename, verbose):
+	with open(keys_filename) as f:
+		citation_keys = [line.strip() for line in f.readlines()]
+	if verbose:
+		print("Read {} keys from {}".format(len(citation_keys), keys_filename))
+
+	with open(bibtex_filename) as f:
+		main_bib = load_bib(f)
+	if verbose:
+		print("Read {} entries from {}".format(len(main_bib.entries), bibtex_filename))
+
+	out_bib = BibDatabase()
+	for key in citation_keys:
+		e = main_bib.entries_dict[key]
+		out_bib.entries.append(e)
+	
+	if verbose:
+		print("Writing {} entries to {}".format(len(out_bib.entries), output_filename))
+	writer = BibTexWriter()
+	with open(output_filename, 'w') as f:
+	    f.write(writer.write(out_bib))
+
+
+if __name__ == '__main__':
+	main()

### 2017-11-03 ms.tex
threshold modifier

@@
-\documentclass[9pt, a4paper, twocolumn]{extarticle}   	% use "amsart" instead of "article" for AMSLaTeX format
-\usepackage[margin=1in]{geometry}                		% See geometry.pdf to learn the layout options. There are lots.
-\geometry{letterpaper}                   		% ... or a4paper or a5paper or ... 
-%\geometry{landscape}                		% Activate for rotated page geometry
-%\usepackage[parfill]{parskip}    		% Activate to begin paragraphs with an empty line rather than an indent
+\documentclass[9pt, a4paper, twocolumn]{extarticle}
+\usepackage[margin=1in]{geometry} % See geometry.pdf to learn the layout options.
+\geometry{a4paper}
+\usepackage[parfill]{parskip} % Activate to begin paragraphs with an empty line rather than an indent
 \usepackage{graphicx} % Use pdf, png, jpg, or eps§ with pdflatex; use eps in DVI mode
 % TeX will automatically convert eps --> pdf in pdflatex
+
 \usepackage{amssymb,amsmath,amsthm}
@@
 
-\date{\today} % Activate to display a given date or no date
+\date{\today}
 
@@
 \citet{Ram2012} stated that their result represents a departure from the \emph{reduction principle}, but did not explain this departure.
-Their analysis was specific to a model that classified individuals by the number of mutant genes in their genotype, similar to those of~\citet{Kimura1966} and~\citet{Haigh1978}.
+Their analysis was specific to a model that classified individuals by the number of mutant genes in their genotype, similar to those of~\citet{Kimura1966a} and~\citet{Haigh1978}.
 
@@
 
-\subsection*{Mutation modification model 1.}
+\subsection*{Mutation modification model 1.}\label{mut_model_1}
 
@@
 
-\subsection*{Mutation modification model 2.}
+\subsection*{Mutation modification model 2.}\label{mut_model_2}
 
@@
 
-\subsection*{Migration modification model.}
+\subsection*{Migration modification model.}\label{mig_model_2}
 
@@
 or in simpler terms,
-\begin{equation}
+\begin{equation}\label{eq:sign_theorem}
 sign\frac{\partial \bar{\omega}^*}{\partial C_k} = 
@@
 
+% Threshold rule
+
 We turn our attention to a modifier that concurrently controls the transition out of a subset of types, $K$.
 For example, \citet[Appendix~B]{Ram2012} considered individuals that are grouped by the number of their accumulated deleterious mutations, $k$, and a modifier that regulates the mutation rates in individuals with at least $\pi$ deleterious mutations.
-
 Similarly, if the transition rates are controlled by a threshold rule 
@@
 \end{equation}
-then the sensitivity of the stable population mean fitness to changes in the
-rate of transition out of types $K$ is
+where $\tau > 0$, then the sensitivity of the stable population mean fitness to changes in the
+rate of transition from types $K$ is
 \begin{equation}
@@
 u\tr \Big( \sum_{k \in \mathbb{K{}}}{\frac{\partial (I - C + MC)D}{\partial \tau}} \Big) v = \\
-\frac{1}{\tau} \sum_{k \in \mathbb{K}}{u_k v_k (\bar{\omega}^* - D_k)}.
+\frac{1}{\tau} \sum_{k \in \mathbb{K}}{u_k v_k (\bar{\omega}^* - D_k)},
 \end{aligned}
 \end{equation}
-
-\textbf{TODO} summarize result in words.
+which can be simplified to
+\begin{multline}
+sign \frac{\partial \bar{\omega}^*}{\partial \tau} = 
+sign \sum_{k \in \mathbb{K}}{u_k v_k (\bar{\omega}^* - D_k)} = \\ 
+sign \Big(\bar{\omega}^* - \frac{\sum_{k \in \mathbb{K}}{u_k v_k D_k}}{\sum_{k \in \mathbb{K}}{u_k v_k}}\Big), 
+\end{multline}
+a generalization of Eq.~\ref{eq:sign_theorem}.
+Therefore, increased transition from types $K$ will increase the stable population mean fitness if and only if the average fitness of individuals descended from types $K$ is below the stable population mean fitness.
 
@@
 \bibliographystyle{plainnat}
-\bibliography{/Users/yoavram/Documents/library}
+\bibliography{ms}
 

### 2017-11-03 ms.tex
discussion

@@
 
-\subsection*{Mutation modification model 1.}\label{mut_model_1}
+\subsection*{Mutation modification model 1.}\label{sec:mut_model_1}
 
@@
 
-\subsection*{Mutation modification model 2.}\label{mut_model_2}
+\subsection*{Mutation modification model 2.}\label{sec:mut_model_2}
 
@@
 
-\subsection*{Migration modification model.}\label{mig_model_2}
+\subsection*{Migration modification model.}\label{sec:mig_model_1}
 
@@
 which can be simplified to
-\begin{multline}
+\begin{multline}\label{eq:sign_theorem_threshold}
 sign \frac{\partial \bar{\omega}^*}{\partial \tau} = 
@@
 \section*{Discussion}
-\textbf{TODO}: Departure from the \emph{reduction principle}
+
+We have shown that under constant-viability selection and in an effectively infinite haploid populations at an equilibrium, the stable population mean fitness increases if below-fitness individuals increase the rate at which variation is generated.
+We have given several examples on variation-generating processes for which this principle applies -- mainly mutation and migration -- but our model is general enough to possibly apply to other processes as well.
+For example, the \emph{reduction principle} appears some ecological models of dispersal, and~\citet{Gueijman2013} have demonstrated that even in homogeneous environments, fitness-associated dispersal increases the population mean fitness and is favored by selection over uniform dispersal.
+Similarly, if the transmission fidelity of culturally-transmitted traits depends on the type of the type or fitness of the transmitting individual, we expect that our \emph{mean fitness principle} will hold.
 
 Eq.~\ref{eq:theorem} is a generalization of a result from \citet[Eq.~4]{Ram2012}.
-Translated to the notation presented here, \citeauthor{Ram2012} modeled the accumulation of deleterious mutations in a population at the mutation-selection balance, assuming no more than one mutation occurs per individual per generation...
-\textbf{TODO} continue this...
+\citeauthor{Ram2012} modeled the accumulation of mutant alleles in a population evolving at the mutation-selection balance (see \nameref{sec:mut_model_2}).
+Using Eq.~\ref{eq:Caswells_formula} in \nameref{sec:AppB} and a recursion on the ratios of the reproductive values \cite[see][eqs.~A5-6]{Ram2012}, they concluded that at the mutation-selection balance, if individuals with below-average fitness ($D_k < \bar{\omega}^*$) increase their mutation rate, then the population mean fitness will increase -- a result generalized by our \emph{mean fitness principle} in Eqs.~\ref{eq:sign_theorem} and~\ref{eq:sign_theorem_threshold}.
 
-Using Eq.~\ref{eq:Caswells_formula} in \nameref{sec:AppB} and a recursion on the ratios of the reproductive values \cite[see][eqs.~A5-6]{Ram2012}, they concluded that at the mutation-selection balance, if individuals with below-average fitness ($D_k < \bar{\omega}^*$) increase their mutation rate, then the population mean fitness will increase.
+Our analysis focuses on populations at equilibrium.
+Nevertheless, it has been demonstrated that a modifier which increases the mutation rate of maladapted individuals can be favored by selection~\citep{Ram2012} and reduce the adaptation rate~\citep{Ram2014}.
+Similar results have been demonstrated for a modifier that increases the recombination rate in maladapted individuals~\citep{Hadany2003a,Hadany2003b}.
+Importantly, there are empirical evidence both for stress-induced mutagenesis~\citep{Rosenberg2012} and for condition-dependent recombination and sex~\citep{Ram2016}.
 
@@
 
-\textbf{TODO}
+General examples of departures from the \emph{reduction principle} are rare~\citep{Altenberg2017}.
+However, it has been demonstrated that higher mutation, recombination, and migration rates are favored by selection in populations evolving under fluctuating selection; see~\citet{Carja2014} and references therein.
+Here we have provided another general example, which suggests that if maladapted or below-average individuals increase the rate at which variation is generated, the population mean fitness will increase.
 
@@
 \bibliographystyle{plainnat}
+%\bibliography{/Users/yoavram/Documents/library}
 \bibliography{ms}

### 2017-11-03 ms.tex
cite Fisher for reproductive value; fix section referencing - was messed by hyperref+titlesec so just switched subsections to paragraph

@@
 \usepackage{amssymb,amsmath,amsthm}
-\usepackage{longtable,booktabs}
+\usepackage{longtable}
+\usepackage[hyphens]{url}
 \PassOptionsToPackage{hyphens}{url} % url is loaded by hyperref
 \usepackage[unicode=true]{hyperref}
+%\usepackage{nameref} % included in hyperref
 \usepackage[auth-sc]{authblk}
-\usepackage{titlesec}
 
-\titleformat{\subsection}[runin]
-  {\bf}{\thesection}{1em}{}
+% messes up nameref
+%\usepackage{titlesec}
+%\titleformat{\subsection}[runin]
+%  {\bf}{\thesection}{1em}{}
 
@@
 
-\subsection*{General model.}
+\paragraph*{General model.}
 Consider a large haploid population with an arbitrary (possibly infinite) set of types $A_1, A_2, \ldots$.
@@
 
-\subsection*{Mutation modification model 1.}\label{sec:mut_model_1}
+\paragraph*{Mutation modification model 1.}\label{sec:mut_model_1}
 
@@
 
-\subsection*{Mutation modification model 2.}\label{sec:mut_model_2}
+\paragraph*{Mutation modification model 2.}\label{sec:mut_model_2}
 
@@
 
-\subsection*{Migration modification model.}\label{sec:mig_model_1}
+\paragraph*{Migration modification model.}\label{sec:mig_model_1}
 
@@
 \end{equation}
-Here, $u_k$ can be regarded as the \emph{reproductive value} of type $k$ (see~\nameref{sec:AppC}), which gives the relative contribution of type $k$ to the long-term population.
+Here, $u_k$ can be regarded as the \emph{reproductive value}~\citep[pg.~27]{Fisher1930} of type $k$ (see~\nameref{sec:AppC}), which gives the relative contribution of type $k$ to the long-term population.
 Consequently, $u_k v_k$ is the \emph{ancestor frequency} \citep{Hermisson2002} of type $k$, which is the fraction of the equilibrium population descended from type $k$.
@@
 \subsection*{Appendix A}\label{sec:AppA}
-\textbf{TODO} reference to Fisher 1930 or Fisher 1958 (Caswell 1978 cites the latter in pg. 217 just above eq. 4).
 
@@
 Let $M$ be a stochastic matrix and $D$ be a diagonal matrix with positive diagonal elements. 
-The elements of the left \emph{Perron} eigenvector $u$ of the matrix $MD$ can be regarded as \emph{Fisher's reproductive values}~\citep{Fisher1930,Grafen2006}.
+The elements of the left \emph{Perron} eigenvector $u$ of the matrix $MD$ can be regarded as \emph{Fisher's reproductive values}~\citep[pg.~27]{Fisher1930}
 \end{frv}
 
-\emph{Fisher's reproductive values} can be understood as follows.
+\emph{Fisher's reproductive values} can be understood as follows~\citep{Grafen2006}.
 Consider the dynamics not of frequencies but of absolute population sizes such that the vector of the number of individuals of each type at time $t$ is $n(t)$ and $f_k(t) = n_k(t) / \sum_i{n_i(t)}$.
@@
 
-Now take the long termlimit $t \goesto \infty$.
+Now take the limit $t \goesto \infty$.
 By assumption, $MD$ is irreducible, so $\lambda_i < \lambda_1$ for all $i > 1$.

### 2017-11-06 ms.tex
remove huge space at begining of paragraphs
first version of modifier invasion proof

@@
 \geometry{a4paper}
-\usepackage[parfill]{parskip} % Activate to begin paragraphs with an empty line rather than an indent
+%\usepackage[parfill]{parskip} % Activate to begin paragraphs with an empty line rather than an indent
 \usepackage{graphicx} % Use pdf, png, jpg, or eps§ with pdflatex; use eps in DVI mode
@@
 \newtheorem*{corollary}{Corollary}
-\newtheorem*{lemma}{Lemma}
+\newtheorem{lemma}{Lemma}
 \newtheorem*{remark}{Remark}
@@
 \newtheorem*{rvp}{Corollary: Reproductive Value Principle}
-\newtheorem*{caswell}{Lemma: Caswell's Formula}
 \newtheorem*{frv}{Fisher's reproductive values}
@@
 % Title page
-\title{Necessity is the mother of genetic invention}
+\title{Variation generation and mean fitness increase: \\ Necessity is the mother of genetic invention}
 
@@
 \begin{abstract}
-Generation of variation is considered detrimental in well-adapted populations evolving under constant selection.
-Therefore, in a constant environment, genetic modifiers that reduce the rate at which variation is generated by processes such as mutation and migration, succeed.
+Generation of variation may be detrimental in well-adapted populations evolving under constant selection.
+In a constant environment, genetic modifiers that reduce the rate at which variation is generated by processes such as mutation and migration, succeed.
 However, departures from this \emph{reduction principle} have been demonstrated.
@@
 This principle applies to phenomena such as stress-induced mutagenesis, condition-dependent dispersal and outcrossing,
-and even to the famous proverb: \emph{"Necessity is the mother of genetic invention"}.
+and even to the proverb: \emph{"Necessity is the mother of genetic invention"}.
 \end{abstract}
@@
 
-Another departure from the \emph{linear variation} assumption of the \emph{reduction principle} for mutation rates involves a mechanism commonly called \emph{stress-induced mutagenesis}, in which the mutation rate increases in individuals with poor fitness.
+Another departure from the \emph{linear variation} assumption of the \emph{reduction principle} for mutation rates involves \emph{stress-induced mutagenesis}, a mechanism in which the mutation rate increases in individuals with poor fitness.
+Empirical evidence suggests \emph{stress-induced mutagenesis} is common in bacteria and yeast, and may be prevalent in plants, flies, and human cancer cells~\citep{Rosenberg2012,Fitzgerald2017b}.
 \citet{Ram2012} demonstrated that modifiers that increase the mutation rate of individuals with below-average fitness actually \textbf{increase} the population mean fitness, rather than decrease it.
-Their analysis assumed an infinite population size and constant-viability selection, such that the only departure from the \emph{reduction principle} assumptions is due to the differential effect the modifier has on different genotypes.
+Their analysis assumed an infinite population size and constant-viability selection, where the only departure from the \emph{reduction principle} assumptions is due to the differential effect the modifier has on different genotypes.
+A similar result has been demonstrated for fitness-associated recombination~\citep{Hadany2003a} and for condition-dependent sexual reproduction~\citep{Hadany2007a}, and evidence suggests that both mechanisms are common in nature~\citep{Ram2016}.
+
 \citet{Ram2012} stated that their result represents a departure from the \emph{reduction principle}, but did not explain this departure.
-Their analysis was specific to a model that classified individuals by the number of mutant genes in their genotype, similar to those of~\citet{Kimura1966a} and~\citet{Haigh1978}.
+Their analysis was specific to a model that classified individuals by the number of mutant genes in their genotype, similar to models studied by~\citet{Kimura1966a} and~\citet{Haigh1978}.
 
 Here, we present a general evolutionary model where the rate at which variation is generated depends on the type of individual, and specifically on the its fitness.
-Our result shows that low fitness individuals should produce more variation than high fitness individuals.
+Our result shows that if below-average individuals produce more variation than above-average individuals, then the population mean fitness increases.
 
@@
 
-\paragraph*{General model.}
+\paragraph*{General model.}\label{sec:general_model}
 Consider a large haploid population with an arbitrary (possibly infinite) set of types $A_1, A_2, \ldots$.
@@
 \begin{equation}\label{eq:model}
-\bar{\omega} f' = (I-C+MC)D f,
+\bar{w} f' = (I-C+MC)D f,
 \end{equation}
@@
 $C_k$ is the probability that an individual of type $A_k$ will transition to another type ($C$ is a diagonal non-negative matrix with diagonal elements $C_k$);
-$M_{i,j}$ is the transition probability from type $A_j$ to type $A_i$ given that an individual of type $A_j$ is transitioning ($M$ is a column-stochastic matrix: $\sum_j {M_{i,j}} = 1$ for any column \emph{j});
+$M_{i,j}$ is the transition probability from type $A_j$ to type $A_i$ given that an individual of type $A_j$ is transitioning ($M$ is a column-stochastic matrix: $\sum_j {M_{i,j}} = 1$ for all \emph{j});
 $D_k$ is the relative fitness of type $A_k$ ($D$ is a diagonal matrix with diagonal elements $D_k$, where we can assume $0 < D_k \le 1$, i.e., fitness values are relative to the fitness of the fittest type);
 $I$ is the identity matrix;
-and $\bar{\omega}$ is a normalizing factor such that $\sum_k{f'_k}=1$ and is equal to the population mean fitness (see~\nameref{sec:AppA}).
+and $\bar{w}$ is the normalizing factor such that $\sum_k{f'_k}=1$ and is equal to the population mean fitness (see~\nameref{sec:AppA}).
 
@@
 (i) the probability to transition \emph{out} of type $A_k$ is determined by $C_k$;
-(ii) given a transition out of type $A_i$, the distribution of the destination types $A_j$ is given by $M_{j,i}$ (note the index order).
+(ii) given a transition out of type $A_k$, the distribution of the destination types $A_j$ is given by $M_{j,k}$ (note the index order).
 We focus on a modifier that has no direct effect on fitness and whose sole function is to determine the rate of transition out of the different types ($C_k$).
@@
 \begin{equation}
-f_k' = (1 - \mu_k) \frac{w_k}{\bar{w}} f_k + \frac{1}{n-1} \sum_{k \ne j}{\mu_j \frac{w_j}{\bar{w}} x_j},
+f_k' = (1 - \mu_k) \frac{w_k}{\bar{w}} f_k + \frac{1}{n-1} \sum_{k \ne j}{\mu_j \frac{w_j}{\bar{w}} f_j},
 \label{eq:mutation_model_1}
@@
 A modifier locus then determines the probability $m_k$ that a mutation occurs in individuals with $k$ mutant alleles.
-In this case, we assume a low mutation rate $m_k$ such that two mutations are unlikely to occur in the same individual in one generation.
-When a mutation occurs it is \emph{deleterious} with probability $\delta$, generating a new mutant allele and moving the individual from $A_k$ to $A_{k+1}$, or it is beneficial with probability $\beta$, moving the individual from $A_k$ to $A_{k-1}$.
+In this case, we assume a low mutation rate $m_k$ so that two mutations are unlikely to occur in the same individual in one generation.
+When a mutation occurs it is \emph{deleterious} with probability $\delta$, generating a mutant allele and converting the individual from $A_k$ to $A_{k+1}$, or it is beneficial with probability $\beta$, converting the individual from $A_k$ to $A_{k-1}$.
 Note that such beneficial mutations can be either compensatory or back-mutations, and that mutations are neutral with probability $1-\delta-\beta$.
@@
 
-To rephrase this model in the terms of the general model, we define
-
+In terms of the general model, we can write
 \begin{equation}
@@
 \vdots & \ddots & \ddots & \ddots 
-\end{bmatrix}
+\end{bmatrix},
 \end{equation}
-so that Eq.~\ref{eq:mutation_model_2} is equivalent to Eq.~\ref{eq:model}.
+in which case Eq.~\ref{eq:mutation_model_2} is a specific case of Eq.~\ref{eq:model}.
 
@@
 Here we assume a large population of haploids that occupies two demes and a modifier locus that determine the migration rates between the demes, $m_k$. 
-Let the fitness of genotypes $A_k$ be $w_k$ in deme 1 and $u_k$ in deme 2 for $k=1,2,\ldots,n$.
-Let the frequencies of $A_k$ in deme 1 be $x_k$ and in deme 2 $y_k$ for $k=1,2,\ldots,n$ with $x_k \ge 0$, $y_k \ge 0$, $\sum_{k=1}^{n}{x_k} = 1$. and $\sum_{k=1}^{n}{y_k} = 1$.
+Let the fitness of genotypes $A_k$ be $w_k$ in deme 1 and $q_k$ in deme 2 for $k=1,2,\ldots,n$.
+Let the frequencies of $A_k$ be $x_k$ in deme 1 and $y_k$ in deme 2 for $k=1,2,\ldots,n$ with $x_k \ge 0$, $y_k \ge 0$, $\sum_{k=1}^{n}{x_k} = 1$, and $\sum_{k=1}^{n}{y_k} = 1$.
 
@@
 \begin{align}
-x_k' = (1-m_k) \frac{w_k}{\bar{w}} x_k + m_k \frac{u_k}{\bar{u}} y_k \\
-y_k' = m_k \frac{w_k}{\bar{w}} x_k + (1 - m_k) \frac{u_k}{\bar{u}} y_k
+x_k' = (1-m_k) \frac{w_k}{\bar{w}} x_k + m_k \frac{q_k}{\bar{q}} y_k \\
+y_k' = m_k \frac{w_k}{\bar{w}} x_k + (1 - m_k) \frac{q_k}{\bar{q}} y_k
 \end{align}
-for $k=1,2,\ldots,n$, where $\bar{w} = \sum_{k \ge 0}{w_k x_k}$ and $\bar{u} = \sum_{k \ge 0}{u_k y_k}$ is the population mean fitness in deme 1 and 2, respectively.
+for $k=1,2,\ldots,n$, where $\bar{w} = \sum_{k \ge 0}{w_k x_k}$ and $\bar{q} = \sum_{k \ge 0}{q_k y_k}$ are the population mean fitnesses in deme 1 and 2, respectively.
 This is an extension of a model presented by~\citet{Altenberg2017}, allowing for the migration rate of an individual, $m_k$, to depend on the identity of the allele $A_k$.
@@
 C = diag[m_1, \ldots, m_n, m_1, \ldots, m_n], \\
-D = diag[w_1, \ldots, w_n, u_1, \ldots, u_n], \\
+D = diag[w_1, \ldots, w_n, q_1, \ldots, q_n], \\
 M = \begin{bmatrix}
@@
 
+% Modifier model
+\paragraph*{Modifier model.}
+We now consider the case of two modifier alleles, $m$ and $M$, in complete linkage to the types $A_k$.
+These modifiers potentially induce different transition probabilities out of types $A_k$, with transition matrices $C=diag[C_1, \ldots, C_n]$ and $\tilde{C}=diag[\tilde{C}_1, \ldots, \tilde{C}_n]$.
+The frequencies of type $A_k$ linked to modifier $m$ or $M$ is $f_k$ and $g_k$, respectively, such that $\sum_k{f_k+g_k}=1$;
+$\bar{w}$ now ensures that $\sum_{k}{f_k'+g_k'}=1$;
+the rest of the model parameters are the same as in Eq.~\ref{eq:model};
+and the frequencies in the next generation, $f'$ and $g'$, are given by
+\begin{equation}
+\begin{cases}
+\bar{w} f' &= (I-C+MC)D f \\
+\bar{w} g' &= (I-\tilde{C}+M\tilde{C})D g
+\end{cases}.
+\label{eq:modifier_model}
+\end{equation}
+Note that Eq.~\ref{eq:model} is a special case for $g_k=0$ for all $k$.
+Here, $\bar{w}=\sum_{k}{(f_k+g_k)D_k}$ is the mean fitness of the entire population (see~\nameref{sec:AppA}).
+
 % Results
@@
 % Mean fitness principle
-With the equilibrium of the frequency vector $f$ in Eq.~\ref{eq:model} as $v$ and the stable population mean fitness as $\bar{\omega}^*$, then
+Write the equilibrium of the frequency vector $f$ in Eq.~\ref{eq:model} as $v$ and the stable population mean fitness as $\bar{w}^*$, then
 \begin{equation}\label{eq:model_equilibrium}
-\bar{\omega}^* v = (I-C+MC)D v
+\bar{w}^* v = (I-C+MC)D v
 \end{equation}
@@
 The following theorem presents a \emph{mean fitness principle} for the
-sensitivity of the stable population mean fitness $\bar{\omega}^*$ to changes
-in $C_k$ the probability for transition out of $A_k$.
+sensitivity of the stable population mean fitness $\bar{w}^*$ to changes
+in $C_k$ the probability of transition from $A_k$.
 
 \begin{mfp}
-Let $\bar{\omega}^*$ be the leading eigenvalue of $(I-C+MC)D$, $u$ and $v$ be the corresponding left and right eigenvectors, and $[M]_k$ the $k$-th column of $M$. Then
+Let $\bar{w}^*$ be the leading eigenvalue of $(I-C+MC)D$, $u$ and $v$ be the corresponding left and right eigenvectors, and $[M]_k$ the $k$-th column of $M$. Then
 \begin{equation}\label{eq:theorem}
-\frac{\partial \bar{\omega}^*}{\partial C_k} = 
-\frac{u_k v_k}{C_k} (\bar{\omega}^* - D_k),
+\frac{\partial \bar{w}^*}{\partial C_k} = 
+\frac{u_k v_k}{C_k} (\bar{w}^* - D_k),
 \end{equation}
@@
 \begin{equation}\label{eq:sign_theorem}
-sign\frac{\partial \bar{\omega}^*}{\partial C_k} = 
-sign(\bar{\omega}^* - D_k), 
+sign\frac{\partial \bar{w}^*}{\partial C_k} = 
+sign(\bar{w}^* - D_k).
 \end{equation}
-and therefore increased transition from type $k$ will increase the stable population mean fitness if and only if the fitness of type $k$ is below the stable population mean fitness.
+
+Therefore increased transition from type $k$ will increase the stable population mean fitness if and only if the fitness of type $k$ is below the stable population mean fitness.
 \end{mfp}
@@
 By the \emph{Perron-Frobenius theorem}~\cite[Appendix~A]{Otto2007},
-$\bar{\omega}^*$ is the leading real positive eigenvalue of $(I-C+MC)D$,
+$\bar{w}^*$ is the leading real positive eigenvalue of $(I-C+MC)D$,
 and the corresponding left and right eigenvectors $u$ and $v$ are both non-negative and uniquely defined by $\sum_k{v_k} = 1$ and $\sum_k{u_k v_k} = 1$.
@@
 \begin{equation}
-\frac{\partial \bar{\omega}^*}{\partial C_k} = 
+\frac{\partial \bar{w}^*}{\partial C_k} = 
 u\tr \frac{\partial (I-C+MC)D}{\partial C_k} v.
@@
 \begin{equation}
-u\tr \bar{\omega}^* = u\tr (I - C + MC) D,
+u\tr \bar{w}^* = u\tr (I - C + MC) D,
 \end{equation}
-which gives us a relation between $\bar{\omega}^*$ and the $k$
+which gives us a relation between $\bar{w}^*$ and the $k$
 element of $u$:
 \begin{equation}
-u_k \bar{\omega}^* = (1-C_k) D_k u_k + C_k D_k u\tr [M]_k.
+u_k \bar{w}^* = (1-C_k) D_k u_k + C_k D_k u\tr [M]_k.
 \end{equation}
@@
 \begin{equation}
-\frac{u_k v_k}{C_k} (\bar{\omega}^* - D_k) = v_k D_k(u\tr [M]_k - u_k).
+\frac{u_k v_k}{C_k} (\bar{w}^* - D_k) = v_k D_k(u\tr [M]_k - u_k),
 \end{equation}
-Plugging this in Eq.~\ref{eq:theorem_intermediate} yields:
+which when substituted into Eq.~\ref{eq:theorem_intermediate} yields:
 \begin{equation}
-\frac{\partial \bar{\omega}^*}{\partial C_k} = 
-\frac{u_k v_k}{C_k} (\bar{\omega}^* - D_k).
+\frac{\partial \bar{w}^*}{\partial C_k} = 
+\frac{u_k v_k}{C_k} (\bar{w}^* - D_k).
 \end{equation}
@@
 \begin{equation}
-sign \frac{\partial \bar{\omega}^*}{\partial C_k} = 
-sign (\bar{\omega}^* - D_k),
+sign \frac{\partial \bar{w}^*}{\partial C_k} = 
+sign (\bar{w}^* - D_k),
 \end{equation}
@@
 
-Note that if the modifier has the same effect on all types (i.e.,
-$C_k = \alpha$ for all $k$) then Eq.~\ref{eq:model_equilibrium}
-becomes $\bar{\omega}^* v = ((1-\alpha)I + \alpha M)Dv$,
-and the \emph{unified reduction principle} is in effect~\cite[eqs. 65, 72]{Altenberg2017}.
-Therefore, we can expect modifiers that reduce $\alpha$ to be favored by natural selection.
-Indeed, setting $C = \alpha I$ in Eq.~\ref{eq:model_equilibrium} and proceeding as in the above proof, we get
-$$
-\frac{\partial \bar{\omega}^*}{\partial \alpha} = 
-\bar{\omega}^* - 1 \le 0,
-$$
-because $\bar{\omega}^* = \sum_i{v_i D_i}$ (see~\nameref{sec:AppA}) and $D_i \le 1$ for all $i$.
+The above theorem provides a condition for the effect of a modifier in control of $C_k$, the probability for transition from $A_k$, on the stable population mean fitness.
+Specifically, if $A_k$ individuals have below-average fitness, then a modifier that increases $C_k$ will increase the population mean fitness.
 
@@
 \begin{aligned}
-\frac{\partial \bar{\omega}^*}{\partial C_k} = 
+\frac{\partial \bar{w}^*}{\partial C_k} = 
 v_k D_k (u\tr[M]_k - u_k) = \\
@@
 Here, $u_k$ can be regarded as the \emph{reproductive value}~\citep[pg.~27]{Fisher1930} of type $k$ (see~\nameref{sec:AppC}), which gives the relative contribution of type $k$ to the long-term population.
-Consequently, $u_k v_k$ is the \emph{ancestor frequency} \citep{Hermisson2002} of type $k$, which is the fraction of the equilibrium population descended from type $k$.
+Consequently, $u_k v_k$ is the \emph{ancestor frequency} \citep{Hermisson2002} of type $k$, namely the fraction of the equilibrium population descended from type $k$.
 The sum $\sum_i{u_i M_{i,k} v_k}$ can be similarly interpreted as the fraction of the equilibrium population descended from individuals that transitioned from type $k$ to another type (via the $k$ column of the transition matrix $M$).
@@
 \begin{equation}
-sign \frac{\partial \bar{\omega}^*}{\partial C_k} = 
+sign \frac{\partial \bar{w}^*}{\partial C_k} = 
 sign (u\tr [M]_k - u_k),
@@
 where $\tau > 0$, then the sensitivity of the stable population mean fitness to changes in the
-rate of transition from types $K$ is
+rate of transition from types $K$ can be written 
 \begin{equation}
 \begin{aligned}
-\frac{\partial \bar{\omega}^*}{\partial \tau} = 
-u\tr \Big( \sum_{k \in \mathbb{K{}}}{\frac{\partial (I - C + MC)D}{\partial \tau}} \Big) v = \\
-\frac{1}{\tau} \sum_{k \in \mathbb{K}}{u_k v_k (\bar{\omega}^* - D_k)},
+\frac{\partial \bar{w}^*}{\partial \tau} = 
+u\tr \Big( \sum_{k \in K}{\frac{\partial (I - C + MC)D}{\partial \tau}} \Big) v = \\
+\frac{1}{\tau} \sum_{k \in K}{u_k v_k (\bar{w}^* - D_k)},
 \end{aligned}
@@
 \begin{multline}\label{eq:sign_theorem_threshold}
-sign \frac{\partial \bar{\omega}^*}{\partial \tau} = 
-sign \sum_{k \in \mathbb{K}}{u_k v_k (\bar{\omega}^* - D_k)} = \\ 
-sign \Big(\bar{\omega}^* - \frac{\sum_{k \in \mathbb{K}}{u_k v_k D_k}}{\sum_{k \in \mathbb{K}}{u_k v_k}}\Big), 
+sign \frac{\partial \bar{w}^*}{\partial \tau} = 
+sign \sum_{k \in K}{u_k v_k (\bar{w}^* - D_k)} = \\ 
+sign \Big(\bar{w}^* - \frac{\sum_{k \in K}{u_k v_k D_k}}{\sum_{k \in K}{u_k v_k}}\Big), 
 \end{multline}
@@
 
+If such a modifier has the same effect on all individuals (i.e., $C_k = \alpha$ for all $k$) then Eq.~\ref{eq:model_equilibrium} $\bar{w}^* v = ((1-\alpha)I + \alpha M)Dv$,
+and the \emph{unified reduction principle} is in effect~\cite[eqs. 65, 72]{Altenberg2017}.
+Therefore, we can expect modifiers that reduce $\alpha$ to be favored by natural selection.
+Indeed, setting $C = \alpha I$ in Eq.~\ref{eq:model_equilibrium} and proceeding as in the above proof, we find
+$$
+\frac{\partial \bar{w}^*}{\partial \alpha} = 
+\bar{w}^* - 1 \le 0,
+$$
+because $\bar{w}^* = \sum_i{v_i D_i}$ (see~\nameref{sec:AppA}), and $D_i \le 1$ for all $i$.
+
+% Evolutionary stability
+
+We determined a condition for a modifier that controls that transition rate $C_k$ from type $A_k$ to increase the stable population mean fitness.
+But could such a modifier increase in frequency when initially rare in the population?
+To answer this question we will analyze the stability of modifier $m$ with transition rates $C_k$ to the invasion of modifier $M$ with rates $\tilde{C}_k$, as described in Eq.~\ref{eq:modifier_model}.
+
+The equilibrium of Eq.~\ref{eq:modifier_model} when $M$ is missing from the population ($g_k=0$ for all $k$) is (modified from \ref{eq:model_equilibrium})
+\begin{equation}\label{eq:modifier_model_equilibrium}
+\bar{w} \cdot \begin{pmatrix}v \\ 0\end{pmatrix} = 
+\Big(\vec{I} - \vec{C} + \vec{M} \vec{C}\Big)\vec{D} \begin{pmatrix}v \\ 0\end{pmatrix},
+\end{equation}
+where $\vec{I}$ is a $2n$-by-$2n$ identity matrix, and
+\begin{equation}
+%\begin{aligned}
+\vec{C} =  \begin{pmatrix}C & 0 \\ 0 & \tilde{C} \end{pmatrix}, \quad
+\vec{M} =  \begin{pmatrix}M & 0 \\ 0 & M \end{pmatrix}, \quad
+\vec{D} =  \begin{pmatrix}D & 0 \\ 0 & D \end{pmatrix}.
+%\end{aligned}
+\end{equation}
+
+The stability of modifier $m$ to invasion by modifier $M$ is determined by the leading eigenvalue $\lambda$ of the external stability matrix $L_{ex}$ of modifier $M$, which is in turn determined by the Jacobian of the system in Eq.~\ref{eq:modifier_model_equilibrium}
+\begin{equation}
+J = \begin{pmatrix}L_{in} & 0 \\ 0 & L_{ex} \end{pmatrix}.
+\end{equation}
+Note the zero block matrices due to complete linkage between the modifier and the types $A_k$.
+
+$L_{ex}$ can be written 
+\begin{equation}\label{eq:Lex}
+L_{ex} = (I - \tilde{C} + M \tilde{C}) D /\bar{w}^*,
+\end{equation}
+where $\bar{w}^*$ is the stable population mean fitness without modifier $M$ (see~\nameref{sec:AppA}).
+
+\begin{theorem}[Evolution of increased genetic variation]
+The sign of the derivative of the leading eigenvalue $\lambda$ of the external stability matrix of $L_{ex}$ of modifier $M$, when evaluated at $\tilde{C}_k=C_k$, is
+\begin{equation}
+sign \frac{\partial \lambda}{\partial \tilde{C}_k}\bigg\rvert_{\tilde{C}_k = C_k} = 
+sign(\bar{w}^* - D_k).
+\end{equation}
+In addition, the leading eigenvalue is $\lambda=1$ when transition rates are equal $\tilde{C}_k=C_k$ for all $k$.
+
+Therefore, an invading modifier $M$ with transition rates slightly different from the resident modifier $m$ can successful invade the population ($\lambda>1$ if $M$ increases the probability of transition from types with below-average fitness.
+\end{theorem}
+
+\begin{proof}
+First, consider the case the $\tilde{C_k} = C_k$ for all $k$.
+Substituting $\tilde{C}=C$ in Eq.~\ref{eq:Lex} and multiplying both sides by $\bar{w}^*$,
+\begin{equation}
+\bar{w}^* L_{ex} = (I-C+MC) D,
+\end{equation}
+and since $\bar{w}^*$ is the leading eigenvalue of the RHS (\nameref{sec:AppA}),
+then the leading eigenvalue of $L_{ex}$ is $\lambda=1$.
+
+Now, applying Eq.~\ref{eq:sign_theorem} on Eq.~\ref{eq:Lex}, the sign of the derivative of $\lambda$ with respect to $\tilde{C_k}$ is
+\begin{equation}
+sign \frac{\partial \lambda}{\partial \tilde{C}_k} =
+sign\Big(\lambda - \frac{D_k}{\bar{w}^*}\Big),
+\end{equation}
+and evaluating this expression at $\lambda=1$ (the case for $\tilde{C}_k = C_k$) gives us
+\begin{equation}
+sign \frac{\partial \lambda}{\partial \tilde{C}_k}\bigg\rvert_{\lambda=1} = 
+sign(\bar{w}^* - D_k),
+\end{equation}
+since $\bar{w}^*>0$.
+This completes the proof.
+\end{proof}
+
 % Discussion
@@
 
-We have shown that under constant-viability selection and in an effectively infinite haploid populations at an equilibrium, the stable population mean fitness increases if below-fitness individuals increase the rate at which variation is generated.
-We have given several examples on variation-generating processes for which this principle applies -- mainly mutation and migration -- but our model is general enough to possibly apply to other processes as well.
-For example, the \emph{reduction principle} appears some ecological models of dispersal, and~\citet{Gueijman2013} have demonstrated that even in homogeneous environments, fitness-associated dispersal increases the population mean fitness and is favored by selection over uniform dispersal.
+We have shown that under constant-viability selection and in an effectively infinite haploid population at mutation-selection or migration-selection equilibrium, the stable population mean fitness increases if below-fitness individuals increase the rate at which variation is generated.
+We have given several examples of variation-generating processes for which this principle applies -- mainly mutation and migration -- but our model may apply to other processes as well.
+For example, the \emph{reduction principle} applies to ecological models of dispersal, and~\citet{Gueijman2013} have demonstrated that even in homogeneous environments, fitness-associated dispersal increases the population mean fitness and is favored by selection over uniform dispersal.
 Similarly, if the transmission fidelity of culturally-transmitted traits depends on the type of the type or fitness of the transmitting individual, we expect that our \emph{mean fitness principle} will hold.
 
-Eq.~\ref{eq:theorem} is a generalization of a result from \citet[Eq.~4]{Ram2012}.
+Eq.~\ref{eq:theorem} is a generalization of a result of~\citet[Eq.~4]{Ram2012}.
 \citeauthor{Ram2012} modeled the accumulation of mutant alleles in a population evolving at the mutation-selection balance (see \nameref{sec:mut_model_2}).
-Using Eq.~\ref{eq:Caswells_formula} in \nameref{sec:AppB} and a recursion on the ratios of the reproductive values \cite[see][eqs.~A5-6]{Ram2012}, they concluded that at the mutation-selection balance, if individuals with below-average fitness ($D_k < \bar{\omega}^*$) increase their mutation rate, then the population mean fitness will increase -- a result generalized by our \emph{mean fitness principle} in Eqs.~\ref{eq:sign_theorem} and~\ref{eq:sign_theorem_threshold}.
+Using Eq.~\ref{eq:Caswells_formula} in \nameref{sec:AppB} and a recursion on the ratios of the reproductive values \cite[see][eqs.~A5-6]{Ram2012}, they concluded that at the mutation-selection balance, if individuals with below-average fitness ($D_k < \bar{w}^*$) increase their mutation rate, then the population mean fitness will increase -- a result generalized by our \emph{mean fitness principle} in Eqs.~\ref{eq:sign_theorem} and~\ref{eq:sign_theorem_threshold}.
 
@@
 Similar results have been demonstrated for a modifier that increases the recombination rate in maladapted individuals~\citep{Hadany2003a,Hadany2003b}.
-Importantly, there are empirical evidence both for stress-induced mutagenesis~\citep{Rosenberg2012} and for condition-dependent recombination and sex~\citep{Ram2016}.
 
@@
 \begin{lemma}
-The normalizing factor $\bar{\omega}$ in Eq.~\ref{eq:model} is the population mean fitness
+The normalizing factor $\bar{w}$ in Eq.~\ref{eq:model} is the population mean fitness
 \begin{equation}
-\bar{\omega} = \sum_k{f_k D_k}.
+\bar{w} = \sum_k{f_k D_k}.
 \end{equation}
@@
 \begin{proof}
-By definition, the normalizing factor $\bar{\omega}$ ensures that $\sum_k{f'_k}=1$.
+By definition, the normalizing factor $\bar{w}$ ensures that $\sum_k{f'_k}=1$.
 Denote $e = (1, ..., 1)$ the vector with ones at all positions and rewrite $e\tr f' = 1$.
@@
 \begin{multline}\label{eq:mean_fitness}
-\bar{\omega} = 
-\bar{\omega} e\tr f' = 
+\bar{w} = 
+\bar{w} e\tr f' = 
 e\tr (I - C + M C) D f = \\
@@
 
-Note that this lemma also ensures that the stable population mean fitness $\bar{\omega}^*=\sum_k{v_k D_k},$ is the leading eigenvalue of $(I - C + M C) D$; see Eq.~\ref{eq:model_equilibrium} and the \emph{Perron-Frobenius theorem}~\cite[Appendix~A]{Otto2007}.
+Note that this lemma also ensures that the stable population mean fitness $\bar{w}^*=\sum_k{v_k D_k}$ is the leading eigenvalue of $(I - C + M C) D$; see Eq.~\ref{eq:model_equilibrium} and the \emph{Perron-Frobenius theorem}~\cite[Appendix~A]{Otto2007}.
 
-% Appendix B: Caswell
+% Appendix B: Caswell's formula
 \subsection*{Appendix B}\label{sec:AppB}
@@
 
-\begin{caswell}
+\begin{lemma}[Caswell's formula]
 Let $T$ be a non-negative matrix with leading eigenvalue $\lambda$ and left and right eigenvectors $u$ and $v$ such that $\sum{v_k}=1$ and $u\tr v = \sum{u_k v_k} = 1$.
@@
 \end{equation}
-\end{caswell}
+\end{lemma}
 
@@
 \emph{Fisher's reproductive values} can be understood as follows~\citep{Grafen2006}.
-Consider the dynamics not of frequencies but of absolute population sizes such that the vector of the number of individuals of each type at time $t$ is $n(t)$ and $f_k(t) = n_k(t) / \sum_i{n_i(t)}$.
+Consider the dynamics not of frequencies but of absolute population sizes such that the vector of the number of individuals of each type at time $t$ is $n(t)$ and the corresponding frequencies are $f_k(t) = n_k(t) / \sum_i{n_i(t)}$.
 The dynamics are
@@
 \end{equation}
-where $V$ is the matrix of right column eigenvectors of $MD$,
-$U\tr$ is the matrix of left row eigenvectors of $MD$,
-$V U\tr = U\tr V = I$, 
+where $V$ is the matrix of right (column) eigenvectors of $MD$,
+$U\tr$ is the transposed matrix of left (row) eigenvectors of $MD$,
+where we can take $V U\tr = U\tr V = I$, 
 and $\Lambda$ is the diagonal matrix of eigenvalues of $A$ 
@@
 \end{multline}
-
 Now take the limit $t \goesto \infty$.
@@
 
-The vector $u$ is the left \emph{Perron} eigenvector of $MD$, and $u_k$ is element $k$ of $u$.
+The vector $u$ is the left \emph{Perron} eigenvector of $MD$, and $u_k$ is $k$-th element of $u$.
 This is why the value $u_k$ can be interpreted as the \emph{reproductive value} of type $i$: it is a weighting for the size of the population generated by a single individual of type $k$.
@@
 
-This manuscript was supported in part by the Stanford Center for Computational, Evolutionary and Human Genomics, and by the Morrison Institute for Population and Resources Studies, Stanford University.
+This research was supported in part by the Stanford Center for Computational, Evolutionary and Human Genomics, and by the Morrison Institute for Population and Resources Studies, Stanford University.
 }
 
-\bibliographystyle{plainnat}
+\bibliographystyle{agsm}
 %\bibliography{/Users/yoavram/Documents/library}

### 2017-11-06 ms.tex
addtional comments by MWF; initial integration of modifier model in ms

@@
 \newcommand{\dspfrac}[2]{\frac{\displaystyle #1}{\displaystyle #2} }
-\newtheorem*{theorem}{Theorem}
-\newtheorem*{corollary}{Corollary}
+\newtheorem{theorem}{Theorem}
+\newtheorem{corollary}{Corollary}
 \newtheorem{lemma}{Lemma}
-\newtheorem*{remark}{Remark}
-\newtheorem*{mfp}{Theorem: Mean Fitness Principle}
-\newtheorem*{rvp}{Corollary: Reproductive Value Principle}
-\newtheorem*{frv}{Fisher's reproductive values}
+\newtheorem{remark}{Remark}
+%\newtheorem*{mfp}{Theorem: Mean Fitness Principle}
+%\newtheorem*{rvp}{Corollary: Reproductive Value Principle}
+%\newtheorem*{frv}{Fisher's reproductive values}
 \renewcommand\qedsymbol{} % no square at end of proof
+\newcommand{\cl}{\mathbf{L}}
+\newcommand{\cj}{\mathbf{J}}
+\newcommand{\ci}{\mathbf{I}}
 
@@
 Here we analyze a general model of evolution under constant selection where the rate at which variation is generated can depend on the fitness of the individual.
-We find a general and simple result:
-if individuals with below-average fitness increase the rate at which variation is generated, then the population mean fitness increases.
+We find a general and clear result:
+if a modifier allele increases the rate at which below-average individuals generate variation, then it will increase in frequency and increase the  population mean fitness.
 This principle applies to phenomena such as stress-induced mutagenesis, condition-dependent dispersal and outcrossing,
@@
 \paragraph*{General model.}\label{sec:general_model}
-Consider a large haploid population with an arbitrary (possibly infinite) set of types $A_1, A_2, \ldots$.
+Consider a large haploid population with an arbitrary set of types $A_1, A_2, \ldots, A_n$.
 The change in the frequencies of these types is described by the transformation $f \to f'$: 
@@
 \end{equation}
-where $f_k$ is the frequency of type $A_k$ ($f=(f_1, f_2, \ldots)$ is a stochastic vector: $\sum_k{f_k} = 1$);
+where $f_k$ is the frequency of type $A_k$ ($f=(f_1, f_2, \ldots, f_n)$ is a stochastic vector: $\sum_{k=1}^n{f_k} = 1$);
 $C_k$ is the probability that an individual of type $A_k$ will transition to another type ($C$ is a diagonal non-negative matrix with diagonal elements $C_k$);
-$M_{i,j}$ is the transition probability from type $A_j$ to type $A_i$ given that an individual of type $A_j$ is transitioning ($M$ is a column-stochastic matrix: $\sum_j {M_{i,j}} = 1$ for all \emph{j});
+$M_{i,j}$ is the transition probability from type $A_j$ to type $A_i$ given that an individual of type $A_j$ is transitioning ($M$ is a column-stochastic matrix: $\sum_{j=1}^n {M_{i,j}} = 1$ for all \emph{j});
 $D_k$ is the relative fitness of type $A_k$ ($D$ is a diagonal matrix with diagonal elements $D_k$, where we can assume $0 < D_k \le 1$, i.e., fitness values are relative to the fitness of the fittest type);
 $I$ is the identity matrix;
-and $\bar{w}$ is the normalizing factor such that $\sum_k{f'_k}=1$ and is equal to the population mean fitness (see~\nameref{sec:AppA}).
+and $\bar{w}$ is the normalizing factor such that $\sum_{k=1}^n{f'_k}=1$ and is equal to the population mean fitness (see~\nameref{sec:AppA}).
 
@@
 (ii) given a transition out of type $A_k$, the distribution of the destination types $A_j$ is given by $M_{j,k}$ (note the index order).
-We focus on a modifier that has no direct effect on fitness and whose sole function is to determine the rate of transition out of the different types ($C_k$).
+
+We focus on a neutral modifier locus in complete linkage to the types $A_k$, with no direct effect on fitness, and whose sole function is to determine the rate of transition out of the different types ($C_k$).
 Importantly, this modifier can determine different rates for different types; that is, $C_j \ne C_i$ for some $i,j$.
@@
 
-In the following section we present three particular examples of this model for the modification of mutation, migration, and learning.
+% Modifier model
+\paragraph*{Modifier model.}
+Considering the case of two modifier alleles, $m$ and $M$, inducing different transition probabilities out of types $A_k$, with transition matrices $C=diag[C_1, \ldots, C_n]$ and $\tilde{C}=diag[\tilde{C}_1, \ldots, \tilde{C}_n]$.
+The frequencies of type $A_k$ linked to modifier $m$ or $M$ are $f_k$ and $g_k$, respectively, such that $\sum_k{f_k+g_k}=1$;
+$\bar{w}$ now ensures that $\sum_{k}{f_k'+g_k'}=1$;
+the rest of the model parameters are the same as in Eq.~\ref{eq:model};
+and the frequencies in the next generation, $f'$ and $g'$, are given by
+\begin{equation}
+\begin{cases}
+\bar{w} f' &= (I-C+MC)D f \\
+\bar{w} g' &= (I-\tilde{C}+M\tilde{C})D g
+\end{cases}.
+\label{eq:modifier_model}
+\end{equation}
+Note that Eq.~\ref{eq:model} is a special case for $g_k=0$ for all $k$.
+Here, $\bar{w}=\sum_{k}{(f_k+g_k)D_k}$ is the mean fitness of the entire population (see~\nameref{sec:AppA}).
 
+In the following section we present three particular examples of the model presented in Eq.~\ref{eq:model} for the modification of mutation and migration.
+
+% Mutation model 1
 \paragraph*{Mutation modification model 1.}\label{sec:mut_model_1}
@@
 
+% Mutation model 2
 \paragraph*{Mutation modification model 2.}\label{sec:mut_model_2}
 
-Here, we consider a large population of haploids in which individuals with genotype $A_k$ are characterized by the number $k$ of deleterious or mutant alleles in their genotype.
+Here, we consider a large population of haploids in which individuals with genotype $A_k$ are characterized by the number $k$ of deleterious or mutant alleles in their genotype, where $0 \le k \le n$.
 A modifier locus then determines the probability $m_k$ that a mutation occurs in individuals with $k$ mutant alleles.
@@
 
-Let the frequency of $A_k$ in the present generation be $f_k$ such that $f_k \ge 0$ and $\sum_k{f_k}=1$.
-Then after selection and mutation $\vec{f'}$ in the next generation is given by
+Let the frequency of $A_k$ in the present generation be $f_k$ such that $f_k \ge 0$ and $\sum_{k=0}^{n}{f_k}=1$.
+Then after selection and mutation $f'$ in the next generation is given by
 \begin{equation}
-f_0' = (1 - \delta m_0) \frac{w_0}{\bar{w}} f_0 + \beta m_{1} \frac{w_{1}}{\bar{w}} f_{1},
+\begin{aligned}
+f_0' &= (1 - \delta m_0) \frac{w_0}{\bar{w}} f_0 + \beta m_{1} \frac{w_{1}}{\bar{w}} f_{1}, \\
+f_n' &= (1 - \beta m_n) \frac{w_n}{\bar{w}} f_n + \delta m_{n-1} \frac{w_{n-1}}{\bar{w}} f_{n-1}, \\
+f_k' &= (1 - m_k(\delta+\beta)) \frac{w_k}{\bar{w}} f_k + \\
+	 & \delta m_{k-1} \frac{w_{k-1}}{\bar{w}} f_{k-1} + 
+	 \beta m_{k+1} \frac{w_{k+1}}{\bar{w}} f_{k+1},
+\end{aligned}
+\label{eq:mutation_model_2}
 \end{equation}
-and
-\begin{multline}
-f_k' = (1 - m_k(\delta+\beta)) \frac{w_k}{\bar{w}} f_k + \\
-\delta m_{k-1} \frac{w_{k-1}}{\bar{w}} f_{k-1} + \beta m_{k+1} \frac{w_{k+1}}{\bar{w}} f_{k+1},
-\label{eq:mutation_model_2}
-\end{multline}
-for all $k=1,2,\ldots$, where $w_k$ is the fitness of an individual with $k$ mutant alleles and $\bar{w} = \sum_{k \ge 0}{w_k f_k}$ is the population mean fitness.
+for $k=1,2,\ldots, n-1$, where $w_k$ is the fitness of an individual with $k$ mutant alleles and $\bar{w} = \sum_{k=0}^{n}{w_k f_k}$ is the population mean fitness.
 
@@
 
-% Modifier model
-\paragraph*{Modifier model.}
-We now consider the case of two modifier alleles, $m$ and $M$, in complete linkage to the types $A_k$.
-These modifiers potentially induce different transition probabilities out of types $A_k$, with transition matrices $C=diag[C_1, \ldots, C_n]$ and $\tilde{C}=diag[\tilde{C}_1, \ldots, \tilde{C}_n]$.
-The frequencies of type $A_k$ linked to modifier $m$ or $M$ is $f_k$ and $g_k$, respectively, such that $\sum_k{f_k+g_k}=1$;
-$\bar{w}$ now ensures that $\sum_{k}{f_k'+g_k'}=1$;
-the rest of the model parameters are the same as in Eq.~\ref{eq:model};
-and the frequencies in the next generation, $f'$ and $g'$, are given by
-\begin{equation}
-\begin{cases}
-\bar{w} f' &= (I-C+MC)D f \\
-\bar{w} g' &= (I-\tilde{C}+M\tilde{C})D g
-\end{cases}.
-\label{eq:modifier_model}
-\end{equation}
-Note that Eq.~\ref{eq:model} is a special case for $g_k=0$ for all $k$.
-Here, $\bar{w}=\sum_{k}{(f_k+g_k)D_k}$ is the mean fitness of the entire population (see~\nameref{sec:AppA}).
-
 % Results
@@
 
-\begin{mfp}
+\begin{lemma}[Mean fitness principle]
 Let $\bar{w}^*$ be the leading eigenvalue of $(I-C+MC)D$, $u$ and $v$ be the corresponding left and right eigenvectors, and $[M]_k$ the $k$-th column of $M$. Then
@@
 
-Therefore increased transition from type $k$ will increase the stable population mean fitness if and only if the fitness of type $k$ is below the stable population mean fitness.
-\end{mfp}
+Therefore increased transition from type $k$ will increase the stable population mean fitness if the fitness of type $k$ is below the stable population mean fitness.
+\end{lemma}
 
@@
 Consequently, $u_k v_k$ is the \emph{ancestor frequency} \citep{Hermisson2002} of type $k$, namely the fraction of the equilibrium population descended from type $k$.
-The sum $\sum_i{u_i M_{i,k} v_k}$ can be similarly interpreted as the fraction of the equilibrium population descended from individuals that transitioned from type $k$ to another type (via the $k$ column of the transition matrix $M$).
+The sum $\sum_i{u_i M_{i,k} v_k}$ can be similarly interpreted as the fraction of the equilibrium population descended from individuals that transitioned from type $k$ to another type (via the $k$ column of the transition matrix $M$), conditioned on transition occuring.
 
@@
 
-\begin{rvp}
+\begin{corollary}[Reproductive value principle]
 In the notation of the theorem,
@@
 \end{equation}
-and therefore increased transition out of type $k$ will increase the stable population mean fitness if and only if the fraction of the population descended from type $k$ is expected to increase due to a transition to another type.
-\end{rvp}
+and therefore increased transition out of type $k$ will increase the stable population mean fitness if the fraction of the population descended from type $k$ is expected to increase due to a transition to another type.
+\end{corollary}
 
@@
 a generalization of Eq.~\ref{eq:sign_theorem}.
-Therefore, increased transition from types $K$ will increase the stable population mean fitness if and only if the average fitness of individuals descended from types $K$ is below the stable population mean fitness.
+Therefore, increased transition from types $K$ will increase the stable population mean fitness if the average fitness of individuals descended from types $K$ is below the stable population mean fitness.
 
@@
 We determined a condition for a modifier that controls that transition rate $C_k$ from type $A_k$ to increase the stable population mean fitness.
-But could such a modifier increase in frequency when initially rare in the population?
-To answer this question we will analyze the stability of modifier $m$ with transition rates $C_k$ to the invasion of modifier $M$ with rates $\tilde{C}_k$, as described in Eq.~\ref{eq:modifier_model}.
+Could such a modifier increase in frequency when initially rare in the population?
+To answer this question we analyze the stability of modifier $m$ with transition rates $C_k$ to the invasion of modifier $M$ with rates $\tilde{C}_k$, as described in Eqs.~\ref{eq:modifier_model}.
 
-The equilibrium of Eq.~\ref{eq:modifier_model} when $M$ is missing from the population ($g_k=0$ for all $k$) is (modified from \ref{eq:model_equilibrium})
-\begin{equation}\label{eq:modifier_model_equilibrium}
-\bar{w} \cdot \begin{pmatrix}v \\ 0\end{pmatrix} = 
-\Big(\vec{I} - \vec{C} + \vec{M} \vec{C}\Big)\vec{D} \begin{pmatrix}v \\ 0\end{pmatrix},
+The equilibrium of Eqs.~\ref{eq:modifier_model} when modifier allele $M$ is absent from the population is $(v, 0)$, where $v$ is given in eq.~\ref{eq:model_equilibrium} and $g_k=0$ for all $k$. 
+The stability of modifier allele $m$ to invasion by allele $M$ is determined by the leading eigenvalue $\lambda$ of the external stability matrix $\cl_{ex}$ of the equilibrium $(v,0)$, which, in turn, is determined by the Jacobian $\cj$ of Eq.~\ref{eq:modifier_model} evaluated at the equilibrium $(v,0)$, where
+\begin{equation}
+\cj = \begin{pmatrix}\cl_{in} & 0 \\ 0 & \cl_{ex} \end{pmatrix}.
 \end{equation}
-where $\vec{I}$ is a $2n$-by-$2n$ identity matrix, and
-\begin{equation}
-%\begin{aligned}
-\vec{C} =  \begin{pmatrix}C & 0 \\ 0 & \tilde{C} \end{pmatrix}, \quad
-\vec{M} =  \begin{pmatrix}M & 0 \\ 0 & M \end{pmatrix}, \quad
-\vec{D} =  \begin{pmatrix}D & 0 \\ 0 & D \end{pmatrix}.
-%\end{aligned}
+The zero block matrices are due to complete linkage between the modifier and the types $A_k$.
+
+$\cl_{ex}$ can be written as
+\begin{equation}\label{eq:Lex}
+\cl_{ex} = (I - \tilde{C} + M \tilde{C}) D /\bar{w}^*,
 \end{equation}
-
-The stability of modifier $m$ to invasion by modifier $M$ is determined by the leading eigenvalue $\lambda$ of the external stability matrix $L_{ex}$ of modifier $M$, which is in turn determined by the Jacobian of the system in Eq.~\ref{eq:modifier_model_equilibrium}
-\begin{equation}
-J = \begin{pmatrix}L_{in} & 0 \\ 0 & L_{ex} \end{pmatrix}.
-\end{equation}
-Note the zero block matrices due to complete linkage between the modifier and the types $A_k$.
-
-$L_{ex}$ can be written 
-\begin{equation}\label{eq:Lex}
-L_{ex} = (I - \tilde{C} + M \tilde{C}) D /\bar{w}^*,
-\end{equation}
-where $\bar{w}^*$ is the stable population mean fitness without modifier $M$ (see~\nameref{sec:AppA}).
+where $\bar{w}^*$ is the stable population mean fitness in the absence of modifier allele $M$ (see~\nameref{sec:AppA}).
 
 \begin{theorem}[Evolution of increased genetic variation]
-The sign of the derivative of the leading eigenvalue $\lambda$ of the external stability matrix of $L_{ex}$ of modifier $M$, when evaluated at $\tilde{C}_k=C_k$, is
+The sign of the derivative of the leading eigenvalue $\lambda$ of the external stability matrix of $\cl_{ex}$ of modifier $M$, when evaluated at $\tilde{C}_k=C_k$, is
 \begin{equation}
@@
 
-Therefore, an invading modifier $M$ with transition rates slightly different from the resident modifier $m$ can successful invade the population ($\lambda>1$ if $M$ increases the probability of transition from types with below-average fitness.
+Therefore, an initially rate modifier allele $M$ with transition rates slightly different from the resident modifier $m$ can successfully invade the population ($\lambda>1$) if $M$ increases the probability of transition from types with below-average fitness, thereby increasing the stable population mean fitness.
 \end{theorem}
@@
 and since $\bar{w}^*$ is the leading eigenvalue of the RHS (\nameref{sec:AppA}),
-then the leading eigenvalue of $L_{ex}$ is $\lambda=1$.
+the leading eigenvalue of $\cl_{ex}$ is $\lambda=1$.
 
-Now, applying Eq.~\ref{eq:sign_theorem} on Eq.~\ref{eq:Lex}, the sign of the derivative of $\lambda$ with respect to $\tilde{C_k}$ is
+Now, applying Eq.~\ref{eq:sign_theorem} to Eq.~\ref{eq:Lex}, the sign of the derivative of $\lambda$ with respect to $\tilde{C_k}$ is
 \begin{equation}
@@
 \subsection*{Appendix C}\label{sec:AppC}
-\begin{frv}
+\begin{remark}[Fisher's reproductive value]
 Let $M$ be a stochastic matrix and $D$ be a diagonal matrix with positive diagonal elements. 
 The elements of the left \emph{Perron} eigenvector $u$ of the matrix $MD$ can be regarded as \emph{Fisher's reproductive values}~\citep[pg.~27]{Fisher1930}
-\end{frv}
+\end{remark}
 

### 2017-11-07 ms.tex
full integration of modifier model; some small revisions

@@
 Their analysis was specific to a model that classified individuals by the number of mutant genes in their genotype, similar to models studied by~\citet{Kimura1966a} and~\citet{Haigh1978}.
+Moreover, their argument was based on expected increase of the population mean fitness by alleles that modify the mutation rate, rather than on the invasion success of such modifiers~\citep[i.e., analysis of \emph{evolutionary genetic stability}, see][]{Lessard1990}.
 
-Here, we present a general evolutionary model where the rate at which variation is generated depends on the type of individual, and specifically on the its fitness.
-Our result shows that if below-average individuals produce more variation than above-average individuals, then the population mean fitness increases.
+Here, we present a general evolutionary model where the rate at which variation is generated depends on the type of the individual, and specifically on its fitness.
+Our results show that the population mean fitness increases if below-average individuals produce more variation than above-average individuals,
+and that modifier alleles that induce below-average to produce more variation are favored by natural selection.
 
@@
 The types $A_k$ can represent a single or multiple genetic loci or non-genetic traits.
-Importantly, type transmission is vertical and uni-parental (the type is transmitted from a single parent to the offspring), independent of the frequencies of the other types, and is fully described by the matrix $M$.
-Specifically, transmission infidelity causes type transitions.
-Therefore, this model precludes processes such as recombination, social learning, sexual outcrossing, and any horizontal or oblique transmission, as these processes are frequency-dependent~\cite[pg.~54]{Cavalli-Sforza1981}.
+Importantly, type transmission is vertical and uni-parental (the type is transmitted from a single parent to the offspring) and is independent of the frequencies of the other types.
+Specifically, this model precludes processes such as recombination, social learning, sexual outcrossing, and any horizontal or oblique transmission, as these processes are frequency-dependent~\cite[pg.~54]{Cavalli-Sforza1981}.
 
@@
 
-We focus on a neutral modifier locus in complete linkage to the types $A_k$, with no direct effect on fitness, and whose sole function is to determine the rate of transition out of the different types ($C_k$).
+We focus on a neutral modifier locus in complete linkage to the types $A_k$, with no direct effect on fitness, and whose sole function is to determine $C_k$ the rates of transition out of the different types.
 Importantly, this modifier can determine different rates for different types; that is, $C_j \ne C_i$ for some $i,j$.
-The case $C_j = C_j$ for all $i,j$ is covered by the \emph{reduction principle}, see~\citet{Altenberg2017}.
+The case $C_j = C_j$ for all $i,j$ is covered by the \emph{reduction principle}~\citep[see][]{Altenberg2017}.
 
@@
 \paragraph*{Modifier model.}
-Considering the case of two modifier alleles, $m$ and $M$, inducing different transition probabilities out of types $A_k$, with transition matrices $C=diag[C_1, \ldots, C_n]$ and $\tilde{C}=diag[\tilde{C}_1, \ldots, \tilde{C}_n]$.
-The frequencies of type $A_k$ linked to modifier $m$ or $M$ are $f_k$ and $g_k$, respectively, such that $\sum_k{f_k+g_k}=1$;
-$\bar{w}$ now ensures that $\sum_{k}{f_k'+g_k'}=1$;
-the rest of the model parameters are the same as in Eq.~\ref{eq:model};
-and the frequencies in the next generation, $f'$ and $g'$, are given by
+Consider the case of two modifier alleles, $m$ and $M$, inducing different transition probabilities $C=\diag{C_1, \ldots, C_n}$ and $\tilde{C}=\diag{\tilde{C}_1, \ldots, \tilde{C}_n}$, respectively.
+The frequencies of type $A_k$ linked to modifier $m$ or $M$ are $f_k$ and $g_k$, respectively, such that $\sum_{k=1}^{n}{(f_k + g_k)}=1$.
+$\bar{w}$ now ensures that $\sum_{k=1}^{n}{(f_k' + g_k')}=1$, 
+and the rest of the model parameters are the same as in Eq.~\ref{eq:model}.
+
+The frequencies in the next generation, $f'$ for allele $m$ and $g'$ for allele $M$, are given by
 \begin{equation}
@@
 \end{equation}
-Note that Eq.~\ref{eq:model} is a special case for $g_k=0$ for all $k$.
-Here, $\bar{w}=\sum_{k}{(f_k+g_k)D_k}$ is the mean fitness of the entire population (see~\nameref{sec:AppA}).
+Note that Eq.~\ref{eq:model} is a special case for the absence of allele $M$, with $g_k=0$ for all $k$.
+Here, $\bar{w}=\sum_{k=1}^{n}{(f_k+g_k)D_k}$ is the mean fitness of the entire population (see~\nameref{sec:AppA}).
 
 In the following section we present three particular examples of the model presented in Eq.~\ref{eq:model} for the modification of mutation and migration.
+For simplicity of the presentation we focus on one modifier allele in these examples.
 
@@
 \begin{equation}\label{eq:model_equilibrium}
-\bar{w}^* v = (I-C+MC)D v
+\bar{w}^* v = (I-C+MC)D v.
 \end{equation}
 
-The following theorem presents a \emph{mean fitness principle} for the
+The following lemma presents a \emph{mean fitness principle} for the
 sensitivity of the stable population mean fitness $\bar{w}^*$ to changes
-in $C_k$ the probability of transition from $A_k$.
+in $C_k$, the probability of transition from $A_k$.
 
-\begin{lemma}[Mean fitness principle]
+\begin{lemma}[Mean fitness principle]\label{lemma:mfp}
 Let $\bar{w}^*$ be the leading eigenvalue of $(I-C+MC)D$, $u$ and $v$ be the corresponding left and right eigenvectors, and $[M]_k$ the $k$-th column of $M$. Then
-\begin{equation}\label{eq:theorem}
+\begin{equation}\label{mfp}
 \frac{\partial \bar{w}^*}{\partial C_k} = 
@@
 or in simpler terms,
-\begin{equation}\label{eq:sign_theorem}
+\begin{equation}\label{mfp_sign}
 sign\frac{\partial \bar{w}^*}{\partial C_k} = 
@@
 Then,
-\begin{equation}\label{eq:theorem_intermediate}
+\begin{equation}\label{mfp_intermediate}
 \begin{aligned}
@@
 \end{equation}
-which when substituted into Eq.~\ref{eq:theorem_intermediate} yields:
+which when substituted into Eq.~\ref{mfp_intermediate} yields:
 \begin{equation}
@@
 
-The above theorem provides a condition for the effect of a modifier in control of $C_k$, the probability for transition from $A_k$, on the stable population mean fitness.
+The above lemma provides a condition for the effect of changing $C_k$, the probability for transition from $A_k$, on the stable population mean fitness.
 Specifically, if $A_k$ individuals have below-average fitness, then a modifier that increases $C_k$ will increase the population mean fitness.
@@
 
-An interesting intermediate result in Eq.~\ref{eq:theorem_intermediate} is
+An interesting intermediate result in Eq.~\ref{mfp_intermediate} is
 
@@
 Here, $u_k$ can be regarded as the \emph{reproductive value}~\citep[pg.~27]{Fisher1930} of type $k$ (see~\nameref{sec:AppC}), which gives the relative contribution of type $k$ to the long-term population.
-Consequently, $u_k v_k$ is the \emph{ancestor frequency} \citep{Hermisson2002} of type $k$, namely the fraction of the equilibrium population descended from type $k$.
+Consequently, $u_k v_k$ is the \emph{ancestor frequency} of type $k$~\citep{Hermisson2002}, namely the fraction of the equilibrium population descended from type $k$.
 The sum $\sum_i{u_i M_{i,k} v_k}$ can be similarly interpreted as the fraction of the equilibrium population descended from individuals that transitioned from type $k$ to another type (via the $k$ column of the transition matrix $M$), conditioned on transition occuring.
 
-Given $0 < D_k \le 1$, from Eq.~\ref{eq:theorem_intermediate} we can get the following corollary.
+Given $0 < D_k \le 1$, from Eq.~\ref{mfp_intermediate} we can get the following corollary.
 
 \begin{corollary}[Reproductive value principle]
-In the notation of the theorem,
+In the notation of Lemma~\ref{lemma:mfp},
 \begin{equation}
@@
 
-We turn our attention to a modifier that concurrently controls the transition out of a subset of types, $K$.
-For example, \citet[Appendix~B]{Ram2012} considered individuals that are grouped by the number of their accumulated deleterious mutations, $k$, and a modifier that regulates the mutation rates in individuals with at least $\pi$ deleterious mutations.
+We turn our attention to a modifier allele that concurrently controls the transition out of a subset of types, $K$.
+For example, \citet[Appendix~B]{Ram2012} considered individuals that are grouped by the number of their accumulated deleterious mutations, $k$, 
+and a modifier allele that regulates the mutation rates in individuals with at least $\pi$ deleterious mutations.
 Similarly, if the transition rates are controlled by a threshold rule 
@@
 which can be simplified to
-\begin{multline}\label{eq:sign_theorem_threshold}
+\begin{multline}\label{mfp_sign_threshold}
 sign \frac{\partial \bar{w}^*}{\partial \tau} = 
@@
 \end{multline}
-a generalization of Eq.~\ref{eq:sign_theorem}.
+a generalization of Eq.~\ref{mfp_sign}.
 Therefore, increased transition from types $K$ will increase the stable population mean fitness if the average fitness of individuals descended from types $K$ is below the stable population mean fitness.
@@
 $$
-because $\bar{w}^* = \sum_i{v_i D_i}$ (see~\nameref{sec:AppA}), and $D_i \le 1$ for all $i$.
+because $\bar{w}^* = \sum_{k=1}^{n}{v_k D_k}$ (see~\nameref{sec:AppA}), and $D_k \le 1$ for all $k$.
 
@@
 
-We determined a condition for a modifier that controls that transition rate $C_k$ from type $A_k$ to increase the stable population mean fitness.
-Could such a modifier increase in frequency when initially rare in the population?
-To answer this question we analyze the stability of modifier $m$ with transition rates $C_k$ to the invasion of modifier $M$ with rates $\tilde{C}_k$, as described in Eqs.~\ref{eq:modifier_model}.
+We determined a condition for a modifier allele that controls that transition rate $C_k$ from type $A_k$ to increase the stable population mean fitness.
+Could such an allele increase in frequency when initially rare in the population?
+To answer this question we analyze the stability of resident modifier allele $m$ with transition rates $C_k$ to the invasion of a modifier allele $M$ with rates $\tilde{C}_k$, as described in Eqs.~\ref{eq:modifier_model}.
 
-The equilibrium of Eqs.~\ref{eq:modifier_model} when modifier allele $M$ is absent from the population is $(v, 0)$, where $v$ is given in eq.~\ref{eq:model_equilibrium} and $g_k=0$ for all $k$. 
-The stability of modifier allele $m$ to invasion by allele $M$ is determined by the leading eigenvalue $\lambda$ of the external stability matrix $\cl_{ex}$ of the equilibrium $(v,0)$, which, in turn, is determined by the Jacobian $\cj$ of Eq.~\ref{eq:modifier_model} evaluated at the equilibrium $(v,0)$, where
+The equilibrium of Eqs.~\ref{eq:modifier_model} when modifier allele $M$ is absent from the population is $(v, 0)$, where $v$ is given in Eq.~\ref{eq:model_equilibrium} and $g_k=0$ for all $k$. 
+The stability of allele $m$ to invasion by allele $M$ is determined by the leading eigenvalue $\lambda$ of the external stability matrix $\cl_{ex}$ of the equilibrium $(v,0)$, which, in turn, is determined by the Jacobian $\cj$ of Eq.~\ref{eq:modifier_model} evaluated at the equilibrium $(v,0)$, where
 \begin{equation}
@@
 \begin{theorem}[Evolution of increased genetic variation]
-The sign of the derivative of the leading eigenvalue $\lambda$ of the external stability matrix of $\cl_{ex}$ of modifier $M$, when evaluated at $\tilde{C}_k=C_k$, is
+The sign of the derivative of the leading eigenvalue $\lambda$ of the external stability matrix $\cl_{ex}$, when evaluated at $\tilde{C}_k=C_k$, is
 \begin{equation}
@@
 \end{equation}
-In addition, the leading eigenvalue is $\lambda=1$ when transition rates are equal $\tilde{C}_k=C_k$ for all $k$.
+In addition, the leading eigenvalue is $\lambda=1$ when transition rates of modifier alleles $m$ and $M$ are equal ($\tilde{C}_k=C_k$ for all $k$).
 
-Therefore, an initially rate modifier allele $M$ with transition rates slightly different from the resident modifier $m$ can successfully invade the population ($\lambda>1$) if $M$ increases the probability of transition from types with below-average fitness, thereby increasing the stable population mean fitness.
+Therefore, an initially rare modifier allele $M$ with transition rates slightly different from the resident modifier $m$ can successfully invade the population ($\lambda>1$) if $M$ increases the probability of transition from types with below-average fitness, thereby increasing the stable population mean fitness.
 \end{theorem}
@@
 
-Now, applying Eq.~\ref{eq:sign_theorem} to Eq.~\ref{eq:Lex}, the sign of the derivative of $\lambda$ with respect to $\tilde{C_k}$ is
+Now, applying Lemma~\ref{lemma:mfp} (eq.~\ref{mfp_sign}) to Eq.~\ref{eq:Lex}, the sign of the derivative of $\lambda$ with respect to $\tilde{C_k}$ is
 \begin{equation}
@@
 \end{equation}
-and evaluating this expression at $\lambda=1$ (the case for $\tilde{C}_k = C_k$) gives us
+and evaluating this expression for $\tilde{C}_k = C_k$), i.e. at $\lambda=1$, gives us
 \begin{equation}
@@
 
-We have shown that under constant-viability selection and in an effectively infinite haploid population at mutation-selection or migration-selection equilibrium, the stable population mean fitness increases if below-fitness individuals increase the rate at which variation is generated.
+We have shown that under constant-viability selection and in an effectively infinite haploid population at mutation-selection or migration-selection equilibrium, the stable population mean fitness increases if below-average individuals increase the rate at which variation is generated.
+Furthermore, modifier alleles that induce increased generation of variation in such individuals are favored by natural evolution.
+
 We have given several examples of variation-generating processes for which this principle applies -- mainly mutation and migration -- but our model may apply to other processes as well.
 For example, the \emph{reduction principle} applies to ecological models of dispersal, and~\citet{Gueijman2013} have demonstrated that even in homogeneous environments, fitness-associated dispersal increases the population mean fitness and is favored by selection over uniform dispersal.
-Similarly, if the transmission fidelity of culturally-transmitted traits depends on the type of the type or fitness of the transmitting individual, we expect that our \emph{mean fitness principle} will hold.
+Similarly, if the transmission fidelity of culturally-transmitted traits depends on the type of the type or fitness of the transmitting individual, we expect that our results will hold.
 
-Eq.~\ref{eq:theorem} is a generalization of a result of~\citet[Eq.~4]{Ram2012}.
+Eq.~\ref{mfp} is a generalization of a result of~\citet[Eq.~4]{Ram2012}.
 \citeauthor{Ram2012} modeled the accumulation of mutant alleles in a population evolving at the mutation-selection balance (see \nameref{sec:mut_model_2}).
-Using Eq.~\ref{eq:Caswells_formula} in \nameref{sec:AppB} and a recursion on the ratios of the reproductive values \cite[see][eqs.~A5-6]{Ram2012}, they concluded that at the mutation-selection balance, if individuals with below-average fitness ($D_k < \bar{w}^*$) increase their mutation rate, then the population mean fitness will increase -- a result generalized by our \emph{mean fitness principle} in Eqs.~\ref{eq:sign_theorem} and~\ref{eq:sign_theorem_threshold}.
+Using Eq.~\ref{eq:Caswells_formula} in \nameref{sec:AppB} and a recursion on the ratios of the reproductive values \cite[see][eqs.~A5-6]{Ram2012}, they concluded that at the mutation-selection balance, if individuals with below-average fitness ($D_k < \bar{w}^*$) increase their mutation rate, then the population mean fitness will increase -- a result generalized by our \emph{mean fitness principle} in Eqs.~\ref{mfp_sign} and~\ref{mfp_sign_threshold}.
 
@@
 However, it has been demonstrated that higher mutation, recombination, and migration rates are favored by selection in populations evolving under fluctuating selection; see~\citet{Carja2014} and references therein.
-Here we have provided another general example, which suggests that if maladapted or below-average individuals increase the rate at which variation is generated, the population mean fitness will increase.
+Here we have provided another general example, which suggests that a modifier allele that prompts maladapted or below-average individuals increase the rate at which variation is generated, will be favored by selection and will lead to increased population mean fitness.
 

### 2017-11-07 ms.tex
reorganization and other comments by MWF

@@
 Here we analyze a general model of evolution under constant selection where the rate at which variation is generated can depend on the fitness of the individual.
-We find a general and clear result:
-if a modifier allele increases the rate at which below-average individuals generate variation, then it will increase in frequency and increase the  population mean fitness.
+We find that if a modifier allele increases the rate at which individuals of below-average fitness generate variation, then it will increase in frequency and increase the  population mean fitness.
 This principle applies to phenomena such as stress-induced mutagenesis, condition-dependent dispersal and outcrossing,
-and even to the proverb: \emph{"Necessity is the mother of genetic invention"}.
+and demonstrates the proverb: \emph{"Necessity is the mother of genetic invention"}.
 \end{abstract}
@@
 (i) effectively infinite population size, (ii) constant-viability selection, (iii) a population at an equilibrium, and (iv) \emph{linear variation} -- the equal scaling of transition probabilities by the modifier.
-A departure from the latter assumption occurs if two variation producing processes interact~\citep{Feldman1980,Altenberg2012}.
+A departure from the latter assumption occurs if two variation-producing processes interact~\citep{Feldman1980,Altenberg2012}.
 Departures from the \emph{reduction principle} have also been demonstrated when conditions (i)-(iii) are not met, see for example~\citet{Holsinger1986} and references therein.
@@
 Their analysis was specific to a model that classified individuals by the number of mutant genes in their genotype, similar to models studied by~\citet{Kimura1966a} and~\citet{Haigh1978}.
-Moreover, their argument was based on expected increase of the population mean fitness by alleles that modify the mutation rate, rather than on the invasion success of such modifiers~\citep[i.e., analysis of \emph{evolutionary genetic stability}, see][]{Lessard1990}.
+Moreover, their argument was based on expected increase of the population mean fitness by alleles that modify the mutation rate, rather than on the invasion success of such modifiers~\citep[i.e., analysis of \emph{evolutionary genetic stability}, see][]{Eshel1982, Lessard1990}.
 
@@
 % Model
-\section*{The Models}
+\section*{Models}
 
@@
 \end{equation}
-where $f_k$ is the frequency of type $A_k$ ($f=(f_1, f_2, \ldots, f_n)$ is a stochastic vector: $\sum_{k=1}^n{f_k} = 1$);
+where $f_k$ is the frequency of type $A_k$: $f=(f_1, f_2, \ldots, f_n)$ is a stochastic vector such that $\sum_{k=1}^n{f_k} = 1$;
 $C_k$ is the probability that an individual of type $A_k$ will transition to another type ($C$ is a diagonal non-negative matrix with diagonal elements $C_k$);
@@
 
-The types $A_k$ can represent a single or multiple genetic loci or non-genetic traits.
+The types $A_k$ can represent a single or multiple haploid genetic loci or non-genetic traits.
 Importantly, type transmission is vertical and uni-parental (the type is transmitted from a single parent to the offspring) and is independent of the frequencies of the other types.
@@
 (ii) given a transition out of type $A_k$, the distribution of the destination types $A_j$ is given by $M_{j,k}$ (note the index order).
-
-We focus on a neutral modifier locus in complete linkage to the types $A_k$, with no direct effect on fitness, and whose sole function is to determine $C_k$ the rates of transition out of the different types.
-Importantly, this modifier can determine different rates for different types; that is, $C_j \ne C_i$ for some $i,j$.
+Importantly, different types can have different rates; that is, $C_j \ne C_i$ for some $i,j$.
 The case $C_j = C_j$ for all $i,j$ is covered by the \emph{reduction principle}~\citep[see][]{Altenberg2017}.
 
-% Modifier model
-\paragraph*{Modifier model.}
-Consider the case of two modifier alleles, $m$ and $M$, inducing different transition probabilities $C=\diag{C_1, \ldots, C_n}$ and $\tilde{C}=\diag{\tilde{C}_1, \ldots, \tilde{C}_n}$, respectively.
-The frequencies of type $A_k$ linked to modifier $m$ or $M$ are $f_k$ and $g_k$, respectively, such that $\sum_{k=1}^{n}{(f_k + g_k)}=1$.
-$\bar{w}$ now ensures that $\sum_{k=1}^{n}{(f_k' + g_k')}=1$, 
-and the rest of the model parameters are the same as in Eq.~\ref{eq:model}.
-
-The frequencies in the next generation, $f'$ for allele $m$ and $g'$ for allele $M$, are given by
-\begin{equation}
-\begin{cases}
-\bar{w} f' &= (I-C+MC)D f \\
-\bar{w} g' &= (I-\tilde{C}+M\tilde{C})D g
-\end{cases}.
-\label{eq:modifier_model}
-\end{equation}
-Note that Eq.~\ref{eq:model} is a special case for the absence of allele $M$, with $g_k=0$ for all $k$.
-Here, $\bar{w}=\sum_{k=1}^{n}{(f_k+g_k)D_k}$ is the mean fitness of the entire population (see~\nameref{sec:AppA}).
-
-In the following section we present three particular examples of the model presented in Eq.~\ref{eq:model} for the modification of mutation and migration.
-For simplicity of the presentation we focus on one modifier allele in these examples.
+In the following section we present three examples of the model presented in Eq.~\ref{eq:model} for mutation and migration.
 
 % Mutation model 1
-\paragraph*{Mutation modification model 1.}\label{sec:mut_model_1}
+\paragraph*{Mutation model 1.}\label{sec:mut_model_1}
 
 Here we consider a large population of haploids and a trait determined by a major locus with $n$ possible alleles $A_1, A_2, ..., A_n$.
-Linked to to this locus is a modifier locus that determines the mutation rates $\mu_k$ of individuals with allele $A_k$.
-Specifically, with probability $1-\mu_k$, the allele $A_k$ does not mutate, and with probability $\frac{\mu_k}{n-1}$, the allele $A_k$ mutates to $A_j$ for any $j \ne k$.
-This is an extension of a model studied by~\citet{Altenberg2017}, that allows for the mutation rate of an individual, $\mu_k$, to depend on the identity of the allele $A_k$.
+The mutation rates $\mu_k$ of individuals with allele $A_k$ are potentially different;
+specifically, with probability $1-\mu_k$, the allele $A_k$ does not mutate, and with probability $\frac{\mu_k}{n-1}$, the allele $A_k$ mutates to $A_j$ for any $j \ne k$.
+This is an extension of a model studied by~\citet{Altenberg2017} that allows for the mutation rate of an individual, $\mu_k$, to depend on the identity of the allele $A_k$.
 
@@
 This model can be rephrased in terms of the general model: 
-Eq.~\ref{eq:mutation_model_1} is equivalent to Eq.~\ref{eq:model} if we define
+Eq.~\ref{eq:mutation_model_1} is equivalent to Eq.~\ref{eq:model} if we set
 \begin{equation} 
@@
 % Mutation model 2
-\paragraph*{Mutation modification model 2.}\label{sec:mut_model_2}
+\paragraph*{Mutation model 2.}\label{sec:mut_model_2}
 
 Here, we consider a large population of haploids in which individuals with genotype $A_k$ are characterized by the number $k$ of deleterious or mutant alleles in their genotype, where $0 \le k \le n$.
-A modifier locus then determines the probability $m_k$ that a mutation occurs in individuals with $k$ mutant alleles.
-In this case, we assume a low mutation rate $m_k$ so that two mutations are unlikely to occur in the same individual in one generation.
+Importantly, the probability $m_k$ that a mutation occurs in individuals with $k$ mutant alleles depends on $k$.
+We assume a low mutation rate $m_k$ so that two mutations are unlikely to occur in the same individual in one generation.
 When a mutation occurs it is \emph{deleterious} with probability $\delta$, generating a mutant allele and converting the individual from $A_k$ to $A_{k+1}$, or it is beneficial with probability $\beta$, converting the individual from $A_k$ to $A_{k-1}$.
@@
 
-\paragraph*{Migration modification model.}\label{sec:mig_model_1}
+% Migration model
+\paragraph*{Migration model.}\label{sec:mig_model_1}
 
-Here we assume a large population of haploids that occupies two demes and a modifier locus that determine the migration rates between the demes, $m_k$. 
+Here we assume a large population of haploids that occupies two demes.
 Let the fitness of genotypes $A_k$ be $w_k$ in deme 1 and $q_k$ in deme 2 for $k=1,2,\ldots,n$.
+The probability that an individual with genotype $A_k$ migrates between the deme is $m_k$; that is, migration rates depend on the genotypes.
 Let the frequencies of $A_k$ be $x_k$ in deme 1 and $y_k$ in deme 2 for $k=1,2,\ldots,n$ with $x_k \ge 0$, $y_k \ge 0$, $\sum_{k=1}^{n}{x_k} = 1$, and $\sum_{k=1}^{n}{y_k} = 1$.
 
-Then following selection and migration, after one generation the new frequencies $x_k'$ and $y_k'$ are given by
+Following selection and migration the new frequencies $x_k'$ and $y_k'$ are given by
 \begin{align}
@@
 for $k=1,2,\ldots,n$, where $\bar{w} = \sum_{k \ge 0}{w_k x_k}$ and $\bar{q} = \sum_{k \ge 0}{q_k y_k}$ are the population mean fitnesses in deme 1 and 2, respectively.
-This is an extension of a model presented by~\citet{Altenberg2017}, allowing for the migration rate of an individual, $m_k$, to depend on the identity of the allele $A_k$.
+This is an extension of a model presented by~\citet{Altenberg2017}.
 
@@
 \section*{Results}
+
+\subsection*{Mean fitness principle}
+
+We first focus on the stable population mean fitness.
+We will show that if the transition rate from types with below-average fitness increases, then the stable population mean fitness increases, too.
 
@@
 The above lemma provides a condition for the effect of changing $C_k$, the probability for transition from $A_k$, on the stable population mean fitness.
-Specifically, if $A_k$ individuals have below-average fitness, then a modifier that increases $C_k$ will increase the population mean fitness.
+Specifically, if $A_k$ individuals have below-average fitness, then increasing $C_k$ will increase the population mean fitness.
 
@@
 
-Given $0 < D_k \le 1$, from Eq.~\ref{mfp_intermediate} we can get the following corollary.
+Given $0 < D_k \le 1$, from Eq.~\ref{mfp_intermediate} we have the following corollary.
 
@@
 sign \frac{\partial \bar{w}^*}{\partial C_k} = 
-sign (u\tr [M]_k - u_k),
+sign (u\tr [M]_k - u_k);
 \end{equation}
-and therefore increased transition out of type $k$ will increase the stable population mean fitness if the fraction of the population descended from type $k$ is expected to increase due to a transition to another type.
+therefore, increased transition from type $k$ will increase the stable population mean fitness if the fraction of the population descended from type $k$ is expected to increase due to a transition to another type.
 \end{corollary}
@@
 
-We turn our attention to a modifier allele that concurrently controls the transition out of a subset of types, $K$.
+We turn our attention to the case where transition from a subset of types, $K$, is concurrently controlled.
 For example, \citet[Appendix~B]{Ram2012} considered individuals that are grouped by the number of their accumulated deleterious mutations, $k$, 
@@
 where $\tau > 0$, then the sensitivity of the stable population mean fitness to changes in the
-rate of transition from types $K$ can be written 
+rate of transition from types in $K$ can be written 
 \begin{equation}
@@
 u\tr \Big( \sum_{k \in K}{\frac{\partial (I - C + MC)D}{\partial \tau}} \Big) v = \\
-\frac{1}{\tau} \sum_{k \in K}{u_k v_k (\bar{w}^* - D_k)},
+\frac{1}{\tau} \sum_{k \in K}{u_k v_k (\bar{w}^* - D_k)}.
 \end{aligned}
 \end{equation}
-which can be simplified to
+Then
 \begin{multline}\label{mfp_sign_threshold}
@@
 a generalization of Eq.~\ref{mfp_sign}.
-Therefore, increased transition from types $K$ will increase the stable population mean fitness if the average fitness of individuals descended from types $K$ is below the stable population mean fitness.
+Therefore, increased transition from types in $K$ will increase the stable population mean fitness if the average fitness of individuals descended from types in $K$ is below the stable population mean fitness.
 
-If such a modifier has the same effect on all individuals (i.e., $C_k = \alpha$ for all $k$) then Eq.~\ref{eq:model_equilibrium} $\bar{w}^* v = ((1-\alpha)I + \alpha M)Dv$,
+\subsection*{Evolutionary genetic stability}
+
+% Modifier model
+We now focus on a neutral modifier locus in complete linkage to the types $A_k$, with no direct effect on fitness, and whose sole function is to determine $C_k$, the rates of transition from the different types.
+We will show that modifier alleles that increase the stable population mean fitness in accordance with Lemma~\ref{lemma:mfp} are favored by natural selection.
+
+\paragraph*{Modifier model.}
+Consider the case of two modifier alleles, $m$ and $M$, inducing different transition probabilities $C=\diag{C_1, \ldots, C_n}$ and $\tilde{C}=\diag{\tilde{C}_1, \ldots, \tilde{C}_n}$, respectively.
+The frequencies of type $A_k$ linked to modifier $m$ or $M$ are $f_k$ and $g_k$, respectively, such that $\sum_{k=1}^{n}{(f_k + g_k)}=1$.
+$\bar{w}$ now ensures that $\sum_{k=1}^{n}{(f_k' + g_k')}=1$, 
+and the rest of the model parameters are the same as in Eq.~\ref{eq:model}.
+
+The frequencies in the next generation, $f'$ for allele $m$ and $g'$ for allele $M$, are given by
+\begin{equation}
+\begin{cases}
+\bar{w} f' &= (I-C+MC)D f \\
+\bar{w} g' &= (I-\tilde{C}+M\tilde{C})D g
+\end{cases}.
+\label{eq:modifier_model}
+\end{equation}
+Here, $\bar{w}=\sum_{k=1}^{n}{(f_k+g_k)D_k}$ is the mean fitness of the entire population (see~\nameref{sec:AppA}). 
+Note that Eq.~\ref{eq:model} is the special case of Eq.~\ref{eq:modifier_model} where allele $M$ is absent; i.e. $g_k=0$ for all $k$.
+
+% Evolutionary stability
+
+We havefound a condition for a modifier allele that controls that transition rate $C_k$ from type $A_k$ to increase the stable population mean fitness (Lemma~\ref{lemma:mfp}).
+Could such an allele increase in frequency when initially rare in the population?
+To answer this we analyze the stability of resident modifier allele $m$ with transition rates $C_k$ to the invasion of a modifier allele $M$ with rates $\tilde{C}_k$~\ref{eq:modifier_model}~\citep[for an introduction to \emph{evolutionary invasion analysis}, see][ch.~12]{Otto2007}
+
+The equilibrium of Eqs.~\ref{eq:modifier_model} when modifier allele $M$ is absent from the population is $(v, 0)$, where $v$ is given in Eq.~\ref{eq:model_equilibrium} and $g_k=0$ for all $k$. 
+The stability of allele $m$ to invasion by allele $M$ is determined by the leading eigenvalue $\lambda_1$ of $\cl_{ex}$ the external stability matrix of the equilibrium $(v,0)$, which, in turn, is determined by the Jacobian $\cj$ of Eq.~\ref{eq:modifier_model} evaluated at the equilibrium $(v,0)$, where
+\begin{equation}
+\cj = \begin{pmatrix}\cl_{in} & 0 \\ 0 & \cl_{ex} \end{pmatrix},
+\end{equation}
+and $\cl_{in}$ is the local stability matrix of the equilibrium $(v,0)$.
+The zero block matrices are due to the complete linkage between the modifier and the types $A_k$ and to the lack of transition (i.e., mutation) between the modifier alleles.
+
+$\cl_{ex}$ can be written as
+\begin{equation}\label{eq:Lex}
+\cl_{ex} = (I - \tilde{C} + M \tilde{C}) D /\bar{w}^*,
+\end{equation}
+where $\bar{w}^*$ is the stable population mean fitness in the absence of modifier allele $M$ (see~\nameref{sec:AppA}).
+Note that $\bar{w}^*$ is not an eigenvalue of $\cl_{ex}$.
+
+\begin{theorem}[Evolution of increased genetic variation]
+Let $\lambda_1$ be the The leading eigenvalue of the external stability matrix $\cl_{ex}$.
+If the transition rates induced by the modifier alleles $m$ and $M$ are equal, i.e., $\tilde{C}_k=C_k$ for all $k$, then the following applies
+\begin{equation}
+\begin{aligned}
+\lambda_1=1, \\
+sign \frac{\partial \lambda_1}{\partial \tilde{C}_k}\bigg\rvert_{\tilde{C}_k = C_k} = 
+sign(\bar{w}^* - D_k).
+\end{aligned}
+\end{equation}
+
+Therefore, an initially rare modifier allele $M$ with transition rates slightly different from the resident allele $m$ can successfully invade the population ($\lambda_1>1$) if $M$ increases the probability of transition from types with below-average fitness, thereby increasing the stable population mean fitness.
+\end{theorem}
+
+\begin{proof}
+Substituting $\tilde{C}=C$ in Eq.~\ref{eq:Lex} and multiplying both sides by $\bar{w}^*$,
+\begin{equation}
+\bar{w}^* L_{ex} = (I-C+MC) D,
+\end{equation}
+and since $\bar{w}^*$ is the leading eigenvalue of the RHS (see Eq.~\ref{eq:model_equilibrium} and \nameref{sec:AppA}),
+the leading eigenvalue of $\cl_{ex}$ is $\lambda_1=1$.
+
+Now, applying Lemma~\ref{lemma:mfp} (eq.~\ref{mfp_sign}) to Eq.~\ref{eq:Lex}, the sign of the derivative of $\lambda_1$ with respect to $\tilde{C_k}$ is
+\begin{equation}
+sign \frac{\partial \lambda_1}{\partial \tilde{C}_k} =
+sign\Big(\lambda_1 - \frac{D_k}{\bar{w}^*}\Big).
+\end{equation}
+Evaluating this expression for $\tilde{C}_k = C_k$, i.e. at $\lambda_1=1$, gives us
+\begin{multline}
+sign \frac{\partial \lambda_1}{\partial \tilde{C}_k}\bigg\rvert_{\tilde{C}_k = C_k} = 
+sign\Big(1 - \frac{D_k}{\bar{w}^*}\Big) =  
+sign(\bar{w}^* - D_k),
+\end{multline}
+since $\bar{w}^*>0$.
+This completes the proof.
+\end{proof}
+
+Note that if the modifier has the same effect on all individuals (i.e., $C_k = \alpha$ and $\tilde{C}_k = \alpha^*$ for all $k$) then $\cl_{ex}$ becomes
+\begin{equation}
+\cl_{ex} = ((1 - \alpha^*)I + \alpha^* M) D /\bar{w}^*,
+\end{equation}  
 and the \emph{unified reduction principle} is in effect~\cite[eqs. 65, 72]{Altenberg2017}.
-Therefore, we can expect modifiers that reduce $\alpha$ to be favored by natural selection.
-Indeed, setting $C = \alpha I$ in Eq.~\ref{eq:model_equilibrium} and proceeding as in the above proof, we find
+Therefore, we can expect that with these modifiers, an allele $M$ that reduces the transition rate, i.e. $\alpha^* < \alpha$, will be favored by natural selection.
+Indeed, setting $C = \alpha I$ in Eq.~\ref{eq:model_equilibrium} and proceeding as in the above, we find
 $$
@@
 
-% Evolutionary stability
-
-We determined a condition for a modifier allele that controls that transition rate $C_k$ from type $A_k$ to increase the stable population mean fitness.
-Could such an allele increase in frequency when initially rare in the population?
-To answer this question we analyze the stability of resident modifier allele $m$ with transition rates $C_k$ to the invasion of a modifier allele $M$ with rates $\tilde{C}_k$, as described in Eqs.~\ref{eq:modifier_model}.
-
-The equilibrium of Eqs.~\ref{eq:modifier_model} when modifier allele $M$ is absent from the population is $(v, 0)$, where $v$ is given in Eq.~\ref{eq:model_equilibrium} and $g_k=0$ for all $k$. 
-The stability of allele $m$ to invasion by allele $M$ is determined by the leading eigenvalue $\lambda$ of the external stability matrix $\cl_{ex}$ of the equilibrium $(v,0)$, which, in turn, is determined by the Jacobian $\cj$ of Eq.~\ref{eq:modifier_model} evaluated at the equilibrium $(v,0)$, where
-\begin{equation}
-\cj = \begin{pmatrix}\cl_{in} & 0 \\ 0 & \cl_{ex} \end{pmatrix}.
-\end{equation}
-The zero block matrices are due to complete linkage between the modifier and the types $A_k$.
-
-$\cl_{ex}$ can be written as
-\begin{equation}\label{eq:Lex}
-\cl_{ex} = (I - \tilde{C} + M \tilde{C}) D /\bar{w}^*,
-\end{equation}
-where $\bar{w}^*$ is the stable population mean fitness in the absence of modifier allele $M$ (see~\nameref{sec:AppA}).
-
-\begin{theorem}[Evolution of increased genetic variation]
-The sign of the derivative of the leading eigenvalue $\lambda$ of the external stability matrix $\cl_{ex}$, when evaluated at $\tilde{C}_k=C_k$, is
-\begin{equation}
-sign \frac{\partial \lambda}{\partial \tilde{C}_k}\bigg\rvert_{\tilde{C}_k = C_k} = 
-sign(\bar{w}^* - D_k).
-\end{equation}
-In addition, the leading eigenvalue is $\lambda=1$ when transition rates of modifier alleles $m$ and $M$ are equal ($\tilde{C}_k=C_k$ for all $k$).
-
-Therefore, an initially rare modifier allele $M$ with transition rates slightly different from the resident modifier $m$ can successfully invade the population ($\lambda>1$) if $M$ increases the probability of transition from types with below-average fitness, thereby increasing the stable population mean fitness.
-\end{theorem}
-
-\begin{proof}
-First, consider the case the $\tilde{C_k} = C_k$ for all $k$.
-Substituting $\tilde{C}=C$ in Eq.~\ref{eq:Lex} and multiplying both sides by $\bar{w}^*$,
-\begin{equation}
-\bar{w}^* L_{ex} = (I-C+MC) D,
-\end{equation}
-and since $\bar{w}^*$ is the leading eigenvalue of the RHS (\nameref{sec:AppA}),
-the leading eigenvalue of $\cl_{ex}$ is $\lambda=1$.
-
-Now, applying Lemma~\ref{lemma:mfp} (eq.~\ref{mfp_sign}) to Eq.~\ref{eq:Lex}, the sign of the derivative of $\lambda$ with respect to $\tilde{C_k}$ is
-\begin{equation}
-sign \frac{\partial \lambda}{\partial \tilde{C}_k} =
-sign\Big(\lambda - \frac{D_k}{\bar{w}^*}\Big),
-\end{equation}
-and evaluating this expression for $\tilde{C}_k = C_k$), i.e. at $\lambda=1$, gives us
-\begin{equation}
-sign \frac{\partial \lambda}{\partial \tilde{C}_k}\bigg\rvert_{\lambda=1} = 
-sign(\bar{w}^* - D_k),
-\end{equation}
-since $\bar{w}^*>0$.
-This completes the proof.
-\end{proof}
-
 % Discussion
@@
 
-We have shown that under constant-viability selection and in an effectively infinite haploid population at mutation-selection or migration-selection equilibrium, the stable population mean fitness increases if below-average individuals increase the rate at which variation is generated.
+We have shown that under constant-viability selection and in an effectively infinite haploid population at mutation-selection or migration-selection equilibrium, the stable population mean fitness increases if individuals with below-average fitness increase the rate at which variation is generated.
 Furthermore, modifier alleles that induce increased generation of variation in such individuals are favored by natural evolution.
 
-We have given several examples of variation-generating processes for which this principle applies -- mainly mutation and migration -- but our model may apply to other processes as well.
+We have given several examples of variation-generating processes for which this principle applies -- namely mutation and migration -- but our model may apply to other processes as well.
 For example, the \emph{reduction principle} applies to ecological models of dispersal, and~\citet{Gueijman2013} have demonstrated that even in homogeneous environments, fitness-associated dispersal increases the population mean fitness and is favored by selection over uniform dispersal.

### 2017-11-07 get_bibtex.py
fix species names

@@ -1,7 +1,12 @@
+import re
+
 from bibtexparser import load as load_bib
 from bibtexparser.bwriter import BibTexWriter
 from bibtexparser.bibdatabase import BibDatabase
 import click
+
+species_name_pattern = re.compile(
+	r"^(.*?)\{\\textless\}i\{\\textgreater\}(.*?)\{\\textless\}\/i\{\\textgreater\}(.*?)$")
 
 @click.command()
 @click.argument('keys_filename', type=click.Path(exists=True, file_okay=True, dir_okay=False, readable=True))
@@ -26,6 +31,13 @@ def main(keys_filename, bibtex_filename, output_filename, verbose):
 	
 	if verbose:
 		print("Writing {} entries to {}".format(len(out_bib.entries), output_filename))
+
+	for ent in out_bib.entries:
+		m = species_name_pattern.match(ent['title'])
+		if m:
+			prefix, species, postfix = m.groups()
+			ent['title'] = prefix + r"\emph{" + species + r"}" + postfix
+
 	writer = BibTexWriter()
 	with open(output_filename, 'w') as f:
 	    f.write(writer.write(out_bib))

### 2017-11-08 ms.tex
MWF comments; UL comment on equilibrium existence

@@
 $C_k$ is the probability that an individual of type $A_k$ will transition to another type ($C$ is a diagonal non-negative matrix with diagonal elements $C_k$);
-$M_{i,j}$ is the transition probability from type $A_j$ to type $A_i$ given that an individual of type $A_j$ is transitioning ($M$ is a column-stochastic matrix: $\sum_{j=1}^n {M_{i,j}} = 1$ for all \emph{j});
+$M_{i,j}$ is the transition probability from type $A_j$ to type $A_i$ given that an individual of type $A_j$ is transitioning ($M$ is a column-stochastic matrix: $M_{i,j} \ge 0$ for all $i,j$ and $\sum_{j=1}^n {M_{i,j}} = 1$ for all $j$);
 $D_k$ is the relative fitness of type $A_k$ ($D$ is a diagonal matrix with diagonal elements $D_k$, where we can assume $0 < D_k \le 1$, i.e., fitness values are relative to the fitness of the fittest type);
@@
 % Mean fitness principle
+
 Write the equilibrium of the frequency vector $f$ in Eq.~\ref{eq:model} as $v$ and the stable population mean fitness as $\bar{w}^*$, then
@@
 \end{equation}
+The existence of the leading eigenvalue $\bar{w}^*$ and its corresponding right eigenvector $v$ are guaranteed by the \emph{Perron-Frobenius theorem}~\cite[Appendix~A]{Otto2007} because $(I-C+MC)D$ is a non-negative matrix
+(the leading eigenvalue $\bar{w}^*$ is identified as the stable population mean fitness in~\nameref{sec:AppA}).
 
@@
 \begin{proof}
-By the \emph{Perron-Frobenius theorem}~\cite[Appendix~A]{Otto2007},
-$\bar{w}^*$ is the leading real positive eigenvalue of $(I-C+MC)D$,
+$\bar{w}^*$ is the leading (real positive) eigenvalue of $(I-C+MC)D$ (see eq.~\ref{eq:model_equilibrium}),
 and the corresponding left and right eigenvectors $u$ and $v$ are both non-negative and uniquely defined by $\sum_k{v_k} = 1$ and $\sum_k{u_k v_k} = 1$.
@@
 
-We turn our attention to the case where transition from a subset of types, $K$, is concurrently controlled.
+We turn our attention to the case where the transition occurs from a subset $K$ of the types.
 For example, \citet[Appendix~B]{Ram2012} considered individuals that are grouped by the number of their accumulated deleterious mutations, $k$, 
@@
 1, & k \not\in K
-\end{cases},
+\end{cases}.
 \end{equation}
-where $\tau > 0$, then the sensitivity of the stable population mean fitness to changes in the
+If $\tau > 0$, then the sensitivity of the stable population mean fitness to changes in the
 rate of transition from types in $K$ can be written 
@@
 \end{equation}
-Then
+Thus
 \begin{multline}\label{mfp_sign_threshold}
@@
 
-We havefound a condition for a modifier allele that controls that transition rate $C_k$ from type $A_k$ to increase the stable population mean fitness (Lemma~\ref{lemma:mfp}).
+We have found a condition for a modifier allele that controls that transition rate $C_k$ from type $A_k$ to increase the stable population mean fitness (Lemma~\ref{lemma:mfp}).
 Could such an allele increase in frequency when initially rare in the population?
-To answer this we analyze the stability of resident modifier allele $m$ with transition rates $C_k$ to the invasion of a modifier allele $M$ with rates $\tilde{C}_k$~\ref{eq:modifier_model}~\citep[for an introduction to \emph{evolutionary invasion analysis}, see][ch.~12]{Otto2007}
+To answer this we analyze the stability of resident modifier allele $m$ with transition rates $C_k$ to the invasion by a modifier allele $M$ with rates $\tilde{C}_k$, see Eq.~\ref{eq:modifier_model}~\citep[for an introduction to \emph{evolutionary invasion analysis}, see][ch.~12]{Otto2007}
 
 The equilibrium of Eqs.~\ref{eq:modifier_model} when modifier allele $M$ is absent from the population is $(v, 0)$, where $v$ is given in Eq.~\ref{eq:model_equilibrium} and $g_k=0$ for all $k$. 
-The stability of allele $m$ to invasion by allele $M$ is determined by the leading eigenvalue $\lambda_1$ of $\cl_{ex}$ the external stability matrix of the equilibrium $(v,0)$, which, in turn, is determined by the Jacobian $\cj$ of Eq.~\ref{eq:modifier_model} evaluated at the equilibrium $(v,0)$, where
+The stability of allele $m$ to invasion by allele $M$ is determined by the leading eigenvalue $\lambda_1$ of $\cl_{ex}$ the external stability matrix of the equilibrium $(v,0)$, which, in turn, is determined by the Jacobian $\cj$ of the system in Eq.~\ref{eq:modifier_model} evaluated at the equilibrium $(v,0)$, where
 \begin{equation}
@@
 \end{equation}
-and $\cl_{in}$ is the local stability matrix of the equilibrium $(v,0)$.
+and $\cl_{in}$ is the local stability matrix of the equilibrium $(v,0)$ in the space $\sum_{k=1}^n{v_k}=1$.
 The zero block matrices are due to the complete linkage between the modifier and the types $A_k$ and to the lack of transition (i.e., mutation) between the modifier alleles.
@@
 where $\bar{w}^*$ is the stable population mean fitness in the absence of modifier allele $M$ (see~\nameref{sec:AppA}).
-Note that $\bar{w}^*$ is not an eigenvalue of $\cl_{ex}$.
+Note that, in general, $\bar{w}^*$ is not an eigenvalue of $\cl_{ex}$.
 
 \begin{theorem}[Evolution of increased genetic variation]
-Let $\lambda_1$ be the The leading eigenvalue of the external stability matrix $\cl_{ex}$.
+Let $\lambda_1$ be the the leading eigenvalue of the external stability matrix $\cl_{ex}$.
 If the transition rates induced by the modifier alleles $m$ and $M$ are equal, i.e., $\tilde{C}_k=C_k$ for all $k$, then the following applies
 \begin{equation}
-\begin{aligned}
-\lambda_1=1, \\
+\lambda_1=1,
+\end{equation}
+and
+\begin{equation}
 sign \frac{\partial \lambda_1}{\partial \tilde{C}_k}\bigg\rvert_{\tilde{C}_k = C_k} = 
 sign(\bar{w}^* - D_k).
-\end{aligned}
 \end{equation}
@@
 \begin{equation}
-\bar{w}^* L_{ex} = (I-C+MC) D,
+\bar{w}^* \cl_{ex}\big\rvert_{\tilde{C}_k = C_k} = (I-C+MC) D,
 \end{equation}
 and since $\bar{w}^*$ is the leading eigenvalue of the RHS (see Eq.~\ref{eq:model_equilibrium} and \nameref{sec:AppA}),
-the leading eigenvalue of $\cl_{ex}$ is $\lambda_1=1$.
+the leading eigenvalue of $\cl_{ex}\big\rvert_{\tilde{C}_k = C_k}$ is $\lambda_1=1$.
 
@@
 \end{equation}
-Evaluating this expression for $\tilde{C}_k = C_k$, i.e. at $\lambda_1=1$, gives us
+Evaluating this expression at $\tilde{C}_k = C_k$, i.e. at $\lambda_1=1$, gives us
 \begin{multline}

### 2017-11-09 ms.tex
inner stability of equilibrium

@@
 \usepackage{amssymb,amsmath,amsthm}
+\usepackage{commath}
 \usepackage{longtable}
@@
 \end{equation}
-The existence of the leading eigenvalue $\bar{w}^*$ and its corresponding right eigenvector $v$ are guaranteed by the \emph{Perron-Frobenius theorem}~\cite[Appendix~A]{Otto2007} because $(I-C+MC)D$ is a non-negative matrix
-(the leading eigenvalue $\bar{w}^*$ is identified as the stable population mean fitness in~\nameref{sec:AppA}).
+Note that (i) the existence of $\bar{w}^*$ and $v$ are guaranteed by the \emph{Perron-Frobenius theorem}~\cite[Appendix~A]{Otto2007} because $(I-C+MC)D$ is a non-negative matrix; 
+(ii) the local stability of this equilibrium is proven in~\nameref{sec:AppD}; and
+(iii) $\bar{w}^*$ is identified as the stable population mean fitness in~\nameref{sec:AppA}.
 
@@
 
+% Appendix D: inner stability
+\subsection*{Appendix D}\label{sec:AppD}
+\begin{lemma}[Stability of the equilibrium $v$]
+The equilibrium $v$ of the system in Eq.~\ref{eq:model_equilibrium} is locally stable.
+In addition,
+\begin{enumerate}
+\item If $M$ is irreducible and $D$ and $C$ are positive, then the equilibrium is globally stable.
+\item If $M$ is irreducible and either $D$ or $C$ are non-negative but not strictly positive, then the equilibrium is globally stable except possibly for initial conditions on the boundary, i.e., when starting with zero elements in the initial frequency vector.
+\end{enumerate}
+\end{lemma}
+\begin{proof}
+Set $A=(I-C+MC)D$ and rewrite Eq.~\ref{eq:model} such that $f(t)$ is the frequency vector at time $t$ and
+\begin{equation}
+f(t) = \frac{Af(t-1)}{\norm{Af(t-1)}},
+\end{equation}
+where $\norm{f(t)} = \sum_{k=1}^{n}f_k(t)$.
+Denote the eigenvalues of $A$ as $\lambda_1, \ldots, \lambda_n$ such that $\abs{\lambda_1} \ge \ldots \ge \abs{\lambda_n}$ and $\lambda_1$ is the \emph{Perron} leading eigenvalue, that is, the largest real eigenvalue with corresponding real non-negative right eigenvector $v$ ($\lambda_1$ is shown in~\nameref{sec:AppA} to be the stable population mean fitness $\bar{w}^*$).
+
+First, note that the normalization $x \goesto \frac{1}{\norm{x}}$ can be applied once instead of $t$ times~\citep[Appendix~A]{Wakano2004}, that is,
+\begin{equation}
+f(t) = \frac{A^t f(0)}{\norm{A^t f(0)}}.
+\label{eq:ftplus1}
+\end{equation}
+
+We write $f(0)$ as a linear combination of eigenvectors $v^k$ of $A$ for $k=1, \ldots, n$:
+\begin{equation}
+f(0) = a_1 v^1 + a_2 v^2 + \ldots + a_n v^n,
+\label{eq:linear_combination}
+\end{equation}
+such that the eigenvector corresponding to the leading eigenvalue is $v=v_1$.
+
+To proceed we need to make sure that $a_1 \ne 0$.
+If $M$ is reducible then $A$ is reducible. 
+In this case, $a_1 \ne 0$ can be guaranteed by setting $f(0)=v + \epsilon$ such that $\epsilon_k$ is small enough and $v_k+\epsilon_k \ge 0$ for all $k$, and $\sum_{k=1}^{n}{\epsilon_k} = 0$.
+
+If $M$ is irreducible and $D$ and $C$ are positive, then $A$ is irreducible and  any choice of $f(0)$ guarantees $a_1 \ne 0$.
+
+If $M$ is reducible and either $D$ or $C$ are non-negative but not strictly positive, than any choice of $f(0)$ such that $f_k(0)>0$ for all $k$ guarantees $a_1 \ne 0$.
+
+We write, using Eq.~\ref{eq:linear_combination},
+\begin{multline}
+A^t f(0) = 
+a_1 A^t v^1 + a_2 A^t v^2 + \ldots + a_n A^t v^n = \\
+a_1 \lambda_1^t v^1 + a_2 \lambda_2^t v^2 + \ldots + a_n \lambda_n^t v^n = \\
+\lambda_1^t \bigg(a_ 1v^1 + a_2 \bigg(\frac{\lambda_2}{\lambda_1}\bigg)^t v^2 + \ldots +  a_n \bigg(\frac{\lambda_n}{\lambda_1}\bigg)^t v^n\bigg).
+\end{multline}
+Therefore,
+\begin{equation}
+\frac{A^t f(0)}{\lambda_1^t} \goesto a_1 v_1
+\end{equation}
+as $t \goesto \infty$, because $\abs{\frac{\lambda_k}{\lambda_1}} < 1$
+for all $k=2, \ldots, n$.
+Finally, 
+\begin{equation}
+f(t) = 
+\frac{A^t f(0)}{\norm{A^t f(0)}} =
+\frac{\frac{A^t f(0)}{\lambda_1^t}}{\norm{\frac{A^t f(0)}{\lambda_1^t}}} \goesto
+\frac{a_1 v^1}{\norm{a_1 v^1}} = v,
+\end{equation}
+as $t \goesto \infty$, which completes the proof.
+\end{proof}
+
 % Acknowledgements
@@
 \bibliographystyle{agsm}
-%\bibliography{/Users/yoavram/Documents/library}
-\bibliography{ms}
+\bibliography{/Users/yoavram/Documents/library}
+%\bibliography{ms}
 

### 2017-11-10 ms.tex
edited OK comments

@@
 We find that if a modifier allele increases the rate at which individuals of below-average fitness generate variation, then it will increase in frequency and increase the  population mean fitness.
-This principle applies to phenomena such as stress-induced mutagenesis, condition-dependent dispersal and outcrossing,
-and demonstrates the proverb: \emph{"Necessity is the mother of genetic invention"}.
+This principle applies to phenomena such as stress-induced mutagenesis, condition-dependent dispersal and fitness-associated outcrossing,
+and exhibits the paraphrase: \emph{"Necessity is the mother of genetic invention."}
 \end{abstract}
@@
 
-According to the \emph{reduction principle}, in populations that evolve near an equilibrium between natural selection and a process that generates variation (i.e.
+According to the \emph{reduction principle}, in populations at a balance between natural selection and a process that generates variation (i.e.
 mutation, migration, or recombination), selection favors neutral modifiers that decrease the rate at which variation is generated.
-
 The \emph{reduction principle} was demonstrated for modifiers of the recombination~\citep{Feldman1972}, mutation, and migration~\citep{Liberman1986a}, and dispersal rates~\citep{Altenberg1987}.
@@
 
-Another departure from the \emph{linear variation} assumption of the \emph{reduction principle} for mutation rates involves \emph{stress-induced mutagenesis}, a mechanism in which the mutation rate increases in individuals with poor fitness.
-Empirical evidence suggests \emph{stress-induced mutagenesis} is common in bacteria and yeast, and may be prevalent in plants, flies, and human cancer cells~\citep{Rosenberg2012,Fitzgerald2017b}.
-\citet{Ram2012} demonstrated that modifiers that increase the mutation rate of individuals with below-average fitness actually \textbf{increase} the population mean fitness, rather than decrease it.
-Their analysis assumed an infinite population size and constant-viability selection, where the only departure from the \emph{reduction principle} assumptions is due to the differential effect the modifier has on different genotypes.
+Another departure from the \emph{linear variation} assumption of the \emph{reduction principle} for mutation rates involves a mechanism in which the mutation rate increases in individuals with poor fitness -- a mechanism first observed in stressed bacteria~\citep{Foster2007}, although not in a constant environment.
+\citet{Ram2012} demonstrated that even in a constant environment, modifiers that increase the mutation rate of individuals with below-average fitness increase the population mean fitness, rather than decrease it.
+Their analysis assumed an infinite population size and constant-viability selection in which fitness is determined by the number of mutant alleles accumulated in the genotype.
+The only departure from the \emph{reduction principle} assumptions was due to the differential effect the modifier has on different genotypes.
 A similar result has been demonstrated for fitness-associated recombination~\citep{Hadany2003a} and for condition-dependent sexual reproduction~\citep{Hadany2007a}, and evidence suggests that both mechanisms are common in nature~\citep{Ram2016}.
@@
 
-Here, we present a general evolutionary model where the rate at which variation is generated depends on the type of the individual, and specifically on its fitness.
-Our results show that the population mean fitness increases if below-average individuals produce more variation than above-average individuals,
-and that modifier alleles that induce below-average to produce more variation are favored by natural selection.
+Here, we present a general evolutionary model where the the type of the individual determines both its fitness and the rate at which it generates variation.
+Our results show that the population mean fitness increases if individuals with below-average fitness produce more variation than individuals with above-average fitness,
+and that modifier alleles that induce below-average individuals to produce more variation are favored by natural selection.
 
@@
 Consider a large haploid population with an arbitrary set of types $A_1, A_2, \ldots, A_n$.
-The change in the frequencies of these types is described by the transformation $f \to f'$: 
+The frequency and fitness of individuals of type $A_k$ are $f_k$ and $D_k$, respectively.
+The probability that an individual of type $A_k$ will transition to some other type is $C_k$, and given a transition occurs, the probability that it will transition to type $A_i$ is $M_{i,k}$.
+Therefore, the change in the frequencies of types $A_k$ is described by the transformation $f \to f'$: 
 \begin{equation}\label{eq:model}
@@
 \end{equation}
-where $f_k$ is the frequency of type $A_k$: $f=(f_1, f_2, \ldots, f_n)$ is a stochastic vector such that $\sum_{k=1}^n{f_k} = 1$;
-$C_k$ is the probability that an individual of type $A_k$ will transition to another type ($C$ is a diagonal non-negative matrix with diagonal elements $C_k$);
-$M_{i,j}$ is the transition probability from type $A_j$ to type $A_i$ given that an individual of type $A_j$ is transitioning ($M$ is a column-stochastic matrix: $M_{i,j} \ge 0$ for all $i,j$ and $\sum_{j=1}^n {M_{i,j}} = 1$ for all $j$);
-$D_k$ is the relative fitness of type $A_k$ ($D$ is a diagonal matrix with diagonal elements $D_k$, where we can assume $0 < D_k \le 1$, i.e., fitness values are relative to the fitness of the fittest type);
+where $f=(f_1, f_2, \ldots, f_n)$ is a stochastic vector such that $\sum_{k=1}^n{f_k} = 1$;
+$D$ is a diagonal matrix with diagonal elements $D_k$, where we can assume $0 < D_k \le 1$, i.e., fitness values are relative to the fitness of the fittest type; 
+$C$ is a diagonal non-negative matrix with diagonal elements $C_k$;
+$M$ is a column-stochastic matrix: $M_{i,k} \ge 0$ for all $i,k$ and $\sum_{k=1}^n {M_{i,k}} = 1$ for all $j$; 
 $I$ is the identity matrix;
@@
 (i) the probability to transition \emph{out} of type $A_k$ is determined by $C_k$;
-(ii) given a transition out of type $A_k$, the distribution of the destination types $A_j$ is given by $M_{j,k}$ (note the index order).
-Importantly, different types can have different rates; that is, $C_j \ne C_i$ for some $i,j$.
-The case $C_j = C_j$ for all $i,j$ is covered by the \emph{reduction principle}~\citep[see][]{Altenberg2017}.
+(ii) given a transition out of type $A_k$, the distribution of the destination types $A_i$ is given by $M_{i,k}$ (note the index order).
+Importantly, different types can have different rates; that is, $C_i \ne C_j$ for some $i,j$.
+The case $C_i = C_j$ for all $i,j$ is covered by the \emph{reduction principle}~\citep[see][]{Altenberg2017}.
 
@@
 \end{equation}
+Note that here $M$ is irreducible and therefore so is $(I-C+MC)D$.
 
@@
 \vdots & \ddots & \ddots & \ddots 
-\end{bmatrix},
+\end{bmatrix}.
 \end{equation}
-in which case Eq.~\ref{eq:mutation_model_2} is a specific case of Eq.~\ref{eq:model}.
+Here, too, $M$ and $(I-C+MC)D$ are irreducible.
 
@@
 where $I_n$ is the n-by-n identity matrix.
+In this case, $M$ is reducible and so is $(I-C+MC)D$.
 
@@
 Our analysis focuses on populations at equilibrium.
-Nevertheless, it has been demonstrated that a modifier which increases the mutation rate of maladapted individuals can be favored by selection~\citep{Ram2012} and reduce the adaptation rate~\citep{Ram2014}.
-Similar results have been demonstrated for a modifier that increases the recombination rate in maladapted individuals~\citep{Hadany2003a,Hadany2003b}.
+Nevertheless, it has been demonstrated that a modifier which increases the mutation rate of maladapted individuals can be favored by selection~\citep{Ram2012} and reduce the adaptation rate~\citep{Ram2014},
+and empirical evidence suggests that \emph{stress-induced mutagenesis} is common in bacteria and yeast, and may be prevalent in plants, flies, and human cancer cells~\citep{Rosenberg2012,Fitzgerald2017b}.
+Similar theoretical results have been demonstrated for a modifier that increases the recombination rate in maladapted individuals~\citep{Hadany2003a,Hadany2003b}.
 
@@
 General examples of departures from the \emph{reduction principle} are rare~\citep{Altenberg2017}.
-However, it has been demonstrated that higher mutation, recombination, and migration rates are favored by selection in populations evolving under fluctuating selection; see~\citet{Carja2014} and references therein.
-Here we have provided another general example, which suggests that a modifier allele that prompts maladapted or below-average individuals increase the rate at which variation is generated, will be favored by selection and will lead to increased population mean fitness.
+For example, it has been demonstrated that higher mutation, recombination, and migration rates are favored by selection in populations evolving under fluctuating selection; see~\citet{Carja2014} and references therein.
+Here we have provided another general example, which suggests that a modifier allele that prompts individuals with below-average fitness to increase the rate at which variation is generated, will be favored by selection and will lead to increased population mean fitness.
 
@@
 The equilibrium $v$ of the system in Eq.~\ref{eq:model_equilibrium} is locally stable.
-In addition,
-\begin{enumerate}
-\item If $M$ is irreducible and $D$ and $C$ are positive, then the equilibrium is globally stable.
-\item If $M$ is irreducible and either $D$ or $C$ are non-negative but not strictly positive, then the equilibrium is globally stable except possibly for initial conditions on the boundary, i.e., when starting with zero elements in the initial frequency vector.
-\end{enumerate}
 \end{lemma}
@@
 \begin{equation}
-f(t) = \frac{A^t f(0)}{\norm{A^t f(0)}}.
+f(t+k) = \frac{A^t f(k)}{\norm{A^t f(k)}}.
 \label{eq:ftplus1}
@@
 
-We write $f(0)$ as a linear combination of eigenvectors $v^k$ of $A$ for $k=1, \ldots, n$:
+To check local stability, we make a perturbation $\epsilon$ to $v$, such that this perturbation is small enough so that when $v+\epsilon$ is written as a linear combination of $\{v^k\}_{k=1}^{n}$ the eigenvalues of $A$, the coefficient of $v=v^1$ is not zero.
+That is, the frequency vector after the perturbation $f(0)=v+\epsilon$ can be written
 \begin{equation}
-f(0) = a_1 v^1 + a_2 v^2 + \ldots + a_n v^n,
+f(0) = a_1 v^1 + \ldots + a_n v^n,
 \label{eq:linear_combination}
 \end{equation}
-such that the eigenvector corresponding to the leading eigenvalue is $v=v_1$.
+with $\sum_{k=1}^{n}{f_k(0)}=1$ and $a_1 \ne 0$.
 
-To proceed we need to make sure that $a_1 \ne 0$.
-If $M$ is reducible then $A$ is reducible. 
-In this case, $a_1 \ne 0$ can be guaranteed by setting $f(0)=v + \epsilon$ such that $\epsilon_k$ is small enough and $v_k+\epsilon_k \ge 0$ for all $k$, and $\sum_{k=1}^{n}{\epsilon_k} = 0$.
-
-If $M$ is irreducible and $D$ and $C$ are positive, then $A$ is irreducible and  any choice of $f(0)$ guarantees $a_1 \ne 0$.
-
-If $M$ is reducible and either $D$ or $C$ are non-negative but not strictly positive, than any choice of $f(0)$ such that $f_k(0)>0$ for all $k$ guarantees $a_1 \ne 0$.
-
+We now check that $f(t) \goesto v$ as $t \goesto \infty$, which will indicate that $v$ is locally stable.
 We write, using Eq.~\ref{eq:linear_combination},
@@
 \bibliographystyle{agsm}
-\bibliography{/Users/yoavram/Documents/library}
-%\bibliography{ms}
+%\bibliography{/Users/yoavram/Documents/library}
+\bibliography{ms}
 

### 2017-11-13 ms.tex
fix migration model; fix appendix D; fix ref to Altenberg2012; add Uri

@@
 \title{Variation generation and mean fitness increase: \\ Necessity is the mother of genetic invention}
+% \title{A General Principle for the Evolution of Fitness-Associated Generation of Variation}
 
@@
 \author[b]{Lee Altenberg}
+\author[c]{Uri Liberman}
 \author[a]{Marcus W. Feldman}
@@
 \affil[b]{Information and Computer Sciences, University of Hawai`i at M{\=a}noa, Honolulu, HI}
+\affil[c]{School of Mathematical Sciences, Tel Aviv University, Israel}
 
@@
 (i) effectively infinite population size, (ii) constant-viability selection, (iii) a population at an equilibrium, and (iv) \emph{linear variation} -- the equal scaling of transition probabilities by the modifier.
-A departure from the latter assumption occurs if two variation-producing processes interact~\citep{Feldman1980,Altenberg2012}.
+A departure from the latter assumption occurs if two variation-producing processes interact~\citep{Feldman1980,Altenberg2012b}.
 Departures from the \emph{reduction principle} have also been demonstrated when conditions (i)-(iii) are not met, see for example~\citet{Holsinger1986} and references therein.
@@
 The frequency and fitness of individuals of type $A_k$ are $f_k$ and $D_k$, respectively.
-The probability that an individual of type $A_k$ will transition to some other type is $C_k$, and given a transition occurs, the probability that it will transition to type $A_i$ is $M_{i,k}$.
-Therefore, the change in the frequencies of types $A_k$ is described by the transformation $f \to f'$: 
-\begin{equation}\label{eq:model}
+The probability that an individual of type $A_k$ will transition to some other type is $C_k$, and given a transition occurs, the probability that it will transition to type $A_j$ is $M_{j,k}$.
+Therefore, the change in the frequencies of type $A_k$ is described by the transformation $f \to f'$: 
+\begin{equation}
+f'_k = (1-C_k)D_k f_k + \sum_{j=1}^{n}{C_j M_{k,j} D_j f_j},
+\end{equation}
+or in matrix form
+\begin{equation}
 \bar{w} f' = (I-C+MC)D f,
+\label{eq:model}
 \end{equation}
@@
 $C$ is a diagonal non-negative matrix with diagonal elements $C_k$;
-$M$ is a column-stochastic matrix: $M_{i,k} \ge 0$ for all $i,k$ and $\sum_{k=1}^n {M_{i,k}} = 1$ for all $j$; 
+$M$ is a column-stochastic matrix: $M_{j,k} \ge 0$ for all $j,k$ and $\sum_{j=1}^n {M_{j,k}} = 1$ for all $k$; 
 $I$ is the identity matrix;
@@
 \paragraph*{Migration model.}\label{sec:mig_model_1}
+% TODO explain why sum x y is 1 and wbar is sum xw yq
+Here we assume a large haploid population of haploids that occupies two demes (the model can be easily extended to an arbitrary number demes).
+Let the frequencies of $A_k$ be $x_k$ in deme 1 and $y_k$ in deme 2 for $k=1,2,\ldots,n$ with $x_k \ge 0$, $y_k \ge 0$, and $\sum_{k=1}^{n}{x_k + y_k} = 1$.
 
-Here we assume a large population of haploids that occupies two demes.
-Let the fitness of genotypes $A_k$ be $w_k$ in deme 1 and $q_k$ in deme 2 for $k=1,2,\ldots,n$.
-The probability that an individual with genotype $A_k$ migrates between the deme is $m_k$; that is, migration rates depend on the genotypes.
-Let the frequencies of $A_k$ be $x_k$ in deme 1 and $y_k$ in deme 2 for $k=1,2,\ldots,n$ with $x_k \ge 0$, $y_k \ge 0$, $\sum_{k=1}^{n}{x_k} = 1$, and $\sum_{k=1}^{n}{y_k} = 1$.
+In this model, the fitness of the individual is determines by both its type and its deme, but the entire population comes together for reproduction, and therefore reproductive success is determined by competition between individuals of both demes.
+Therefore, the fitness of genotypes $A_k$ is $w_{1,k}$ in deme 1 and $w_{2,k}$ in deme 2 for $k=1,2,\ldots,n$, and the population mean fitnesses of the entire population is $\bar{w} = \sum_{k \ge 0}{w_{1,k} x_k + w_{2,k} y_k}$.
+In addition, we assume that the fitness of type $k$ is different in deme 1 and 2, that is, $w_{1,k} \ne w_{2,k}$ for any $k=1,\ldots,n$.
+
+The probability that offspring of an individual from deme $a$ with type $A_k$ return to their parent home deme is $1-m_{j,k}$.
+That is, migration rates also depend on both the type and the deme.
 
@@
 \begin{align}
-x_k' = (1-m_k) \frac{w_k}{\bar{w}} x_k + m_k \frac{q_k}{\bar{q}} y_k \\
-y_k' = m_k \frac{w_k}{\bar{w}} x_k + (1 - m_k) \frac{q_k}{\bar{q}} y_k
+x_k' = (1-m_{1,k}) \frac{w_{1,k}}{\bar{w}} x_k + m_{2,k} \frac{w_{2,k}}{\bar{w}} y_k \\
+y_k' = m_{1,k} \frac{w_{1,k}}{\bar{w}} x_k + (1 - m_{2,k}) \frac{w_{2,k}}{\bar{w}} y_k
 \end{align}
-for $k=1,2,\ldots,n$, where $\bar{w} = \sum_{k \ge 0}{w_k x_k}$ and $\bar{q} = \sum_{k \ge 0}{q_k y_k}$ are the population mean fitnesses in deme 1 and 2, respectively.
-This is an extension of a model presented by~\citet{Altenberg2017}.
+for $k=1,2,\ldots,n$.
 
@@
 f = (x_1, \ldots, x_n, y_1, \ldots, y_n)\tr, \\
-C = diag[m_1, \ldots, m_n, m_1, \ldots, m_n], \\
-D = diag[w_1, \ldots, w_n, q_1, \ldots, q_n], \\
+C = diag[m_{1,1}, \ldots, m_{1,n}, m_{2,1}, \ldots, m_{2,n}], \\
+D = diag[w_{1,1}, \ldots, w_{1,n}, w_{2,1}, \ldots, w_{2,n}], \\
 M = \begin{bmatrix}
@@
 where $I_n$ is the n-by-n identity matrix.
-In this case, $M$ is reducible and so is $(I-C+MC)D$.
+In this case, $M$ is block matrix with each block an irreducible 2-by-2 matrix, and so is $(I-C+MC)D$.
 
@@
 Note that (i) the existence of $\bar{w}^*$ and $v$ are guaranteed by the \emph{Perron-Frobenius theorem}~\cite[Appendix~A]{Otto2007} because $(I-C+MC)D$ is a non-negative matrix; 
-(ii) the local stability of this equilibrium is proven in~\nameref{sec:AppD}; and
+(ii) the global stability of this equilibrium is proven in~\nameref{sec:AppD}; and
 (iii) $\bar{w}^*$ is identified as the stable population mean fitness in~\nameref{sec:AppA}.
@@
 \begin{lemma}[Mean fitness principle]\label{lemma:mfp}
-Let $\bar{w}^*$ be the leading eigenvalue of $(I-C+MC)D$, $u$ and $v$ be the corresponding left and right eigenvectors, and $[M]_k$ the $k$-th column of $M$. Then
+Let $\bar{w}^*$ be the leading eigenvalue of $(I-C+MC)D$, $u$ and $v$ be the corresponding left and right eigenvectors, and $[M]_k$ the $k$-th column of $M$.
+%TODO If $MCD$ is primitive, then
+Then,
 \begin{equation}\label{mfp}
@@
 
-Using \emph{Caswell's formula} (Eq.~\ref{eq:Caswells_formula} in~\nameref{sec:AppB}),
+Using the formula in~\citet{Caswell1978} (see eq.~\ref{eq:Caswells_formula} in~\nameref{sec:AppB}),
 \begin{equation}
@@
 
+% see Altenberg 2012 Ecological Monographs for a condition-dependent migration model
 We have given several examples of variation-generating processes for which this principle applies -- namely mutation and migration -- but our model may apply to other processes as well.
 For example, the \emph{reduction principle} applies to ecological models of dispersal, and~\citet{Gueijman2013} have demonstrated that even in homogeneous environments, fitness-associated dispersal increases the population mean fitness and is favored by selection over uniform dispersal.
-Similarly, if the transmission fidelity of culturally-transmitted traits depends on the type of the type or fitness of the transmitting individual, we expect that our results will hold.
+% TODO see cultural transmission model in Lee's thesis
+Similarly, if the transmission fidelity of culturally-transmitted traits depends on the type or fitness of the transmitting individual, we expect that our results will hold.
 
@@
 
-% Appendix B: Caswell's formula
+% Appendix B: Caswell formula
 \subsection*{Appendix B}\label{sec:AppB}
@@
 In this formula, the \emph{population growth rate} is the leading eigenvalue of the population transformation matrix $T$, the \emph{life history parameters} are elements of $T$, and the \emph{sensitivity} is the derivative of the former with respect to the latter.
-This is a useful formula~\citep{Caswell1978,Hermisson2002,Ram2012}, and therefore we reproduce it here.
+This is a useful formula~\citep[ch.~10]{Caswell1978,Hermisson2002,Ram2012,Otto2007}, and therefore we reproduce it here.
 
-\begin{lemma}[Caswell's formula]
-Let $T$ be a non-negative matrix with leading eigenvalue $\lambda$ and left and right eigenvectors $u$ and $v$ such that $\sum{v_k}=1$ and $u\tr v = \sum{u_k v_k} = 1$.
+\begin{lemma}
+$T$ be a non-negative matrix with leading eigenvalue $\lambda$ and left and right eigenvectors $u$ and $v$ such that $\sum{v_k}=1$ and $u\tr v = \sum{u_k v_k} = 1$.
 Then the sensitivity of $\lambda$ to changes in any element $t$ of the matrix $T$ is
@@
 
-\emph{Fisher's reproductive values} can be understood as follows~\citep{Grafen2006}.
+\emph{Fisher's reproductive values} can be understood as follows~\citep[ch.~10]{Grafen2006,Otto2007}.
 Consider the dynamics not of frequencies but of absolute population sizes such that the vector of the number of individuals of each type at time $t$ is $n(t)$ and the corresponding frequencies are $f_k(t) = n_k(t) / \sum_i{n_i(t)}$.
@@
 \begin{lemma}[Stability of the equilibrium $v$]
-The equilibrium $v$ of the system in Eq.~\ref{eq:model_equilibrium} is locally stable.
-\end{lemma}
+If $C$ and $D$ are positive diagonal matrices, and $M$ is block irreducible, then
+the equilibrium $v$ of the system in Eq.~\ref{eq:model_equilibrium} is globally stable. \end{lemma}
 \begin{proof}
-Set $A=(I-C+MC)D$ and rewrite Eq.~\ref{eq:model} such that $f(t)$ is the frequency vector at time $t$ and
+Let $A=(I-C+MC)D$ and denote the leading eigenvalue of $A$ as $\bar{w}^*$.
+
+According to the \emph{Perron-Frobenius theorem}, 
 \begin{equation}
-f(t) = \frac{Af(t-1)}{\norm{Af(t-1)}},
+\bigg(\frac{A}{\bar{w}^*}\bigg)^t \goesto \frac{v u\tr}{u\tr v} = v u\tr,
 \end{equation}
-where $\norm{f(t)} = \sum_{k=1}^{n}f_k(t)$.
-Denote the eigenvalues of $A$ as $\lambda_1, \ldots, \lambda_n$ such that $\abs{\lambda_1} \ge \ldots \ge \abs{\lambda_n}$ and $\lambda_1$ is the \emph{Perron} leading eigenvalue, that is, the largest real eigenvalue with corresponding real non-negative right eigenvector $v$ ($\lambda_1$ is shown in~\nameref{sec:AppA} to be the stable population mean fitness $\bar{w}^*$).
+where $u$ and $v$ are the left and right Perron eigenvectors of $A$ such that 
+\begin{equation}
+\sum_{k=1}^{n}{v_k}=1, \quad
+u\tr v=\sum_{k=1}^{n}{u_k v_k}=1,
+\label{eq:sum_uv}
+\end{equation}
+and $v u\tr$ is the \emph{Perron projection} into the eigenspace corresponding to $\bar{w}^*$.
 
-First, note that the normalization $x \goesto \frac{1}{\norm{x}}$ can be applied once instead of $t$ times~\citep[Appendix~A]{Wakano2004}, that is,
+Therefore, for any positive frequency vector $f$
 \begin{equation}
-f(t+k) = \frac{A^t f(k)}{\norm{A^t f(k)}}.
-\label{eq:ftplus1}
+\bigg(\frac{A}{\bar{w}^*}\bigg)^t \cdot f \goesto v u\tr \cdot f = \alpha v,
+\label{eq:perron_projection}
 \end{equation}
+where $\alpha \in \mathbb{R}$.
+Using eqs.~\ref{eq:sum_uv} and~\ref{eq:perron_projection}
+\begin{equation}
+\alpha = \alpha u\tr v = u\tr \alpha v = u\tr v u\tr f = u\tr f = \sum_{k=1}^{n}{u_k f_k} > 0,
+\label{eq:alpha0}
+\end{equation}
+because $f$ is positive and $u$ is non-negative.
 
-To check local stability, we make a perturbation $\epsilon$ to $v$, such that this perturbation is small enough so that when $v+\epsilon$ is written as a linear combination of $\{v^k\}_{k=1}^{n}$ the eigenvalues of $A$, the coefficient of $v=v^1$ is not zero.
-That is, the frequency vector after the perturbation $f(0)=v+\epsilon$ can be written
+Now, let $\norm{x} = \sum_{k=1}^{n}{x_k}$, and therefore $x_t \goesto x$ implies
 \begin{equation}
-f(0) = a_1 v^1 + \ldots + a_n v^n,
-\label{eq:linear_combination}
+\norm{x_t} \goesto \norm{x}
+\label{eq:norm_limit}
 \end{equation}
-with $\sum_{k=1}^{n}{f_k(0)}=1$ and $a_1 \ne 0$.
+as $t \goesto \infty$.
 
-We now check that $f(t) \goesto v$ as $t \goesto \infty$, which will indicate that $v$ is locally stable.
-We write, using Eq.~\ref{eq:linear_combination},
+Rewrite Eq.~\ref{eq:model} as 
+\begin{equation}
+f' = \frac{A f}{\norm{A f}} 
+\label{eq:model_norm}
+\end{equation}
+Note that when applying eq.~\ref{eq:model_norm} $t$ times, the normalization can be applied once~\cite[Appendix~A]{Wakano2004}.
+
+Therefore, for a positive frequency vector $f$, we can apply $f \goesto f'$ (Eq.~\ref{eq:model_norm}) $t$ times, and using eqs.~\ref{eq:perron_projection} and ~\ref{eq:norm_limit}, we have
 \begin{multline}
-A^t f(0) = 
-a_1 A^t v^1 + a_2 A^t v^2 + \ldots + a_n A^t v^n = \\
-a_1 \lambda_1^t v^1 + a_2 \lambda_2^t v^2 + \ldots + a_n \lambda_n^t v^n = \\
-\lambda_1^t \bigg(a_ 1v^1 + a_2 \bigg(\frac{\lambda_2}{\lambda_1}\bigg)^t v^2 + \ldots +  a_n \bigg(\frac{\lambda_n}{\lambda_1}\bigg)^t v^n\bigg).
+\frac{A^t f}{\norm{A^t f}} = 
+\frac{\frac{1}{(\bar{w}^*)^t} A^t f}{\frac{1}{(\bar{w}^*)^t} \norm{A^t f}} = 
+\frac{\big(\frac{A}{\bar{w}^*}\big)^t \cdot f }{ \norm{ \big(\frac{A}{\bar{w}^*}\big)^t \cdot f }} \goesto
+\frac{\alpha v}{\norm{\alpha v}} = v,
 \end{multline}
-Therefore,
-\begin{equation}
-\frac{A^t f(0)}{\lambda_1^t} \goesto a_1 v_1
-\end{equation}
-as $t \goesto \infty$, because $\abs{\frac{\lambda_k}{\lambda_1}} < 1$
-for all $k=2, \ldots, n$.
-Finally, 
-\begin{equation}
-f(t) = 
-\frac{A^t f(0)}{\norm{A^t f(0)}} =
-\frac{\frac{A^t f(0)}{\lambda_1^t}}{\norm{\frac{A^t f(0)}{\lambda_1^t}}} \goesto
-\frac{a_1 v^1}{\norm{a_1 v^1}} = v,
-\end{equation}
-as $t \goesto \infty$, which completes the proof.
+as $t \goesto \infty$, as $\norm{v}=1$ (eq.~\ref{eq:sum_uv}) and $\alpha \ne 0$ (eq.~\ref{eq:alpha0}).
+This completes the proof.
 \end{proof}

### 2017-11-13 ms.tex
hard selection refs

@@
 \paragraph*{Migration model.}\label{sec:mig_model_1}
-% TODO explain why sum x y is 1 and wbar is sum xw yq
+
 Here we assume a large haploid population of haploids that occupies two demes (the model can be easily extended to an arbitrary number demes).
@@
 
-In this model, the fitness of the individual is determines by both its type and its deme, but the entire population comes together for reproduction, and therefore reproductive success is determined by competition between individuals of both demes.
+In this model, the fitness of the individual is determines by both its type and its deme, but the entire population comes together for reproduction, and therefore reproductive success is determined by competition between individuals of both demes -- this has been termed \emph{hard selection}~\citep{Wallace1975,Karlin1982}.
 Therefore, the fitness of genotypes $A_k$ is $w_{1,k}$ in deme 1 and $w_{2,k}$ in deme 2 for $k=1,2,\ldots,n$, and the population mean fitnesses of the entire population is $\bar{w} = \sum_{k \ge 0}{w_{1,k} x_k + w_{2,k} y_k}$.
-In addition, we assume that the fitness of type $k$ is different in deme 1 and 2, that is, $w_{1,k} \ne w_{2,k}$ for any $k=1,\ldots,n$.
+In addition, we assume that the fitness of individuals of type $k$ is different in deme 1 and 2, that is, $w_{1,k} \ne w_{2,k}$ for any $k=1,\ldots,n$.
 

### 2017-11-14 ms.tex
acks for LA

@@
 
-% Migration model
-\paragraph*{Migration model.}\label{sec:mig_model_1}
+% Migration model 1
+\paragraph*{Migration model 1.}\label{sec:mig_model_1}
 
@@
 
-This research was supported in part by the Stanford Center for Computational, Evolutionary and Human Genomics, and by the Morrison Institute for Population and Resources Studies, Stanford University.
+This work was supported in part by the Department of Information and Computer Sciences at the University of Hawai`i at M{\=a}noa, the Konrad Lorenz Institute for Evolution and Cognition Research, the Mathematical Biosciences Institute through National Science Foundation Award \#DMS 0931642, the Stanford Center for Computational, Evolutionary and Human Genomics, by the Morrison Institute for Population and Resources Studies Stanford University.
 }

### 2017-11-15 ms.tex
revised migration model some more

--- 2017-11-13 ms.tex
@@
 We find that if a modifier allele increases the rate at which individuals of below-average fitness generate variation, then it will increase in frequency and increase the  population mean fitness.
-This principle applies to phenomena such as stress-induced mutagenesis, condition-dependent dispersal and fitness-associated outcrossing,
+This principle applies to phenomena such as stress-induced mutagenesis and condition-dependent dispersal,
 and exhibits the paraphrase: \emph{"Necessity is the mother of genetic invention."}
@@
 \begin{equation}
-f'_k = (1-C_k)D_k f_k + \sum_{j=1}^{n}{C_j M_{k,j} D_j f_j},
+\bar{w} f'_k = (1-C_k)D_k f_k + \sum_{j=1}^{n}{C_j M_{k,j} D_j f_j},
+\label{eq:model_sum}
 \end{equation}
@@
 
-Here, we consider a large population of haploids in which individuals with genotype $A_k$ are characterized by the number $k$ of deleterious or mutant alleles in their genotype, where $0 \le k \le n$.
+Here we consider a large population of haploids in which individuals with genotype $A_k$ are characterized by the number $k$ of deleterious or mutant alleles in their genotype, where $0 \le k \le n$.
 Importantly, the probability $m_k$ that a mutation occurs in individuals with $k$ mutant alleles depends on $k$.
@@
 \begin{equation}
-M_{i,j} = \begin{bmatrix}
+M = \begin{bmatrix}
 1-\delta & \beta & 0 &  \cdots \\
@@
 
-% Migration model
-\paragraph*{Migration model.}\label{sec:mig_model_1}
-% TODO explain why sum x y is 1 and wbar is sum xw yq
-Here we assume a large haploid population of haploids that occupies two demes (the model can be easily extended to an arbitrary number demes).
-Let the frequencies of $A_k$ be $x_k$ in deme 1 and $y_k$ in deme 2 for $k=1,2,\ldots,n$ with $x_k \ge 0$, $y_k \ge 0$, and $\sum_{k=1}^{n}{x_k + y_k} = 1$.
+% Migration model 1
+\paragraph*{Migration model 1.}\label{sec:mig_model_1}
 
-In this model, the fitness of the individual is determines by both its type and its deme, but the entire population comes together for reproduction, and therefore reproductive success is determined by competition between individuals of both demes.
-Therefore, the fitness of genotypes $A_k$ is $w_{1,k}$ in deme 1 and $w_{2,k}$ in deme 2 for $k=1,2,\ldots,n$, and the population mean fitnesses of the entire population is $\bar{w} = \sum_{k \ge 0}{w_{1,k} x_k + w_{2,k} y_k}$.
-In addition, we assume that the fitness of type $k$ is different in deme 1 and 2, that is, $w_{1,k} \ne w_{2,k}$ for any $k=1,\ldots,n$.
+Here we assume a large haploid population of haploids that occupies $n$ demes, $A_1, \ldots, A_n$. In this case we directly use the notation of Eq.~\ref{eq:model}.
+Let the frequencies of individuals in deme $A_k$ be $f_k$ with $f_k \ge 0$ and $\sum_{k=1}^{n}{f_k} = 1$.
+The fitness of individuals in deme $A_k$ is $D_k$, but the entire population comes together for reproduction, and therefore reproductive success is determined by competition between individuals of both demes -- this has been termed \emph{hard selection}~\citep{Wallace1975,Karlin1982}.
+The population mean fitness is $\bar{w} = \sum_{k=1}^{n}{D_k f_k}$.
 
-The probability that offspring of an individual from deme $a$ with type $A_k$ return to their parent home deme is $1-m_{j,k}$.
-That is, migration rates also depend on both the type and the deme.
+After reproduction, offspring of individuals from deme $A_k$ return to their parental deme with probability $1-C_k$, or migrate to a different deme $A_j$ with probability $C_k M_{k,j}$, where $C_k \ge 0$, $M_{k,j} \ge 0$ and $\sum_{k=1}^{n}{M_{k,j}} = 1$ for all $j=1, \ldots, n$.
+Therefore, $1-C_k$ can be interpreted as a \emph{homing rate}.
 
-Following selection and migration the new frequencies $x_k'$ and $y_k'$ are given by
-\begin{align}
-x_k' = (1-m_{1,k}) \frac{w_{1,k}}{\bar{w}} x_k + m_{2,k} \frac{w_{2,k}}{\bar{w}} y_k \\
-y_k' = m_{1,k} \frac{w_{1,k}}{\bar{w}} x_k + (1 - m_{2,k}) \frac{w_{2,k}}{\bar{w}} y_k
-\end{align}
-for $k=1,2,\ldots,n$.
-
-To rephrase this model in terms of the general model, we define
+Following selection and migration the new frequencies $x'_k$ are given by Eq.~\ref{eq:model_sum}.
+If the rows of $M$ are equal, that is, $M_{k,j}=m_k$ for all $j$,
+then $m_k$ can be considered the relative population size of deme $A_k$;
+this is the non-homogeneous extension to Deakin's \emph{homing model}~\citep{Deakin1966,Karlin1982}.
+The transition matrix $M$ can account for other factors that determine the probability of non-homing juveniles from deme $A_j$ to migrate to deme $A_k$ -- for example, deme adjacency:
+if demes are arranged in a circle, 
+then $M$ can take the form
 \begin{equation}
-\begin{aligned}
-f = (x_1, \ldots, x_n, y_1, \ldots, y_n)\tr, \\
-C = diag[m_{1,1}, \ldots, m_{1,n}, m_{2,1}, \ldots, m_{2,n}], \\
-D = diag[w_{1,1}, \ldots, w_{1,n}, w_{2,1}, \ldots, w_{2,n}], \\
 M = \begin{bmatrix}
-0 & I_n \\
-I_n & 0
-\end{bmatrix}
-\end{aligned}
+p_0 & p_1 & \ldots & p_n \\
+p_n & p_0 & \ldots & p_{n-1} \\
+\vdots & \vdots & \ddots & \vdots \\
+p_1 & p_n & \ldots & p_0
+\end{bmatrix}.
 \end{equation}
-where $I_n$ is the n-by-n identity matrix.
-In this case, $M$ is block matrix with each block an irreducible 2-by-2 matrix, and so is $(I-C+MC)D$.
+where $p_{-k} = p_{n-k}$, $p_k \ge 0$, and $\sum_{k=1}^{n}{p_k}=1$.
 
@@
 
-This research was supported in part by the Stanford Center for Computational, Evolutionary and Human Genomics, and by the Morrison Institute for Population and Resources Studies, Stanford University.
+This work was supported in part by 
+the Department of Information and Computer Sciences at the University of Hawai`i at M{\=a}noa, 
+the Konrad Lorenz Institute for Evolution and Cognition Research, 
+the Mathematical Biosciences Institute through National Science Foundation Award \#DMS 0931642, 
+the Stanford Center for Computational, Evolutionary and Human Genomics, 
+and the Morrison Institute for Population and Resources Studies, Stanford University.
 }

### 2017-11-15 ms.tex
revised model examples and intro

@@
 mutation, migration, or recombination), selection favors neutral modifiers that decrease the rate at which variation is generated.
-The \emph{reduction principle} was demonstrated for modifiers of the recombination~\citep{Feldman1972}, mutation, and migration~\citep{Liberman1986a}, and dispersal rates~\citep{Altenberg1987}.
+The \emph{reduction principle} was demonstrated for modifiers of the recombination~\citep{Feldman1972}, mutation~\citep{Liberman1986a}, migration~\citep{Feldman1986}, and dispersal rates~\citep{Altenberg1987}.
 These separate principles were recently unified by~\citet{Altenberg2017}
@@
 \citet{Ram2012} demonstrated that even in a constant environment, modifiers that increase the mutation rate of individuals with below-average fitness increase the population mean fitness, rather than decrease it.
-Their analysis assumed an infinite population size and constant-viability selection in which fitness is determined by the number of mutant alleles accumulated in the genotype.
+Their analysis assumed an infinite population in which fitness is determined by the number of mutant alleles accumulated in the genotype.
 The only departure from the \emph{reduction principle} assumptions was due to the differential effect the modifier has on different genotypes.
@@
 \begin{equation}
-\bar{w} f' = (I-C+MC)D f,
+\bar{w} f' = (\ci - C + MC)D f,
 \label{eq:model}
@@
 where $f=(f_1, f_2, \ldots, f_n)$ is a stochastic vector such that $\sum_{k=1}^n{f_k} = 1$;
-$D$ is a diagonal matrix with diagonal elements $D_k$, where we can assume $0 < D_k \le 1$, i.e., fitness values are relative to the fitness of the fittest type; 
+$D$ is a diagonal matrix with diagonal elements $D_k$ such that $D_k \ne D_j$ for $k \ne j$, and we can assume $0 < D_k \le 1$, i.e., fitness values are relative to the fitness of the fittest type; 
 $C$ is a diagonal non-negative matrix with diagonal elements $C_k$;
 $M$ is a column-stochastic matrix: $M_{j,k} \ge 0$ for all $j,k$ and $\sum_{j=1}^n {M_{j,k}} = 1$ for all $k$; 
-$I$ is the identity matrix;
+$\ci$ is the identity matrix;
 and $\bar{w}$ is the normalizing factor such that $\sum_{k=1}^n{f'_k}=1$ and is equal to the population mean fitness (see~\nameref{sec:AppA}).
@@
 Importantly, type transmission is vertical and uni-parental (the type is transmitted from a single parent to the offspring) and is independent of the frequencies of the other types.
-Specifically, this model precludes processes such as recombination, social learning, sexual outcrossing, and any horizontal or oblique transmission, as these processes are frequency-dependent~\cite[pg.~54]{Cavalli-Sforza1981}.
+Specifically, this model precludes processes such as recombination, social learning, sexual outcrossing, and horizontal or oblique transmission, as these processes are frequency-dependent~\cite[pg.~54]{Cavalli-Sforza1981}.
 
@@
 (ii) given a transition out of type $A_k$, the distribution of the destination types $A_i$ is given by $M_{i,k}$ (note the index order).
-Importantly, different types can have different rates; that is, $C_i \ne C_j$ for some $i,j$.
-The case $C_i = C_j$ for all $i,j$ is covered by the \emph{reduction principle}~\citep[see][]{Altenberg2017}.
+Importantly, different types can have different rates.
+That is, $C_i \ne C_j$ for some $i,j$. The case $C_i = C_j$ for all $i,j$ is covered by the \emph{reduction principle}~\citep[see][]{Altenberg2017}.
 
-In the following section we present three examples of the model presented in Eq.~\ref{eq:model} for mutation and migration.
+In the following section we present three examples of the model (Eq.~\ref{eq:model}) for mutation and migration.
 
@@
 
-Here we consider a large population of haploids and a trait determined by a major locus with $n$ possible alleles $A_1, A_2, ..., A_n$.
-The mutation rates $\mu_k$ of individuals with allele $A_k$ are potentially different;
-specifically, with probability $1-\mu_k$, the allele $A_k$ does not mutate, and with probability $\frac{\mu_k}{n-1}$, the allele $A_k$ mutates to $A_j$ for any $j \ne k$.
-This is an extension of a model studied by~\citet{Altenberg2017} that allows for the mutation rate of an individual, $\mu_k$, to depend on the identity of the allele $A_k$.
+Here we consider a large population of haploids and a trait determined by a single genetic locus with $n$ possible alleles $A_1, A_2, \ldots, A_n$ and corresponding fitness values $D_1, D_2, \ldots, D_n$.
+The mutation rates $C_k$ of individuals with allele $A_k$ are potentially different;
+specifically, with probability $1-C_k$, the allele $A_k$ does not mutate, and with probability $\frac{C_k}{n-1}$, the allele $A_k$ mutates to $A_j$ for any $j \ne k$.
+This is an extension of a model studied by~\citet{Altenberg2017} that allows for the mutation rate of an individual, $C_k$, to depend on the identity of the allele $A_k$.
 
-Let the frequency of $A_k$ in the present generation be $f_k$ with $f_k \ge 0$ and $\sum_k{f_k}=1$.
+Let the frequency of $A_k$ in the present generation be $f_k$ with $f_k \ge 0$ and $\sum_{k=1}^{n}{f_k}=1$.
 Then after selection and mutation, $f'_k$ in the next generation is given by
 \begin{equation}
-f_k' = (1 - \mu_k) \frac{w_k}{\bar{w}} f_k + \frac{1}{n-1} \sum_{k \ne j}{\mu_j \frac{w_j}{\bar{w}} f_j},
+f_k' = \big(1 - C_k\big) \frac{D_k}{\bar{w}} f_k + \frac{1}{n-1} \sum_{k \ne j}{C_j \frac{D_j}{\bar{w}} f_j},
 \label{eq:mutation_model_1}
 \end{equation}
-for all $k=1,2,\ldots,n$, where $\bar{w} = \sum_{k=1}^{n}{w_k f_k}$ is the population mean fitness.
+for all $k=1,2,\ldots,n$, where $\bar{w} = \sum_{k=1}^{n}{D_k f_k}$ is the population mean fitness.
 
-This model can be rephrased in terms of the general model: 
-Eq.~\ref{eq:mutation_model_1} is equivalent to Eq.~\ref{eq:model} if we set
+This model is a special case of the general model (Eq.~\ref{eq:model}) where
 \begin{equation} 
-C_k = \mu_k, \quad
-D_k = w_k, \quad
-M_{i,j} = \begin{cases}
-\frac{1}{n-1}, & i \ne j \\
-0, & i=j
-\end{cases}.
+{M} = \frac{1}{n-1} \begin{bmatrix}
+0 & 1 & 1 & \ldots & 1 \\
+1 & 0 & 1 & \ldots & \vdots \\
+1 & 1 & 0 & \ldots & \vdots \\
+\vdots & \vdots & \vdots & \ddots & \vdots \\
+1 & \ldots & \ldots & \ldots & 0
+\end{bmatrix},
 \end{equation}
-Note that here $M$ is irreducible and therefore so is $(I-C+MC)D$.
+with zeros on the diagonal and $\frac{1}{n-1}$ elsewhere.
+Note that here ${M}$ is irreducible.
 
@@
 
-Here we consider a large population of haploids in which individuals with genotype $A_k$ are characterized by the number $k$ of deleterious or mutant alleles in their genotype, where $0 \le k \le n$.
-Importantly, the probability $m_k$ that a mutation occurs in individuals with $k$ mutant alleles depends on $k$.
-We assume a low mutation rate $m_k$ so that two mutations are unlikely to occur in the same individual in one generation.
-When a mutation occurs it is \emph{deleterious} with probability $\delta$, generating a mutant allele and converting the individual from $A_k$ to $A_{k+1}$, or it is beneficial with probability $\beta$, converting the individual from $A_k$ to $A_{k-1}$.
+Again, we consider a large population of haploids, but here individuals with genotype $A_k$ are characterized by the number $k$ of deleterious or mutant alleles in their genotype, where $0 \le k \le n$.
+Specifically, the fitness of individuals with $k$ mutant alleles is $D_k$,
+And the probability $C_k$ that a mutation occurs in individuals with $k$ mutant alleles depends on $k$.
+We assume a low mutation rate $C_k$ so that two mutations are unlikely to occur in the same individual in one generation.
+When a mutation occurs it is \emph{deleterious} with probability $\delta$, generating a mutant allele and converting the individual from $A_k$ to $A_{k+1}$,
+or it is \emph{beneficial} with probability $\beta$, converting the individual from $A_k$ to $A_{k-1}$.
 Note that such beneficial mutations can be either compensatory or back-mutations, and that mutations are neutral with probability $1-\delta-\beta$.
@@
 \begin{aligned}
-f_0' &= (1 - \delta m_0) \frac{w_0}{\bar{w}} f_0 + \beta m_{1} \frac{w_{1}}{\bar{w}} f_{1}, \\
-f_n' &= (1 - \beta m_n) \frac{w_n}{\bar{w}} f_n + \delta m_{n-1} \frac{w_{n-1}}{\bar{w}} f_{n-1}, \\
-f_k' &= (1 - m_k(\delta+\beta)) \frac{w_k}{\bar{w}} f_k + \\
-	 & \delta m_{k-1} \frac{w_{k-1}}{\bar{w}} f_{k-1} + 
-	 \beta m_{k+1} \frac{w_{k+1}}{\bar{w}} f_{k+1},
+f_0' &= \big(1 - \delta C_0\big) \frac{D_0}{\bar{w}} f_0 + \beta C_{1} \frac{D_{1}}{\bar{w}} f_{1}, \\
+f_n' &= \big(1 - \beta C_n\big) \frac{D_n}{\bar{w}} f_n + \delta C_{n-1} \frac{D_{n-1}}{\bar{w}} f_{n-1}, \\
+f_k' &= \big(1 - C_k(\delta+\beta)\big) \frac{D_k}{\bar{w}} f_k + \\
+	 & \delta C_{k-1} \frac{D_{k-1}}{\bar{w}} f_{k-1} + 
+	 \beta C_{k+1} \frac{D_{k+1}}{\bar{w}} f_{k+1},
 \end{aligned}
@@
 \end{equation}
-for $k=1,2,\ldots, n-1$, where $w_k$ is the fitness of an individual with $k$ mutant alleles and $\bar{w} = \sum_{k=0}^{n}{w_k f_k}$ is the population mean fitness.
+for $k=1,2,\ldots, n-1$.
 
-In terms of the general model, we can write
+Therefore, setting 
 \begin{equation}
-C_k = m_k, \quad
-D_k = w_k,
-%M_{i,j} = \begin{cases}
-%\beta, & i = j-1 \\
-%1-\beta-\delta, & i = j \\
-%\delta, & i = j+1 \\
-%1-\delta, & i=j=0 \\
-%0, & \text{otherwise}
-%\end{cases}
+{M} = \begin{bmatrix}
+1-\delta & \beta & 0 &  \cdots & 0\\
+\delta & 1-\delta-\beta & \beta &  \ddots & 0\\
+0 & \delta & 1-\delta-\beta & \ddots & 0 \\
+\vdots & \ddots & \ddots & \ddots & \beta \\
+0 & 0 & 0 & \delta & 1-\beta
+\end{bmatrix},
 \end{equation}
-and
-\begin{equation}
-M = \begin{bmatrix}
-1-\delta & \beta & 0 &  \cdots \\
-\delta & 1-\delta-\beta & \beta &  \ddots \\
-0 & \delta & 1-\delta-\beta & \ddots \\
-\vdots & \ddots & \ddots & \ddots 
-\end{bmatrix}.
-\end{equation}
-Here, too, $M$ and $(I-C+MC)D$ are irreducible.
+with the beneficial, neutral, and deleterious mutation probabilities on the three main digonals and zeros elsewhere, 
+Eq.~\ref{eq:mutation_model_2} can be viewed as a special case of Eq.~\ref{eq:model}.
+Here, too, ${M}$ is irreducible.
 
@@
 
-Here we assume a large haploid population of haploids that occupies $n$ demes, $A_1, \ldots, A_n$. In this case we directly use the notation of Eq.~\ref{eq:model}.
+In this case we consider a large population of haploids that occupy $n$ demes, $A_1, \ldots, A_n$.
 Let the frequencies of individuals in deme $A_k$ be $f_k$ with $f_k \ge 0$ and $\sum_{k=1}^{n}{f_k} = 1$.
 The fitness of individuals in deme $A_k$ is $D_k$, but the entire population comes together for reproduction, and therefore reproductive success is determined by competition between individuals of both demes -- this has been termed \emph{hard selection}~\citep{Wallace1975,Karlin1982}.
-The population mean fitness is $\bar{w} = \sum_{k=1}^{n}{D_k f_k}$.
 
-After reproduction, offspring of individuals from deme $A_k$ return to their parental deme with probability $1-C_k$, or migrate to a different deme $A_j$ with probability $C_k M_{k,j}$, where $C_k \ge 0$, $M_{k,j} \ge 0$ and $\sum_{k=1}^{n}{M_{k,j}} = 1$ for all $j=1, \ldots, n$.
+After reproduction, offspring of individuals from deme $A_k$ return to their parental deme with probability $1-C_k$, or migrate to a different deme $A_j$ with probability $C_k \cdot M_{j,k}$, where $C_k > 0$, $M_{j,k} \ge 0$, and $\sum_{k=1}^{n}{M_{j,k}} = 1$ for all $j=1, \ldots, n$.
 Therefore, $1-C_k$ can be interpreted as a \emph{homing rate}.
 
-Following selection and migration the new frequencies $x'_k$ are given by Eq.~\ref{eq:model_sum}.
-If the rows of $M$ are equal, that is, $M_{k,j}=m_k$ for all $j$,
-then $m_k$ can be considered the relative population size of deme $A_k$;
-this is the non-homogeneous extension to Deakin's \emph{homing model}~\citep{Deakin1966,Karlin1982}.
-The transition matrix $M$ can account for other factors that determine the probability of non-homing juveniles from deme $A_j$ to migrate to deme $A_k$ -- for example, deme adjacency:
-if demes are arranged in a circle, 
-then $M$ can take the form
+Following selection and migration the new frequencies $f'_k$ are given exactly by Eq.~\ref{eq:model_sum}.
+If columns of ${M}$ are identical
 \begin{equation}
-M = \begin{bmatrix}
-p_0 & p_1 & \ldots & p_n \\
-p_n & p_0 & \ldots & p_{n-1} \\
+{M} = \begin{bmatrix}
+m_1 & m_1 & \ldots & m_1 \\
+m_2 & m_2 & \ldots & m_2 \\
+\vdots & \vdots & \vdots & \vdots \\
+m_n & m_n & \ldots & m_n
+\end{bmatrix},
+\end{equation}
+with $\sum_{k=1}^{n}{m_k}=1$,
+then $m_k$ can be considered the relative population size of deme $A_k$ --
+this is the non-homogeneous extension of Deakin's \emph{homing model}~\citep{Deakin1966,Karlin1982}.
+
+Similarly, if demes are arranged in a circle, for example around a lake, 
+then we can denote the probability $p_k$ of migrating $k$ demes away from the parental deme (conditioned on migration which occurs with probability $C_k$)
+and ${M}$ takes the form
+\begin{equation}
+{M} = \begin{bmatrix}
+p_0 & p_1 & \ldots & p_{n-1} \\
+p_{n-1} & p_0 & \ldots & p_{n-2} \\
 \vdots & \vdots & \ddots & \vdots \\
-p_1 & p_n & \ldots & p_0
-\end{bmatrix}.
+p_1 & p_2 & \ldots & p_0
+\end{bmatrix},
 \end{equation}
-where $p_{-k} = p_{n-k}$, $p_k \ge 0$, and $\sum_{k=1}^{n}{p_k}=1$.
+where $p_k \ge 0$ and $\sum_{k=1}^{n}{p_k}=1$.
 
@@
 \begin{lemma}[Mean fitness principle]\label{lemma:mfp}
-Let $\bar{w}^*$ be the leading eigenvalue of $(I-C+MC)D$, $u$ and $v$ be the corresponding left and right eigenvectors, and $[M]_k$ the $k$-th column of $M$.
+Let $\bar{w}^*$ be the leading eigenvalue of $(I-C+MC)D$,
+and $u$ and $v$ be the corresponding left and right eigenvectors.
 %TODO If $MCD$ is primitive, then
@@
 sign \frac{\partial \bar{w}^*}{\partial C_k} = 
-sign (u\tr [M]_k - u_k);
+sign (u\tr [M]_k - u_k),
 \end{equation}
-therefore, increased transition from type $k$ will increase the stable population mean fitness if the fraction of the population descended from type $k$ is expected to increase due to a transition to another type.
+where $[M]_k$ is the $k$-th column of $M$.
+
+Therefore, increased transition from type $k$ will increase the stable population mean fitness if the fraction of the population descended from type $k$ is expected to increase due to a transition to another type.
 \end{corollary}

### 2017-11-16 ms.tex
MWF revision

@@
 This principle applies to phenomena such as stress-induced mutagenesis and condition-dependent dispersal,
-and exhibits the paraphrase: \emph{"Necessity is the mother of genetic invention."}
+and exemplifies \emph{"Necessity is the mother of genetic invention."}
 \end{abstract}
@@
 
-Another departure from the \emph{linear variation} assumption of the \emph{reduction principle} for mutation rates involves a mechanism in which the mutation rate increases in individuals with poor fitness -- a mechanism first observed in stressed bacteria~\citep{Foster2007}, although not in a constant environment.
+Another departure from the \emph{linear variation} assumption of the \emph{reduction principle} for mutation rates involves a mechanism by which the mutation rate increases in individuals of low fitness -- a mechanism first observed in stressed bacteria~\citep{Foster2007}, although not in a constant environment.
 \citet{Ram2012} demonstrated that even in a constant environment, modifiers that increase the mutation rate of individuals with below-average fitness increase the population mean fitness, rather than decrease it.
-Their analysis assumed an infinite population in which fitness is determined by the number of mutant alleles accumulated in the genotype.
-The only departure from the \emph{reduction principle} assumptions was due to the differential effect the modifier has on different genotypes.
+Their analysis assumed  infinite population size and fitness  determined by the number of mutant alleles accumulated in the genotype.
+In their models, the only departure from the \emph{reduction principle} assumptions was due to the differential effect the modifier has on different genotypes.
 A similar result has been demonstrated for fitness-associated recombination~\citep{Hadany2003a} and for condition-dependent sexual reproduction~\citep{Hadany2007a}, and evidence suggests that both mechanisms are common in nature~\citep{Ram2016}.
@@
 Their analysis was specific to a model that classified individuals by the number of mutant genes in their genotype, similar to models studied by~\citet{Kimura1966a} and~\citet{Haigh1978}.
-Moreover, their argument was based on expected increase of the population mean fitness by alleles that modify the mutation rate, rather than on the invasion success of such modifiers~\citep[i.e., analysis of \emph{evolutionary genetic stability}, see][]{Eshel1982, Lessard1990}.
+Moreover, their argument was based on the expected increase of the population mean fitness by alleles that modify the mutation rate, rather than on the invasion success of such modifiers~\citep[i.e., analysis of \emph{evolutionary genetic stability}, see][]{Eshel1982, Lessard1990}.
 
-Here, we present a general evolutionary model where the the type of the individual determines both its fitness and the rate at which it generates variation.
+Here, we present an evolutionary model where the the type of the individual determines both its fitness and the rate at which it generates variation.
 Our results show that the population mean fitness increases if individuals with below-average fitness produce more variation than individuals with above-average fitness,
@@
 Transition between types is determined by a combination of two effects:
-(i) the probability to transition \emph{out} of type $A_k$ is determined by $C_k$;
+(i) the probability of transitioning \emph{out} of type $A_k$ is determined by $C_k$;
 (ii) given a transition out of type $A_k$, the distribution of the destination types $A_i$ is given by $M_{i,k}$ (note the index order).
@@
 
-In the following section we present three examples of the model (Eq.~\ref{eq:model}) for mutation and migration.
+In the following section we present three examples of the model (Eq.~\ref{eq:model}) that apply to mutation and migration.
 
@@
 specifically, with probability $1-C_k$, the allele $A_k$ does not mutate, and with probability $\frac{C_k}{n-1}$, the allele $A_k$ mutates to $A_j$ for any $j \ne k$.
-This is an extension of a model studied by~\citet{Altenberg2017} that allows for the mutation rate of an individual, $C_k$, to depend on the identity of the allele $A_k$.
+This is an extension of a model studied by~\citet{Altenberg2017} that allows for the mutation rate of $A_k$, $C_k$, to depend on the identity of the allele $A_k$.
 
@@
 
-Let the frequency of $A_k$ in the present generation be $f_k$ such that $f_k \ge 0$ and $\sum_{k=0}^{n}{f_k}=1$.
+Let the frequency of $A_k$ in the present generation be $f_k$ with $f_k \ge 0$ and $\sum_{k=0}^{n}{f_k}=1$.
 Then after selection and mutation $f'$ in the next generation is given by
@@
 for $k=1,2,\ldots, n-1$.
+Here $\bar w=\sum_{k=0}^nw_kf_n$ is the mean fitness.
 
@@
 
+%ARE THERE OTHER MIGRATION MODELS (MODEL 1)?
 In this case we consider a large population of haploids that occupy $n$ demes, $A_1, \ldots, A_n$.
 Let the frequencies of individuals in deme $A_k$ be $f_k$ with $f_k \ge 0$ and $\sum_{k=1}^{n}{f_k} = 1$.
-The fitness of individuals in deme $A_k$ is $D_k$, but the entire population comes together for reproduction, and therefore reproductive success is determined by competition between individuals of both demes -- this has been termed \emph{hard selection}~\citep{Wallace1975,Karlin1982}.
+The fitness of individuals in deme $A_k$ is $D_k$, but the entire population comes together for reproduction, and therefore reproductive success is determined by competition among individuals of all demes -- this has been termed \emph{hard selection}~\citep{Wallace1975,Karlin1982}.
 
@@
 
-Let $e_k$ and $e\tr_k$ be the column and row vectors with 1 at position $k$ and 0 elsewhere, $Z_k = e_k e\tr_k$ the matrix with 1 at position $(k,k)$ and 0 elsewhere, and $[M]_k$ the $k$-th column of $M$.
+Let $e_k$ and $e\tr_k$ be the column and row vectors with 1 at position $k$ and 0 elsewhere, $Z_k = e_k e\tr_k$ be the matrix with 1 at position $(k,k)$ and 0 elsewhere, and $[M]_k$ be the $k$-th column of $M$.
 
@@
 and a modifier allele that regulates the mutation rates in individuals with at least $\pi$ deleterious mutations.
-Similarly, if the transition rates are controlled by a threshold rule 
+If the transition rates are controlled by a threshold rule, then 
 \begin{equation}
@@
 Consider the case of two modifier alleles, $m$ and $M$, inducing different transition probabilities $C=\diag{C_1, \ldots, C_n}$ and $\tilde{C}=\diag{\tilde{C}_1, \ldots, \tilde{C}_n}$, respectively.
-The frequencies of type $A_k$ linked to modifier $m$ or $M$ are $f_k$ and $g_k$, respectively, such that $\sum_{k=1}^{n}{(f_k + g_k)}=1$.
+The frequencies of type $A_k$ linked to modifier $m$ or $M$ are $f_k$ and $g_k$, respectively, where $\sum_{k=1}^{n}{(f_k + g_k)}=1$.
 $\bar{w}$ now ensures that $\sum_{k=1}^{n}{(f_k' + g_k')}=1$, 
@@
 We have shown that under constant-viability selection and in an effectively infinite haploid population at mutation-selection or migration-selection equilibrium, the stable population mean fitness increases if individuals with below-average fitness increase the rate at which variation is generated.
-Furthermore, modifier alleles that induce increased generation of variation in such individuals are favored by natural evolution.
+Furthermore, modifier alleles that  increase generation of variation in such individuals are favored by natural evolution.
 
@@
 For example, it has been demonstrated that higher mutation, recombination, and migration rates are favored by selection in populations evolving under fluctuating selection; see~\citet{Carja2014} and references therein.
-Here we have provided another general example, which suggests that a modifier allele that prompts individuals with below-average fitness to increase the rate at which variation is generated, will be favored by selection and will lead to increased population mean fitness.
+Here we have provided another general example, which suggests that a modifier allele that causes individuals with below-average fitness to increase the rate at which variation is generated, will be favored by selection and will lead to increased population mean fitness.
 
@@
 
-Now, let $\norm{x} = \sum_{k=1}^{n}{x_k}$, and therefore $x_t \goesto x$ implies
+Now, let $\norm{x} = \sum_{k=1}^{n}{x_k}$, so that $x_t \goesto x$ implies
 \begin{equation}
@@
 \end{multline}
-as $t \goesto \infty$, as $\norm{v}=1$ (eq.~\ref{eq:sum_uv}) and $\alpha \ne 0$ (eq.~\ref{eq:alpha0}).
+since $t \goesto \infty$, as $\norm{v}=1$ (eq.~\ref{eq:sum_uv}) and $\alpha \ne 0$ (eq.~\ref{eq:alpha0}).
 This completes the proof.

### 2017-11-16 ms.tex
D_k -> w_k; typos

@@
 Consider a large haploid population with an arbitrary set of types $A_1, A_2, \ldots, A_n$.
-The frequency and fitness of individuals of type $A_k$ are $f_k$ and $D_k$, respectively.
+The frequency and fitness of individuals of type $A_k$ are $f_k$ and $w_k$, respectively.
 The probability that an individual of type $A_k$ will transition to some other type is $C_k$, and given a transition occurs, the probability that it will transition to type $A_j$ is $M_{j,k}$.
@@
 \begin{equation}
-\bar{w} f'_k = (1-C_k)D_k f_k + \sum_{j=1}^{n}{C_j M_{k,j} D_j f_j},
+\bar{w} f'_k = (1-C_k) w_k f_k + \sum_{j=1}^{n}{C_j M_{k,j} w_j f_j},
 \label{eq:model_sum}
@@
 where $f=(f_1, f_2, \ldots, f_n)$ is a stochastic vector such that $\sum_{k=1}^n{f_k} = 1$;
-$D$ is a diagonal matrix with diagonal elements $D_k$ such that $D_k \ne D_j$ for $k \ne j$, and we can assume $0 < D_k \le 1$, i.e., fitness values are relative to the fitness of the fittest type; 
+$D$ is a diagonal matrix with diagonal elements $w_k$ such that $w_k \ne w_j$ for $k \ne j$, and we can assume $0 < w_k \le 1$, i.e., fitness values are relative to the fitness of the fittest type; 
 $C$ is a diagonal non-negative matrix with diagonal elements $C_k$;
@@
 
-Here we consider a large population of haploids and a trait determined by a single genetic locus with $n$ possible alleles $A_1, A_2, \ldots, A_n$ and corresponding fitness values $D_1, D_2, \ldots, D_n$.
+Here we consider a large population of haploids and a trait determined by a single genetic locus with $n$ possible alleles $A_1, A_2, \ldots, A_n$ and corresponding fitness values $w_1, w_2, \ldots, w_n$.
 The mutation rates $C_k$ of individuals with allele $A_k$ are potentially different;
@@
 \begin{equation}
-f_k' = \big(1 - C_k\big) \frac{D_k}{\bar{w}} f_k + \frac{1}{n-1} \sum_{k \ne j}{C_j \frac{D_j}{\bar{w}} f_j},
+f_k' = \big(1 - C_k\big) \frac{w_k}{\bar{w}} f_k + \frac{1}{n-1} \sum_{k \ne j}{C_j \frac{w_j}{\bar{w}} f_j},
 \label{eq:mutation_model_1}
 \end{equation}
-for all $k=1,2,\ldots,n$, where $\bar{w} = \sum_{k=1}^{n}{D_k f_k}$ is the population mean fitness.
+for all $k=1, 2, \ldots, n$, where $\bar{w} = \sum_{k=1}^{n}{f_k w_k}$ is the population mean fitness.
 
@@
 Again, we consider a large population of haploids, but here individuals with genotype $A_k$ are characterized by the number $k$ of deleterious or mutant alleles in their genotype, where $0 \le k \le n$.
-Specifically, the fitness of individuals with $k$ mutant alleles is $D_k$,
-And the probability $C_k$ that a mutation occurs in individuals with $k$ mutant alleles depends on $k$.
+Specifically, the fitness of individuals with $k$ mutant alleles is $w_k$ ($w_0 \ge w_1 \ge \ldots > w_n$),
+and the probability $C_k$ that a mutation occurs in individuals with $k$ mutant alleles depends on $k$.
 We assume a low mutation rate $C_k$ so that two mutations are unlikely to occur in the same individual in one generation.
@@
 
-Let the frequency of $A_k$ in the present generation be $f_k$ with $f_k \ge 0$ and $\sum_{k=0}^{n}{f_k}=1$.
+Let the frequency of $A_k$ in the present generation be $f_k$ with $f_k \ge 0$ and $\sum_{k=1}^{n}{f_k}=1$.
 Then after selection and mutation $f'$ in the next generation is given by
@@
 \begin{aligned}
-f_0' &= \big(1 - \delta C_0\big) \frac{D_0}{\bar{w}} f_0 + \beta C_{1} \frac{D_{1}}{\bar{w}} f_{1}, \\
-f_n' &= \big(1 - \beta C_n\big) \frac{D_n}{\bar{w}} f_n + \delta C_{n-1} \frac{D_{n-1}}{\bar{w}} f_{n-1}, \\
-f_k' &= \big(1 - C_k(\delta+\beta)\big) \frac{D_k}{\bar{w}} f_k + \\
-	 & \delta C_{k-1} \frac{D_{k-1}}{\bar{w}} f_{k-1} + 
-	 \beta C_{k+1} \frac{D_{k+1}}{\bar{w}} f_{k+1},
+f_0' &= \big(1 - \delta C_0\big) \frac{w_0}{\bar{w}} f_0 + \beta C_{1} \frac{w_{1}}{\bar{w}} f_{1}, \\
+f_n' &= \big(1 - \beta C_n\big) \frac{w_n}{\bar{w}} f_n + \delta C_{n-1} \frac{w_{n-1}}{\bar{w}} f_{n-1}, \\
+f_k' &= \big(1 - C_k(\delta+\beta)\big) \frac{w_k}{\bar{w}} f_k + \\
+	 & \delta C_{k-1} \frac{w_{k-1}}{\bar{w}} f_{k-1} + 
+	 \beta C_{k+1} \frac{w_{k+1}}{\bar{w}} f_{k+1},
 \end{aligned}
@@
 for $k=1,2,\ldots, n-1$.
-Here $\bar w=\sum_{k=0}^nw_kf_n$ is the mean fitness.
+Here $\bar{w}=\sum_{k=1}^{n}{f_k w_k}$ is the population mean fitness.
 
@@
 \end{equation}
-with the beneficial, neutral, and deleterious mutation probabilities on the three main digonals and zeros elsewhere, 
+with the beneficial, neutral, and deleterious mutation probabilities on the three main diagonals and zeros elsewhere, 
 Eq.~\ref{eq:mutation_model_2} can be viewed as a special case of Eq.~\ref{eq:model}.
@@
 
-% Migration model 1
-\paragraph*{Migration model 1.}\label{sec:mig_model_1}
+% Migration model
+\paragraph*{Migration model.}\label{sec:mig_model}
 
-%ARE THERE OTHER MIGRATION MODELS (MODEL 1)?
 In this case we consider a large population of haploids that occupy $n$ demes, $A_1, \ldots, A_n$.
 Let the frequencies of individuals in deme $A_k$ be $f_k$ with $f_k \ge 0$ and $\sum_{k=1}^{n}{f_k} = 1$.
-The fitness of individuals in deme $A_k$ is $D_k$, but the entire population comes together for reproduction, and therefore reproductive success is determined by competition among individuals of all demes -- this has been termed \emph{hard selection}~\citep{Wallace1975,Karlin1982}.
+The fitness of individuals in deme $A_k$ is $w_k$, but the entire population comes together for reproduction, and therefore reproductive success is determined by competition among individuals of all demes -- this has been termed \emph{hard selection}~\citep{Wallace1975,Karlin1982}.
 
@@
 \frac{\partial \bar{w}^*}{\partial C_k} = 
-\frac{u_k v_k}{C_k} (\bar{w}^* - D_k),
+\frac{u_k v_k}{C_k} (\bar{w}^* - w_k),
 \end{equation}
@@
 sign\frac{\partial \bar{w}^*}{\partial C_k} = 
-sign(\bar{w}^* - D_k).
+sign(\bar{w}^* - w_k).
 \end{equation}
@@
 u\tr (0 - Z_k + M Z_k)D v = \\
--v_k u_k D_k + v_k D_k u\tr[M]_k = \\
-v_k D_k (u\tr[M]_k - u_k). 
+-v_k u_k w_k + v_k w_k u\tr[M]_k = \\
+v_k w_k (u\tr[M]_k - u_k). 
 \end{aligned}
@@
 \begin{equation}
-u_k \bar{w}^* = (1-C_k) D_k u_k + C_k D_k u\tr [M]_k.
+u_k \bar{w}^* = (1-C_k) w_k u_k + C_k w_k u\tr [M]_k.
 \end{equation}
-Multiplying both sides by $v_k$ (which is positive if $D_k>0$) and rearranging, we get 
+Multiplying both sides by $v_k$ (which is positive if $w_k>0$) and rearranging, we get 
 \begin{equation}
-\frac{u_k v_k}{C_k} (\bar{w}^* - D_k) = v_k D_k(u\tr [M]_k - u_k),
+\frac{u_k v_k}{C_k} (\bar{w}^* - w_k) = v_k w_k (u\tr [M]_k - u_k),
 \end{equation}
@@
 \frac{\partial \bar{w}^*}{\partial C_k} = 
-\frac{u_k v_k}{C_k} (\bar{w}^* - D_k).
+\frac{u_k v_k}{C_k} (\bar{w}^* - w_k).
 \end{equation}
@@
 sign \frac{\partial \bar{w}^*}{\partial C_k} = 
-sign (\bar{w}^* - D_k),
+sign (\bar{w}^* - w_k),
 \end{equation}
@@
 \frac{\partial \bar{w}^*}{\partial C_k} = 
-v_k D_k (u\tr[M]_k - u_k) = \\
-D_k (\sum_i{u_i M_{i,k} v_k} - u_k v_k).
+v_k w_k (u\tr[M]_k - u_k) = \\
+w_k (\sum_i{u_i M_{i,k} v_k} - u_k v_k).
 \end{aligned}
@@
 Consequently, $u_k v_k$ is the \emph{ancestor frequency} of type $k$~\citep{Hermisson2002}, namely the fraction of the equilibrium population descended from type $k$.
-The sum $\sum_i{u_i M_{i,k} v_k}$ can be similarly interpreted as the fraction of the equilibrium population descended from individuals that transitioned from type $k$ to another type (via the $k$ column of the transition matrix $M$), conditioned on transition occuring.
+The sum $\sum_i{u_i M_{i,k} v_k}$ can be similarly interpreted as the fraction of the equilibrium population descended from individuals that transitioned from type $k$ to another type (via the $k$ column of the transition matrix $M$), conditioned on transition occurring.
 
-Given $0 < D_k \le 1$, from Eq.~\ref{mfp_intermediate} we have the following corollary.
+Given $0 < w_k \le 1$, from Eq.~\ref{mfp_intermediate} we have the following corollary.
 
@@
 u\tr \Big( \sum_{k \in K}{\frac{\partial (I - C + MC)D}{\partial \tau}} \Big) v = \\
-\frac{1}{\tau} \sum_{k \in K}{u_k v_k (\bar{w}^* - D_k)}.
+\frac{1}{\tau} \sum_{k \in K}{u_k v_k (\bar{w}^* - w_k)}.
 \end{aligned}
@@
 sign \frac{\partial \bar{w}^*}{\partial \tau} = 
-sign \sum_{k \in K}{u_k v_k (\bar{w}^* - D_k)} = \\ 
-sign \Big(\bar{w}^* - \frac{\sum_{k \in K}{u_k v_k D_k}}{\sum_{k \in K}{u_k v_k}}\Big), 
+sign \sum_{k \in K}{u_k v_k (\bar{w}^* - w_k)} = \\ 
+sign \Big(\bar{w}^* - \frac{\sum_{k \in K}{u_k v_k w_k}}{\sum_{k \in K}{u_k v_k}}\Big), 
 \end{multline}
@@
 \end{equation}
-Here, $\bar{w}=\sum_{k=1}^{n}{(f_k+g_k)D_k}$ is the mean fitness of the entire population (see~\nameref{sec:AppA}). 
+Here, $\bar{w}=\sum_{k=1}^{n}{(f_k + g_k) w_k}$ is the mean fitness of the entire population (see~\nameref{sec:AppA}). 
 Note that Eq.~\ref{eq:model} is the special case of Eq.~\ref{eq:modifier_model} where allele $M$ is absent; i.e. $g_k=0$ for all $k$.
@@
 sign \frac{\partial \lambda_1}{\partial \tilde{C}_k}\bigg\rvert_{\tilde{C}_k = C_k} = 
-sign(\bar{w}^* - D_k).
+sign(\bar{w}^* - w_k).
 \end{equation}
@@
 sign \frac{\partial \lambda_1}{\partial \tilde{C}_k} =
-sign\Big(\lambda_1 - \frac{D_k}{\bar{w}^*}\Big).
+sign\Big(\lambda_1 - \frac{w_k}{\bar{w}^*}\Big).
 \end{equation}
@@
 sign \frac{\partial \lambda_1}{\partial \tilde{C}_k}\bigg\rvert_{\tilde{C}_k = C_k} = 
-sign\Big(1 - \frac{D_k}{\bar{w}^*}\Big) =  
-sign(\bar{w}^* - D_k),
+sign\Big(1 - \frac{w_k}{\bar{w}^*}\Big) =  
+sign(\bar{w}^* - w_k),
 \end{multline}
@@
 $$
-because $\bar{w}^* = \sum_{k=1}^{n}{v_k D_k}$ (see~\nameref{sec:AppA}), and $D_k \le 1$ for all $k$.
+because $\bar{w}^* = \sum_{k=1}^{n}{v_k w_k}$ (see~\nameref{sec:AppA}), and $w_k \le 1$ for all $k$.
 
@@
 \begin{equation}
-\bar{w} = \sum_k{f_k D_k}.
+\bar{w} = \sum_k{f_k w_k}.
 \end{equation}
@@
 \diag{D} f = 
-\sum_k{f_k D_k},
+\sum_k{f_k w_k},
 \end{multline}
@@
 
-Note that this lemma also ensures that the stable population mean fitness $\bar{w}^*=\sum_k{v_k D_k}$ is the leading eigenvalue of $(I - C + M C) D$; see Eq.~\ref{eq:model_equilibrium} and the \emph{Perron-Frobenius theorem}~\cite[Appendix~A]{Otto2007}.
+Note that this lemma also ensures that the stable population mean fitness $\bar{w}^*=\sum_k{v_k w_k}$ is the leading eigenvalue of $(I - C + M C) D$; see Eq.~\ref{eq:model_equilibrium} and the \emph{Perron-Frobenius theorem}~\cite[Appendix~A]{Otto2007}.
 
@@
 \subsection*{Appendix D}\label{sec:AppD}
+
 \begin{lemma}[Stability of the equilibrium $v$]
-If $C$ and $D$ are positive diagonal matrices, and $M$ is block irreducible, then
-the equilibrium $v$ of the system in Eq.~\ref{eq:model_equilibrium} is globally stable. \end{lemma}
+If $C$ and $D$ are positive diagonal matrices, and $M$ is block irreducible, 
+then the equilibrium $v$ of the system in Eq.~\ref{eq:model_equilibrium} is globally stable.
+\end{lemma}
+
 \begin{proof}
@@
 \begin{equation}
-\norm{x_t} \goesto \norm{x}
+\norm{x_t} \goesto \norm{x},
 \label{eq:norm_limit}
 \end{equation}
-as $t \goesto \infty$.
-
-Rewrite Eq.~\ref{eq:model} as 
+as $t \goesto \infty$ and rewrite Eq.~\ref{eq:model} as 
 \begin{equation}
-f' = \frac{A f}{\norm{A f}} 
+f' = \frac{A f}{\norm{A f}}. 
 \label{eq:model_norm}
 \end{equation}
-Note that when applying eq.~\ref{eq:model_norm} $t$ times, the normalization can be applied once~\cite[Appendix~A]{Wakano2004}.
 
-Therefore, for a positive frequency vector $f$, we can apply $f \goesto f'$ (Eq.~\ref{eq:model_norm}) $t$ times, and using eqs.~\ref{eq:perron_projection} and ~\ref{eq:norm_limit}, we have
+Therefore, for a positive frequency vector $f$, we can apply $f \goesto f'$ (Eq.~\ref{eq:model_norm}) $t$ times, 
+and as $f \goesto Af$ is linear, one normalization ($x \goesto x/\norm{x}$) suffices~\citep[see][Appendix~A]{Wakano2004}.
+Using eqs.~\ref{eq:perron_projection} and ~\ref{eq:norm_limit}, we have
 \begin{multline}

### 2017-11-16 ms.tex
learning model

@@
 % Model
-\section*{Models}
+\section*{Models}\label{sec:models}
 
@@
 
-In the following section we present three examples of the model (Eq.~\ref{eq:model}) that apply to mutation and migration.
+In the following section we present four examples of the model (Eq.~\ref{eq:model}) that apply to mutation, migration, and learning.
 
@@
 where $p_k \ge 0$ and $\sum_{k=1}^{n}{p_k}=1$.
+
+% Learning model
+\paragraph*{Learning model.}\label{sec:learn_model}
+
+In our final example, we consider a large population and an integer phenotype $k$ where $1 \le k \le n$.
+Individuals are characterized by their initial and mature phenotypes~\citep[pg.~94]{Boyd1985}.
+Fitness is determined by the difference between the individual mature phenotype and the optimal phenotype:
+denoting the optimal phenotype as $h$, the fitness of an individual with mature phenotype $k$ is $w_{|k-h|}$.
+
+The initial phenotype is acquired by learning the mature phenotype of one of the parents (i.e. uni-parental vertical transmission).
+In individuals with initial phenotype $k$, the mature phenotype is the same as the initial phenotype with probability $1-C_k$, and is modified by individual learning or \emph{exploration}~\citep{Borenstein2008} with probability $C_k$.
+Such individual exploratory learning, which can be considered intentional or the result of transmission infidelity, modifies initial phenotype $k$ to mature phenotype $j$ with probability $M_{k,j}$.
+
+For example, the case of \emph{symmetric individual learning}~\citep{Borenstein2008} can be represented by
+\begin{equation}
+M_{k,j} = \begin{cases}
+\frac{1}{2b+1}, & \abs{k-j} \le b \\
+0, & otherwise
+\end{cases},
+\end{equation}
+where the breadth of exploration is $b \ge 0$, and phenotypes $0$ and $n$ "absorb" transitions to $k<0$ and $k>n$, respectively. 
+For instance, for $n=5$ and $b=1$ we have
+\begin{equation}
+M = \begin{bmatrix}
+2/3 & 1/3 & 0 & 0 & 0 \\
+1/3 & 1/3 & 1/3 & 0 & 0 \\
+0 & 1/3 & 1/3 & 1/3 & 0 \\
+0 & 0 & 1/3 & 1/3 & 1/3 \\
+0 & 0 & 0 & 1/3 & 2/3
+\end{bmatrix}.
+\end{equation}
+
+Therefore, if the frequency of individuals with mature phenotype $k$ in the current generation is $f_k$, then the frequency in the next generation $f'_k$ is
+\begin{equation}
+f'_k = (1-C_k) f_k \frac{w_{\abs{k-h}}}{\bar{w}} \sum_{j=1}^{n}{C_j M_{k,j} \frac{w_{\abs{j-h}}}{\bar{w}} f_j },
+\end{equation}
+where $\bar{w} = \sum_{k=1}^{n}{f_k w_{\abs{k-h}}}$ is the population mean fitness.
 
@@
 % see Altenberg 2012 Ecological Monographs for a condition-dependent migration model
-We have given several examples of variation-generating processes for which this principle applies -- namely mutation and migration -- but our model may apply to other processes as well.
-For example, the \emph{reduction principle} applies to ecological models of dispersal, and~\citet{Gueijman2013} have demonstrated that even in homogeneous environments, fitness-associated dispersal increases the population mean fitness and is favored by selection over uniform dispersal.
-% TODO see cultural transmission model in Lee's thesis
-Similarly, if the transmission fidelity of culturally-transmitted traits depends on the type or fitness of the transmitting individual, we expect that our results will hold.
+We have given several examples of variation-generating processes for which this principle applies -- namely mutation, migration, and learning, see ~\nameref{sec:models} -- but our model may apply to other processes as well.
+For example, the \emph{reduction principle} applies to ecological models of dispersal, and~\citet{Gueijman2013} have demonstrated that even in homogeneous environments, fitness-associated dispersal increases the mean fitness of diploid populations and is favored by selection over uniform dispersal.
+Similarly, if the transmission fidelity of culturally-transmitted traits depends on the type or fitness of the transmitting individual, we expect that our results will hold (see~\nameref{sec:learn_model}).
 

### 2017-11-16 ms.tex
conditional dispersal remark

@@
 In their models, the only departure from the \emph{reduction principle} assumptions was due to the differential effect the modifier has on different genotypes.
-A similar result has been demonstrated for fitness-associated recombination~\citep{Hadany2003a} and for condition-dependent sexual reproduction~\citep{Hadany2007a}, and evidence suggests that both mechanisms are common in nature~\citep{Ram2016}.
+A similar result has been demonstrated for conditional dispersal~\citep[Th.~39]{Altenberg2012b}, fitness-associated recombination~\citep{Hadany2003a} and for condition-dependent sexual reproduction~\citep{Hadany2007a}, and evidence suggests that both mechanisms are common in nature~\citep{Ram2016}.
 

### 2017-11-16 ms.tex
primitive condition; I typeface; nameref emph

@@
 \newcommand{\cj}{\mathbf{J}}
-\newcommand{\ci}{\mathbf{I}}
+\newcommand{\ci}{I}
 
@@
 where $f=(f_1, f_2, \ldots, f_n)$ is a stochastic vector such that $\sum_{k=1}^n{f_k} = 1$;
-$D$ is a diagonal matrix with diagonal elements $w_k$ such that $w_k \ne w_j$ for $k \ne j$, and we can assume $0 < w_k \le 1$, i.e., fitness values are relative to the fitness of the fittest type; 
-$C$ is a diagonal non-negative matrix with diagonal elements $C_k$;
-$M$ is a column-stochastic matrix: $M_{j,k} \ge 0$ for all $j,k$ and $\sum_{j=1}^n {M_{j,k}} = 1$ for all $k$; 
+$D$ is a diagonal matrix with diagonal elements $w_k$ such that $w_k \ne w_j$ for $k \ne j$, and we can assume $0 < w_k \le 1$, i.e., fitness values are positive and relative to the fitness of the fittest type;
+$C$ is a diagonal positive matrix with diagonal elements $C_k$;
+$M$ is a primitive column-stochastic matrix: $M_{j,k} \ge 0$ for all $j,k$, $\sum_{j=1}^n {M_{j,k}} = 1$ for all $k$, and $M^n_{j,k} > 0$ for all $j,k$ for some natural $n$.
 $\ci$ is the identity matrix;
@@
 
-The initial phenotype is acquired by learning the mature phenotype of one of the parents (i.e. uni-parental vertical transmission).
+The initial phenotype is acquired by learning the mature phenotype of a parent.
 In individuals with initial phenotype $k$, the mature phenotype is the same as the initial phenotype with probability $1-C_k$, and is modified by individual learning or \emph{exploration}~\citep{Borenstein2008} with probability $C_k$.
-Such individual exploratory learning, which can be considered intentional or the result of transmission infidelity, modifies initial phenotype $k$ to mature phenotype $j$ with probability $M_{k,j}$.
+Such individual exploratory learning, which can be considered either intentional or the result of incorrect learning, modifies initial phenotype $k$ to mature phenotype $j$ with probability $M_{k,j}$.
 
@@
 where the breadth of exploration is $b \ge 0$, and phenotypes $0$ and $n$ "absorb" transitions to $k<0$ and $k>n$, respectively. 
-For instance, for $n=5$ and $b=1$ we have
+For instance, with $n=5$ and $b=1$ we have
 \begin{equation}
@@
 \begin{equation}
-f'_k = (1-C_k) f_k \frac{w_{\abs{k-h}}}{\bar{w}} \sum_{j=1}^{n}{C_j M_{k,j} \frac{w_{\abs{j-h}}}{\bar{w}} f_j },
+f'_k = (1-C_k) f_k \frac{w_{\abs{k-h}}}{\bar{w}} + \sum_{j=1}^{n}{C_j M_{k,j} \frac{w_{\abs{j-h}}}{\bar{w}} f_j },
 \end{equation}
@@
 and $u$ and $v$ be the corresponding left and right eigenvectors.
-%TODO If $MCD$ is primitive, then
 Then,
@@
 
-% see Altenberg 2012 Ecological Monographs for a condition-dependent migration model
-We have given several examples of variation-generating processes for which this principle applies -- namely mutation, migration, and learning, see ~\nameref{sec:models} -- but our model may apply to other processes as well.
+We have given several examples of variation-generating processes for which this principle applies -- namely mutation, migration, and learning (see \emph{\nameref{sec:models}} section) -- but our model may apply to other processes as well.
 For example, the \emph{reduction principle} applies to ecological models of dispersal, and~\citet{Gueijman2013} have demonstrated that even in homogeneous environments, fitness-associated dispersal increases the mean fitness of diploid populations and is favored by selection over uniform dispersal.
-Similarly, if the transmission fidelity of culturally-transmitted traits depends on the type or fitness of the transmitting individual, we expect that our results will hold (see~\nameref{sec:learn_model}).
+Similarly, if the transmission fidelity of culturally-transmitted traits depends on the type or fitness of the transmitting individual, we expect that our results will hold (see \emph{\nameref{sec:learn_model}} section).
 
 Eq.~\ref{mfp} is a generalization of a result of~\citet[Eq.~4]{Ram2012}.
-\citeauthor{Ram2012} modeled the accumulation of mutant alleles in a population evolving at the mutation-selection balance (see \nameref{sec:mut_model_2}).
+\citeauthor{Ram2012} modeled the accumulation of mutant alleles in a population evolving at the mutation-selection balance (see \emph{\nameref{sec:mut_model_2}} section).
 Using Eq.~\ref{eq:Caswells_formula} in \nameref{sec:AppB} and a recursion on the ratios of the reproductive values \cite[see][eqs.~A5-6]{Ram2012}, they concluded that at the mutation-selection balance, if individuals with below-average fitness ($D_k < \bar{w}^*$) increase their mutation rate, then the population mean fitness will increase -- a result generalized by our \emph{mean fitness principle} in Eqs.~\ref{mfp_sign} and~\ref{mfp_sign_threshold}.
@@
 and empirical evidence suggests that \emph{stress-induced mutagenesis} is common in bacteria and yeast, and may be prevalent in plants, flies, and human cancer cells~\citep{Rosenberg2012,Fitzgerald2017b}.
-Similar theoretical results have been demonstrated for a modifier that increases the recombination rate in maladapted individuals~\citep{Hadany2003a,Hadany2003b}.
+Similar theoretical results have been demonstrated for a modifier that increases the recombination rate in maladapted individuals~\citep{Hadany2003b,Hadany2003a}.
 

### 2017-11-19 ms.tex
fixed another sum
MWF changes

@@
 \paragraph*{General model.}\label{sec:general_model}
-Consider a large haploid population with an arbitrary set of types $A_1, A_2, \ldots, A_n$.
+Consider a large  population with an arbitrary set of types $A_1, A_2, \ldots, A_n$.
 The frequency and fitness of individuals of type $A_k$ are $f_k$ and $w_k$, respectively.
@@
 \end{equation}
-where $f=(f_1, f_2, \ldots, f_n)$ is a stochastic vector such that $\sum_{k=1}^n{f_k} = 1$;
-$D$ is a diagonal matrix with diagonal elements $w_k$ such that $w_k \ne w_j$ for $k \ne j$, and we can assume $0 < w_k \le 1$, i.e., fitness values are positive and relative to the fitness of the fittest type;
-$C$ is a diagonal positive matrix with diagonal elements $C_k$;
-$M$ is a primitive column-stochastic matrix: $M_{j,k} \ge 0$ for all $j,k$, $\sum_{j=1}^n {M_{j,k}} = 1$ for all $k$, and $M^n_{j,k} > 0$ for all $j,k$ for some natural $n$.
-$\ci$ is the identity matrix;
-and $\bar{w}$ is the normalizing factor such that $\sum_{k=1}^n{f'_k}=1$ and is equal to the population mean fitness (see~\nameref{sec:AppA}).
+where $f=(f_1, f_2, \ldots, f_n)$ is a frequency vector with $f_k\ge 0$ and  $\sum_{k=1}^n{f_k} = 1$;
+$D$ is a positive diagonal matrix with  elements $w_k$ such that $w_k \ne w_j$ for some $k \ne j$, and we can assume $0 < w_k \le 1$, i.e., fitness values are positive and relative to the fitness of the fittest type;
+$C$ is a positive diagonal  matrix with  elements $C_k$;
+$M$ is a primitive column-stochastic matrix: $M_{j,k} \ge 0$ for all $j,k$, $\sum_{j=1}^n {M_{j,k}} = 1$ for all $k$, and $(M^n)_{j,k} > 0$ for all $j,k$ for some natural $n$.
+$\ci$ is the $n\times n$ identity matrix;
+and $\bar{w}$ is the normalizing factor such that $\sum_{k=1}^n{f'_k}=1$ and is equal to the population mean fitness.
 
@@
 \begin{equation}
-f_k' = \big(1 - C_k\big) \frac{w_k}{\bar{w}} f_k + \frac{1}{n-1} \sum_{k \ne j}{C_j \frac{w_j}{\bar{w}} f_j},
+f_k' = \big(1 - C_k\big) \frac{w_k}{\bar{w}} f_k + \frac{1}{n-1} \sum_{j \ne k}{C_j \frac{w_j}{\bar{w}} f_j},
 \label{eq:mutation_model_1}
@@
 with zeros on the diagonal and $\frac{1}{n-1}$ elsewhere.
-Note that here ${M}$ is irreducible.
+Note that here ${M}$ is irreducible and primitive.
 
@@
 Again, we consider a large population of haploids, but here individuals with genotype $A_k$ are characterized by the number $k$ of deleterious or mutant alleles in their genotype, where $0 \le k \le n$.
-Specifically, the fitness of individuals with $k$ mutant alleles is $w_k$ ($w_0 \ge w_1 \ge \ldots > w_n$),
+Specifically, the fitness of individuals with $k$ mutant alleles is $w_k$ ($w_0 \ge w_1 \ge \ldots \ge w_n$),
 and the probability $C_k$ that a mutation occurs in individuals with $k$ mutant alleles depends on $k$.
@@
 Eq.~\ref{eq:mutation_model_2} can be viewed as a special case of Eq.~\ref{eq:model}.
-Here, too, ${M}$ is irreducible.
+Here, too, ${M}$ is irreducible and primitive.
 
@@
 
-After reproduction, offspring of individuals from deme $A_k$ return to their parental deme with probability $1-C_k$, or migrate to a different deme $A_j$ with probability $C_k \cdot M_{j,k}$, where $C_k > 0$, $M_{j,k} \ge 0$, and $\sum_{k=1}^{n}{M_{j,k}} = 1$ for all $j=1, \ldots, n$.
+After reproduction, offspring of individuals from deme $A_k$ return to their parental deme with probability $1-C_k$, or migrate to a different deme $A_j$ with probability $C_k \cdot M_{j,k}$, where matrix $M$ is primitive, $C_k > 0$, $M_{j,k} \ge 0$, and $\sum_{k=1}^{n}{M_{j,k}} = 1$ for all $j=1, \ldots, n$.
 Therefore, $1-C_k$ can be interpreted as a \emph{homing rate}.
@@
 \end{equation}
-with $\sum_{k=1}^{n}{m_k}=1$,
+with $m_k>0$ and $\sum_{k=1}^{n}{m_k}=1$,
 then $m_k$ can be considered the relative population size of deme $A_k$ --
@@
 \end{equation}
-where $p_k \ge 0$ and $\sum_{k=1}^{n}{p_k}=1$.
+where $p_k > 0$ and $\sum_{k=1}^{n}{p_k}=1$.
 
@@
 
-Write the equilibrium of the frequency vector $f$ in Eq.~\ref{eq:model} as $v$ and the stable population mean fitness as $\bar{w}^*$, then
+Write the equilibrium  frequency vector $f$ in Eq.~\ref{eq:model} as $\hat v$ and the stable population mean fitness as $\hat{\bar w}$, then
 \begin{equation}\label{eq:model_equilibrium}
-\bar{w}^* v = (I-C+MC)D v.
+\hat{\bar w} \hat v = (I-C+MC)D \hat v.
 \end{equation}
-Note that (i) the existence of $\bar{w}^*$ and $v$ are guaranteed by the \emph{Perron-Frobenius theorem}~\cite[Appendix~A]{Otto2007} because $(I-C+MC)D$ is a non-negative matrix; 
-(ii) the global stability of this equilibrium is proven in~\nameref{sec:AppD}; and
-(iii) $\bar{w}^*$ is identified as the stable population mean fitness in~\nameref{sec:AppA}.
+Note that (i) the existence and uniqueness of $\hat{\bar w}$ and $\hat v$ are guaranteed by the \emph{Perron-Frobenius theorem}~(\cite{Otto2007}) because $(I-C+MC)D$ is a non-negative primitive matrix;
+(ii) the global stability of this equilibrium is proven in~\nameref{sec:AppC}.
 
-The following lemma presents a \emph{mean fitness principle} for the
-sensitivity of the stable population mean fitness $\bar{w}^*$ to changes
+The following result presents a \emph{mean fitness principle} for the
+sensitivity of the equilbrium mean fitness $\hat{\bar w}$ to changes
 in $C_k$, the probability of transition from $A_k$.
+\medskip
 
-\begin{lemma}[Mean fitness principle]\label{lemma:mfp}
-Let $\bar{w}^*$ be the leading eigenvalue of $(I-C+MC)D$,
-and $u$ and $v$ be the corresponding left and right eigenvectors.
+\noindent{\bf Result 1. Mean fitness principle.}
+{\sl Let $\hat{\bar w}$ be the leading eigenvalue of $(I-C+MC)D$,
+and $\hat u$ and $\hat v$ be the corresponding left and right eigenvectors.
 Then,
 \begin{equation}\label{mfp}
-\frac{\partial \bar{w}^*}{\partial C_k} = 
-\frac{u_k v_k}{C_k} (\bar{w}^* - w_k),
+\frac{\partial \hat{\bar w}}{\partial C_k} = 
+\frac{\hat u_k \hat v_k}{C_k} (\hat{\bar w} - w_k),
 \end{equation}
@@
 \begin{equation}\label{mfp_sign}
-sign\frac{\partial \bar{w}^*}{\partial C_k} = 
-sign(\bar{w}^* - w_k).
+sign\frac{\partial \hat{\bar w}}{\partial C_k} = 
+sign(\hat{\bar w} - w_k).
 \end{equation}
 
-Therefore increased transition from type $k$ will increase the stable population mean fitness if the fitness of type $k$ is below the stable population mean fitness.
-\end{lemma}
+Therefore increased transition from type $k$ will increase the stable population mean fitness if the fitness of type $k$ is below the stable population mean fitness.}
 
 \begin{proof}
-$\bar{w}^*$ is the leading (real positive) eigenvalue of $(I-C+MC)D$ (see eq.~\ref{eq:model_equilibrium}),
-and the corresponding left and right eigenvectors $u$ and $v$ are both non-negative and uniquely defined by $\sum_k{v_k} = 1$ and $\sum_k{u_k v_k} = 1$.
+$\hat{\bar w}$ is the leading (real positive) eigenvalue of $(I-C+MC)D$ (see eq.~\ref{eq:model_equilibrium}),
+and the corresponding left and right eigenvectors $u$ and $v$ are both non-negative and uniquely defined by $\sum_k{\hat v_k} = 1$ and $\sum_k{\hat u_k \hat v_k} = 1$.
 
-Using the formula in~\citet{Caswell1978} (see eq.~\ref{eq:Caswells_formula} in~\nameref{sec:AppB}),
+Using the formula in~\citet{Caswell1978} (see eq.~\ref{eq:Caswells_formula} in~\nameref{sec:AppA}),
 \begin{equation}
-\frac{\partial \bar{w}^*}{\partial C_k} = 
-u\tr \frac{\partial (I-C+MC)D}{\partial C_k} v.
+\frac{\partial \hat{\bar w}}{\partial C_k} = 
+\hat u\tr \frac{\partial (I-C+MC)D}{\partial C_k} \hat v.
 \end{equation}
@@
 \begin{aligned}
-u\tr \frac{\partial (I-C+MC)D}{\partial C_k} v = 
-u\tr (0 - Z_k + M Z_k)D v = \\
--v_k u_k w_k + v_k w_k u\tr[M]_k = \\
-v_k w_k (u\tr[M]_k - u_k). 
+\hat u\tr \frac{\partial (I-C+MC)D}{\partial C_k} \hat v = 
+\hat u\tr (0 - Z_k + M Z_k)D \hat v = \\
+-\hat v_k \hat u_k w_k + \hat v_k w_k \hat u\tr[M]_k = \\
+\hat v_k w_k (\hat u\tr[M]_k - \hat u_k). 
 \end{aligned}
@@
 \begin{equation}
-u\tr \bar{w}^* = u\tr (I - C + MC) D,
+\hat u\tr \hat{\bar w} = \hat u\tr (I - C + MC) D,
 \end{equation}
-which gives us a relation between $\bar{w}^*$ and the $k$
+which gives us a relation between $\hat{\bar w}$ and the $k$
 element of $u$:
 \begin{equation}
-u_k \bar{w}^* = (1-C_k) w_k u_k + C_k w_k u\tr [M]_k.
+\hat u_k \hat{\bar w} = (1-C_k) w_k \hat u_k + C_k w_k \hat u\tr [M]_k.
 \end{equation}
-Multiplying both sides by $v_k$ (which is positive if $w_k>0$) and rearranging, we get 
+Multiplying both sides by $\hat v_k$  and rearranging, we get 
 \begin{equation}
-\frac{u_k v_k}{C_k} (\bar{w}^* - w_k) = v_k w_k (u\tr [M]_k - u_k),
+\frac{\hat u_k \hat v_k}{C_k} (\hat{\bar w} - w_k) = \hat v_k w_k (\hat u\tr [M]_k - \hat u_k),
 \end{equation}
@@
 \begin{equation}
-\frac{\partial \bar{w}^*}{\partial C_k} = 
-\frac{u_k v_k}{C_k} (\bar{w}^* - w_k).
+\frac{\partial \hat{\bar w}}{\partial C_k} = 
+\frac{\hat u_k \hat v_k}{C_k} (\hat{\bar w} - w_k).
 \end{equation}
-Finally, note that $u_k, v_k, C_k \ge 0$ and therefore
+Finally, note that $\hat u_k, \hat v_k, C_k \ge 0$ and therefore
 \begin{equation}
-sign \frac{\partial \bar{w}^*}{\partial C_k} = 
-sign (\bar{w}^* - w_k),
+sign \frac{\partial \hat{\bar w}}{\partial C_k} = 
+sign (\hat{\bar w} - w_k),
 \end{equation}
@@
 
-The above lemma provides a condition for the effect of changing $C_k$, the probability for transition from $A_k$, on the stable population mean fitness.
+The above result provides a condition for the effect of changing $C_k$, the probability for transition from $A_k$, on the stable population mean fitness.
 Specifically, if $A_k$ individuals have below-average fitness, then increasing $C_k$ will increase the population mean fitness.
@@
 \begin{aligned}
-\frac{\partial \bar{w}^*}{\partial C_k} = 
-v_k w_k (u\tr[M]_k - u_k) = \\
-w_k (\sum_i{u_i M_{i,k} v_k} - u_k v_k).
+\frac{\partial \hat{\bar w}}{\partial C_k} = 
+\hat v_k w_k (\hat u\tr[M]_k - \hat u_k) = \\
+w_k (\sum_{j=1}^{n}{\hat u_j M_{j,k} \hat v_k} - \hat u_k \hat v_k).
 \end{aligned}
 \end{equation}
-Here, $u_k$ can be regarded as the \emph{reproductive value}~\citep[pg.~27]{Fisher1930} of type $k$ (see~\nameref{sec:AppC}), which gives the relative contribution of type $k$ to the long-term population.
-Consequently, $u_k v_k$ is the \emph{ancestor frequency} of type $k$~\citep{Hermisson2002}, namely the fraction of the equilibrium population descended from type $k$.
-The sum $\sum_i{u_i M_{i,k} v_k}$ can be similarly interpreted as the fraction of the equilibrium population descended from individuals that transitioned from type $k$ to another type (via the $k$ column of the transition matrix $M$), conditioned on transition occurring.
+Here, $\hat u_k$ can be regarded as the \emph{reproductive value}~\citep[pg.~27]{Fisher1930} of type $k$ (see~\nameref{sec:AppB}), which gives the relative contribution of type $k$ to the long-term population.
+Consequently, $\hat u_k \hat v_k$ is the \emph{ancestor frequency} of type $k$~\citep{Hermisson2002}, namely the fraction of the equilibrium population descended from type $k$.
+The sum $\sum_i{\hat u_i M_{i,k} \hat v_k}$ can be similarly interpreted as the fraction of the equilibrium population descended from individuals that transitioned from type $k$ to another type (via the $k$ column of the transition matrix $M$), conditioned on transition occurring.
 
-Given $0 < w_k \le 1$, from Eq.~\ref{mfp_intermediate} we have the following corollary.
+Given $0 < w_k\le 1$, from Eq.~\ref{mfp_intermediate} we have the following corollary.
 
 \begin{corollary}[Reproductive value principle]
-In the notation of Lemma~\ref{lemma:mfp},
+In the notation of Result 1,
 \begin{equation}
-sign \frac{\partial \bar{w}^*}{\partial C_k} = 
-sign (u\tr [M]_k - u_k),
+sign \frac{\partial \hat{\bar w}}{\partial C_k} = 
+sign (\hat u\tr [M]_k - \hat u_k),
 \end{equation}
@@
 We turn our attention to the case where the transition occurs from a subset $K$ of the types.
-For example, \citet[Appendix~B]{Ram2012} considered individuals that are grouped by the number of their accumulated deleterious mutations, $k$, 
+For example, \citet[Appendix~A]{Ram2012} considered individuals that are grouped by the number of their accumulated deleterious mutations, $k$, 
 and a modifier allele that regulates the mutation rates in individuals with at least $\pi$ deleterious mutations.
@@
 \begin{aligned}
-\frac{\partial \bar{w}^*}{\partial \tau} = 
-u\tr \Big( \sum_{k \in K}{\frac{\partial (I - C + MC)D}{\partial \tau}} \Big) v = \\
-\frac{1}{\tau} \sum_{k \in K}{u_k v_k (\bar{w}^* - w_k)}.
+\frac{\partial \hat{\bar w}}{\partial \tau} = 
+\hat u\tr \Big( \sum_{k \in K}{\frac{\partial (I - C + MC)D}{\partial \tau}} \Big) \hat v = \\
+\frac{1}{\tau} \sum_{k \in K}{\hat u_k \hat v_k (\hat{\bar w} - w_k)}.
 \end{aligned}
@@
 \begin{multline}\label{mfp_sign_threshold}
-sign \frac{\partial \bar{w}^*}{\partial \tau} = 
-sign \sum_{k \in K}{u_k v_k (\bar{w}^* - w_k)} = \\ 
-sign \Big(\bar{w}^* - \frac{\sum_{k \in K}{u_k v_k w_k}}{\sum_{k \in K}{u_k v_k}}\Big), 
+sign \frac{\partial \hat{\bar w}}{\partial \tau} = 
+sign \sum_{k \in K}{\hat u_k \hat v_k (\hat{\bar w} - w_k)} = \\ 
+sign \Big(\hat{\bar w} - \frac{\sum_{k \in K}{\hat u_k \hat v_k w_k}}{\sum_{k \in K}{\hat u_k \hat v_k}}\Big), 
 \end{multline}
@@
 We now focus on a neutral modifier locus in complete linkage to the types $A_k$, with no direct effect on fitness, and whose sole function is to determine $C_k$, the rates of transition from the different types.
-We will show that modifier alleles that increase the stable population mean fitness in accordance with Lemma~\ref{lemma:mfp} are favored by natural selection.
+We will show that modifier alleles that increase the stable population mean fitness in accordance with Result 1 are favored by natural selection.
 
@@
 \end{equation}
-Here, $\bar{w}=\sum_{k=1}^{n}{(f_k + g_k) w_k}$ is the mean fitness of the entire population (see~\nameref{sec:AppA}). 
+Here, $\bar{w}=\sum_{k=1}^{n}{(f_k + g_k) w_k}$ is the mean fitness of the entire population. 
 Note that Eq.~\ref{eq:model} is the special case of Eq.~\ref{eq:modifier_model} where allele $M$ is absent; i.e. $g_k=0$ for all $k$.
@@
 
-We have found a condition for a modifier allele that controls that transition rate $C_k$ from type $A_k$ to increase the stable population mean fitness (Lemma~\ref{lemma:mfp}).
+We have found a condition for a modifier allele that controls that transition rate $C_k$ from type $A_k$ to increase the stable population mean fitness (Result 1).
 Could such an allele increase in frequency when initially rare in the population?
@@
 \end{equation}
-and $\cl_{in}$ is the local stability matrix of the equilibrium $(v,0)$ in the space $\sum_{k=1}^n{v_k}=1$.
+and $\cl_{in}$ is the local stability matrix of the equilibrium $(v,0)$ in the space $\sum_{k=1}^n{\hat v_k}=1$.
 The zero block matrices are due to the complete linkage between the modifier and the types $A_k$ and to the lack of transition (i.e., mutation) between the modifier alleles.
@@
 \begin{equation}\label{eq:Lex}
-\cl_{ex} = (I - \tilde{C} + M \tilde{C}) D /\bar{w}^*,
+\cl_{ex} = (I - \tilde{C} + M \tilde{C}) D /\hat{\bar w},
 \end{equation}
-where $\bar{w}^*$ is the stable population mean fitness in the absence of modifier allele $M$ (see~\nameref{sec:AppA}).
-Note that, in general, $\bar{w}^*$ is not an eigenvalue of $\cl_{ex}$.
+where $\hat{\bar w}$ is the stable population mean fitness in the absence of modifier allele $M$.
+Note that, in general, $\hat{\bar w}$ is not an eigenvalue of $\cl_{ex}$.
 
@@
 sign \frac{\partial \lambda_1}{\partial \tilde{C}_k}\bigg\rvert_{\tilde{C}_k = C_k} = 
-sign(\bar{w}^* - w_k).
+sign(\hat{\bar w} - w_k).
 \end{equation}
@@
 \begin{proof}
-Substituting $\tilde{C}=C$ in Eq.~\ref{eq:Lex} and multiplying both sides by $\bar{w}^*$,
+Substituting $\tilde{C}=C$ in Eq.~\ref{eq:Lex} and multiplying both sides by $\hat{\bar w}$,
 \begin{equation}
-\bar{w}^* \cl_{ex}\big\rvert_{\tilde{C}_k = C_k} = (I-C+MC) D,
+\hat{\bar w} \cl_{ex}\big\rvert_{\tilde{C}_k = C_k} = (I-C+MC) D,
 \end{equation}
-and since $\bar{w}^*$ is the leading eigenvalue of the RHS (see Eq.~\ref{eq:model_equilibrium} and \nameref{sec:AppA}),
+and since $\hat{\bar w}$ is the leading eigenvalue of the RHS (see Eq.~\ref{eq:model_equilibrium}),
 the leading eigenvalue of $\cl_{ex}\big\rvert_{\tilde{C}_k = C_k}$ is $\lambda_1=1$.
 
-Now, applying Lemma~\ref{lemma:mfp} (eq.~\ref{mfp_sign}) to Eq.~\ref{eq:Lex}, the sign of the derivative of $\lambda_1$ with respect to $\tilde{C_k}$ is
+Now, applying Result 1 (eq.~\ref{mfp_sign}) to Eq.~\ref{eq:Lex}, the sign of the derivative of $\lambda_1$ with respect to $\tilde{C_k}$ is
 \begin{equation}
 sign \frac{\partial \lambda_1}{\partial \tilde{C}_k} =
-sign\Big(\lambda_1 - \frac{w_k}{\bar{w}^*}\Big).
+sign\Big(\lambda_1 - \frac{w_k}{\hat{\bar w}}\Big).
 \end{equation}
-Evaluating this expression at $\tilde{C}_k = C_k$, i.e. at $\lambda_1=1$, gives us
+Thus
 \begin{multline}
 sign \frac{\partial \lambda_1}{\partial \tilde{C}_k}\bigg\rvert_{\tilde{C}_k = C_k} = 
-sign\Big(1 - \frac{w_k}{\bar{w}^*}\Big) =  
-sign(\bar{w}^* - w_k),
+sign\Big(1 - \frac{w_k}{\hat{\bar w}}\Big) =  
+sign(\hat{\bar w} - w_k),
 \end{multline}
-since $\bar{w}^*>0$.
+since $\hat{\bar w}>0$.
 This completes the proof.
@@
 
-Note that if the modifier has the same effect on all individuals (i.e., $C_k = \alpha$ and $\tilde{C}_k = \alpha^*$ for all $k$) then $\cl_{ex}$ becomes
+Note that if the modifier has the same effect on all individuals (i.e., $C_k = \hat\alpha$ and $\tilde{C}_k = \alpha$ for all $k$) then $\cl_{ex}$ becomes
 \begin{equation}
-\cl_{ex} = ((1 - \alpha^*)I + \alpha^* M) D /\bar{w}^*,
+\cl_{ex} = ((1 - \hat\alpha)I + \hat\alpha M) D /\hat{\bar w},
 \end{equation}  
 and the \emph{unified reduction principle} is in effect~\cite[eqs. 65, 72]{Altenberg2017}.
-Therefore, we can expect that with these modifiers, an allele $M$ that reduces the transition rate, i.e. $\alpha^* < \alpha$, will be favored by natural selection.
+Therefore, we can expect that with these modifiers, an allele $M$ that reduces the transition rate, i.e. $\alpha < \hat\alpha$, will be favored by natural selection.
 Indeed, setting $C = \alpha I$ in Eq.~\ref{eq:model_equilibrium} and proceeding as in the above, we find
 $$
-\frac{\partial \bar{w}^*}{\partial \alpha} = 
-\bar{w}^* - 1 \le 0,
+\frac{\partial \hat{\bar w}}{\partial \alpha} = 
+\hat{\bar w} - 1 \le 0,
 $$
-because $\bar{w}^* = \sum_{k=1}^{n}{v_k w_k}$ (see~\nameref{sec:AppA}), and $w_k \le 1$ for all $k$.
+because $\hat{\bar w} = \sum_{k=1}^{n}{\hat v_k w_k}$, and $w_k \le 1$ for all $k$.
 
@@
 \citeauthor{Ram2012} modeled the accumulation of mutant alleles in a population evolving at the mutation-selection balance (see \emph{\nameref{sec:mut_model_2}} section).
-Using Eq.~\ref{eq:Caswells_formula} in \nameref{sec:AppB} and a recursion on the ratios of the reproductive values \cite[see][eqs.~A5-6]{Ram2012}, they concluded that at the mutation-selection balance, if individuals with below-average fitness ($D_k < \bar{w}^*$) increase their mutation rate, then the population mean fitness will increase -- a result generalized by our \emph{mean fitness principle} in Eqs.~\ref{mfp_sign} and~\ref{mfp_sign_threshold}.
+Using Eq.~\ref{eq:Caswells_formula} in \nameref{sec:AppA} and a recursion on the ratios of the reproductive values \cite[see][eqs.~A5-6]{Ram2012}, they concluded that at the mutation-selection balance, if individuals with below-average fitness ($D_k < \hat{\bar w}$) increase their mutation rate, then the population mean fitness will increase -- a result generalized by our \emph{mean fitness principle} in Eqs.~\ref{mfp_sign} and~\ref{mfp_sign_threshold}.
 
@@
 % Appendix A: mean fitness
-\subsection*{Appendix A}\label{sec:AppA}
+%\subsection*{Appendix A}\label{sec:AppA}
 
-\begin{lemma}
-The normalizing factor $\bar{w}$ in Eq.~\ref{eq:model} is the population mean fitness
-\begin{equation}
-\bar{w} = \sum_k{f_k w_k}.
-\end{equation}
-\end{lemma}
+%\begin{lemma}
+%The normalizing factor $\bar{w}$ in Eq.~\ref{eq:model} is the population mean fitness
+%\begin{equation}
+%\bar{w} = \sum_k{f_k w_k}.
+%\end{equation}
+%\end{lemma}
 
-\begin{proof}
-By definition, the normalizing factor $\bar{w}$ ensures that $\sum_k{f'_k}=1$.
-Denote $e = (1, ..., 1)$ the vector with ones at all positions and rewrite $e\tr f' = 1$.
-Therefore, 
-\begin{multline}\label{eq:mean_fitness}
-\bar{w} = 
-\bar{w} e\tr f' = 
-e\tr (I - C + M C) D f = \\
-(e\tr I - e\tr C + e\tr M C) D f = \\
-(e\tr - e\tr C + e\tr C) D f = \\
-e\tr D f = 
-\diag{D} f = 
-\sum_k{f_k w_k},
-\end{multline}
-using the fact that $e\tr M = e\tr$ and $e\tr I = e\tr$ because $M$ and $I$ are stochastic matrices.
-\end{proof}
+%\begin{proof}
+%By definition, the normalizing factor $\bar{w}$ ensures that $\sum_k{f'_k}=1$.
+%Denote $e = (1, ..., 1)$ the vector with ones at all positions and rewrite $e\tr f' = 1$.
+%Therefore, 
+%\begin{multline}\label{eq:mean_fitness}
+%\bar{w} = 
+%\bar{w} e\tr f' = 
+%e\tr (I - C + M C) D f = \\
+%(e\tr I - e\tr C + e\tr M C) D f = \\
+%(e\tr - e\tr C + e\tr C) D f = \\
+%e\tr D f = 
+%\diag{D} f = 
+%\sum_k{f_k w_k},
+%\end{multline}
+%using the fact that $e\tr M = e\tr$ and $e\tr I = e\tr$ because $M$ and $I$ are stochastic matrices.
+%\end{proof}
 
-Note that this lemma also ensures that the stable population mean fitness $\bar{w}^*=\sum_k{v_k w_k}$ is the leading eigenvalue of $(I - C + M C) D$; see Eq.~\ref{eq:model_equilibrium} and the \emph{Perron-Frobenius theorem}~\cite[Appendix~A]{Otto2007}.
+%Note that this lemma also ensures that the stable population mean fitness $\hat{\bar w}=\sum_k{v_k w_k}$ is the leading eigenvalue of $(I - C + M %C) D$; see Eq.~\ref{eq:model_equilibrium} and the \emph{Perron-Frobenius theorem}~\cite[Appendix~A]{Otto2007}.
 
 % Appendix B: Caswell formula
-\subsection*{Appendix B}\label{sec:AppB}
+\subsection*{Appendix A}\label{sec:AppA}
 
@@
 \begin{lemma}
-$T$ be a non-negative matrix with leading eigenvalue $\lambda$ and left and right eigenvectors $u$ and $v$ such that $\sum{v_k}=1$ and $u\tr v = \sum{u_k v_k} = 1$.
+$T$ be a non-negative matrix with leading eigenvalue $\lambda$ and left and right eigenvectors $\hat u$ and $\hat v$ such that $\sum{\hat v_k}=1$ and $\hat u\tr \hat v = \sum{\hat u_k \hat v_k} = 1$.
 Then the sensitivity of $\lambda$ to changes in any element $t$ of the matrix $T$ is
@@
 \frac{\partial \lambda}{\partial t} = 
-u\tr \frac{\partial T}{\partial t} v
+\hat u\tr \frac{\partial T}{\partial t} \hat v
 \end{equation}
@@
 Using the lemma assumptions,
-$\lambda = \lambda u\tr v = u\tr \lambda v = u\tr T v$ and differentiating both sides we get $\partial \lambda = \partial (u\tr T v)$.
+$\lambda = \lambda \hat u\tr \hat v = \hat u\tr \lambda \hat v = \hat u\tr T \hat v$ and differentiating both sides we get $\partial \lambda = \partial (\hat \hat u\tr T \hat v)$.
 Using the product rule (once in each direction),
@@
 \begin{multline}
-\partial (u\tr T v) = 
-\partial u\tr T v + u\tr \partial T v + u\tr T \partial v = \\
-u\tr \partial T v + \partial u\tr \lambda v  + \lambda u\tr \partial v = \\
-u\tr \partial T v + \lambda(\partial u\tr v  + u\tr \partial v) = \\
-u\tr \partial T v + \lambda \partial(u\tr v).
+\partial (\hat u\tr T \hat v) = 
+\partial \hat u\tr T \hat v + \hat u\tr \partial T \hat v + \hat u\tr T \partial \hat v = \\
+\hat u\tr \partial T \hat v + \partial \hat u\tr \lambda \hat v  + \lambda \hat u\tr \partial \hat v = \\
+\hat u\tr \partial T \hat v + \lambda(\partial \hat u\tr \hat v  + \hat u\tr \partial \hat v) = \\
+\hat u\tr \partial T \hat v + \lambda \partial(\hat u\tr \hat v).
 \end{multline}
 
-Because $u\tr v = 1$,
-we have $\partial (u\tr v) = 0$ and
-$\partial \lambda = u\tr \partial T v$.
+Because $\hat u\tr \hat v = 1$,
+we have $\partial (\hat u\tr \hat v) = 0$ and
+$\partial \lambda = \hat u\tr \partial T \hat v$.
 \end{proof}
 
-% Appendix C: Fisher's reproductive value
-\subsection*{Appendix C}\label{sec:AppC}
+% Appendix B: Fisher's reproductive value
+\subsection*{Appendix B}\label{sec:AppB}
 \begin{remark}[Fisher's reproductive value]
@@
 
-If we begin with a population at the equilibrium distribution $v$,
+If we begin with a population at the equilibrium distribution $\hat v$,
 and ask what fraction of long-term descendants descended from type $k$ at that time,
-we weight the equilibrium frequency $v_k$ by the reproductive value $u_k$, to get $u_k v_k$.
-$\{u_k v_k\}_{k}$ is a probability distribution, since
+we weight the equilibrium frequency $\hat v_k$ by the reproductive value $\hat u_k$, to get $\hat u_k \hat v_k$.
+$\{\hat u_k \hat v_k\}_{k}$ is a probability distribution, since
 \begin{equation}
-\sum_k {u_k v_k} = u\tr v = 1.
+\sum_k {\hat u_k \hat v_k} = \hat u\tr v = 1.
 \end{equation}
@@
 
-% Appendix D: inner stability
-\subsection*{Appendix D}\label{sec:AppD}
+% Appendix C: inner stability
+\subsection*{Appendix C}\label{sec:AppC}
 
-\begin{lemma}[Stability of the equilibrium $v$]
-If $C$ and $D$ are positive diagonal matrices, and $M$ is block irreducible, 
+\begin{lemma}[Stability of the equilibrium $\hat v$]
+If $C$ and $D$ are positive diagonal matrices, and $M$ is primitive, 
 then the equilibrium $v$ of the system in Eq.~\ref{eq:model_equilibrium} is globally stable.
@@
 \begin{proof}
-Let $A=(I-C+MC)D$ and denote the leading eigenvalue of $A$ as $\bar{w}^*$.
+Let $A=(I-C+MC)D$ and denote the leading eigenvalue of $A$ as $\hat{\bar w}$.
 
@@
 \begin{equation}
-\bigg(\frac{A}{\bar{w}^*}\bigg)^t \goesto \frac{v u\tr}{u\tr v} = v u\tr,
+\bigg(\frac{A}{\hat{\bar w}}\bigg)^t \goesto \frac{\hat v \hat u\tr}{\hat u\tr \hat v} = \hat v \hat u\tr,
 \end{equation}
-where $u$ and $v$ are the left and right Perron eigenvectors of $A$ such that 
+where $\hat u$ and $\hat v$ are the left and right Perron eigenvectors of $A$ such that 
 \begin{equation}
 \sum_{k=1}^{n}{v_k}=1, \quad
-u\tr v=\sum_{k=1}^{n}{u_k v_k}=1,
+\hat u\tr \hat v=\sum_{k=1}^{n}{\hat u_k \hat v_k}=1,
 \label{eq:sum_uv}
 \end{equation}
-and $v u\tr$ is the \emph{Perron projection} into the eigenspace corresponding to $\bar{w}^*$.
+and $\hat v \hat u\tr$ is the \emph{Perron projection} into the eigenspace corresponding to $\hat{\bar w}$.
 
@@
 \begin{equation}
-\bigg(\frac{A}{\bar{w}^*}\bigg)^t \cdot f \goesto v u\tr \cdot f = \alpha v,
+\bigg(\frac{A}{\hat{\bar w}}\bigg)^t \cdot f \goesto \hat v \hat u\tr \cdot f = \alpha \hat v,
 \label{eq:perron_projection}
@@
 \begin{equation}
-\alpha = \alpha u\tr v = u\tr \alpha v = u\tr v u\tr f = u\tr f = \sum_{k=1}^{n}{u_k f_k} > 0,
+\alpha = \alpha \hat u\tr \hat v = \hat u\tr \alpha \hat v = \hat u\tr \hat v \hat u\tr f = \hat u\tr f = \sum_{k=1}^{n}{\hat u_k f_k} > 0,
 \label{eq:alpha0}
 \end{equation}
-because $f$ is positive and $u$ is non-negative.
+because $f$  $u$ are positive vectors.
 
-Now, let $\norm{x} = \sum_{k=1}^{n}{x_k}$, so that $x_t \goesto x$ implies
+Now, let $\norm{x} = \sum_{k=1}^{n}\norm{x_k}$, so that $x_t \goesto x$ implies
 \begin{equation}
@@
 
-Therefore, for a positive frequency vector $f$, we can apply $f \goesto f'$ (Eq.~\ref{eq:model_norm}) $t$ times, 
-and as $f \goesto Af$ is linear, one normalization ($x \goesto x/\norm{x}$) suffices~\citep[see][Appendix~A]{Wakano2004}.
-Using eqs.~\ref{eq:perron_projection} and ~\ref{eq:norm_limit}, we have
+It is easily seen that $f^t =A^tf/\|A^tf\|$. Hence 
 \begin{multline}
 \frac{A^t f}{\norm{A^t f}} = 
-\frac{\frac{1}{(\bar{w}^*)^t} A^t f}{\frac{1}{(\bar{w}^*)^t} \norm{A^t f}} = 
-\frac{\big(\frac{A}{\bar{w}^*}\big)^t \cdot f }{ \norm{ \big(\frac{A}{\bar{w}^*}\big)^t \cdot f }} \goesto
-\frac{\alpha v}{\norm{\alpha v}} = v,
+\frac{\frac{1}{(\hat{\bar w})^t} A^t f}{\frac{1}{(\hat{\bar w})^t} \norm{A^t f}} = 
+\frac{\big(\frac{A}{\hat{\bar w}}\big)^t \cdot f }{ \norm{ \big(\frac{A}{\hat{\bar w}}\big)^t \cdot f }} \goesto
+\frac{\alpha \hat v}{\norm{\alpha \hat v}} = \hat v,
 \end{multline}
-since $t \goesto \infty$, as $\norm{v}=1$ (eq.~\ref{eq:sum_uv}) and $\alpha \ne 0$ (eq.~\ref{eq:alpha0}).
+as $t \goesto \infty$, since $\norm{v}=1$ (eq.~\ref{eq:sum_uv}) and $\alpha \ne 0$ (eq.~\ref{eq:alpha0}).
 This completes the proof.

### 2017-11-19 ms.tex
typos

@@
 We turn our attention to the case where the transition occurs from a subset $K$ of the types.
-For example, \citet[Appendix~A]{Ram2012} considered individuals that are grouped by the number of their accumulated deleterious mutations, $k$, 
+For example, \citet[Appendix~B]{Ram2012} considered individuals that are grouped by the number of their accumulated deleterious mutations, $k$, 
 and a modifier allele that regulates the mutation rates in individuals with at least $\pi$ deleterious mutations.
@@
 \citeauthor{Ram2012} modeled the accumulation of mutant alleles in a population evolving at the mutation-selection balance (see \emph{\nameref{sec:mut_model_2}} section).
-Using Eq.~\ref{eq:Caswells_formula} in \nameref{sec:AppA} and a recursion on the ratios of the reproductive values \cite[see][eqs.~A5-6]{Ram2012}, they concluded that at the mutation-selection balance, if individuals with below-average fitness ($D_k < \hat{\bar w}$) increase their mutation rate, then the population mean fitness will increase -- a result generalized by our \emph{mean fitness principle} in Eqs.~\ref{mfp_sign} and~\ref{mfp_sign_threshold}.
+Using Eq.~\ref{eq:Caswells_formula} in \nameref{sec:AppA} and a recursion on the ratios of the reproductive values \cite[see][eqs.~A5-6]{Ram2012}, they concluded that at the mutation-selection balance, if individuals with below-average fitness ($w_k < \hat{\bar w}$) increase their mutation rate, then the population mean fitness will increase -- a result generalized by our \emph{mean fitness principle} in Eqs.~\ref{mfp_sign} and~\ref{mfp_sign_threshold}.
 

### 2017-11-21 ms.tex
UL changes

@@
 where $f=(f_1, f_2, \ldots, f_n)$ is a frequency vector with $f_k\ge 0$ and  $\sum_{k=1}^n{f_k} = 1$;
-$D$ is a positive diagonal matrix with  elements $w_k$ such that $w_k \ne w_j$ for some $k \ne j$, and we can assume $0 < w_k \le 1$, i.e., fitness values are positive and relative to the fitness of the fittest type;
+$D$ is a positive diagonal matrix with  elements $w_k$ such that $w_k \ne w_j$ for some $k \ne j$; 
+%and we can assume $0 < w_k \le 1$, i.e., fitness values are positive and relative to the fitness of the fittest type; 
 $C$ is a positive diagonal  matrix with  elements $C_k$;
@@
 $\ci$ is the $n\times n$ identity matrix;
-and $\bar{w}$ is the normalizing factor such that $\sum_{k=1}^n{f'_k}=1$ and is equal to the population mean fitness.
+and $\bar{w}$ is the normalizing factor such that $\sum_{k=1}^n{f'_k}=1$ and is equal to the population mean fitness $\bar w=\sum_{k=1}^n f_kw_k$.
 
@@
 Let the frequency of $A_k$ in the present generation be $f_k$ with $f_k \ge 0$ and $\sum_{k=1}^{n}{f_k}=1$.
-Then after selection and mutation $f'$ in the next generation is given by
+Then after selection and mutation $f'_k$ in the next generation is given by
 \begin{equation}
@@
 
-After reproduction, offspring of individuals from deme $A_k$ return to their parental deme with probability $1-C_k$, or migrate to a different deme $A_j$ with probability $C_k \cdot M_{j,k}$, where matrix $M$ is primitive, $C_k > 0$, $M_{j,k} \ge 0$, and $\sum_{k=1}^{n}{M_{j,k}} = 1$ for all $j=1, \ldots, n$.
+After reproduction, offspring of individuals from deme $A_k$ return to their parental deme with probability $1-C_k$, or migrate to a different deme $A_j$ with probability $C_k \cdot M_{j,k}$, where the matrix $M$ is primitive, $C_k > 0$, $M_{j,k} \ge 0$, and $\sum_{k=1}^{n}{M_{j,k}} = 1$ for all $j=1, \ldots, n$.
 Therefore, $1-C_k$ can be interpreted as a \emph{homing rate}.
@@
 Following selection and migration the new frequencies $f'_k$ are given exactly by Eq.~\ref{eq:model_sum}.
-If columns of ${M}$ are identical
+If the columns of ${M}$ are identical
 \begin{equation}
@@
 {\sl Let $\hat{\bar w}$ be the leading eigenvalue of $(I-C+MC)D$,
-and $\hat u$ and $\hat v$ be the corresponding left and right eigenvectors.
+and $\hat u$ and $\hat v$ be the corresponding left and right eigenvectors, such that $\sum_k\hat v_k=1$ and $\sum_k\hat u_k\hat v_k=1$..
 Then,
@@
 \begin{proof}
-$\hat{\bar w}$ is the leading (real positive) eigenvalue of $(I-C+MC)D$ (see eq.~\ref{eq:model_equilibrium}),
-and the corresponding left and right eigenvectors $u$ and $v$ are both non-negative and uniquely defined by $\sum_k{\hat v_k} = 1$ and $\sum_k{\hat u_k \hat v_k} = 1$.
-
+%$\hat{\bar w}$ is the leading (real positive) eigenvalue of $(I-C+MC)D$ (see eq.~\ref{eq:model_equilibrium}),
+%and the corresponding left and right eigenvectors $u$ and $v$ are both non-negative and uniquely defined by $\sum_k{\hat v_k} = 1$ and $\sum_k{\hat u_k \hat v_k} = 1$.
 Using the formula in~\citet{Caswell1978} (see eq.~\ref{eq:Caswells_formula} in~\nameref{sec:AppA}),
@@
 \end{equation}
-Finally, note that $\hat u_k, \hat v_k, C_k \ge 0$ and therefore
+Finally, note that $\hat u_k, \hat v_k, C_k > 0$ and therefore
 \begin{equation}
@@
 
-Given $0 < w_k\le 1$, from Eq.~\ref{mfp_intermediate} we have the following corollary.
+Given $w_k>0$, from Eq.~\ref{mfp_intermediate} we have the following corollary.
 
@@
 \end{equation}
-and $\cl_{in}$ is the local stability matrix of the equilibrium $(v,0)$ in the space $\sum_{k=1}^n{\hat v_k}=1$.
+and $\cl_{in}$ is the local stability matrix of the equilibrium $(\hat v,0)$ in the space $\sum_{k=1}^n{v_k}=1$.
 The zero block matrices are due to the complete linkage between the modifier and the types $A_k$ and to the lack of transition (i.e., mutation) between the modifier alleles.
@@
 \begin{equation}
-\cl_{ex} = ((1 - \hat\alpha)I + \hat\alpha M) D /\hat{\bar w},
+\cl_{ex} = ((1 - \alpha)I + \alpha M) D /\hat{\bar w},
 \end{equation}  
@@
 Therefore, we can expect that with these modifiers, an allele $M$ that reduces the transition rate, i.e. $\alpha < \hat\alpha$, will be favored by natural selection.
-Indeed, setting $C = \alpha I$ in Eq.~\ref{eq:model_equilibrium} and proceeding as in the above, we find
+Setting $C = \alpha I$ in Eq.~\ref{eq:model_equilibrium} and proceeding as in Eq.\ 25, we find
 $$
-\frac{\partial \hat{\bar w}}{\partial \alpha} = 
-\hat{\bar w} - 1 \le 0,
+\frac{\partial \bar w}{\partial \alpha} = 
+\frac{1}{\alpha}\left[\bar w-\sum_{k=1}^nu_kv_kw_k\right]
 $$
-because $\hat{\bar w} = \sum_{k=1}^{n}{\hat v_k w_k}$, and $w_k \le 1$ for all $k$.
+(following~\citet{Altenberg2017}, $\frac{\partial\bar w}{\partial \alpha}<0$).
 
@@
 
-Now, let $\norm{x} = \sum_{k=1}^{n}\norm{x_k}$, so that $x_t \goesto x$ implies
+Now, let $\norm{x} = \sum_{k=1}^{n}|x_k|$, so that $x_t \goesto x$ implies
 \begin{equation}
@@
 \end{multline}
-as $t \goesto \infty$, since $\norm{v}=1$ (eq.~\ref{eq:sum_uv}) and $\alpha \ne 0$ (eq.~\ref{eq:alpha0}).
+since $\norm{v}=1$ (eq.~\ref{eq:sum_uv}) and $\alpha \ne 0$ (eq.~\ref{eq:alpha0}).
 This completes the proof.

### 2017-11-21 ms.tex
more UL changes

@@
 for $k=1,2,\ldots, n-1$.
-Here $\bar{w}=\sum_{k=1}^{n}{f_k w_k}$ is the population mean fitness.
+Here $\bar{w}=\sum_{k=0}^{n}{f_k w_k}$ is the population mean fitness.
 
@@
 \end{equation}
-where $p_k > 0$ and $\sum_{k=1}^{n}{p_k}=1$.
+where $p_k > 0$ and $\sum_{k=0}^{n-1}{p_k}=1$.
 
@@
 Note that, in general, $\hat{\bar w}$ is not an eigenvalue of $\cl_{ex}$.
+\medskip
 
-\begin{theorem}[Evolution of increased genetic variation]
-Let $\lambda_1$ be the the leading eigenvalue of the external stability matrix $\cl_{ex}$.
+\noindent{\bf Result 2. Evolution of increased genetic variation.}
+{\sl Let $\lambda_1$ be the the leading eigenvalue of the external stability matrix $\cl_{ex}$.
 If the transition rates induced by the modifier alleles $m$ and $M$ are equal, i.e., $\tilde{C}_k=C_k$ for all $k$, then the following applies
@@
 
-Therefore, an initially rare modifier allele $M$ with transition rates slightly different from the resident allele $m$ can successfully invade the population ($\lambda_1>1$) if $M$ increases the probability of transition from types with below-average fitness, thereby increasing the stable population mean fitness.
-\end{theorem}
+Therefore, an initially rare modifier allele $M$ with transition rates slightly different from the resident allele $m$ can successfully invade the population ($\lambda_1>1$) if $M$ increases the probability of transition from types with below-average fitness, thereby increasing the stable population mean fitness.}
 
@@
 Eq.~\ref{mfp} is a generalization of a result of~\citet[Eq.~4]{Ram2012}.
-\citeauthor{Ram2012} modeled the accumulation of mutant alleles in a population evolving at the mutation-selection balance (see \emph{\nameref{sec:mut_model_2}} section).
+\citeauthor{Ram2012} modeled the accumulation of mutant alleles in a population evolving at the mutation-selection balance (see \nameref{sec:mut_model_2} section).
 Using Eq.~\ref{eq:Caswells_formula} in \nameref{sec:AppA} and a recursion on the ratios of the reproductive values \cite[see][eqs.~A5-6]{Ram2012}, they concluded that at the mutation-selection balance, if individuals with below-average fitness ($w_k < \hat{\bar w}$) increase their mutation rate, then the population mean fitness will increase -- a result generalized by our \emph{mean fitness principle} in Eqs.~\ref{mfp_sign} and~\ref{mfp_sign_threshold}.

### 2017-11-22 ms.tex
fixed lots of index and hats; added paragraph about mutational loss; fixed symmetric learning example

--- 2017-11-21 ms.tex
@@
 \newtheorem{remark}{Remark}
+\newtheorem{result}{Result}
 %\newtheorem*{mfp}{Theorem: Mean Fitness Principle}
@@
 $C$ is a positive diagonal  matrix with  elements $C_k$;
-$M$ is a primitive column-stochastic matrix: $M_{j,k} \ge 0$ for all $j,k$, $\sum_{j=1}^n {M_{j,k}} = 1$ for all $k$, and $(M^n)_{j,k} > 0$ for all $j,k$ for some natural $n$.
+$M$ is a primitive column-stochastic matrix: $M_{j,k} \ge 0$ for all $j,k$, $\sum_{j=1}^n {M_{j,k}} = 1$ for all $k$, and $(M^n)_{j,k} > 0$ for all $j,k$ for some natural $n$;
 $\ci$ is the $n\times n$ identity matrix;
@@
 specifically, with probability $1-C_k$, the allele $A_k$ does not mutate, and with probability $\frac{C_k}{n-1}$, the allele $A_k$ mutates to $A_j$ for any $j \ne k$.
-This is an extension of a model studied by~\citet{Altenberg2017} that allows for the mutation rate of $A_k$, $C_k$, to depend on the identity of the allele $A_k$.
+This is an extension of a model studied by~\citet{Altenberg2017} that allows for the mutation rate of $A_k$, $C_k$, to depend on properties of the allele $A_k$.
 
@@
 
-Let the frequency of $A_k$ in the present generation be $f_k$ with $f_k \ge 0$ and $\sum_{k=1}^{n}{f_k}=1$.
+Let the frequency of $A_k$ in the present generation be $f_k$ with $f_k \ge 0$ and $\sum_{k=0}^{n}{f_k}=1$.
 Then after selection and mutation $f'_k$ in the next generation is given by
@@
 f_0' &= \big(1 - \delta C_0\big) \frac{w_0}{\bar{w}} f_0 + \beta C_{1} \frac{w_{1}}{\bar{w}} f_{1}, \\
-f_n' &= \big(1 - \beta C_n\big) \frac{w_n}{\bar{w}} f_n + \delta C_{n-1} \frac{w_{n-1}}{\bar{w}} f_{n-1}, \\
-f_k' &= \big(1 - C_k(\delta+\beta)\big) \frac{w_k}{\bar{w}} f_k + \\
+f_k' &= \big(1 - (\delta+\beta) C_k\big) \frac{w_k}{\bar{w}} f_k + \\
 	 & \delta C_{k-1} \frac{w_{k-1}}{\bar{w}} f_{k-1} + 
-	 \beta C_{k+1} \frac{w_{k+1}}{\bar{w}} f_{k+1},
+	 \beta C_{k+1} \frac{w_{k+1}}{\bar{w}} f_{k+1}, \\
+f_n' &= \big(1 - \beta C_n\big) \frac{w_n}{\bar{w}} f_n + \delta C_{n-1} \frac{w_{n-1}}{\bar{w}} f_{n-1},	 
 \end{aligned}
@@
 
-After reproduction, offspring of individuals from deme $A_k$ return to their parental deme with probability $1-C_k$, or migrate to a different deme $A_j$ with probability $C_k \cdot M_{j,k}$, where the matrix $M$ is primitive, $C_k > 0$, $M_{j,k} \ge 0$, and $\sum_{k=1}^{n}{M_{j,k}} = 1$ for all $j=1, \ldots, n$.
+After reproduction, offspring of individuals from deme $A_k$ return to their parental deme with probability $1-C_k$, or migrate to a different deme $A_j$ with probability $C_k M_{j,k}$, where the matrix $M$ is primitive, $C_k > 0$, $M_{j,k} \ge 0$, and $\sum_{j=1}^{n}{M_{j,k}} = 1$ for all $k=1, \ldots, n$.
 Therefore, $1-C_k$ can be interpreted as a \emph{homing rate}.
@@
 Fitness is determined by the difference between the individual mature phenotype and the optimal phenotype:
-denoting the optimal phenotype as $h$, the fitness of an individual with mature phenotype $k$ is $w_{|k-h|}$.
+denoting the optimal phenotype as $h$, the fitness of an individual with mature phenotype $k$ is $w_{\abs{k-h}}$.
 
@@
 In individuals with initial phenotype $k$, the mature phenotype is the same as the initial phenotype with probability $1-C_k$, and is modified by individual learning or \emph{exploration}~\citep{Borenstein2008} with probability $C_k$.
-Such individual exploratory learning, which can be considered either intentional or the result of incorrect learning, modifies initial phenotype $k$ to mature phenotype $j$ with probability $M_{k,j}$.
+Such individual exploratory learning, which can be considered either intentional or the result of incorrect learning, modifies initial phenotype $k$ to mature phenotype $j$ with probability $M_{j,k}$.
 
-For example, the case of \emph{symmetric individual learning}~\citep{Borenstein2008} can be represented by
-\begin{equation}
-M_{k,j} = \begin{cases}
-\frac{1}{2b+1}, & \abs{k-j} \le b \\
-0, & otherwise
-\end{cases},
-\end{equation}
-where the breadth of exploration is $b \ge 0$, and phenotypes $0$ and $n$ "absorb" transitions to $k<0$ and $k>n$, respectively. 
+For example, in the case of \emph{symmetric individual learning}~\citep{Borenstein2008}, the mature phenotype is randomly drawn from $2b+1$ phenotypes symmetrically and uniformly distributed around the initial phenotype, with the limitation that any "spillover" of phenotypes below $1$ or above $n$ is "absorbed" by those boundaries.
+This ensures that $M$ is primitive and column-stochastic.
+
 For instance, with $n=5$ and $b=1$ we have
@@
 0 & 0 & 0 & 1/3 & 2/3
+\end{bmatrix},
+\end{equation}
+and with $n=6$ and $b=2$ we have
+\begin{equation}
+M = \begin{bmatrix}
+3/5 & 2/5 & 1/5 & 0 & 0 & 0 \\
+1/5 & 1/5 & 1/5 & 1/5 & 0 & 0 \\
+1/5 & 1/5 & 1/5 & 1/5 & 1/5 & 0 \\
+0 & 1/5 & 1/5 & 1/5 & 1/5 & 1/5 \\
+0 & 0 & 1/5 & 1/5 & 1/5 & 1/5 \\
+0 & 0 & 0 & 1/5 & 2/5 & 3/5
 \end{bmatrix}.
@@
 
-\noindent{\bf Result 1. Mean fitness principle.}
-{\sl Let $\hat{\bar w}$ be the leading eigenvalue of $(I-C+MC)D$,
-and $\hat u$ and $\hat v$ be the corresponding left and right eigenvectors, such that $\sum_k\hat v_k=1$ and $\sum_k\hat u_k\hat v_k=1$..
+\begin{result}[Mean fitness principle]\label{result:mfp}
+Let $\hat{\bar w}$ be the leading eigenvalue of $(I-C+MC)D$,
+and $\hat u$ and $\hat v$ be the corresponding left and right eigenvectors, such that $\sum_{k=1}^{n}\hat v_k=1$ and $\sum_{k=1}^{n}\hat u_k\hat v_k=1$.
 Then,
-\begin{equation}\label{mfp}
+\begin{equation}\label{eq:mfp}
 \frac{\partial \hat{\bar w}}{\partial C_k} = 
@@
 or in simpler terms,
-\begin{equation}\label{mfp_sign}
+\begin{equation}\label{eq:mfp_sign}
 sign\frac{\partial \hat{\bar w}}{\partial C_k} = 
@@
 
-Therefore increased transition from type $k$ will increase the stable population mean fitness if the fitness of type $k$ is below the stable population mean fitness.}
+Therefore increased transition from type $k$ will increase the stable population mean fitness if the fitness of type $k$ is below the stable population mean fitness.
+\end{result}
 
@@
 Then,
-\begin{equation}\label{mfp_intermediate}
+\begin{equation}\label{eq:mfp_intermediate}
 \begin{aligned}
@@
 which gives us a relation between $\hat{\bar w}$ and the $k$
-element of $u$:
+element of $\hat u$:
 \begin{equation}
@@
 \end{equation}
-which when substituted into Eq.~\ref{mfp_intermediate} yields:
+which when substituted into Eq.~\ref{eq:mfp_intermediate} yields:
 \begin{equation}
@@
 
-An interesting intermediate result in Eq.~\ref{mfp_intermediate} is
+An interesting intermediate result in Eq.~\ref{eq:mfp_intermediate} is
 
@@
 \hat v_k w_k (\hat u\tr[M]_k - \hat u_k) = \\
-w_k (\sum_{j=1}^{n}{\hat u_j M_{j,k} \hat v_k} - \hat u_k \hat v_k).
+w_k \Big(\sum_{j=1}^{n}{\hat u_j M_{j,k} \hat v_k} - \hat u_k \hat v_k\Big).
 \end{aligned}
 \end{equation}
-Here, $\hat u_k$ can be regarded as the \emph{reproductive value}~\citep[pg.~27]{Fisher1930} of type $k$ (see~\nameref{sec:AppB}), which gives the relative contribution of type $k$ to the long-term population.
+Here, $\hat u_k$ can be regarded as the \emph{reproductive value} of type $k$~\citep[pg.~27]{Fisher1930}, which gives the relative contribution of type $k$ to the long-term population (see~\nameref{sec:AppB}).
 Consequently, $\hat u_k \hat v_k$ is the \emph{ancestor frequency} of type $k$~\citep{Hermisson2002}, namely the fraction of the equilibrium population descended from type $k$.
-The sum $\sum_i{\hat u_i M_{i,k} \hat v_k}$ can be similarly interpreted as the fraction of the equilibrium population descended from individuals that transitioned from type $k$ to another type (via the $k$ column of the transition matrix $M$), conditioned on transition occurring.
+The sum $\sum_{j=1}^{n}{\hat u_j M_{j,k} \hat v_k}$ can be similarly interpreted as the fraction of the equilibrium population descended from individuals that transitioned from type $k$ to another type (via the $k$ column of the transition matrix $M$), conditioned on transition occurring.
 
-Given $w_k>0$, from Eq.~\ref{mfp_intermediate} we have the following corollary.
+Since $w_k>0$, from Eq.~\ref{eq:mfp_intermediate} we have the following corollary.
 
 \begin{corollary}[Reproductive value principle]
-In the notation of Result 1,
+In the notation of Result~\ref{result:mfp},
 \begin{equation}
@@
 rate of transition from types in $K$ can be written 
-\begin{equation}
+\begin{equation}\label{eq:mfp_K}
 \begin{aligned}
@@
 Thus
-\begin{multline}\label{mfp_sign_threshold}
+\begin{multline}\label{eq:mfp_sign_threshold}
 sign \frac{\partial \hat{\bar w}}{\partial \tau} = 
@@
 \end{multline}
-a generalization of Eq.~\ref{mfp_sign}.
+a generalization of Eq.~\ref{eq:mfp_sign}.
 Therefore, increased transition from types in $K$ will increase the stable population mean fitness if the average fitness of individuals descended from types in $K$ is below the stable population mean fitness.
@@
 We now focus on a neutral modifier locus in complete linkage to the types $A_k$, with no direct effect on fitness, and whose sole function is to determine $C_k$, the rates of transition from the different types.
-We will show that modifier alleles that increase the stable population mean fitness in accordance with Result 1 are favored by natural selection.
+We will show that modifier alleles that increase the stable population mean fitness in accordance with Result~\ref{result:mfp} are favored by natural selection.
 
@@
 Here, $\bar{w}=\sum_{k=1}^{n}{(f_k + g_k) w_k}$ is the mean fitness of the entire population. 
-Note that Eq.~\ref{eq:model} is the special case of Eq.~\ref{eq:modifier_model} where allele $M$ is absent; i.e. $g_k=0$ for all $k$.
+Note that Eq.~\ref{eq:model} is the special case of Eq.~\ref{eq:modifier_model} where allele $M$ is absent, i.e. $g_k=0$ for all $k$.
 
@@
 
-We have found a condition for a modifier allele that controls that transition rate $C_k$ from type $A_k$ to increase the stable population mean fitness (Result 1).
+We have found a condition for a modifier allele that controls that transition rate $C_k$ from type $A_k$ to increase the stable population mean fitness (Result~\ref{result:mfp}).
 Could such an allele increase in frequency when initially rare in the population?
@@
 
-The equilibrium of Eqs.~\ref{eq:modifier_model} when modifier allele $M$ is absent from the population is $(v, 0)$, where $v$ is given in Eq.~\ref{eq:model_equilibrium} and $g_k=0$ for all $k$. 
-The stability of allele $m$ to invasion by allele $M$ is determined by the leading eigenvalue $\lambda_1$ of $\cl_{ex}$ the external stability matrix of the equilibrium $(v,0)$, which, in turn, is determined by the Jacobian $\cj$ of the system in Eq.~\ref{eq:modifier_model} evaluated at the equilibrium $(v,0)$, where
+The equilibrium of Eqs.~\ref{eq:modifier_model} when modifier allele $M$ is absent from the population is $(\hat v,0)$, where $\hat v$ is given in Eq.~\ref{eq:model_equilibrium} and $g_k=0$ for all $k$.
+The stability of allele $m$ to invasion by allele $M$ is determined by the leading eigenvalue $\lambda_1$ of $\cl_{ex}$ the external stability matrix of the equilibrium $(\hat v,0)$, which, in turn, is determined by the Jacobian $\cj$ of the system in Eq.~\ref{eq:modifier_model} evaluated at the equilibrium $(\hat v,0)$, where
 \begin{equation}
@@
 Note that, in general, $\hat{\bar w}$ is not an eigenvalue of $\cl_{ex}$.
+\medskip
 
-\begin{theorem}[Evolution of increased genetic variation]
+\begin{result}[Evolution of increased genetic variation]
 Let $\lambda_1$ be the the leading eigenvalue of the external stability matrix $\cl_{ex}$.
@@
 Therefore, an initially rare modifier allele $M$ with transition rates slightly different from the resident allele $m$ can successfully invade the population ($\lambda_1>1$) if $M$ increases the probability of transition from types with below-average fitness, thereby increasing the stable population mean fitness.
-\end{theorem}
+\end{result}
 
@@
 
-Now, applying Result 1 (eq.~\ref{mfp_sign}) to Eq.~\ref{eq:Lex}, the sign of the derivative of $\lambda_1$ with respect to $\tilde{C_k}$ is
+Now, applying Result~\ref{result:mfp} (eq.~\ref{eq:mfp_sign}) to Eq.~\ref{eq:Lex}, the sign of the derivative of $\lambda_1$ with respect to $\tilde{C_k}$ is
 \begin{equation}
@@
 Therefore, we can expect that with these modifiers, an allele $M$ that reduces the transition rate, i.e. $\alpha < \hat\alpha$, will be favored by natural selection.
-Setting $C = \alpha I$ in Eq.~\ref{eq:model_equilibrium} and proceeding as in Eq.\ 25, we find
-$$
+Setting $C = \alpha I$ in Eq.~\ref{eq:model_equilibrium} and proceeding as in Eq.~\ref{eq:mfp_K}, we find
+\begin{equation}\label{eq:mutational_loss}
 \frac{\partial \bar w}{\partial \alpha} = 
-\frac{1}{\alpha}\left[\bar w-\sum_{k=1}^nu_kv_kw_k\right]
-$$
-(following~\citet{Altenberg2017}, $\frac{\partial\bar w}{\partial \alpha}<0$).
+-\frac{1}{\alpha}\Big(\sum_{k=1}^{n} \hat u_k \hat v_k w_k - \bar w \Big).
+\end{equation}
+Eq.~\ref{eq:mutational_loss} has been derived by~\citet[eq.~24]{Hermisson2002}.
+Furthermore, the term in the parentheses of Eq.~\ref{eq:mutational_loss} is the difference between the 
+\emph{ancestral mean fitness} and the \emph{stable population mean fitness}, and has been termed the \emph{mutational loss} by~\citet[eq.~24]{Hermisson2002} -- "the linear response of the equilibrium mean fitness to changes in the mutation rates".
+Following~\citet{Altenberg2017}, $\frac{\partial\bar w}{\partial \alpha}<0$, and therefore we can conclude that the \emph{mutational loss} is positive and that the \emph{ancestral mean fitness} is higher then the \emph{stable population mean fitness}.
 
@@
 For example, the \emph{reduction principle} applies to ecological models of dispersal, and~\citet{Gueijman2013} have demonstrated that even in homogeneous environments, fitness-associated dispersal increases the mean fitness of diploid populations and is favored by selection over uniform dispersal.
-Similarly, if the transmission fidelity of culturally-transmitted traits depends on the type or fitness of the transmitting individual, we expect that our results will hold (see \emph{\nameref{sec:learn_model}} section).
+Similarly, if the transmission fidelity of culturally-transmitted traits depends on the type or fitness of the transmitting individual, we expect that our results will hold (see \nameref{sec:learn_model}).
 
-Eq.~\ref{mfp} is a generalization of a result of~\citet[Eq.~4]{Ram2012}.
-\citeauthor{Ram2012} modeled the accumulation of mutant alleles in a population evolving at the mutation-selection balance (see \emph{\nameref{sec:mut_model_2}} section).
-Using Eq.~\ref{eq:Caswells_formula} in \nameref{sec:AppA} and a recursion on the ratios of the reproductive values \cite[see][eqs.~A5-6]{Ram2012}, they concluded that at the mutation-selection balance, if individuals with below-average fitness ($w_k < \hat{\bar w}$) increase their mutation rate, then the population mean fitness will increase -- a result generalized by our \emph{mean fitness principle} in Eqs.~\ref{mfp_sign} and~\ref{mfp_sign_threshold}.
+Eq.~\ref{eq:mfp} is a generalization of a result of~\citet[Eq.~4]{Ram2012}.
+\citeauthor{Ram2012} modeled the accumulation of mutant alleles in a population evolving at the mutation-selection balance (see \nameref{sec:mut_model_2}).
+Using Eq.~\ref{eq:Caswells_formula} in \nameref{sec:AppA} and a recursion on the ratios of the reproductive values \cite[see][eqs.~A5-6]{Ram2012}, they concluded that at the mutation-selection balance, if individuals with below-average fitness ($w_k < \hat{\bar w}$) increase their mutation rate, then the population mean fitness will increase -- a result generalized by our \emph{\nameref{result:mfp}} in Eqs.~\ref{eq:mfp_sign} and~\ref{eq:mfp_sign_threshold}.
 
@@
 \begin{remark}[Fisher's reproductive value]
-Let $M$ be a stochastic matrix and $D$ be a diagonal matrix with positive diagonal elements. 
-The elements of the left \emph{Perron} eigenvector $u$ of the matrix $MD$ can be regarded as \emph{Fisher's reproductive values}~\citep[pg.~27]{Fisher1930}
+Let $M$ be an irreducible column-stochastic matrix and $D$ be a positive diagonal matrix. 
+The elements of the left \emph{Perron} eigenvector $\hat u$ of the matrix $MD$ can be regarded as \emph{Fisher's reproductive values}~\citep[pg.~27]{Fisher1930}
 \end{remark}
@@
 \begin{equation}
-\frac{N(j,t)}{N(k,t)} = \frac{e\tr (MD)^t e_j}{e\tr (MD)^t e_k}.
+\frac{N(j,t)}{N(k,t)} = 
+\frac{e\tr (MD)^t e_j}{e\tr (MD)^t e_k}.
 \end{equation}
@@
 \frac{e\tr V (e_1 e_1\tr) U\tr e_j} { e\tr V (e_1 e_1\tr) U\tr e_k} = 
-\frac{ e\tr v u_j} { e\tr v  u_k} =
-\frac{u_j} {u_k}.
+\frac{ e\tr \hat v \hat u_j} { e\tr \hat v  \hat u_k} =
+\frac{\hat u_j} {\hat u_k}.
 \end{equation}
 
-The vector $u$ is the left \emph{Perron} eigenvector of $MD$, and $u_k$ is $k$-th element of $u$.
-This is why the value $u_k$ can be interpreted as the \emph{reproductive value} of type $i$: it is a weighting for the size of the population generated by a single individual of type $k$.
+The vector $\hat u$ is the left \emph{Perron} eigenvector of $MD$, and $\hat u_k$ is $k$-th element of $\hat u$.
+This is why the value $\hat u_k$ can be interpreted as the \emph{reproductive value} of type $k$: it is a weighting for the size of the population generated by a single individual of type $k$.
 
@@
 \begin{equation}
-\sum_k {\hat u_k \hat v_k} = \hat u\tr v = 1.
+\sum_k {\hat u_k \hat v_k} = \hat u\tr \hat v = 1.
 \end{equation}
@@
 If $C$ and $D$ are positive diagonal matrices, and $M$ is primitive, 
-then the equilibrium $v$ of the system in Eq.~\ref{eq:model_equilibrium} is globally stable.
+then the equilibrium $\hat v$ of the system in Eq.~\ref{eq:model_equilibrium} is globally stable.
 \end{lemma}
@@
 \begin{equation}
-\sum_{k=1}^{n}{v_k}=1, \quad
+\sum_{k=1}^{n}{\hat v_k}=1, \quad
 \hat u\tr \hat v=\sum_{k=1}^{n}{\hat u_k \hat v_k}=1,
@@
 \end{equation}
-because $f$  $u$ are positive vectors.
+because $f$ and $\hat u$ are positive vectors.
 
-Now, let $\norm{x} = \sum_{k=1}^{n}|x_k|$, so that $x_t \goesto x$ implies
+Now, let $\norm{x} = \sum_{k=1}^{n}\abs{x_k}$, so that $x_t \goesto x$ implies
 \begin{equation}
@@
 \end{multline}
-since $\norm{v}=1$ (eq.~\ref{eq:sum_uv}) and $\alpha \ne 0$ (eq.~\ref{eq:alpha0}).
+since $\norm{\hat v}=1$ (eq.~\ref{eq:sum_uv}) and $\alpha \ne 0$ (eq.~\ref{eq:alpha0}).
 This completes the proof.

### 2017-11-22 ms.tex
moved result on multiple types to after result 1; changes subsections to paragraphs; formatting

@@
 \newtheorem{result}{Result}
-%\newtheorem*{mfp}{Theorem: Mean Fitness Principle}
-%\newtheorem*{rvp}{Corollary: Reproductive Value Principle}
-%\newtheorem*{frv}{Fisher's reproductive values}
 \renewcommand\qedsymbol{} % no square at end of proof
@@
 
-\subsection*{Mean fitness principle}
+% Mean fitness principle
+\paragraph*{Mean fitness principle.}
 
@@
 We will show that if the transition rate from types with below-average fitness increases, then the stable population mean fitness increases, too.
-
-% Mean fitness principle
 
@@
 \begin{proof}
-%$\hat{\bar w}$ is the leading (real positive) eigenvalue of $(I-C+MC)D$ (see eq.~\ref{eq:model_equilibrium}),
-%and the corresponding left and right eigenvectors $u$ and $v$ are both non-negative and uniquely defined by $\sum_k{\hat v_k} = 1$ and $\sum_k{\hat u_k \hat v_k} = 1$.
 Using the formula in~\citet{Caswell1978} (see eq.~\ref{eq:Caswells_formula} in~\nameref{sec:AppA}),
@@
 
+% Multiple types
+\paragraph*{}
+We turn our attention to the case where the transition rates from a subset $K$ of the types are correlated, that is, $C_j=C_i$ for $i,j \in K$.
+In this case, the following directly results from Eq.~\ref{eq:mfp}.
+
+\begin{corollary}
+The sensitivity of the stable population mean fitness to change in the rate of transition $\tau$ from types in $K$ is
+\begin{equation}\label{eq:mfp_K}
+\begin{aligned}
+\frac{\partial \hat{\bar w}}{\partial \tau} = 
+\hat u\tr \Big( \sum_{k \in K}{\frac{\partial (I - C + MC)D}{\partial \tau}} \Big) \hat v = \\
+\frac{1}{\tau} \sum_{k \in K}{\hat u_k \hat v_k (\hat{\bar w} - w_k)},
+\end{aligned}
+\end{equation}
+and
+\begin{multline}\label{eq:mfp_sign_threshold}
+sign \frac{\partial \hat{\bar w}}{\partial \tau} = 
+sign \sum_{k \in K}{\hat u_k \hat v_k (\hat{\bar w} - w_k)} = \\ 
+sign \Big(\hat{\bar w} - \frac{\sum_{k \in K}{\hat u_k \hat v_k w_k}}{\sum_{k \in K}{\hat u_k \hat v_k}}\Big).
+\end{multline}
+\end{corollary}
+
+Therefore, increased transition from types in $K$ will increase the stable population mean fitness if the average fitness of individuals descended from types in $K$ is below the stable population mean fitness.
+For example, \citet[Appendix~B]{Ram2012} considered individuals that are grouped by the number of their accumulated mutant alleles, $k$ (see \nameref{sec:mut_model_2}), 
+and a modifier allele that increases the mutation rates in individuals with at least $\pi$ mutant alleles.
+According to Eq.~\ref{eq:mfp_sign_threshold}, this modifier will increase the stable population mean fitness if individuals with $\pi$ or more mutant alleles have below-average fitness.
+
 % Reproductive value principle
+\paragraph*{Reproductive value principle.}
 
 An interesting intermediate result in Eq.~\ref{eq:mfp_intermediate} is
-
 \begin{equation}
-\begin{aligned}
 \frac{\partial \hat{\bar w}}{\partial C_k} = 
-\hat v_k w_k (\hat u\tr[M]_k - \hat u_k) = \\
+\hat v_k w_k (\hat u\tr[M]_k - \hat u_k) = 
 w_k \Big(\sum_{j=1}^{n}{\hat u_j M_{j,k} \hat v_k} - \hat u_k \hat v_k\Big).
-\end{aligned}
 \end{equation}
@@
 
-% Threshold rule
-
-We turn our attention to the case where the transition occurs from a subset $K$ of the types.
-For example, \citet[Appendix~B]{Ram2012} considered individuals that are grouped by the number of their accumulated deleterious mutations, $k$, 
-and a modifier allele that regulates the mutation rates in individuals with at least $\pi$ deleterious mutations.
-If the transition rates are controlled by a threshold rule, then 
-\begin{equation}
-C_k = \begin{cases}
-\tau, & k \in K \\
-1, & k \not\in K
-\end{cases}.
-\end{equation}
-If $\tau > 0$, then the sensitivity of the stable population mean fitness to changes in the
-rate of transition from types in $K$ can be written 
-\begin{equation}\label{eq:mfp_K}
-\begin{aligned}
-\frac{\partial \hat{\bar w}}{\partial \tau} = 
-\hat u\tr \Big( \sum_{k \in K}{\frac{\partial (I - C + MC)D}{\partial \tau}} \Big) \hat v = \\
-\frac{1}{\tau} \sum_{k \in K}{\hat u_k \hat v_k (\hat{\bar w} - w_k)}.
-\end{aligned}
-\end{equation}
-Thus
-\begin{multline}\label{eq:mfp_sign_threshold}
-sign \frac{\partial \hat{\bar w}}{\partial \tau} = 
-sign \sum_{k \in K}{\hat u_k \hat v_k (\hat{\bar w} - w_k)} = \\ 
-sign \Big(\hat{\bar w} - \frac{\sum_{k \in K}{\hat u_k \hat v_k w_k}}{\sum_{k \in K}{\hat u_k \hat v_k}}\Big), 
-\end{multline}
-a generalization of Eq.~\ref{eq:mfp_sign}.
-Therefore, increased transition from types in $K$ will increase the stable population mean fitness if the average fitness of individuals descended from types in $K$ is below the stable population mean fitness.
-
-\subsection*{Evolutionary genetic stability}
+% Evolutionary stability
+\paragraph*{Evolutionary genetic stability.}
 
@@
 Note that Eq.~\ref{eq:model} is the special case of Eq.~\ref{eq:modifier_model} where allele $M$ is absent, i.e. $g_k=0$ for all $k$.
-
-% Evolutionary stability
 
@@
 
+% Reduction principle & mutational loss
+\paragraph*{Reduction principle and mutational loss.}
+
 Note that if the modifier has the same effect on all individuals (i.e., $C_k = \hat\alpha$ and $\tilde{C}_k = \alpha$ for all $k$) then $\cl_{ex}$ becomes
@@
 Therefore, we can expect that with these modifiers, an allele $M$ that reduces the transition rate, i.e. $\alpha < \hat\alpha$, will be favored by natural selection.
-Setting $C = \alpha I$ in Eq.~\ref{eq:model_equilibrium} and proceeding as in Eq.~\ref{eq:mfp_K}, we find
+Setting $C = \alpha I$ in Eq.~\ref{eq:model_equilibrium} and proceeding as in Eq.~\ref{eq:mfp_K}, we have
 \begin{equation}\label{eq:mutational_loss}
-\frac{\partial \bar w}{\partial \alpha} = 
--\frac{1}{\alpha}\Big(\sum_{k=1}^{n} \hat u_k \hat v_k w_k - \bar w \Big).
+\frac{\partial \hat{\bar{w}}}{\partial \alpha} = 
+-\frac{1}{\alpha}\Big(\sum_{k=1}^{n} \hat u_k \hat v_k w_k - \hat{\bar{w}} \Big),
 \end{equation}
-Eq.~\ref{eq:mutational_loss} has been derived by~\citet[eq.~24]{Hermisson2002}.
-Furthermore, the term in the parentheses of Eq.~\ref{eq:mutational_loss} is the difference between the 
-\emph{ancestral mean fitness} and the \emph{stable population mean fitness}, and has been termed the \emph{mutational loss} by~\citet[eq.~24]{Hermisson2002} -- "the linear response of the equilibrium mean fitness to changes in the mutation rates".
-Following~\citet{Altenberg2017}, $\frac{\partial\bar w}{\partial \alpha}<0$, and therefore we can conclude that the \emph{mutational loss} is positive and that the \emph{ancestral mean fitness} is higher then the \emph{stable population mean fitness}.
+which has been derived by~\citet[eq.~24]{Hermisson2002}.
+They called the term in the parentheses of Eq.~\ref{eq:mutational_loss}, which is equal to the difference between the \emph{ancestral mean fitness} $\sum_{k=1}^{n} \hat u_k \hat v_k w_k$ and the \emph{stable population mean fitness} $\hat{\bar{w}}$, the \emph{mutational loss}.
+Following~\citet{Altenberg2017}, $\frac{\partial\bar w}{\partial \alpha}<0$, and we can conclude that the \emph{mutational loss} is positive and that the \emph{ancestral mean fitness} is higher then the \emph{stable population mean fitness}.
 
@@
 % Conclusions
-\subsection*{Conclusions}
+\paragraph*{Conclusions.}
 
@@
 \section*{Appendices}
-% Appendix A: mean fitness
-%\subsection*{Appendix A}\label{sec:AppA}
-
-%\begin{lemma}
-%The normalizing factor $\bar{w}$ in Eq.~\ref{eq:model} is the population mean fitness
-%\begin{equation}
-%\bar{w} = \sum_k{f_k w_k}.
-%\end{equation}
-%\end{lemma}
-
-%\begin{proof}
-%By definition, the normalizing factor $\bar{w}$ ensures that $\sum_k{f'_k}=1$.
-%Denote $e = (1, ..., 1)$ the vector with ones at all positions and rewrite $e\tr f' = 1$.
-%Therefore, 
-%\begin{multline}\label{eq:mean_fitness}
-%\bar{w} = 
-%\bar{w} e\tr f' = 
-%e\tr (I - C + M C) D f = \\
-%(e\tr I - e\tr C + e\tr M C) D f = \\
-%(e\tr - e\tr C + e\tr C) D f = \\
-%e\tr D f = 
-%\diag{D} f = 
-%\sum_k{f_k w_k},
-%\end{multline}
-%using the fact that $e\tr M = e\tr$ and $e\tr I = e\tr$ because $M$ and $I$ are stochastic matrices.
-%\end{proof}
-
-%Note that this lemma also ensures that the stable population mean fitness $\hat{\bar w}=\sum_k{v_k w_k}$ is the leading eigenvalue of $(I - C + M %C) D$; see Eq.~\ref{eq:model_equilibrium} and the \emph{Perron-Frobenius theorem}~\cite[Appendix~A]{Otto2007}.
 

### 2017-11-23 ms.tex
comments
revised Reduction principle & mutational loss after speaking with UL

@@
 
-Note that if the modifier has the same effect on all individuals (i.e., $C_k = \hat\alpha$ and $\tilde{C}_k = \alpha$ for all $k$) then $\cl_{ex}$ becomes
+Note that if the modifier has the same effect on all types,
+then we can substitute $C = \mu I$ (with $\mu > 0$) in Eq.~\ref{eq:model_equilibrium}, and proceeding as in Eq.~\ref{eq:mfp_K}, we have
+\begin{equation}\label{eq:mutational_loss}
+\frac{\partial \hat{\bar{w}}}{\partial \mu} = 
+-\frac{1}{\mu}\Big(\sum_{k=1}^{n} \hat u_k \hat v_k w_k - \hat{\bar{w}} \Big) = -\frac{1}{\mu} G.
+\end{equation}
+This relationship been derived by~\citet[eq.~24]{Hermisson2002}.
+They called $G$ the \emph{mutational loss}: it is the difference between the \emph{ancestral mean fitness} $(\sum_{k=1}^{n} \hat u_k \hat v_k w_k)$ and the \emph{stable population mean fitness} $(\hat{\bar{w}})$, .
+
+If an invading modifier allele $M$ reduces the transition rate compared to the resident allele $m$, i.e.
+$C_k = \hat \mu$ and $\tilde{C}_k = \mu$ for all $k$ in Eq.~\ref{eq:Lex} and $\mu < \hat \mu$, then $\cl_{ex}$ becomes
 \begin{equation}
-\cl_{ex} = ((1 - \alpha)I + \alpha M) D /\hat{\bar w},
+\cl_{ex} = ((1 - \mu)I + \mu M) D /\hat{\bar w},
 \end{equation} 
-and the \emph{unified reduction principle} is in effect~\cite[eqs. 65, 72]{Altenberg2017}.
-Therefore, we can expect that with these modifiers, an allele $M$ that reduces the transition rate, i.e. $\alpha < \hat\alpha$, will be favored by natural selection.
-Setting $C = \alpha I$ in Eq.~\ref{eq:model_equilibrium} and proceeding as in Eq.~\ref{eq:mfp_K}, we have
-\begin{equation}\label{eq:mutational_loss}
-\frac{\partial \hat{\bar{w}}}{\partial \alpha} = 
--\frac{1}{\alpha}\Big(\sum_{k=1}^{n} \hat u_k \hat v_k w_k - \hat{\bar{w}} \Big),
-\end{equation}
-which has been derived by~\citet[eq.~24]{Hermisson2002}.
-They called the term in the parentheses of Eq.~\ref{eq:mutational_loss}, which is equal to the difference between the \emph{ancestral mean fitness} $\sum_{k=1}^{n} \hat u_k \hat v_k w_k$ and the \emph{stable population mean fitness} $\hat{\bar{w}}$, the \emph{mutational loss}.
-Following~\citet{Altenberg2017}, $\frac{\partial\bar w}{\partial \alpha}<0$, and we can conclude that the \emph{mutational loss} is positive and that the \emph{ancestral mean fitness} is higher then the \emph{stable population mean fitness}.
+and we can we can expect that $M$ will be favored by natural selection.
+That is, the \emph{unified reduction principle} is in effect~\cite[eqs. 65, 72]{Altenberg2017}.
+Thus, we can conclude that the \emph{mutational loss} $G$ is positive and that the \emph{ancestral mean fitness} is higher then the \emph{stable population mean fitness}.
 
@@
 
-% Appendix B: Caswell formula
+% Appendix: Caswell formula
 \subsection*{Appendix A}\label{sec:AppA}
@@
 
-% Appendix B: Fisher's reproductive value
+% Appendix: Fisher's reproductive value
 \subsection*{Appendix B}\label{sec:AppB}
@@
 
-% Appendix C: inner stability
+% Appendix: inner stability
 \subsection*{Appendix C}\label{sec:AppC}

### 2017-11-23 ms.tex
revised the learning model description of exploration

@@
 Individuals are characterized by their initial and mature phenotypes~\citep[pg.~94]{Boyd1985}.
-Fitness is determined by the difference between the individual mature phenotype and the optimal phenotype:
-denoting the optimal phenotype as $h$, the fitness of an individual with mature phenotype $k$ is $w_{\abs{k-h}}$.
+Fitness is determined by the mature phenotype: the fitness of an individual with mature phenotype $k$ is $w_k$.
 
-The initial phenotype is acquired by learning the mature phenotype of a parent.
+An offspring's initial phenotype is acquired by learning the mature phenotype of its parent (assuming uni-parental transmission).
 In individuals with initial phenotype $k$, the mature phenotype is the same as the initial phenotype with probability $1-C_k$, and is modified by individual learning or \emph{exploration}~\citep{Borenstein2008} with probability $C_k$.
@@
 
-For example, in the case of \emph{symmetric individual learning}~\citep{Borenstein2008}, the mature phenotype is randomly drawn from $2b+1$ phenotypes symmetrically and uniformly distributed around the initial phenotype, with the limitation that any "spillover" of phenotypes below $1$ or above $n$ is "absorbed" by those boundaries.
-This ensures that $M$ is primitive and column-stochastic.
+Therefore, if the frequency of individuals with mature phenotype $k$ in the current generation is $f_k$, then the frequency in the next generation $f'_k$ is
+\begin{equation}
+f'_k = (1-C_k) f_k \frac{w_{k}}{\bar{w}} + \sum_{j=1}^{n}{C_j M_{k,j} \frac{w_{j}}{\bar{w}} f_j },
+\end{equation}
+where $\bar{w} = \sum_{k=1}^{n}{f_k w_{k}}$ is the population mean fitness.
+
+For example, in the case of \emph{symmetric individual learning}~\citep{Borenstein2008}, learning is parameterized by its breadth of exploration $b$ and the mature phenotype $j$ is randomly drawn from $2b+1$ phenotypes symmetrically and uniformly distributed around the initial phenotype $k$, with the limitation that any "spillover" of phenotypes below $1$ or above $n$ is "absorbed" by those boundaries.
+This "absorption" ensures $M$ is column-stochastic. 
+In other words, given initial phenotype $k$, the probability for maturation to phenotype $j$ such that $k-b \le j \le k+b$ is $1/(2b+1)$, but any phenotype $j<1$ actually becomes $j=1$ and any phenotype $j>n$ actually becomes $j=n$.
+The probability for maturation to other phenotypes is $0$.
 
@@
 \end{equation}
-
-Therefore, if the frequency of individuals with mature phenotype $k$ in the current generation is $f_k$, then the frequency in the next generation $f'_k$ is
-\begin{equation}
-f'_k = (1-C_k) f_k \frac{w_{\abs{k-h}}}{\bar{w}} + \sum_{j=1}^{n}{C_j M_{k,j} \frac{w_{\abs{j-h}}}{\bar{w}} f_j },
-\end{equation}
-where $\bar{w} = \sum_{k=1}^{n}{f_k w_{\abs{k-h}}}$ is the population mean fitness.
 

### 2017-11-26 ms.tex
merged LA's changes to abstract and intro

@@
 % Title page
-\title{Variation generation and mean fitness increase: \\ Necessity is the mother of genetic invention}
+\title{Conditional Generation of Variation and Mean Fitness Increase: \\ Necessity is the Mother of Genetic Invention}
 % \title{A General Principle for the Evolution of Fitness-Associated Generation of Variation}
@@
 However, departures from this \emph{reduction principle} have been demonstrated.
-Here we analyze a general model of evolution under constant selection where the rate at which variation is generated can depend on the fitness of the individual.
+Here we analyze a general model of evolution under constant selection where the rate at which variation is generated depends on the individual.
 We find that if a modifier allele increases the rate at which individuals of below-average fitness generate variation, then it will increase in frequency and increase the  population mean fitness.
-This principle applies to phenomena such as stress-induced mutagenesis and condition-dependent dispersal,
-and exemplifies \emph{"Necessity is the mother of genetic invention."}
+This principle applies to phenomena such as stress-induced mutagenesis and condition-dependent dispersal, and exemplifies \emph{``Necessity is the mother of genetic invention.''}
 \end{abstract}
@@
 
-According to the \emph{reduction principle}, in populations at a balance between natural selection and a process that generates variation (i.e.
+The \emph{reduction principle} is the widely general phenomenon that, in populations at a balance between natural selection and a process that generates variation (i.e.
 mutation, migration, or recombination), selection favors neutral modifiers that decrease the rate at which variation is generated.
-The \emph{reduction principle} was demonstrated for modifiers of the recombination~\citep{Feldman1972}, mutation~\citep{Liberman1986a}, migration~\citep{Feldman1986}, and dispersal rates~\citep{Altenberg1987}.
-These separate principles were recently unified by~\citet{Altenberg2017}
+The \emph{reduction principle} was demonstrated for modifiers of the recombination~\citep{Feldman1972}, mutation~\citep{Liberman1986a}, migration~\citep{Feldman1986}, and these results were generalized in a unified treatment in a series of studies ~\citep{Altenberg:1984:Book,Altenberg1987,Altenberg:2009:Linear,Altenberg:2012:Resolvent-Positive,Altenberg2012b,Altenberg2017}.
 
-The assumptions and conditions for the validity of the \emph{unified reduction principle} by neutral genetic modifiers are~\citep{Altenberg2017}:
+The latter studies establish that that conditions for a \emph{unified reduction principle} for neutral genetic modifiers are:
+%are ~\citep{Altenberg:1984:Book,Altenberg1987,Altenberg:2009:Linear,Altenberg2017}:
 (i) effectively infinite population size, (ii) constant-viability selection, (iii) a population at an equilibrium, and (iv) \emph{linear variation} -- the equal scaling of transition probabilities by the modifier.
@@
 \end{equation}
-This relationship been derived by~\citet[eq.~24]{Hermisson2002}.
+This relationship has been derived by~\citet[eq.~24]{Hermisson2002}.
 They called $G$ the \emph{mutational loss}: it is the difference between the \emph{ancestral mean fitness} $(\sum_{k=1}^{n} \hat u_k \hat v_k w_k)$ and the \emph{stable population mean fitness} $(\hat{\bar{w}})$, .
@@
 %\bibliography{/Users/yoavram/Documents/library}
-\bibliography{ms}
+\bibliography{ms+LA}
 

### 2017-11-26 ms.tex
revised LS's changes

@@
 
-The \emph{reduction principle} is the widely general phenomenon that, in populations at a balance between natural selection and a process that generates variation (i.e.
+According to the \emph{reduction principle}, in populations at a balance between natural selection and a process that generates variation (i.e.
 mutation, migration, or recombination), selection favors neutral modifiers that decrease the rate at which variation is generated.
-The \emph{reduction principle} was demonstrated for modifiers of the recombination~\citep{Feldman1972}, mutation~\citep{Liberman1986a}, migration~\citep{Feldman1986}, and these results were generalized in a unified treatment in a series of studies ~\citep{Altenberg:1984:Book,Altenberg1987,Altenberg:2009:Linear,Altenberg:2012:Resolvent-Positive,Altenberg2012b,Altenberg2017}.
+The \emph{reduction principle} was demonstrated for modifiers of the recombination~\citep{Feldman1972}, mutation~\citep{Liberman1986a}, and migration~\citep{Feldman1986}.
+These results were unified in a series of studies ~\citep{Altenberg1984,Altenberg1987,Altenberg2009,Altenberg2012b,Altenberg2012a,Altenberg2017}.
 
-The latter studies establish that that conditions for a \emph{unified reduction principle} for neutral genetic modifiers are:
-%are ~\citep{Altenberg:1984:Book,Altenberg1987,Altenberg:2009:Linear,Altenberg2017}:
+The latter studies have established the conditions for a \emph{unified reduction principle} by neutral genetic modifiers:
 (i) effectively infinite population size, (ii) constant-viability selection, (iii) a population at an equilibrium, and (iv) \emph{linear variation} -- the equal scaling of transition probabilities by the modifier.
@@
 %\bibliography{/Users/yoavram/Documents/library}
-\bibliography{ms+LA}
+\bibliography{ms}
 

### 2017-11-26 ms.tex
revised mutational loss section after MWF comments

@@
 Note that if the modifier has the same effect on all types,
-then we can substitute $C = \mu I$ (with $\mu > 0$) in Eq.~\ref{eq:model_equilibrium}, and proceeding as in Eq.~\ref{eq:mfp_K}, we have
+then we can substitute $C = \mu I$ (with $\mu > 0$) in Eq.~\ref{eq:model_equilibrium}, and proceeding as in Eq.~\ref{eq:mfp_K}, we find a relationship previously described by~\citet[eq.~24]{Hermisson2002},
 \begin{equation}\label{eq:mutational_loss}
 \frac{\partial \hat{\bar{w}}}{\partial \mu} = 
--\frac{1}{\mu}\Big(\sum_{k=1}^{n} \hat u_k \hat v_k w_k - \hat{\bar{w}} \Big) = -\frac{1}{\mu} G.
+-\frac{1}{\mu}\Big(\sum_{k=1}^{n} \hat u_k \hat v_k w_k - \hat{\bar{w}} \Big) = -\frac{1}{\mu} G,
 \end{equation}
-This relationship has been derived by~\citet[eq.~24]{Hermisson2002}.
-They called $G$ the \emph{mutational loss}: it is the difference between the \emph{ancestral mean fitness} $(\sum_{k=1}^{n} \hat u_k \hat v_k w_k)$ and the \emph{stable population mean fitness} $(\hat{\bar{w}})$, .
+where $G$, equal to the difference between the \emph{ancestral mean fitness} $(\sum_{k=1}^{n} \hat u_k \hat v_k w_k)$ and the \emph{stable population mean fitness} $(\hat{\bar{w}})$, is called the \emph{mutational loss}~\citep{Hermisson2002}.
 
-If an invading modifier allele $M$ reduces the transition rate compared to the resident allele $m$, i.e.
+If an invading modifier allele $M$ reduces the transition probability compared to the resident allele $m$, i.e.
 $C_k = \hat \mu$ and $\tilde{C}_k = \mu$ for all $k$ in Eq.~\ref{eq:Lex} and $\mu < \hat \mu$, then $\cl_{ex}$ becomes

### 2017-11-26 ms.tex
some MWF changes over the phone

@@
 Another departure from the \emph{linear variation} assumption of the \emph{reduction principle} for mutation rates involves a mechanism by which the mutation rate increases in individuals of low fitness -- a mechanism first observed in stressed bacteria~\citep{Foster2007}, although not in a constant environment.
-\citet{Ram2012} demonstrated that even in a constant environment, modifiers that increase the mutation rate of individuals with below-average fitness increase the population mean fitness, rather than decrease it.
+\citet{Ram2012} demonstrated that even in a constant environment, increasing the mutation rate of individuals with below-average fitness increases the population mean fitness, rather than decreases it.
 Their analysis assumed  infinite population size and fitness  determined by the number of mutant alleles accumulated in the genotype.
@@
 
-The following result presents a \emph{mean fitness principle} for the
-sensitivity of the equilbrium mean fitness $\hat{\bar w}$ to changes
+The following result constitutes a \emph{mean fitness principle} for the
+sensitivity of the equilibrium mean fitness $\hat{\bar w}$ to changes
 in $C_k$, the probability of transition from $A_k$.
@@
 Let $\hat{\bar w}$ be the leading eigenvalue of $(I-C+MC)D$,
-and $\hat u$ and $\hat v$ be the corresponding left and right eigenvectors, such that $\sum_{k=1}^{n}\hat v_k=1$ and $\sum_{k=1}^{n}\hat u_k\hat v_k=1$.
+and $\hat u$ and $\hat v$ be the corresponding positive left and right eigenvectors, such that $\sum_{k=1}^{n}\hat v_k=1$ and $\sum_{k=1}^{n}\hat u_k\hat v_k=1$.
 Then,
@@
 \end{equation}
-Finally, note that $\hat u_k, \hat v_k, C_k > 0$ and therefore
+Finally, since $\hat u_k, \hat v_k, C_k > 0$, we have
 \begin{equation}
@@
 We turn our attention to the case where the transition rates from a subset $K$ of the types are correlated, that is, $C_j=C_i$ for $i,j \in K$.
-In this case, the following directly results from Eq.~\ref{eq:mfp}.
+In this case, Eq.~\ref{eq:mfp} leads directly to the following.
 
@@
 
-An interesting intermediate result in Eq.~\ref{eq:mfp_intermediate} is
+An interesting interpretation of Eq.~\ref{eq:mfp_intermediate} is
 \begin{equation}
@@
 % Modifier model
-We now focus on a neutral modifier locus in complete linkage to the types $A_k$, with no direct effect on fitness, and whose sole function is to determine $C_k$, the rates of transition from the different types.
+We now focus on a neutral modifier locus completely linked to the types $A_k$, with no direct effect on fitness, and whose sole function is to determine $C_k$, the rates of transition from the different types.
 We will show that modifier alleles that increase the stable population mean fitness in accordance with Result~\ref{result:mfp} are favored by natural selection.
@@
 Could such an allele increase in frequency when initially rare in the population?
-To answer this we analyze the stability of resident modifier allele $m$ with transition rates $C_k$ to the invasion by a modifier allele $M$ with rates $\tilde{C}_k$, see Eq.~\ref{eq:modifier_model}~\citep[for an introduction to \emph{evolutionary invasion analysis}, see][ch.~12]{Otto2007}
+To answer this we analyze the stability of resident modifier allele $m$ with transition rates $C_k$ to the invasion by a modifier allele $M$ with rates $\tilde{C}_k$, see Eqs.~\ref{eq:modifier_model}.
 
 The equilibrium of Eqs.~\ref{eq:modifier_model} when modifier allele $M$ is absent from the population is $(\hat v,0)$, where $\hat v$ is given in Eq.~\ref{eq:model_equilibrium} and $g_k=0$ for all $k$.
-The stability of allele $m$ to invasion by allele $M$ is determined by the leading eigenvalue $\lambda_1$ of $\cl_{ex}$ the external stability matrix of the equilibrium $(\hat v,0)$, which, in turn, is determined by the Jacobian $\cj$ of the system in Eq.~\ref{eq:modifier_model} evaluated at the equilibrium $(\hat v,0)$, where
+The stability of allele $m$ to invasion by allele $M$ is determined by the leading eigenvalue $\lambda_1$ of $\cl_{ex}$ the external stability matrix of the equilibrium $(\hat v,0)$, which, in turn, is determined by the Jacobian $\cj$ of the system in Eqs.~\ref{eq:modifier_model} evaluated at the equilibrium $(\hat v,0)$, where
 \begin{equation}
@@
 Let $\lambda_1$ be the the leading eigenvalue of the external stability matrix $\cl_{ex}$.
-If the transition rates induced by the modifier alleles $m$ and $M$ are equal, i.e., $\tilde{C}_k=C_k$ for all $k$, then the following applies
+If the transition rates induced by the modifier alleles $m$ and $M$ are equal, i.e., $\tilde{C}_k=C_k$ for all $k$, then 
 \begin{equation}

### 2017-11-27 ms.tex
why M should be primitive; removed conditional from title; don't talk about modifiers when discussing Ram2012; better single mutation explanation in mutation model 2

@@
 % Title page
-\title{Conditional Generation of Variation and Mean Fitness Increase: \\ Necessity is the Mother of Genetic Invention}
+\title{Generation of Variation and Mean Fitness Increase: \\ Necessity is the Mother of Genetic Invention}
+%\title{Conditional Generation of Variation and Mean Fitness Increase: \\ Necessity is the Mother of Genetic Invention}
 % \title{A General Principle for the Evolution of Fitness-Associated Generation of Variation}
@@
 Their analysis assumed  infinite population size and fitness  determined by the number of mutant alleles accumulated in the genotype.
-In their models, the only departure from the \emph{reduction principle} assumptions was due to the differential effect the modifier has on different genotypes.
+In their models, the only departure from the \emph{reduction principle} assumptions was the unequal scaling of mutation probabilities between different genotypes introduced by the correlation between the mutation rate and fitness.
 A similar result has been demonstrated for conditional dispersal~\citep[Th.~39]{Altenberg2012b}, fitness-associated recombination~\citep{Hadany2003a} and for condition-dependent sexual reproduction~\citep{Hadany2007a}, and evidence suggests that both mechanisms are common in nature~\citep{Ram2016}.
@@
 \citet{Ram2012} stated that their result represents a departure from the \emph{reduction principle}, but did not explain this departure.
-Their analysis was specific to a model that classified individuals by the number of mutant genes in their genotype, similar to models studied by~\citet{Kimura1966a} and~\citet{Haigh1978}.
-Moreover, their argument was based on the expected increase of the population mean fitness by alleles that modify the mutation rate, rather than on the invasion success of such modifiers~\citep[i.e., analysis of \emph{evolutionary genetic stability}, see][]{Eshel1982, Lessard1990}.
+Their analysis was specific to a model that classified individuals by the number of mutant alleles in their genotype, similar to models studied by~\citet{Kimura1966a} and~\citet{Haigh1978}.
+Moreover, their argument was based on the expected increase of the stable population mean fitness, rather than on the invasion success of modifier alleles that modify the mutation rate~\citep[i.e., analysis of \emph{evolutionary genetic stability}, see][]{Eshel1982, Lessard1990}.
 
@@
 Again, we consider a large population of haploids, but here individuals with genotype $A_k$ are characterized by the number $k$ of deleterious or mutant alleles in their genotype, where $0 \le k \le n$.
-Specifically, the fitness of individuals with $k$ mutant alleles is $w_k$ ($w_0 \ge w_1 \ge \ldots \ge w_n$),
+Specifically, the fitness of individuals with $k$ mutant alleles is $w_k$ ($w_0 > w_1 > \ldots > w_n$),
 and the probability $C_k$ that a mutation occurs in individuals with $k$ mutant alleles depends on $k$.
-We assume a low mutation rate $C_k$ so that two mutations are unlikely to occur in the same individual in one generation.
 When a mutation occurs it is \emph{deleterious} with probability $\delta$, generating a mutant allele and converting the individual from $A_k$ to $A_{k+1}$,
@@
 Note that such beneficial mutations can be either compensatory or back-mutations, and that mutations are neutral with probability $1-\delta-\beta$.
+We assume that both the deleterious and the beneficial mutation rates are low, such that two mutations are unlikely to occur in the same individual in one generation: $C_k(\delta + \beta) \ll 1$ for all $k=1, \ldots, n$.
 This model has been analyzed by~\citet{Ram2012}.
@@
 Eq.~\ref{eq:mutation_model_2} can be viewed as a special case of Eq.~\ref{eq:model}.
-Here, too, ${M}$ is irreducible and primitive.
+Here, too, ${M}$ is irreducible and primitive as long as $\delta, \beta > 0$.
 
@@
 
-\begin{corollary}[Reproductive value principle]
+\begin{corollary}[Reproductive value principle]\label{cor:rvp}
 In the notation of Result~\ref{result:mfp},
@@
 \end{corollary}
+
+Corollary~\ref{cor:rvp} sheds light on why we require $M$ to be primitive.
+If $M$ is primitive then individuals of type $k$ can transition into any other type in a finite number of generations.
+So individuals with below-average fitness can have descendants with above-average fitness, and increased generation of variation in these individuals will increase the stable population mean fitness.
+In contrast, if $M$ is imprimitive, individuals with below-average fitness are "doomed" and increasing the generation of variation in these individuals can only hasten their removal from the population.
+For example, if we set $\beta=0$ in \nameref{sec:mut_model_2}, $M$ becomes triangular and imprimitive, and the stable mean fitness becomes $\bar{w} = (1-\delta C_0)w_0$, which is not affected by changes in $C_k$ for $k \ge 1$~\citep[see also][Fig.~1A]{Agrawal2002,Ram2012}.
 
@@
 We have shown that under constant-viability selection and in an effectively infinite haploid population at mutation-selection or migration-selection equilibrium, the stable population mean fitness increases if individuals with below-average fitness increase the rate at which variation is generated.
-Furthermore, modifier alleles that  increase generation of variation in such individuals are favored by natural evolution.
+Furthermore, modifier alleles that  increase generation of variation in such individuals are favored by natural selection.
+These results apply as long as there is a chance for the variation-generating process to transform an individual with below-average fitness into one with above-average fitness (e.g.
+$M$ in Eq.~\ref{eq:model} is primitive).
 
@@
 Eq.~\ref{eq:mfp} is a generalization of a result of~\citet[Eq.~4]{Ram2012}.
-\citeauthor{Ram2012} modeled the accumulation of mutant alleles in a population evolving at the mutation-selection balance (see \nameref{sec:mut_model_2}).
+\citeauthor{Ram2012} modeled the accumulation of mutant alleles in a population (see \nameref{sec:mut_model_2}).
 Using Eq.~\ref{eq:Caswells_formula} in \nameref{sec:AppA} and a recursion on the ratios of the reproductive values \cite[see][eqs.~A5-6]{Ram2012}, they concluded that at the mutation-selection balance, if individuals with below-average fitness ($w_k < \hat{\bar w}$) increase their mutation rate, then the population mean fitness will increase -- a result generalized by our \emph{\nameref{result:mfp}} in Eqs.~\ref{eq:mfp_sign} and~\ref{eq:mfp_sign_threshold}.
@@
 Our analysis focuses on populations at equilibrium.
-Nevertheless, it has been demonstrated that a modifier which increases the mutation rate of maladapted individuals can be favored by selection~\citep{Ram2012} and reduce the adaptation rate~\citep{Ram2014},
+Nevertheless, it has been demonstrated that during adaptive evolution (i.e., in non-equilibrium populations), a modifier that increases the mutation rate of maladapted individuals can be favored by selection~\citep{Ram2012} and increase the adaptation rate~\citep{Ram2014},
 and empirical evidence suggests that \emph{stress-induced mutagenesis} is common in bacteria and yeast, and may be prevalent in plants, flies, and human cancer cells~\citep{Rosenberg2012,Fitzgerald2017b}.

### 2017-11-28 ms.tex
remove another modifier statement

@@
 For example, \citet[Appendix~B]{Ram2012} considered individuals that are grouped by the number of their accumulated mutant alleles, $k$ (see \nameref{sec:mut_model_2}), 
-and a modifier allele that increases the mutation rates in individuals with at least $\pi$ mutant alleles.
-According to Eq.~\ref{eq:mfp_sign_threshold}, this modifier will increase the stable population mean fitness if individuals with $\pi$ or more mutant alleles have below-average fitness.
+and the effect of increasing the mutation rate in individuals with at least $\pi$ mutant alleles.
+According to Eq.~\ref{eq:mfp_sign_threshold}, this will result in increased stable population mean fitness if individuals with $\pi$ or more mutant alleles have below-average fitness.
 

### 2017-11-28 ms.tex
merged MWF's changes

@@
 mutation, migration, or recombination), selection favors neutral modifiers that decrease the rate at which variation is generated.
-The \emph{reduction principle} was demonstrated for modifiers of the recombination~\citep{Feldman1972}, mutation~\citep{Liberman1986a}, and migration~\citep{Feldman1986}.
+The \emph{reduction principle} was demonstrated for modifiers of  recombination~\citep{Feldman1972}, mutation~\citep{Liberman1986a}, and migration~\citep{Feldman1986}.
 These results were unified in a series of studies ~\citep{Altenberg1984,Altenberg1987,Altenberg2009,Altenberg2012b,Altenberg2012a,Altenberg2017}.
@@
 
-Here, we present an evolutionary model where the the type of the individual determines both its fitness and the rate at which it generates variation.
+Here, we present an evolutionary model in which the type of the individual determines both its fitness and the rate at which it generates variation.
 Our results show that the population mean fitness increases if individuals with below-average fitness produce more variation than individuals with above-average fitness,
@@
 $C$ is a positive diagonal  matrix with  elements $C_k$;
-$M$ is a primitive column-stochastic matrix: $M_{j,k} \ge 0$ for all $j,k$, $\sum_{j=1}^n {M_{j,k}} = 1$ for all $k$, and $(M^n)_{j,k} > 0$ for all $j,k$ for some natural $n$;
+$M$ is a primitive column-stochastic matrix: $M_{j,k} \ge 0$ for all $j,k$, $\sum_{j=1}^n {M_{j,k}} = 1$ for all $k$, and $(M^{\ell})_{j,k} > 0$ for all $j,k$ for some positive integer $\ell$;
 $\ci$ is the $n\times n$ identity matrix;
@@
 Importantly, type transmission is vertical and uni-parental (the type is transmitted from a single parent to the offspring) and is independent of the frequencies of the other types.
-Specifically, this model precludes processes such as recombination, social learning, sexual outcrossing, and horizontal or oblique transmission, as these processes are frequency-dependent~\cite[pg.~54]{Cavalli-Sforza1981}.
+This model precludes processes such as recombination, social learning, sexual outcrossing, and horizontal or oblique transmission, as these processes are frequency-dependent~\citep[pg.~54]{Cavalli-Sforza1981}.
 
@@
 \end{equation}
-for all $k=1, 2, \ldots, n$, where $\bar{w} = \sum_{k=1}^{n}{f_k w_k}$ is the population mean fitness.
+for  $k=1, 2, \ldots, n$, where $\bar{w} = \sum_{k=1}^{n}{f_k w_k}$ is the population mean fitness.
 
@@
 Note that such beneficial mutations can be either compensatory or back-mutations, and that mutations are neutral with probability $1-\delta-\beta$.
-We assume that both the deleterious and the beneficial mutation rates are low, such that two mutations are unlikely to occur in the same individual in one generation: $C_k(\delta + \beta) \ll 1$ for all $k=1, \ldots, n$.
+We assume that both the deleterious and the beneficial mutation rates are low enough that two mutations are unlikely to occur in the same individual in one generation: $C_k(\delta + \beta) \ll 1$ for all $k=1, \ldots, n$.
 This model has been analyzed by~\citet{Ram2012}.
@@
 This "absorption" ensures $M$ is column-stochastic. 
-In other words, given initial phenotype $k$, the probability for maturation to phenotype $j$ such that $k-b \le j \le k+b$ is $1/(2b+1)$, but any phenotype $j<1$ actually becomes $j=1$ and any phenotype $j>n$ actually becomes $j=n$.
+In other words, given initial phenotype $k$, the probability of maturation to phenotype $j$, where $k-b \le j \le k+b$ is $1/(2b+1)$, but any phenotype $j<1$ actually becomes $j=1$ and any phenotype $j>n$ actually becomes $j=n$.
 The probability for maturation to other phenotypes is $0$.
@@
 We first focus on the stable population mean fitness.
-We will show that if the transition rate from types with below-average fitness increases, then the stable population mean fitness increases, too.
+We  show that if the transition rate from types with below-average fitness increases, then the stable population mean fitness increases, too.
 
@@
 So individuals with below-average fitness can have descendants with above-average fitness, and increased generation of variation in these individuals will increase the stable population mean fitness.
-In contrast, if $M$ is imprimitive, individuals with below-average fitness are "doomed" and increasing the generation of variation in these individuals can only hasten their removal from the population.
+In contrast, if $M$ is  not primitive, individuals with below-average fitness are "doomed" and increasing the generation of variation in these individuals can only hasten their removal from the population.
 For example, if we set $\beta=0$ in \nameref{sec:mut_model_2}, $M$ becomes triangular and imprimitive, and the stable mean fitness becomes $\bar{w} = (1-\delta C_0)w_0$, which is not affected by changes in $C_k$ for $k \ge 1$~\citep[see also][Fig.~1A]{Agrawal2002,Ram2012}.
@@
 
-We have found a condition for a modifier allele that controls that transition rate $C_k$ from type $A_k$ to increase the stable population mean fitness (Result~\ref{result:mfp}).
-Could such an allele increase in frequency when initially rare in the population?
-To answer this we analyze the stability of resident modifier allele $m$ with transition rates $C_k$ to the invasion by a modifier allele $M$ with rates $\tilde{C}_k$, see Eqs.~\ref{eq:modifier_model}.
+Result~\ref{result:mfp} provides a condition under which increasing the transition rate $C_k$ from type $A_k$ will increase the stable population mean fitness.
+Could a modifier allele that increases $C_k$ increase in frequency when initially rare in the population?
+To answer this we analyze the stability of resident modifier allele $m$ with transition rates $C_k$ to  invasion by a modifier allele $M$ with rates $\tilde{C}_k$ under Eqs.~\ref{eq:modifier_model}.
 
@@
 \begin{equation}\label{eq:Lex}
-\cl_{ex} = (I - \tilde{C} + M \tilde{C}) D /\hat{\bar w},
+\cl_{ex} = (I - \tilde{C} + M \tilde{C}) \tilde{D} ,
 \end{equation}
-where $\hat{\bar w}$ is the stable population mean fitness in the absence of modifier allele $M$.
-Note that, in general, $\hat{\bar w}$ is not an eigenvalue of $\cl_{ex}$.
-\medskip
+where $\tilde{D} =D/\hat{\bar{w}}$ is the diagonal matrix with elements $w_k/\hat{\bar{w}}$ for all $k$, and $\hat{\bar{w}}$ is the stable population mean fitness in the absence of the modifier allele $M$.
+$\lambda_1$, the leading eigenvalue of $\cl_{ex}$, coincides with $\tilde{\bar{w}}$, the maximal mean fitness associated with (28).
+Thus we can apply Result 1 to $\lambda_1$ and obtain the following result. %TODO refs, maximal?
 
@@
 
-Therefore, an initially rare modifier allele $M$ with transition rates slightly different from the resident allele $m$ can successfully invade the population ($\lambda_1>1$) if $M$ increases the probability of transition from types with below-average fitness, thereby increasing the stable population mean fitness.
+Therefore, an initially rare modifier allele $M$ with transition rates slightly different from the resident allele $m$ can successfully invade the population ($\lambda_1>1$) if $M$ increases the probability of transition from types with below-average fitness, thereby increasing the  mean fitness $\tilde{\bar w}$.
 \end{result}
@@
 \end{equation}
-where $G$, equal to the difference between the \emph{ancestral mean fitness} $(\sum_{k=1}^{n} \hat u_k \hat v_k w_k)$ and the \emph{stable population mean fitness} $(\hat{\bar{w}})$, is called the \emph{mutational loss}~\citep{Hermisson2002}.
+where $G$, equal to the difference between the \emph{ancestral mean fitness} $(\sum_{k=1}^{n} \hat u_k \hat v_k w_k)$ and the \emph{stable population mean fitness} $\hat{\bar{w}}$ when $C_k=\mu$, is  called the \emph{mutational loss}~\citep{Hermisson2002}.
 
-If an invading modifier allele $M$ reduces the transition probability compared to the resident allele $m$, i.e.
-$C_k = \hat \mu$ and $\tilde{C}_k = \mu$ for all $k$ in Eq.~\ref{eq:Lex} and $\mu < \hat \mu$, then $\cl_{ex}$ becomes
+If an invading  allele $M$ reduces the transition probability compared to the resident allele $m$, i.e. $C_k = \mu$ and $\tilde{C}_k = \tilde\mu$ with $\tilde\mu<\mu$,  then $\cl_{ex}$ becomes
 \begin{equation}
-\cl_{ex} = ((1 - \mu)I + \mu M) D /\hat{\bar w},
+\cl_{ex} = \bigl[(1 - \tilde\mu)I + \tilde\mu M\bigr] \bar D,
 \end{equation} 
-and we can we can expect that $M$ will be favored by natural selection.
-That is, the \emph{unified reduction principle} is in effect~\cite[eqs. 65, 72]{Altenberg2017}.
-Thus, we can conclude that the \emph{mutational loss} $G$ is positive and that the \emph{ancestral mean fitness} is higher then the \emph{stable population mean fitness}.
+where $\bar D =D/\bar w$ is  the diagonal matrix with entries $w_k/\bar w$ for $k=1,2,\dots,n$.
+Using the unified reduction principle \citep{Altenberg2017} we know that the leading eigenvalue $\lambda_1$  of $\cl_{ex}$ in Eq.\ (35) satisfies $\partial\lambda_1/\partial\tilde\mu<0$.
+Thus we can conclude that the mutational loss $G$ is positive and the ancestral mean fitness is higher than the population mean fitness.
+% TODO check above for refs, hats, etc.
 
@@
 
-General examples of departures from the \emph{reduction principle} are rare~\citep{Altenberg2017}.
-For example, it has been demonstrated that higher mutation, recombination, and migration rates are favored by selection in populations evolving under fluctuating selection; see~\citet{Carja2014} and references therein.
+Departures from the \emph{reduction principle} for mutation, recombination, and migration rates usually involve fluctuating selection,  non-equilibrium dynamics, or departures from random mating (see~\cite{Carja2014} and references therein).
 Here we have provided another general example, which suggests that a modifier allele that causes individuals with below-average fitness to increase the rate at which variation is generated, will be favored by selection and will lead to increased population mean fitness.

### 2017-11-28 get_bibtex.py
remove unused keys

@@ -37,6 +37,9 @@ def main(keys_filename, bibtex_filename, output_filename, verbose):
 		if m:
 			prefix, species, postfix = m.groups()
 			ent['title'] = prefix + r"\emph{" + species + r"}" + postfix
+		for key in ['file', 'abstract', 'mendeley-tags', 'keyword']:
+			if key in ent:
+				ent.pop(key)		
 
 	writer = BibTexWriter()
 	with open(output_filename, 'w') as f:

### 2017-11-28 ms.tex
cite Lukacisinova2017

@@
 where $f=(f_1, f_2, \ldots, f_n)$ is a frequency vector with $f_k\ge 0$ and  $\sum_{k=1}^n{f_k} = 1$;
-$D$ is a positive diagonal matrix with  elements $w_k$ such that $w_k \ne w_j$ for some $k \ne j$; 
-%and we can assume $0 < w_k \le 1$, i.e., fitness values are positive and relative to the fitness of the fittest type; 
-$C$ is a positive diagonal  matrix with  elements $C_k$;
+$D$ is a positive diagonal matrix with entries $w_k$ such that $w_k \ne w_j$ for some $k \ne j$; 
+$C$ is a positive diagonal  matrix with entries $C_k$;
 $M$ is a primitive column-stochastic matrix: $M_{j,k} \ge 0$ for all $j,k$, $\sum_{j=1}^n {M_{j,k}} = 1$ for all $k$, and $(M^{\ell})_{j,k} > 0$ for all $j,k$ for some positive integer $\ell$;
@@
 \end{equation}
-and $\cl_{in}$ is the local stability matrix of the equilibrium $(\hat v,0)$ in the space $\sum_{k=1}^n{v_k}=1$.
+and $\cl_{in}$ is the local stability matrix of the equilibrium $(\hat v,0)$ in the space $\sum_{k=1}^n{f_k}=1$.
 The zero block matrices are due to the complete linkage between the modifier and the types $A_k$ and to the lack of transition (i.e., mutation) between the modifier alleles.
@@
 \end{equation}
-where $\tilde{D} =D/\hat{\bar{w}}$ is the diagonal matrix with elements $w_k/\hat{\bar{w}}$ for all $k$, and $\hat{\bar{w}}$ is the stable population mean fitness in the absence of the modifier allele $M$.
-$\lambda_1$, the leading eigenvalue of $\cl_{ex}$, coincides with $\tilde{\bar{w}}$, the maximal mean fitness associated with (28).
-Thus we can apply Result 1 to $\lambda_1$ and obtain the following result. %TODO refs, maximal?
+where $\tilde{D} =D/\hat{\bar{w}}$ is the diagonal matrix with entries $w_k/\hat{\bar{w}}$ for all $k$, and $\hat{\bar{w}}$ is the stable population mean fitness in the absence of the modifier allele $M$.
+$\lambda_1$, the leading eigenvalue of $\cl_{ex}$, coincides with $\tilde{\bar{w}}$, the maximal mean fitness associated with Eq.~\ref{eq:Lex}.
+Thus we can apply Result~\ref{result:mfp} to $\lambda_1$ and obtain the following result.
 
@@
 \end{equation}
-where $G$, equal to the difference between the \emph{ancestral mean fitness} $(\sum_{k=1}^{n} \hat u_k \hat v_k w_k)$ and the \emph{stable population mean fitness} $\hat{\bar{w}}$ when $C_k=\mu$, is  called the \emph{mutational loss}~\citep{Hermisson2002}.
+where $G$, equal to the difference between the \emph{ancestral mean fitness} $(\sum_{k=1}^{n} \hat u_k \hat v_k w_k)$ and the \emph{stable population mean fitness} $(\hat{\bar{w}})$ when $C_k=\mu$, is  called the \emph{mutational loss}~\citep{Hermisson2002}.
 
-If an invading  allele $M$ reduces the transition probability compared to the resident allele $m$, i.e. $C_k = \mu$ and $\tilde{C}_k = \tilde\mu$ with $\tilde\mu<\mu$,  then $\cl_{ex}$ becomes
-\begin{equation}
-\cl_{ex} = \bigl[(1 - \tilde\mu)I + \tilde\mu M\bigr] \bar D,
+If an invading  allele $M$ reduces the transition probability compared to the resident allele $m$, i.e., $C_k = \mu$ and $\tilde{C}_k = \tilde{\mu}$ for all $k$ with $\tilde{\mu}<\mu$, then $\cl_{ex}$ becomes
+\begin{equation} \label{eq:Lex_mu}
+\cl_{ex} = \bigl[(1 - \tilde\mu)I + \tilde\mu M\bigr] \tilde{D},
 \end{equation} 
-where $\bar D =D/\bar w$ is  the diagonal matrix with entries $w_k/\bar w$ for $k=1,2,\dots,n$.
-Using the unified reduction principle \citep{Altenberg2017} we know that the leading eigenvalue $\lambda_1$  of $\cl_{ex}$ in Eq.\ (35) satisfies $\partial\lambda_1/\partial\tilde\mu<0$.
-Thus we can conclude that the mutational loss $G$ is positive and the ancestral mean fitness is higher than the population mean fitness.
-% TODO check above for refs, hats, etc.
+where $\tilde{D} =D/\hat{\bar{w}}$ is  the diagonal matrix with entries $w_k/\hat{\bar{w}}$ for all $k$.
+Using the \emph{unified reduction principle}~\citep{Altenberg2017} we know that the leading eigenvalue $\lambda_1$  of $\cl_{ex}$ in Eq.~\ref{eq:Lex_mu} satisfies $\partial\lambda_1 / \partial\tilde{\mu} < 0$.
+Thus we can conclude that the mutational loss $G$ is positive and the ancestral mean fitness is higher than the stable population mean fitness.
 
@@
 Our analysis focuses on populations at equilibrium.
-Nevertheless, it has been demonstrated that during adaptive evolution (i.e., in non-equilibrium populations), a modifier that increases the mutation rate of maladapted individuals can be favored by selection~\citep{Ram2012} and increase the adaptation rate~\citep{Ram2014},
+Nevertheless, it has been demonstrated that during adaptive evolution (i.e., in non-equilibrium populations), a modifier that increases the mutation rate of maladapted individuals can be favored by selection~\citep{Ram2012,Lukacisinova2017} and increase the adaptation rate~\citep{Ram2014},
 and empirical evidence suggests that \emph{stress-induced mutagenesis} is common in bacteria and yeast, and may be prevalent in plants, flies, and human cancer cells~\citep{Rosenberg2012,Fitzgerald2017b}.
@@
 \citet{Caswell1978} gave a \emph{formula for the sensitivity of the population growth rate to changes in life history parameters}.
-In this formula, the \emph{population growth rate} is the leading eigenvalue of the population transformation matrix $T$, the \emph{life history parameters} are elements of $T$, and the \emph{sensitivity} is the derivative of the former with respect to the latter.
+In this formula, the \emph{population growth rate} is the leading eigenvalue of the population transformation matrix $T$, the \emph{life history parameters} are entries of $T$, and the \emph{sensitivity} is the derivative of the former with respect to the latter.
 This is a useful formula~\citep[ch.~10]{Caswell1978,Hermisson2002,Ram2012,Otto2007}, and therefore we reproduce it here.
@@
 Let $M$ be an irreducible column-stochastic matrix and $D$ be a positive diagonal matrix. 
-The elements of the left \emph{Perron} eigenvector $\hat u$ of the matrix $MD$ can be regarded as \emph{Fisher's reproductive values}~\citep[pg.~27]{Fisher1930}
+The entries of the left \emph{Perron} eigenvector $\hat u$ of the matrix $MD$ can be regarded as \emph{Fisher's reproductive values}~\citep[pg.~27]{Fisher1930}
 \end{remark}

### 2017-11-29 ms.tex
small change by MWF

@@
 sign \frac{\partial \lambda_1}{\partial \tilde{C}_k} =
-sign\Big(\lambda_1 - \frac{w_k}{\hat{\bar w}}\Big).
+sign\left(\lambda_1 - \frac{w_k}{\hat{\bar w}}\right).
 \end{equation}
@@
 \begin{multline}
-sign \frac{\partial \lambda_1}{\partial \tilde{C}_k}\bigg\rvert_{\tilde{C}_k = C_k} = 
-sign\Big(1 - \frac{w_k}{\hat{\bar w}}\Big) =  
+sign \left.\frac{\partial \lambda_1}{\partial \tilde{C}_k}\right\rvert_{\tilde{C}_k = C_k} = 
+sign\left(1 - \frac{w_k}{\hat{\bar w}}\right) =  
 sign(\hat{\bar w} - w_k),
@@
 \frac{\partial \hat{\bar{w}}}{\partial \mu} = 
--\frac{1}{\mu}\Big(\sum_{k=1}^{n} \hat u_k \hat v_k w_k - \hat{\bar{w}} \Big) = -\frac{1}{\mu} G,
+-\frac{1}{\mu}\left(\sum_{k=1}^{n} \hat u_k \hat v_k w_k - \hat{\bar{w}} \right) = -\frac{1}{\mu} G,
 \end{equation}
-where $G$, equal to the difference between the \emph{ancestral mean fitness} $(\sum_{k=1}^{n} \hat u_k \hat v_k w_k)$ and the \emph{stable population mean fitness} $(\hat{\bar{w}})$ when $C_k=\mu$, is  called the \emph{mutational loss}~\citep{Hermisson2002}.
+where $\hat u_k$ and $\hat v_k$ are computed at the equilibrium associated with $\mu$ and $\hat{\bar{w}}$ is the mean fitness at that equilibrium.
+$G$, the difference between the \emph{ancestral mean fitness} $(\sum_{k=1}^{n} \hat u_k \hat v_k w_k)$ and the \emph{stable population mean fitness} $(\hat{\bar{w}})$ when $C_k=\mu$, is  called the \emph{mutational loss}~\citep{Hermisson2002}.
 

### 2017-12-01 ms.tex
remove hats after eq 34

@@
 \begin{equation}\label{eq:mutational_loss}
-\frac{\partial \hat{\bar{w}}}{\partial \mu} = 
--\frac{1}{\mu}\left(\sum_{k=1}^{n} \hat u_k \hat v_k w_k - \hat{\bar{w}} \right) = -\frac{1}{\mu} G,
+\frac{\partial {\bar{w}}}{\partial \mu} = 
+-\frac{1}{\mu}\left(\sum_{k=1}^{n}  u_k  v_k w_k - {\bar{w}} \right) = -\frac{1}{\mu} G,
 \end{equation}
-where $\hat u_k$ and $\hat v_k$ are computed at the equilibrium associated with $\mu$ and $\hat{\bar{w}}$ is the mean fitness at that equilibrium.
-$G$, the difference between the \emph{ancestral mean fitness} $(\sum_{k=1}^{n} \hat u_k \hat v_k w_k)$ and the \emph{stable population mean fitness} $(\hat{\bar{w}})$ when $C_k=\mu$, is  called the \emph{mutational loss}~\citep{Hermisson2002}.
+where $u_k,v_k$ are computed at the equilibrium associated with $\mu$ and $\bar w$ is the mean fitness at that equilibrium.
+$G$ is the difference between the \emph{ancestral mean fitness} $(\sum_{k=1}^{n}  u_k  v_k w_k)$ and the \emph{stable population mean fitness} $({\bar{w}})$ when $C_k=\mu$, is  called the \emph{mutational loss}~\citep{Hermisson2002}.
 

### 2017-12-04 ms.tex
MWF and UL changed mutational loss paragraph

@@
 
-If an invading  allele $M$ reduces the transition probability compared to the resident allele $m$, i.e., $C_k = \mu$ and $\tilde{C}_k = \tilde{\mu}$ for all $k$ with $\tilde{\mu}<\mu$, then $\cl_{ex}$ becomes
+If an invading  allele $M$ changes the transition probability from that of the resident allele $m$, i.e., $C_k=\mu$ and $\tilde C_k=\tilde\mu$, then $\cl_{ex}$ becomes
 \begin{equation} \label{eq:Lex_mu}
@@
 \end{equation} 
-where $\tilde{D} =D/\hat{\bar{w}}$ is  the diagonal matrix with entries $w_k/\hat{\bar{w}}$ for all $k$.
-Using the \emph{unified reduction principle}~\citep{Altenberg2017} we know that the leading eigenvalue $\lambda_1$  of $\cl_{ex}$ in Eq.~\ref{eq:Lex_mu} satisfies $\partial\lambda_1 / \partial\tilde{\mu} < 0$.
-Thus we can conclude that the mutational loss $G$ is positive and the ancestral mean fitness is higher than the stable population mean fitness.
+where $\tilde{D} =D/{\bar{w}}$ is  the diagonal matrix with entries $w_k/{\bar{w}}$ for all $k$.
+Using the \emph{unified reduction principle}~\citep{Altenberg2017} the leading eigenvalue $\lambda_1=\lambda(\tilde\mu)$ of $\cl_{ex}$ satisfies $\partial\lambda_1/\partial\tilde\mu<0$, $\lambda_1(\tilde\mu)=1$ when $\tilde\mu=\mu$, and $\lambda_1(\tilde\mu)=\bar w(\tilde\mu)$.
+Thus we can conclude that the mutational loss $G$ is positive, the reduction principle holds, and the mean fitness is a decreasing function of $\tilde\mu$.
 

### 2018-01-30 ms.tex
changes following genetics review

@@
 \usepackage[hyphens]{url}
+\usepackage[dvipsnames]{xcolor}
+\usepackage[unicode=true,colorlinks=true,urlcolor=CadetBlue,citecolor=black,linkcolor=black]{hyperref}
 \PassOptionsToPackage{hyphens}{url} % url is loaded by hyperref
-\usepackage[unicode=true]{hyperref}
-%\usepackage{nameref} % included in hyperref
 \usepackage[auth-sc]{authblk}
@@
 The latter studies have established the conditions for a \emph{unified reduction principle} by neutral genetic modifiers:
-(i) effectively infinite population size, (ii) constant-viability selection, (iii) a population at an equilibrium, and (iv) \emph{linear variation} -- the equal scaling of transition probabilities by the modifier.
+(i) effectively infinite population size, (ii) constant-viability selection, (iii) a population at an equilibrium, and (iv) \emph{linear variation} -- the equal scaling of transition probabilities by the modifier, i.e., the modifier has the same effect on all individuals.
 A departure from the latter assumption occurs if two variation-producing processes interact~\citep{Feldman1980,Altenberg2012b}.
@@
 Their analysis assumed  infinite population size and fitness  determined by the number of mutant alleles accumulated in the genotype.
-In their models, the only departure from the \emph{reduction principle} assumptions was the unequal scaling of mutation probabilities between different genotypes introduced by the correlation between the mutation rate and fitness.
-A similar result has been demonstrated for conditional dispersal~\citep[Th.~39]{Altenberg2012b}, fitness-associated recombination~\citep{Hadany2003a} and for condition-dependent sexual reproduction~\citep{Hadany2007a}, and evidence suggests that both mechanisms are common in nature~\citep{Ram2016}.
+In their models, the only departure from the \emph{reduction principle} conditions was the unequal scaling of mutation probabilities between different genotypes introduced by the correlation between the mutation rate and fitness.
+A similar result has been demonstrated for conditional dispersal~\citep[Th.~39]{Altenberg2012b}, fitness-associated recombination~\citep{Hadany2003a} and for condition-dependent sexual reproduction~\citep{Hadany2007a}, and empirical evidence suggests these mechanisms are common in nature~\citep{Ram2016}.
 
@@
 $C$ is a positive diagonal  matrix with entries $C_k$;
-$M$ is a primitive column-stochastic matrix: $M_{j,k} \ge 0$ for all $j,k$, $\sum_{j=1}^n {M_{j,k}} = 1$ for all $k$, and $(M^{\ell})_{j,k} > 0$ for all $j,k$ for some positive integer $\ell$;
+$M$ is a primitive column-stochastic matrix: $M_{j,k} \ge 0$ for all $j,k$, $\sum_{j=1}^n {M_{j,k}} = 1$ for all $k$, and $(M^{\ell})_{j,k} > 0$ for all $j,k$ for some positive integer $\ell$ (the latter is the condition for primitivity~\citep[see][Appendix 3]{Otto2007});
 $\ci$ is the $n\times n$ identity matrix;
@@
 \end{equation}
-Note that (i) the existence and uniqueness of $\hat{\bar w}$ and $\hat v$ are guaranteed by the \emph{Perron-Frobenius theorem}~(\cite{Otto2007}) because $(I-C+MC)D$ is a non-negative primitive matrix;
+Note that (i) the existence and uniqueness of $\hat{\bar w}$ and $\hat v$ are guaranteed by the \emph{Perron-Frobenius theorem}~~\citep[][Appendix 3]{Otto2007} because $(I-C+MC)D$ is a non-negative primitive matrix;
 (ii) the global stability of this equilibrium is proven in~\nameref{sec:AppC}.
@@
 \paragraph*{}
-We turn our attention to the case where the transition rates from a subset $K$ of the types are correlated, that is, $C_j=C_i$ for $i,j \in K$.
-In this case, Eq.~\ref{eq:mfp} leads directly to the following.
+We turn our attention to the case where the transition rates from a subset $K$ of the types are equal, that is, $C_j \equiv C_i$ for $i,j \in K$.
+In this case, Eq.~\ref{eq:mfp} leads directly to the following by setting $\tau \coloneqq C_i$.
 
@@
 
-An interesting interpretation of Eq.~\ref{eq:mfp_intermediate} is
+An interesting consequence of Eq.~\ref{eq:mfp_intermediate} is
 \begin{equation}
@@
 So individuals with below-average fitness can have descendants with above-average fitness, and increased generation of variation in these individuals will increase the stable population mean fitness.
-In contrast, if $M$ is  not primitive, individuals with below-average fitness are "doomed" and increasing the generation of variation in these individuals can only hasten their removal from the population.
+In contrast, if $M$ is not primitive, individuals with below-average fitness are "doomed" and increasing the generation of variation in these individuals can only hasten their removal from the population, affecting the rate at which the population converges to the stable equilibrium, but not necessarily the stable population mean fitness.
 For example, if we set $\beta=0$ in \nameref{sec:mut_model_2}, $M$ becomes triangular and imprimitive, and the stable mean fitness becomes $\bar{w} = (1-\delta C_0)w_0$, which is not affected by changes in $C_k$ for $k \ge 1$~\citep[see also][Fig.~1A]{Agrawal2002,Ram2012}.
@@
 Furthermore, modifier alleles that  increase generation of variation in such individuals are favored by natural selection.
-These results apply as long as there is a chance for the variation-generating process to transform an individual with below-average fitness into one with above-average fitness (e.g.
-$M$ in Eq.~\ref{eq:model} is primitive).
-
-We have given several examples of variation-generating processes for which this principle applies -- namely mutation, migration, and learning (see \emph{\nameref{sec:models}} section) -- but our model may apply to other processes as well.
-For example, the \emph{reduction principle} applies to ecological models of dispersal, and~\citet{Gueijman2013} have demonstrated that even in homogeneous environments, fitness-associated dispersal increases the mean fitness of diploid populations and is favored by selection over uniform dispersal.
-Similarly, if the transmission fidelity of culturally-transmitted traits depends on the type or fitness of the transmitting individual, we expect that our results will hold (see \nameref{sec:learn_model}).
+These results apply as long as there is a chance for the variation-generating process to transform an individual with below-average fitness into one with above-average fitness.
+This condition is satisfied by the requirement that $M$ in Eq.~\ref{eq:model} be primitive. Further work will be required to find a sufficient and \emph{necessary} condition on $M$.
 
@@
 Using Eq.~\ref{eq:Caswells_formula} in \nameref{sec:AppA} and a recursion on the ratios of the reproductive values \cite[see][eqs.~A5-6]{Ram2012}, they concluded that at the mutation-selection balance, if individuals with below-average fitness ($w_k < \hat{\bar w}$) increase their mutation rate, then the population mean fitness will increase -- a result generalized by our \emph{\nameref{result:mfp}} in Eqs.~\ref{eq:mfp_sign} and~\ref{eq:mfp_sign_threshold}.
+
+Besides mutation, our principle applies to other variation-generating processes such as migration and learning (see \emph{\nameref{sec:models}} section).
+For example, the \emph{reduction principle} applies to ecological models of dispersal, and~\citet{Gueijman2013} have demonstrated that even in homogeneous environments, fitness-associated dispersal increases the mean fitness of diploid populations and is favored by selection over uniform dispersal.
+Similarly, if the transmission fidelity of culturally-transmitted traits depends on the type or fitness of the transmitting individual, we expect that our results will hold (see \nameref{sec:learn_model}).
 

### 2018-01-30 ms.tex
quorom sensing and rival assessment

@@
 In our final example, we consider a large population and an integer phenotype $k$ where $1 \le k \le n$.
-Individuals are characterized by their initial and mature phenotypes~\citep[pg.~94]{Boyd1985}.
+Individuals are characterized by their initial and mature phenotypes~\citep[pg.~94]{Boyd1988}.
 Fitness is determined by the mature phenotype: the fitness of an individual with mature phenotype $k$ is $w_k$.
@@
 
+How can individuals estimate their fitness relative to the population mean fitness? 
+In microbes, empirical evidence suggests that the mutation rate is negatively-correlated with population density~\citep{Krasovec2017} and is probably regulated by quorum sensing mechanisms~\citep{Krasovec2014,Dellus-Gur2017}.
+Animals can assess the well-being of other individuals, for example via rival and mutual assessment~\citep{Parker1974,Arnott2017,Reichert2017}, and possibly integrate these assessments over a population sample.
+Our results suggest that these assessment might affect the tendency of individuals to generate variation, e.g. explore new environments and behaviors.
+
 Eq.~\ref{eq:mfp} is a generalization of a result of~\citet[Eq.~4]{Ram2012}.

### 2018-02-09 ms.tex
reformat margins, font size, equation overflows
removed titlesec, messed up nameref
MWF and LA revisions for TPB

@@
-\documentclass[9pt, a4paper, twocolumn]{extarticle}
-\usepackage[margin=1in]{geometry} % See geometry.pdf to learn the layout options.
-\geometry{a4paper}
+\documentclass[12pt, twocolumn]{extarticle}
+\usepackage{geometry}
+\geometry{
+a4paper,
+total={170mm,257mm},
+left=20mm,
+top=20mm,
+headheight=12pt
+}
+
 %\usepackage[parfill]{parskip} % Activate to begin paragraphs with an empty line rather than an indent
@@
 \usepackage[auth-sc]{authblk}
-
-% messes up nameref
-%\usepackage{titlesec}
-%\titleformat{\subsection}[runin]
-%  {\bf}{\thesection}{1em}{}
 
@@
 % Title page
-\title{Generation of Variation and Mean Fitness Increase: \\ Necessity is the Mother of Genetic Invention}
+\title{Generation of Variation and a Modified Mean Fitness Principle: \\ Necessity is the Mother of Genetic Invention}
 %\title{Conditional Generation of Variation and Mean Fitness Increase: \\ Necessity is the Mother of Genetic Invention}
@@
 Another departure from the \emph{linear variation} assumption of the \emph{reduction principle} for mutation rates involves a mechanism by which the mutation rate increases in individuals of low fitness -- a mechanism first observed in stressed bacteria~\citep{Foster2007}, although not in a constant environment.
+A modifier gene that changes transition rates in only a subset of types exemplifies the \emph{principle of partial control} \citep{Altenberg1984,Altenberg2012a} a general statement about non-linear variation, which proposes that, ``When the modifier gene has only partial control over the
+transformations occurring at loci under selection, then it may be possible for the part which it controls to evolve an increase in rates.''
+
 \citet{Ram2012} demonstrated that even in a constant environment, increasing the mutation rate of individuals with below-average fitness increases the population mean fitness, rather than decreases it.
@@
 \end{equation}
-or in matrix form
+or in matrix form (introduced to describe conditional dispersal in \citet{Altenberg2012a})
 \begin{equation}
@@
 $D$ is a positive diagonal matrix with entries $w_k$ such that $w_k \ne w_j$ for some $k \ne j$; 
-$C$ is a positive diagonal  matrix with entries $C_k$;
-$M$ is a primitive column-stochastic matrix: $M_{j,k} \ge 0$ for all $j,k$, $\sum_{j=1}^n {M_{j,k}} = 1$ for all $k$, and $(M^{\ell})_{j,k} > 0$ for all $j,k$ for some positive integer $\ell$ (the latter is the condition for primitivity~\citep[see][Appendix 3]{Otto2007});
+$C$ is a positive diagonal  matrix with entries $C_k$, where $C_k<1$ for at least one $k$;
+$M$ is an irreducible  column-stochastic matrix: $M_{j,k} \ge 0$ for all $j,k$, $\sum_{j=1}^n {M_{j,k}} = 1$ for all $k$, and $(M^{\ell})_{j,k} > 0$ for all $j,k$ for some positive integer $\ell$ (the latter is the condition for primitivity~\citep[see][Appendix 3]{Otto2007});
 $\ci$ is the $n\times n$ identity matrix;
@@
 Importantly, different types can have different rates.
-That is, $C_i \ne C_j$ for some $i,j$. The case $C_i = C_j$ for all $i,j$ is covered by the \emph{reduction principle}~\citep[see][]{Altenberg2017}.
+That is, $C_i \ne C_j$ for some $i,j$. The case $C_i = C_j$ for all $i,j$ is covered by the \emph{reduction principle}~(see \citealt{Altenberg1984,Altenberg1987,Altenberg2009,Altenberg2017}).
 
@@
 with zeros on the diagonal and $\frac{1}{n-1}$ elsewhere.
-Note that here ${M}$ is irreducible and primitive.
-
-% Mutation model 2
+Note that here ${M}$ is primitive (i.e., irreducible and aperiodic).% Mutation model 2
 \paragraph*{Mutation model 2.}\label{sec:mut_model_2}
@@
 Eq.~\ref{eq:mutation_model_2} can be viewed as a special case of Eq.~\ref{eq:model}.
-Here, too, ${M}$ is irreducible and primitive as long as $\delta, \beta > 0$.
+Here, too, ${M}$ is primitive as long as $\delta, \beta > 0$.
 
@@
 
-After reproduction, offspring of individuals from deme $A_k$ return to their parental deme with probability $1-C_k$, or migrate to a different deme $A_j$ with probability $C_k M_{j,k}$, where the matrix $M$ is primitive, $C_k > 0$, $M_{j,k} \ge 0$, and $\sum_{j=1}^{n}{M_{j,k}} = 1$ for all $k=1, \ldots, n$.
-Therefore, $1-C_k$ can be interpreted as a \emph{homing rate}.
+After reproduction, offspring of individuals from deme $A_k$ return to their parental deme with probability $1-C_k$, or migrate to a different deme $A_j$ with probability $C_k M_{j,k}$, where the matrix $M$ is irreducible, $C_k > 0$, $C\ne I$, $M_{j,k} \ge 0$, and $\sum_{j=1}^{n}{M_{j,k}} = 1$ for all $k=1, \ldots, n$.
+Therefore, $1-C_k$ can be interpreted as a \emph{homing rate}. When the homing rate differs for different demes, it has been called \emph{conditional dispersal} \citep{McPeek1992,Kirkland2006}.
 
@@
 
-For example, in the case of \emph{symmetric individual learning}~\citep{Borenstein2008}, learning is parameterized by its breadth of exploration $b$ and the mature phenotype $j$ is randomly drawn from $2b+1$ phenotypes symmetrically and uniformly distributed around the initial phenotype $k$, with the limitation that any "spillover" of phenotypes below $1$ or above $n$ is "absorbed" by those boundaries.
+When $C_k = \tau$ for all $k$ and fitnesses are not all equal, then a culturally transmitted variant in $\tau$ will spread if and only if $\tau$ is reduced, the first manifestation of the reduction principle shown in cultural evolution \citep[pp.
+203--206]{Altenberg1984}.
+
+In a more specific example, for the case of \emph{symmetric individual learning}~\citep{Borenstein2008}, learning is parameterized by its breadth of exploration $b$ and the mature phenotype $j$ is randomly drawn from $2b+1$ phenotypes symmetrically and uniformly distributed around the initial phenotype $k$, with the limitation that any "spillover" of phenotypes below $1$ or above $n$ is "absorbed" by those boundaries.
 This "absorption" ensures $M$ is column-stochastic. 
@@
 % Mean fitness principle
-\paragraph*{Mean fitness principle.}
+\paragraph*{A modified mean-fitness principle.}
 
@@
 We  show that if the transition rate from types with below-average fitness increases, then the stable population mean fitness increases, too.
+We refer to this as a ``modified'' mean fitness principle (MMFP) because for modifiers of mutation and migration rates the standard mean fitness principle involves a decrease in the mean fitness with increasing rates.
 
@@
 
-The following result constitutes a \emph{mean fitness principle} for the
-sensitivity of the equilibrium mean fitness $\hat{\bar w}$ to changes
+The following result constitutes an MMFP for the
+direction of the sensitivity of the equilibrium mean fitness $\hat{\bar w}$ to changes
 in $C_k$, the probability of transition from $A_k$.
@@
 
-\begin{result}[Mean fitness principle]\label{result:mfp}
+\begin{result}[Modified mean fitness principle]\label{result:mfp}
 Let $\hat{\bar w}$ be the leading eigenvalue of $(I-C+MC)D$,
@@
 \begin{proof}
-Using the formula in~\citet{Caswell1978} (see eq.~\ref{eq:Caswells_formula} in~\nameref{sec:AppA}),
+Following the analysis in \citet [Th.~39]{Altenberg2012a}, we apply to \eqref {eq:model_equilibrium} the formula in~\citet{Caswell1978} (see eq.~\ref{eq:Caswells_formula} in~\nameref{sec:AppA}),
 \begin{equation}
@@
 Then,
-\begin{equation}\label{eq:mfp_intermediate}
-\begin{aligned}
+\begin{multline}\label{eq:mfp_intermediate}
+%\begin{aligned}
 \hat u\tr \frac{\partial (I-C+MC)D}{\partial C_k} \hat v = 
@@
 \hat v_k w_k (\hat u\tr[M]_k - \hat u_k). 
-\end{aligned}
-\end{equation}
+%\end{aligned}
+\end{multline}
 
@@
 We turn our attention to the case where the transition rates from a subset $K$ of the types are equal, that is, $C_j \equiv C_i$ for $i,j \in K$.
-In this case, Eq.~\ref{eq:mfp} leads directly to the following by setting $\tau \coloneqq C_i$.
+In this case, Eq.~\ref{eq:mfp} leads directly to the following by setting $\tau = C_i$ for $i\in K$.
 
@@
 The sensitivity of the stable population mean fitness to change in the rate of transition $\tau$ from types in $K$ is
-\begin{equation}\label{eq:mfp_K}
-\begin{aligned}
+\begin{multline}\label{eq:mfp_K}
+%\begin{aligned}
 \frac{\partial \hat{\bar w}}{\partial \tau} = 
@@
 \frac{1}{\tau} \sum_{k \in K}{\hat u_k \hat v_k (\hat{\bar w} - w_k)},
-\end{aligned}
-\end{equation}
+%\end{aligned}
+\end{multline}
 and
@@
 
-An interesting consequence of Eq.~\ref{eq:mfp_intermediate} is
+An interesting consequence of Eq.~\ref{eq:mfp_intermediate} observed in \citet [Th.~39]{Altenberg2012a} is
 \begin{equation}
+\begin{aligned}
 \frac{\partial \hat{\bar w}}{\partial C_k} = 
-\hat v_k w_k (\hat u\tr[M]_k - \hat u_k) = 
+\hat v_k w_k (\hat u\tr[M]_k - \hat u_k) = \\
 w_k \Big(\sum_{j=1}^{n}{\hat u_j M_{j,k} \hat v_k} - \hat u_k \hat v_k\Big).
+\end{aligned}
 \end{equation}
 Here, $\hat u_k$ can be regarded as the \emph{reproductive value} of type $k$~\citep[pg.~27]{Fisher1930}, which gives the relative contribution of type $k$ to the long-term population (see~\nameref{sec:AppB}).
-Consequently, $\hat u_k \hat v_k$ is the \emph{ancestor frequency} of type $k$~\citep{Hermisson2002}, namely the fraction of the equilibrium population descended from type $k$.
+Consequently, $\hat u_k \hat v_k$ is what ~\citet{Hermisson2002} call the \emph{ancestor frequency} of type $k$, which is better described as the fraction of the population in the distant future that descended from the individuals of type $k$ present at a particular moment in a population at equilibrium.
 The sum $\sum_{j=1}^{n}{\hat u_j M_{j,k} \hat v_k}$ can be similarly interpreted as the fraction of the equilibrium population descended from individuals that transitioned from type $k$ to another type (via the $k$ column of the transition matrix $M$), conditioned on transition occurring.
@@
 
-Corollary~\ref{cor:rvp} sheds light on why we require $M$ to be primitive.
-If $M$ is primitive then individuals of type $k$ can transition into any other type in a finite number of generations.
+Corollary~\ref{cor:rvp} sheds light on why we require $M$ to be irreducible.
+If $M$ is irreducible then individuals of type $k$ can transition into any other type in a finite number of generations.
 So individuals with below-average fitness can have descendants with above-average fitness, and increased generation of variation in these individuals will increase the stable population mean fitness.
-In contrast, if $M$ is not primitive, individuals with below-average fitness are "doomed" and increasing the generation of variation in these individuals can only hasten their removal from the population, affecting the rate at which the population converges to the stable equilibrium, but not necessarily the stable population mean fitness.
-For example, if we set $\beta=0$ in \nameref{sec:mut_model_2}, $M$ becomes triangular and imprimitive, and the stable mean fitness becomes $\bar{w} = (1-\delta C_0)w_0$, which is not affected by changes in $C_k$ for $k \ge 1$~\citep[see also][Fig.~1A]{Agrawal2002,Ram2012}.
+In contrast, if $M$ is not irreducible, it is possible that some individuals with below-average fitness are ``doomed'' since they cannot produce descendants with above-average fitness, so increasing the generation of variation in these individuals can only hasten their removal from the population, affecting the rate at which the population converges to the stable equilibrium, but not necessarily the stable population mean fitness.
+For example, if we set $\beta=0$ in \nameref{sec:mut_model_2}, $M$ becomes triangular and reducible, and the stable mean fitness becomes $\bar{w} = (1-\delta C_0)w_0$, which is not affected by changes in $C_k$ for $k \ge 1$~\citep[see also][Fig.~1A]{Agrawal2002,Ram2012}.
 
@@
 \paragraph*{Modifier model.}
-Consider the case of two modifier alleles, $m$ and $M$, inducing different transition probabilities $C=\diag{C_1, \ldots, C_n}$ and $\tilde{C}=\diag{\tilde{C}_1, \ldots, \tilde{C}_n}$, respectively.
-The frequencies of type $A_k$ linked to modifier $m$ or $M$ are $f_k$ and $g_k$, respectively, where $\sum_{k=1}^{n}{(f_k + g_k)}=1$.
+Consider the case of two modifier alleles, $b$ and $B$, inducing different transition probabilities $C=\diag{C_1, \ldots, C_n}$ and $\tilde{C}=\diag{\tilde{C}_1, \ldots, \tilde{C}_n}$, respectively.
+The frequencies of type $A_k$ linked to modifiers $b$ or $B$ are $f_k$ and $g_k$, respectively, where $\sum_{k=1}^{n}{(f_k + g_k)}=1$.
 $\bar{w}$ now ensures that $\sum_{k=1}^{n}{(f_k' + g_k')}=1$, 
@@
 
-The frequencies in the next generation, $f'$ for allele $m$ and $g'$ for allele $M$, are given by
+The frequencies in the next generation, $f'$ for allele $b$ and $g'$ for allele $B$, are given by
 \begin{equation}
@@
 Could a modifier allele that increases $C_k$ increase in frequency when initially rare in the population?
-To answer this we analyze the stability of resident modifier allele $m$ with transition rates $C_k$ to  invasion by a modifier allele $M$ with rates $\tilde{C}_k$ under Eqs.~\ref{eq:modifier_model}.
+To answer this we analyze the stability of resident modifier allele $b$ with transition rates $C_k$ to  invasion by a modifier allele $B$ with rates $\tilde{C}_k$ under Eqs.~\ref{eq:modifier_model}.
 
-The equilibrium of Eqs.~\ref{eq:modifier_model} when modifier allele $M$ is absent from the population is $(\hat v,0)$, where $\hat v$ is given in Eq.~\ref{eq:model_equilibrium} and $g_k=0$ for all $k$.
-The stability of allele $m$ to invasion by allele $M$ is determined by the leading eigenvalue $\lambda_1$ of $\cl_{ex}$ the external stability matrix of the equilibrium $(\hat v,0)$, which, in turn, is determined by the Jacobian $\cj$ of the system in Eqs.~\ref{eq:modifier_model} evaluated at the equilibrium $(\hat v,0)$, where
+The equilibrium of Eqs.~\ref{eq:modifier_model} when modifier allele $B$ is absent from the population is $(\hat v,0)$, where $\hat v$ is given in Eq.~\ref{eq:model_equilibrium} and $g_k=0$ for all $k$.
+The stability of allele $b$ to invasion by allele $B$ is determined by the leading eigenvalue $\lambda_1$ of $\cl_{ex}$ the external stability matrix of the equilibrium $(\hat v,0)$, which, in turn, is determined by the Jacobian $\cj$ of the system in Eqs.~\ref{eq:modifier_model} evaluated at the equilibrium $(\hat v,0)$, where
 \begin{equation}
@@
 \end{equation}
-where $\tilde{D} =D/\hat{\bar{w}}$ is the diagonal matrix with entries $w_k/\hat{\bar{w}}$ for all $k$, and $\hat{\bar{w}}$ is the stable population mean fitness in the absence of the modifier allele $M$.
+where $\tilde{D} =D/\hat{\bar{w}}$ is the diagonal matrix with entries $w_k/\hat{\bar{w}}$ for all $k$, and $\hat{\bar{w}}$ is the stable population mean fitness in the absence of the modifier allele $B$.
 $\lambda_1$, the leading eigenvalue of $\cl_{ex}$, coincides with $\tilde{\bar{w}}$, the maximal mean fitness associated with Eq.~\ref{eq:Lex}.
@@
 \begin{result}[Evolution of increased genetic variation]
-Let $\lambda_1$ be the the leading eigenvalue of the external stability matrix $\cl_{ex}$.
-If the transition rates induced by the modifier alleles $m$ and $M$ are equal, i.e., $\tilde{C}_k=C_k$ for all $k$, then 
+Let $\lambda_1$ be the leading eigenvalue of the external stability matrix $\cl_{ex}$.
+If the transition rates induced by the modifier alleles $b$ and $B$ are equal, i.e., $\tilde{C}_k=C_k$ for all $k$, then 
 \begin{equation}
@@
 
-Therefore, an initially rare modifier allele $M$ with transition rates slightly different from the resident allele $m$ can successfully invade the population ($\lambda_1>1$) if $M$ increases the probability of transition from types with below-average fitness, thereby increasing the  mean fitness $\tilde{\bar w}$.
+Therefore, an initially rare modifier allele $B$ with transition rates slightly different from the resident allele $m$ can successfully invade the population ($\lambda_1>1$) if $B$ increases the probability of transition from types with below-average fitness, thereby increasing the  mean fitness $\tilde{\bar w}$.
 \end{result}
@@
 Thus
-\begin{multline}
+\begin{equation}
+\begin{aligned}
 sign \left.\frac{\partial \lambda_1}{\partial \tilde{C}_k}\right\rvert_{\tilde{C}_k = C_k} = 
-sign\left(1 - \frac{w_k}{\hat{\bar w}}\right) =  
+sign\left(1 - \frac{w_k}{\hat{\bar w}}\right) = \\
 sign(\hat{\bar w} - w_k),
-\end{multline}
+\end{aligned}
+\end{equation}
 since $\hat{\bar w}>0$.
@@
 
-If an invading  allele $M$ changes the transition probability from that of the resident allele $m$, i.e., $C_k=\mu$ and $\tilde C_k=\tilde\mu$, then $\cl_{ex}$ becomes
+If an invading  allele $B$ changes the transition probability from that of the resident allele $b$, i.e., $C_k=\mu$ and $\tilde C_k=\tilde\mu$, then $\cl_{ex}$ becomes
 \begin{equation} \label{eq:Lex_mu}
@@
 where $\tilde{D} =D/{\bar{w}}$ is  the diagonal matrix with entries $w_k/{\bar{w}}$ for all $k$.
-Using the \emph{unified reduction principle}~\citep{Altenberg2017} the leading eigenvalue $\lambda_1=\lambda(\tilde\mu)$ of $\cl_{ex}$ satisfies $\partial\lambda_1/\partial\tilde\mu<0$, $\lambda_1(\tilde\mu)=1$ when $\tilde\mu=\mu$, and $\lambda_1(\tilde\mu)=\bar w(\tilde\mu)$.
+Using the \emph{unified reduction principle}~\citep{Altenberg1984,Altenberg1987,Altenberg2009,Altenberg2017} the leading eigenvalue $\lambda_1=\lambda(\tilde\mu)$ of $\cl_{ex}$ satisfies $\partial\lambda_1/\partial\tilde\mu<0$, $\lambda_1(\tilde\mu)=1$ when $\tilde\mu=\mu$, and $\lambda_1(\tilde\mu)=\bar w(\tilde\mu)$.
 Thus we can conclude that the mutational loss $G$ is positive, the reduction principle holds, and the mean fitness is a decreasing function of $\tilde\mu$.
@@
 These results apply as long as there is a chance for the variation-generating process to transform an individual with below-average fitness into one with above-average fitness. 
-This condition is satisfied by the requirement that $M$ in Eq.~\ref{eq:model} be primitive. Further work will be required to find a sufficient and \emph{necessary} condition on $M$.
+This condition is satisfied by the requirement that $M$ in Eq.~\ref{eq:model} be irreducible. Further work will be required to find a sufficient and \emph{necessary} condition on $M$.
+We refer to this as a ``modified mean-fitness principle'' because, unlike the situation with mutation and migration modifiers, the mean fitness here can increase as rates increase.
 
@@
 Animals can assess the well-being of other individuals, for example via rival and mutual assessment~\citep{Parker1974,Arnott2017,Reichert2017}, and possibly integrate these assessments over a population sample.
-Our results suggest that these assessment might affect the tendency of individuals to generate variation, e.g. explore new environments and behaviors.
+Our results suggest that such assessments might affect the tendency of individuals to generate variation, e.g. explore new genetic backgrounds, environments, and behaviors.
 
@@
 \citeauthor{Ram2012} modeled the accumulation of mutant alleles in a population (see \nameref{sec:mut_model_2}).
-Using Eq.~\ref{eq:Caswells_formula} in \nameref{sec:AppA} and a recursion on the ratios of the reproductive values \cite[see][eqs.~A5-6]{Ram2012}, they concluded that at the mutation-selection balance, if individuals with below-average fitness ($w_k < \hat{\bar w}$) increase their mutation rate, then the population mean fitness will increase -- a result generalized by our \emph{\nameref{result:mfp}} in Eqs.~\ref{eq:mfp_sign} and~\ref{eq:mfp_sign_threshold}.
+Using Eq.~\ref{eq:Caswells_formula} in \nameref{sec:AppA} and a recursion on the ratios of the reproductive values \cite[see][eqs.~A5-6]{Ram2012}, they concluded that at the mutation-selection balance, if individuals with below-average fitness ($w_k < \hat{\bar w}$) increase their mutation rate, then the population mean fitness will increase -- a result generalized by our  \emph{modified mean-fitness principle} in Eqs.~\ref{eq:mfp_sign} and~\ref{eq:mfp_sign_threshold}.
 
-Besides mutation, our principle applies to other variation-generating processes such as migration and learning (see \emph{\nameref{sec:models}} section).
-For example, the \emph{reduction principle} applies to ecological models of dispersal, and~\citet{Gueijman2013} have demonstrated that even in homogeneous environments, fitness-associated dispersal increases the mean fitness of diploid populations and is favored by selection over uniform dispersal.
+Besides mutation, our principle applies to other variation-generating processes such as conditional dispersal and migration, and learning (see \emph{\nameref{sec:models}} section).
+For example, the \emph{reduction principle} applies to ecological models of dispersal~\citep{Hastings1983, Hutson1995,Dockery1998,Cantrell2010,Altenberg2012a}, and~\citet{Gueijman2013} have demonstrated that even in homogeneous environments, fitness-associated dispersal increases the mean fitness of diploid populations and is favored by selection over uniform dispersal.
+These results parallel similar results found for reaction-diffusion models for the evolution of conditional dispersal in continuous time and space \cite{Chen2008,Hambrock2009,Cosner2009,Cantrell2010Ideal,Lam2014}.
 Similarly, if the transmission fidelity of culturally-transmitted traits depends on the type or fitness of the transmitting individual, we expect that our results will hold (see \nameref{sec:learn_model}).
@@
 Using the product rule (once in each direction),
-
-\begin{multline}
+\begin{equation}
+\begin{aligned}
 \partial (\hat u\tr T \hat v) = 
@@
 \hat u\tr \partial T \hat v + \lambda \partial(\hat u\tr \hat v).
-\end{multline}
-
+\end{aligned}
+\end{equation}
 Because $\hat u\tr \hat v = 1$,
@@
 \begin{remark}[Fisher's reproductive value]
-Let $M$ be an irreducible column-stochastic matrix and $D$ be a positive diagonal matrix. 
+Let $M$ be a primitive column-stochastic matrix and $D$ be a positive diagonal matrix. 
 The entries of the left \emph{Perron} eigenvector $\hat u$ of the matrix $MD$ can be regarded as \emph{Fisher's reproductive values}~\citep[pg.~27]{Fisher1930}
@@
 Now take the limit $t \goesto \infty$.
-By assumption, $MD$ is irreducible, so $\lambda_i < \lambda_1$ for all $i > 1$.
+By assumption, $MD$ is primitive, so $\left|\lambda_i\right| < \lambda_1$ for all $i > 1$.
 Therefore,
@@
 \begin{equation}
+\begin{aligned}
 \lim_{t \goesto \infty}\frac{N(j,t)}{N(k,t)} = 
-\frac{e\tr V (e_1 e_1\tr) U\tr e_j} { e\tr V (e_1 e_1\tr) U\tr e_k} = 
+\frac{e\tr V (e_1 e_1\tr) U\tr e_j} { e\tr V (e_1 e_1\tr) U\tr e_k} = \\
 \frac{ e\tr \hat v \hat u_j} { e\tr \hat v  \hat u_k} =
 \frac{\hat u_j} {\hat u_k}.
+\end{aligned}
 \end{equation}
@@
 \begin{lemma}[Stability of the equilibrium $\hat v$]
-If $C$ and $D$ are positive diagonal matrices, and $M$ is primitive, 
+If $C$ and $D$ are positive diagonal matrices, $C\ne I$, and $M$ is irreducible, 
 then the equilibrium $\hat v$ of the system in Eq.~\ref{eq:model_equilibrium} is globally stable.
@@
 \begin{proof}
-Let $A=(I-C+MC)D$ and denote the leading eigenvalue of $A$ as $\hat{\bar w}$.
+Let $E=(I-C+MC)D$ and denote the spectral radius of $E$ as $\hat{\bar w}$.
+
+Since $D$ is a positive diagonal matrix, and by assumption $C_k < 1$ for at least one $k$, $E$ has at least one positive diagonal element, making it aperiodic.
+Further more, since $D$ is a positive diagonal matrix and $M$ is irreducible, $E$ is also irreducible.
+Hence $E$ is primitive.
 
@@
 \begin{equation}
-\bigg(\frac{A}{\hat{\bar w}}\bigg)^t \goesto \frac{\hat v \hat u\tr}{\hat u\tr \hat v} = \hat v \hat u\tr,
+\bigg(\frac{E}{\hat{\bar w}}\bigg)^t \goesto \frac{\hat v \hat u\tr}{\hat u\tr \hat v} = \hat v \hat u\tr,
 \end{equation}
@@
 \begin{equation}
-\bigg(\frac{A}{\hat{\bar w}}\bigg)^t \cdot f \goesto \hat v \hat u\tr \cdot f = \alpha \hat v,
+\bigg(\frac{E}{\hat{\bar w}}\bigg)^t \cdot f \goesto \hat v \hat u\tr \cdot f = \alpha \hat v,
 \label{eq:perron_projection}
@@
 \begin{equation}
-\alpha = \alpha \hat u\tr \hat v = \hat u\tr \alpha \hat v = \hat u\tr \hat v \hat u\tr f = \hat u\tr f = \sum_{k=1}^{n}{\hat u_k f_k} > 0,
+\begin{aligned}
+\alpha = \alpha \hat u\tr \hat v = 
+\hat u\tr \alpha \hat v = 
+\hat u\tr \hat v \hat u\tr f = \\
+\hat u\tr f = 
+\sum_{k=1}^{n}{\hat u_k f_k} > 0,
 \label{eq:alpha0}
+\end{aligned}
 \end{equation}
@@
 \begin{equation}
-f' = \frac{A f}{\norm{A f}}. 
+f' = \frac{E f}{\norm{E f}}. 
 \label{eq:model_norm}
@@
 
-It is easily seen that $f^t =A^tf/\|A^tf\|$. Hence 
+It is easily seen that $f^t =E^tf/\|E^tf\|$. Hence 
 \begin{multline}
-\frac{A^t f}{\norm{A^t f}} = 
-\frac{\frac{1}{(\hat{\bar w})^t} A^t f}{\frac{1}{(\hat{\bar w})^t} \norm{A^t f}} = 
-\frac{\big(\frac{A}{\hat{\bar w}}\big)^t \cdot f }{ \norm{ \big(\frac{A}{\hat{\bar w}}\big)^t \cdot f }} \goesto
+\frac{E^t f}{\norm{E^t f}} = 
+\frac{\frac{1}{(\hat{\bar w})^t} E^t f}{\frac{1}{(\hat{\bar w})^t} \norm{E^t f}} = 
+\frac{\big(\frac{E}{\hat{\bar w}}\big)^t \cdot f }{ \norm{ \big(\frac{E}{\hat{\bar w}}\big)^t \cdot f }} \goesto \\
 \frac{\alpha \hat v}{\norm{\alpha \hat v}} = \hat v,
@@
 %\bibliography{/Users/yoavram/Documents/library}
-\bibliography{ms}
+\bibliography{ms+Lee2}
 

### 2018-02-09 ms.tex
typos and formatting

@@
 %SetFonts
+
+% less space before sections 
+% \@startsection {NAME}{LEVEL}{INDENT}{BEFORESKIP}{AFTERSKIP}{STYLE} 
+%            optional * [ALTHEADING]{HEADING} 
+\makeatletter
+ \renewcommand\section{\@startsection {section}{1}{\z@}%
+     {-2.5ex \@plus -1ex \@minus -.2ex}%
+     {1.3ex \@plus.2ex}%
+    {\Large\bfseries}}
 
@@
 \citet{Ram2012} demonstrated that even in a constant environment, increasing the mutation rate of individuals with below-average fitness increases the population mean fitness, rather than decreases it.
-Their analysis assumed  infinite population size and fitness  determined by the number of mutant alleles accumulated in the genotype.
+Their analysis assumed effectively infinite population size and fitness determined by the number of mutant alleles accumulated in the genotype.
 In their models, the only departure from the \emph{reduction principle} conditions was the unequal scaling of mutation probabilities between different genotypes introduced by the correlation between the mutation rate and fitness.
@@
 The probability that an individual of type $A_k$ will transition to some other type is $C_k$, and given a transition occurs, the probability that it will transition to type $A_j$ is $M_{j,k}$.
-Therefore, the change in the frequencies of type $A_k$ is described by the transformation $f \to f'$: 
+Therefore, the change in the frequencies of type $A_k$ is described by the transformation $f \to f'$
 \begin{equation}
@@
 \end{equation}
-or in matrix form (introduced to describe conditional dispersal in \citet{Altenberg2012a})
+or in matrix form 
+%(introduced to describe conditional dispersal in \citet{Altenberg2012a})
 \begin{equation}
@@
 $C$ is a positive diagonal  matrix with entries $C_k$, where $C_k<1$ for at least one $k$;
-$M$ is an irreducible  column-stochastic matrix: $M_{j,k} \ge 0$ for all $j,k$, $\sum_{j=1}^n {M_{j,k}} = 1$ for all $k$, and $(M^{\ell})_{j,k} > 0$ for all $j,k$ for some positive integer $\ell$ (the latter is the condition for primitivity~\citep[see][Appendix 3]{Otto2007});
+$M$ is an irreducible  column-stochastic matrix: $M_{j,k} \ge 0$ for all $j,k$, $\sum_{j=1}^n {M_{j,k}} = 1$ for all $k$, and $(M^{\ell})_{j,k} > 0$ for all $j,k$ for some positive integer $\ell$ (the latter is the condition for primitivity, \citealp[see][Appendix 3]{Otto2007});
 $\ci$ is the $n\times n$ identity matrix;
@@
 Importantly, different types can have different rates.
-That is, $C_i \ne C_j$ for some $i,j$. The case $C_i = C_j$ for all $i,j$ is covered by the \emph{reduction principle}~(see \citealt{Altenberg1984,Altenberg1987,Altenberg2009,Altenberg2017}).
+That is, $C_i \ne C_j$ for some $i,j$. The case $C_i = C_j$ for all $i,j$ is covered by the \emph{reduction principle}~\citep[see][]{Altenberg2017}.
 
@@
 specifically, with probability $1-C_k$, the allele $A_k$ does not mutate, and with probability $\frac{C_k}{n-1}$, the allele $A_k$ mutates to $A_j$ for any $j \ne k$.
-This is an extension of a model studied by~\citet{Altenberg2017} that allows for the mutation rate of $A_k$, $C_k$, to depend on properties of the allele $A_k$.
+This is an extension of a model studied by~\citet{Altenberg2017} that allows for $C_k$, the mutation rate of $A_k$, to depend on properties of the allele $A_k$.
 
@@
 
-When $C_k = \tau$ for all $k$ and fitnesses are not all equal, then a culturally transmitted variant in $\tau$ will spread if and only if $\tau$ is reduced, the first manifestation of the reduction principle shown in cultural evolution \citep[pp.
-203--206]{Altenberg1984}.
+When $C_k = \tau$ for all $k$ and fitnesses are not all equal, then a culturally transmitted variant in $\tau$ will spread if and only if $\tau$ is reduced, the first manifestation of the reduction principle shown in cultural evolution \citep[pg.~203--206]{Altenberg1984}.
 
@@
 % Mean fitness principle
-\paragraph*{A modified mean-fitness principle.}
+\paragraph*{Modified mean-fitness principle.}
 
@@
 \end{equation}
-Note that (i) the existence and uniqueness of $\hat{\bar w}$ and $\hat v$ are guaranteed by the \emph{Perron-Frobenius theorem}~~\citep[][Appendix 3]{Otto2007} because $(I-C+MC)D$ is a non-negative primitive matrix;
+Note that (i) the existence and uniqueness of $\hat{\bar w}$ and $\hat v$ are guaranteed by the \emph{Perron-Frobenius theorem}~\citep[][Appendix 3]{Otto2007} because $(I-C+MC)D$ is a non-negative primitive matrix;
 (ii) the global stability of this equilibrium is proven in~\nameref{sec:AppC}.
@@
 The corresponding equation to Eq.~\ref{eq:model_equilibrium} for the left
-eigenvector $u$ is 
+eigenvector $\hat u$ is 
 \begin{equation}
@@
 Here, $\hat u_k$ can be regarded as the \emph{reproductive value} of type $k$~\citep[pg.~27]{Fisher1930}, which gives the relative contribution of type $k$ to the long-term population (see~\nameref{sec:AppB}).
-Consequently, $\hat u_k \hat v_k$ is what ~\citet{Hermisson2002} call the \emph{ancestor frequency} of type $k$, which is better described as the fraction of the population in the distant future that descended from the individuals of type $k$ present at a particular moment in a population at equilibrium.
+Consequently, $\hat u_k \hat v_k$ is what ~\citet{Hermisson2002} call the \emph{ancestor frequency} of type $k$, which is the fraction of the population in the distant future that descended from individuals of type $k$ present at a particular moment in a population at equilibrium.
 The sum $\sum_{j=1}^{n}{\hat u_j M_{j,k} \hat v_k}$ can be similarly interpreted as the fraction of the equilibrium population descended from individuals that transitioned from type $k$ to another type (via the $k$ column of the transition matrix $M$), conditioned on transition occurring.
@@
 Here, $\bar{w}=\sum_{k=1}^{n}{(f_k + g_k) w_k}$ is the mean fitness of the entire population. 
-Note that Eq.~\ref{eq:model} is the special case of Eq.~\ref{eq:modifier_model} where allele $M$ is absent, i.e. $g_k=0$ for all $k$.
+Note that Eq.~\ref{eq:model} is the special case of Eq.~\ref{eq:modifier_model} where allele $B$ is absent, i.e. $g_k=0$ for all $k$.
 
@@
 
-Now, applying Result~\ref{result:mfp} (eq.~\ref{eq:mfp_sign}) to Eq.~\ref{eq:Lex}, the sign of the derivative of $\lambda_1$ with respect to $\tilde{C_k}$ is
+Now, applying Result~\ref{result:mfp} (Eq.~\ref{eq:mfp_sign}) to Eq.~\ref{eq:Lex}, the sign of the derivative of $\lambda_1$ with respect to $\tilde{C_k}$ is
 \begin{equation}
@@
 Note that if the modifier has the same effect on all types,
-then we can substitute $C = \mu I$ (with $\mu > 0$) in Eq.~\ref{eq:model_equilibrium}, and proceeding as in Eq.~\ref{eq:mfp_K}, we find a relationship previously described by~\citet[eq.~24]{Hermisson2002},
+then we can substitute $C = \mu I$ (with $\mu > 0$) in Eq.~\ref{eq:model_equilibrium}, and proceeding as in Eq.~\ref{eq:mfp_K}, we find a relationship previously described by~\citet[Eq.~24]{Hermisson2002},
 \begin{equation}\label{eq:mutational_loss}
@@
 where $u_k,v_k$ are computed at the equilibrium associated with $\mu$ and $\bar w$ is the mean fitness at that equilibrium.
-$G$ is the difference between the \emph{ancestral mean fitness} $(\sum_{k=1}^{n}  u_k  v_k w_k)$ and the \emph{stable population mean fitness} $({\bar{w}})$ when $C_k=\mu$, is  called the \emph{mutational loss}~\citep{Hermisson2002}.
+$G$ is the difference between the \emph{ancestral mean fitness} $(\sum_{k=1}^{n}  u_k  v_k w_k)$ and the \emph{stable population mean fitness} $({\bar{w}})$ when $C_k=\mu$, called the \emph{mutational loss}~\citep{Hermisson2002}.
 
@@
 where $\tilde{D} =D/{\bar{w}}$ is  the diagonal matrix with entries $w_k/{\bar{w}}$ for all $k$.
-Using the \emph{unified reduction principle}~\citep{Altenberg1984,Altenberg1987,Altenberg2009,Altenberg2017} the leading eigenvalue $\lambda_1=\lambda(\tilde\mu)$ of $\cl_{ex}$ satisfies $\partial\lambda_1/\partial\tilde\mu<0$, $\lambda_1(\tilde\mu)=1$ when $\tilde\mu=\mu$, and $\lambda_1(\tilde\mu)=\bar w(\tilde\mu)$.
+Using the \emph{unified reduction principle}~\citep{Altenberg2017} the leading eigenvalue $\lambda_1=\lambda(\tilde\mu)$ of $\cl_{ex}$ satisfies $\partial\lambda_1/\partial\tilde\mu<0$, $\lambda_1(\tilde\mu)=1$ when $\tilde\mu=\mu$, and $\lambda_1(\tilde\mu)=\bar w(\tilde\mu)$.
 Thus we can conclude that the mutational loss $G$ is positive, the reduction principle holds, and the mean fitness is a decreasing function of $\tilde\mu$.
@@
 This condition is satisfied by the requirement that $M$ in Eq.~\ref{eq:model} be irreducible. Further work will be required to find a sufficient and \emph{necessary} condition on $M$.
-We refer to this as a ``modified mean-fitness principle'' because, unlike the situation with mutation and migration modifiers, the mean fitness here can increase as rates increase.
+We refer to this as a ``modified mean-fitness principle'' because, unlike the situation with mutation and migration modifiers that uniformly affect all individuals, the mean fitness here can increase as rates increase.
 
@@
 Animals can assess the well-being of other individuals, for example via rival and mutual assessment~\citep{Parker1974,Arnott2017,Reichert2017}, and possibly integrate these assessments over a population sample.
-Our results suggest that such assessments might affect the tendency of individuals to generate variation, e.g. explore new genetic backgrounds, environments, and behaviors.
+Our results suggest that such assessments might affect the tendency of individuals to generate variation, e.g., explore new genetic backgrounds, environments, and behaviors.
 
@@
 For example, the \emph{reduction principle} applies to ecological models of dispersal~\citep{Hastings1983, Hutson1995,Dockery1998,Cantrell2010,Altenberg2012a}, and~\citet{Gueijman2013} have demonstrated that even in homogeneous environments, fitness-associated dispersal increases the mean fitness of diploid populations and is favored by selection over uniform dispersal.
-These results parallel similar results found for reaction-diffusion models for the evolution of conditional dispersal in continuous time and space \cite{Chen2008,Hambrock2009,Cosner2009,Cantrell2010Ideal,Lam2014}.
+These results parallel similar results found for reaction-diffusion models for the evolution of conditional dispersal in continuous time and space \citep{Chen2008,Hambrock2009,Cosner2009,Cantrell2010Ideal,Lam2014}.
 Similarly, if the transmission fidelity of culturally-transmitted traits depends on the type or fitness of the transmitting individual, we expect that our results will hold (see \nameref{sec:learn_model}).
@@
 \begin{lemma}
-$T$ be a non-negative matrix with leading eigenvalue $\lambda$ and left and right eigenvectors $\hat u$ and $\hat v$ such that $\sum{\hat v_k}=1$ and $\hat u\tr \hat v = \sum{\hat u_k \hat v_k} = 1$.
+Let $T$ be a non-negative matrix with leading eigenvalue $\lambda$ and left and right eigenvectors $\hat u$ and $\hat v$ such that $\sum{\hat v_k}=1$ and $\hat u\tr \hat v = \sum{\hat u_k \hat v_k} = 1$.
 Then the sensitivity of $\lambda$ to changes in any element $t$ of the matrix $T$ is
@@
 \frac{\partial \lambda}{\partial t} = 
-\hat u\tr \frac{\partial T}{\partial t} \hat v
+\hat u\tr \frac{\partial T}{\partial t} \hat v.
 \end{equation}
@@
 Using the lemma assumptions,
-$\lambda = \lambda \hat u\tr \hat v = \hat u\tr \lambda \hat v = \hat u\tr T \hat v$ and differentiating both sides we get $\partial \lambda = \partial (\hat \hat u\tr T \hat v)$.
+$\lambda = \lambda \hat u\tr \hat v = \hat u\tr \lambda \hat v = \hat u\tr T \hat v$ and differentiating both sides we get $\partial \lambda = \partial (\hat u\tr T \hat v)$.
 Using the product rule (once in each direction),
@@
 
-It is easily seen that $f^t =E^tf/\|E^tf\|$. Hence 
+It can be seen that $f^t =E^tf/\|E^tf\|$. Hence 
 \begin{multline}
@@
 \bibliographystyle{agsm}
-%\bibliography{/Users/yoavram/Documents/library}
-\bibliography{ms+Lee2}
+\bibliography{ms}
 

### 2018-02-11 ms.tex
MWF editing

@@
 \end{equation}
-where $f=(f_1, f_2, \ldots, f_n)$ is a frequency vector with $f_k\ge 0$ and  $\sum_{k=1}^n{f_k} = 1$;
+\citep{Kirkland2006,Altenberg2012b} where $f=(f_1, f_2, \ldots, f_n)$ is a frequency vector with $f_k\ge 0$ and  $\sum_{k=1}^n{f_k} = 1$;
 $D$ is a positive diagonal matrix with entries $w_k$ such that $w_k \ne w_j$ for some $k \ne j$; 
@@
 
-When $C_k = \tau$ for all $k$ and fitnesses are not all equal, then a culturally transmitted variant in $\tau$ will spread if and only if $\tau$ is reduced, the first manifestation of the reduction principle shown in cultural evolution \citep[pg.~203--206]{Altenberg1984}.
+When $C_k = \tau$ for all $k$ and fitnesses are not all equal, then a culturally transmitted variant in $\tau$ will spread if and only if $\tau$ is reduced, the first manifestation of the reduction principle shown in cultural evolution \citep[pp.~203--206]{Altenberg1984}.
 
-In a more specific example, for the case of \emph{symmetric individual learning}~\citep{Borenstein2008}, learning is parameterized by its breadth of exploration $b$ and the mature phenotype $j$ is randomly drawn from $2b+1$ phenotypes symmetrically and uniformly distributed around the initial phenotype $k$, with the limitation that any "spillover" of phenotypes below $1$ or above $n$ is "absorbed" by those boundaries.
-This "absorption" ensures $M$ is column-stochastic. 
+In a more specific example, for the case of \emph{symmetric individual learning}~\citep{Borenstein2008}, learning is parameterized by its breadth of exploration $b$ and the mature phenotype $j$ is randomly drawn from $2b+1$ phenotypes symmetrically and uniformly distributed around the initial phenotype $k$, with the limitation that any ``spillover'' of phenotypes below $1$ or above $n$ is ``absorbed'' by those boundaries.
+This ``absorption'' ensures $M$ is column-stochastic. 
 In other words, given initial phenotype $k$, the probability of maturation to phenotype $j$, where $k-b \le j \le k+b$ is $1/(2b+1)$, but any phenotype $j<1$ actually becomes $j=1$ and any phenotype $j>n$ actually becomes $j=n$.
@@
 We  show that if the transition rate from types with below-average fitness increases, then the stable population mean fitness increases, too.
-We refer to this as a ``modified'' mean fitness principle (MMFP) because for modifiers of mutation and migration rates the standard mean fitness principle involves a decrease in the mean fitness with increasing rates.
+We refer to this as a ``modified'' mean fitness principle (MMFP) because for modifiers of mutation and migration rates the standard mean fitness principle~\citep{Karlin1972}  involves a decrease in the mean fitness with increasing rates.
 
@@
 \begin{equation}\label{eq:mfp_sign}
-sign\frac{\partial \hat{\bar w}}{\partial C_k} = 
-sign(\hat{\bar w} - w_k).
+{\rm sign}\frac{\partial \hat{\bar w}}{\partial C_k} = 
+{\rm sign}(\hat{\bar w} - w_k).
 \end{equation}
@@
 \begin{equation}
-sign \frac{\partial \hat{\bar w}}{\partial C_k} = 
-sign (\hat{\bar w} - w_k),
+{\rm sign} \frac{\partial \hat{\bar w}}{\partial C_k} = 
+{\rm sign} (\hat{\bar w} - w_k),
 \end{equation}
@@
 \begin{multline}\label{eq:mfp_sign_threshold}
-sign \frac{\partial \hat{\bar w}}{\partial \tau} = 
-sign \sum_{k \in K}{\hat u_k \hat v_k (\hat{\bar w} - w_k)} = \\ 
-sign \Big(\hat{\bar w} - \frac{\sum_{k \in K}{\hat u_k \hat v_k w_k}}{\sum_{k \in K}{\hat u_k \hat v_k}}\Big).
+{\rm sign} \frac{\partial \hat{\bar w}}{\partial \tau} = 
+{\rm sign} \sum_{k \in K}{\hat u_k \hat v_k (\hat{\bar w} - w_k)} = \\ 
+{\rm sign} \Big(\hat{\bar w} - \frac{\sum_{k \in K}{\hat u_k \hat v_k w_k}}{\sum_{k \in K}{\hat u_k \hat v_k}}\Big).
 \end{multline}
@@
 \begin{equation}
-sign \frac{\partial \hat{\bar w}}{\partial C_k} = 
-sign (\hat u\tr [M]_k - \hat u_k),
+{\rm sign} \frac{\partial \hat{\bar w}}{\partial C_k} = 
+{\rm sign} (\hat u\tr [M]_k - \hat u_k),
 \end{equation}
@@
 \begin{equation}
-sign \frac{\partial \lambda_1}{\partial \tilde{C}_k}\bigg\rvert_{\tilde{C}_k = C_k} = 
-sign(\hat{\bar w} - w_k).
+{\rm sign} \frac{\partial \lambda_1}{\partial \tilde{C}_k}\bigg\rvert_{\tilde{C}_k = C_k} = 
+{\rm sign}(\hat{\bar w} - w_k).
 \end{equation}
@@
 \begin{equation}
-sign \frac{\partial \lambda_1}{\partial \tilde{C}_k} =
-sign\left(\lambda_1 - \frac{w_k}{\hat{\bar w}}\right).
+{\rm sign} \frac{\partial \lambda_1}{\partial \tilde{C}_k} =
+{\rm sign}\left(\lambda_1 - \frac{w_k}{\hat{\bar w}}\right).
 \end{equation}
@@
 \begin{aligned}
-sign \left.\frac{\partial \lambda_1}{\partial \tilde{C}_k}\right\rvert_{\tilde{C}_k = C_k} = 
-sign\left(1 - \frac{w_k}{\hat{\bar w}}\right) = \\
-sign(\hat{\bar w} - w_k),
+{\rm sign} \left.\frac{\partial \lambda_1}{\partial \tilde{C}_k}\right\rvert_{\tilde{C}_k = C_k} = 
+{\rm sign}\left(1 - \frac{w_k}{\hat{\bar w}}\right) = \\
+{\rm sign}(\hat{\bar w} - w_k),
 \end{aligned}
@@
 This condition is satisfied by the requirement that $M$ in Eq.~\ref{eq:model} be irreducible. Further work will be required to find a sufficient and \emph{necessary} condition on $M$.
-We refer to this as a ``modified mean-fitness principle'' because, unlike the situation with mutation and migration modifiers that uniformly affect all individuals, the mean fitness here can increase as rates increase.
+We refer to our result as a ``modified mean-fitness principle'' because, unlike the situation with mutation and migration modifiers that uniformly affect all individuals, the mean fitness here can increase as rates increase.
 
@@
 Since $D$ is a positive diagonal matrix, and by assumption $C_k < 1$ for at least one $k$, $E$ has at least one positive diagonal element, making it aperiodic.
-Further more, since $D$ is a positive diagonal matrix and $M$ is irreducible, $E$ is also irreducible.
+Furthermore, since $D$ is a positive diagonal matrix and $M$ is irreducible, $E$ is also irreducible.
 Hence $E$ is primitive.
